---
title: "Single_Cell_analysis"
author: "Alos Diallo"
date: "2025-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading data previously processed:
```{r}
combined_data = readRDS(file= "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/Single_cell/Single_cell_samples/Single\ Cell/combined_data.rds")

analysis_outputs <- "/Users/adiallo/Desktop/Thesis/Results"
```



Build biologically-meaningful "program" gene sets
```{r}
###############################################################################
# PURPOSE
# -------
# Build atlas-driven program gene sets and score with UCell; export
# patient-level summaries. Then derive Tex↑/Tex↓ within CD8 **without**
# pseudotime and **without** mixed models by using within-patient Spearman
# correlations and Stouffer meta-analysis across patients.
#
# NEW:
# - Adds Treg program(s):
#     1) Treg_suppressive   (atlas-driven from cTNI08/cTNI09)
#     2) Treg_cytokine_IL10_TGFB_IL35 (curated fallback panel; optional)
###############################################################################
suppressPackageStartupMessages({
  library(Seurat); library(dplyr); library(purrr); library(stringr)
  library(tibble); library(readr); library(UCell); library(Matrix); library(stats)
})

# ---------------------------
# 0) Inputs, labels, QC covars
# ---------------------------
srt <- combined_data
srt$atlas_label <- srt$predicted.id

# Broad class mapping (unchanged; truncated to save space)
label_to_class <- c(
  "cB2 (B GC-like)"="B cells","cB3 (B CD40+ GC-like)"="B cells","cB1 (B IGD+IgM+)"="B cells",
  "cP1 (Plasma IgA)"="Plasma cells","cP2 (Plasma IgG)"="Plasma cells","cP3 (Plasma IgG prolif)"="Plasma cells",
  "cMA01 (Mast)"="Mast cells","cM01 (Monocyte)"="Monocytes","cM02 (Macrophage-like)"="Macrophages",
  "cM03 (DC1)"="Dendritic cells","cM04 (DC2)"="Dendritic cells","cM05 (DC2 C1Q+)"="Dendritic cells",
  "cM06 (DC IL22RA2)"="Dendritic cells","cM07 (pDC)"="Dendritic cells","cM08 (AS-DC)"="Dendritic cells",
  "cM09 (mregDC)"="Dendritic cells","cM10 (Granulocyte)"="Granulocytes",
  "cTNI01 (CD4+ IL7R+)"="CD4+ T-cells","cTNI02 (CD4+ IL7R+SELL+)"="CD4+ T-cells",
  "cTNI03 (CD4+ IL7R+HSP+)"="CD4+ T-cells","cTNI04 (CD4+ IL7R+CCL5+)"="CD4+ T-cells",
  "cTNI05 (CD4+ IL17+)"="CD4+ T-cells","cTNI06 (CD4+ TFH)"="TFH cells","cTNI07 (CD4+ CXCL13+)"="CD4+ T-cells",
  "cTNI08 (CD4+ Treg)"="Treg","cTNI09 (CD4+ Treg prolif)"="Treg",
  "cTNI10 (CD8+ IL7R+)"="CD8+ T-cells","cTNI11 (CD8+GZMK+)"="CD8+ T-cells",
  "cTNI12 (CD8+ IL7R+)"="CD8+ T-cells","cTNI13 (CD8+ T IL17+)"="CD8+ T-cells",
  "cTNI14 (CD8+ CXCL13+)"="CD8+ T-cells","cTNI15 (CD8+ CXCL13+ HSP+)"="CD8+ T-cells",
  "cTNI16 (CD8+ CXCL13+ prolif)"="CD8+ T-cells","cTNI17 (gd-like T)"="γδ T-cells",
  "cTNI18 (gd-like T PDCD1+)"="γδ T-cells","cTNI19 (gd-like T prolif)"="γδ T-cells",
  "cTNI20 (PLZF+ T)"="iNKT cells","cTNI21 (PLZF+ T prolif)"="MAIT cells","cTNI22 (cTNI22)"="T cells (other)",
  "cTNI23 (NK CD16A+)"="NK cells","cTNI24 (NK GZMK+)"="NK cells","cTNI25 (NK XCL1+)"="NK cells",
  "cTNI26 (ILC3)"="Innate lymphoid cells",
  "cE01 (Stem/TA-like)"="Tumor cells","cE02 (Stem/TA-like/Immature Goblet)"="Tumor cells",
  "cE03 (Stem/TA-like prolif)"="Tumor cells","cE04 (Enterocyte 1)"="Tumor cells","cE05 (Enterocyte 2)"="Tumor cells",
  "cE06 (Immature Goblet)"="Tumor cells","cE07 (Goblet/Enterocyte)"="Goblet cells","cE08 (Goblet)"="Goblet cells",
  "cE09 (Best4)"="Goblet cells","cE10 (Tuft)"="Tuft cells","cE11 (Enteroendocrine)"="Enteroendocrine cells",
  "cS01 (Endo arterial)"="Endothelial cells","cS02 (Endo capillary)"="Endothelial cells",
  "cS03 (Endo capillary)"="Endothelial cells","cS04 (Endo)"="Endothelial cells","cS05 (Endo venous)"="Endothelial cells",
  "cS06 (Endo lymphatic)"="Endothelial cells","cS07 (Endo capillary-like)"="Endothelial cells",
  "cS08 (Endo arterial-like)"="Endothelial cells","cS09 (Endo)"="Endothelial cells","cS10 (Endo tip cells)"="Endothelial cells",
  "cS11 (Endo proif)"="Endothelial cells","cS12 (Endo)"="Endothelial cells","cS13 (Endo venous-like)"="Endothelial cells",
  "cS14 (Endo)"="Endothelial cells","cS15 (Pericyte)"="Pericytes","cS16 (Pericyte)"="Pericytes","cS17 (Pericyte)"="Pericytes",
  "cS18 (Pericyte)"="Pericytes","cS19 (Pericyte)"="Pericytes","cS20 (Pericyte prolif)"="Pericytes",
  "cS21 (Fibro stem cell niche)"="CAF_stem_niche","cS22 (Fibro stem cell niche)"="CAF_stem_niche",
  "cS23 (Fibro BMP-producing)"="CAF_stem_niche","cS24 (Fibro BMP-producing)"="CAF_stem_niche",
  "cS25 (Fibro CCL8+)"="iCAF","cS26 (Myofibro)"="mCAF","cS27 (CXCL14+ CAF)"="CAF",
  "cS28 (GREM1+ CAF)"="mCAF","cS29 (MMP3+ CAF)"="mCAF","cS30 (CAF CCL8 Fibro-like)"="iCAF",
  "cS31 (CAF stem niche Fibro-like)"="CAF_stem_niche","cS32 (Smooth Muscle)"="Smooth muscle cells","cS33 (Schwann)"="Schwann cells"
)

srt$cell_class <- ifelse(srt$atlas_label %in% names(label_to_class),
                         label_to_class[srt$atlas_label], "Unmapped")

# Patient/metastasis and batch proxy
srt$patient_id <- ifelse(grepl("^patient_\\d+$", srt$orig.ident),
                         sub("^patient_", "", srt$orig.ident), srt$orig.ident)
srt$mets <- ifelse(tolower(srt$dataset_origin) %in% c("metastasis","mets","ln","distant"), 1L, 0L)
if (!"batch" %in% colnames(srt@meta.data)) srt$batch <- srt$orig.ident

# QC covariates
DefaultAssay(srt) <- "RNA"
srt[["percent.mt"]] <- PercentageFeatureSet(srt, pattern = "^MT-")
cc <- Seurat::cc.genes.updated.2019
srt <- CellCycleScoring(srt, s.features = cc$s.genes, g2m.features = cc$g2m.genes, set.ident = FALSE)

# ------------------------------------
# 1) Markers within each broad class
# ------------------------------------
min_cells_per_label <- 50; min_labels_in_class <- 2
get_markers_within_class <- function(srt_class){
  Idents(srt_class) <- factor(srt_class$atlas_label)
  keep_levels <- names(which(table(Idents(srt_class)) >= min_cells_per_label))
  srt_class <- subset(srt_class, idents = keep_levels)
  if (length(unique(Idents(srt_class))) < min_labels_in_class) return(tibble())
  mk <- Seurat::FindAllMarkers(srt_class, only.pos=TRUE, min.pct=0.1, logfc.threshold=0.25, test.use="wilcox") |> tibble::as_tibble()
  nm <- names(mk)
  if ("cluster" %in% nm) mk <- dplyr::rename(mk, label=.data$cluster)
  else if ("group" %in% nm) mk <- dplyr::rename(mk, label=.data$group)
  else if ("ident" %in% nm) mk <- dplyr::rename(mk, label=.data$ident)
  mk
}
classes_target <- c("Fibroblasts","CD8+ T-cells","Tumor cells","Monocytes","Macrophages","Dendritic cells","Treg")
classes_keep   <- intersect(classes_target, unique(srt$cell_class))
markers_tbl <- map(classes_keep, function(cls){
  srt_sub <- subset(srt, subset = cell_class == cls)
  if (ncol(srt_sub) == 0) return(tibble())
  out <- get_markers_within_class(srt_sub)
  if (nrow(out) == 0) return(tibble())
  mutate(out, cell_class = cls)
}) |> bind_rows()
if (nrow(markers_tbl) == 0) stop("No markers found—lower thresholds or check class mapping.")

# -------------------------
# 2) Program definitions
# -------------------------
program_map <- list(
  CAF_myofibroblast = c("cS26 (Myofibro)", "cS28 (GREM1+ CAF)", "cS29 (MMP3+ CAF)"),
  CAF_inflammatory  = c("cS25 (Fibro CCL8+)", "cS30 (CAF CCL8 Fibro-like)"),
  CAF_stem_niche    = c("cS21 (Fibro stem cell niche)", "cS22 (Fibro stem cell niche)",
                        "cS31 (CAF stem niche Fibro-like)", "cS23 (Fibro BMP-producing)", "cS24 (Fibro BMP-producing)"),
  CD8_cytotoxic     = c("cTNI11 (CD8+GZMK+)"),  # NK removed
  CD8_exhausted     = c("cTNI14 (CD8+ CXCL13+)", "cTNI15 (CD8+ CXCL13+ HSP+)",
                        "cTNI16 (CD8+ CXCL13+ prolif)"),
  CD8_naive_memory  = c("cTNI10 (CD8+ IL7R+)", "cTNI12 (CD8+ IL7R+)"),
  Tumor_epithelial  = c("cE01 (Stem/TA-like)", "cE02 (Stem/TA-like/Immature Goblet)",
                        "cE03 (Stem/TA-like prolif)", "cE04 (Enterocyte 1)", "cE05 (Enterocyte 2)"),
  Myeloid_inflammatory = c("cM01 (Monocyte)", "cM02 (Macrophage-like)"),
  # NEW: Treg atlas-driven program (suppression/activation markers come from markers_tbl)
  Treg_suppressive  = c("cTNI08 (CD4+ Treg)", "cTNI09 (CD4+ Treg prolif)")
)

# Optional curated Treg cytokine panel to ensure IL10/TGFB/IL35 coverage (added after atlas sets)
curated_Treg_cytokine <- c(
  "FOXP3","IL2RA","IKZF2","CTLA4","TIGIT","LAG3","TNFRSF18","TNFRSF4","ENTPD1","NT5E",
  "IL10","TGFB1","TGFB2","TGFB3","IL12A","EBI3"  # IL-35 = IL12A + EBI3
)

# -----------------------------------------------
# 3) Build program gene sets from marker tables
# -----------------------------------------------
topN <- 80
rm_patterns <- c("^RPL","^RPS","^MT-")
cleanup <- function(genes) genes[!Reduce(`|`, lapply(rm_patterns, function(p) grepl(p, genes)))]

program_gene_sets <- imap(program_map, function(lbls, prog){
  lbls_present <- intersect(lbls, unique(markers_tbl$label))
  if (length(lbls_present) == 0) return(NULL)
  markers_tbl |>
    filter(label %in% lbls_present) |>
    group_by(label) |>
    slice_max(order_by = avg_log2FC, n = topN, with_ties = FALSE) |>
    ungroup() |>
    pull(gene) |>
    unique() |>
    cleanup()
}) |> discard(is.null)

# Append curated Treg cytokine program if genes exist in the object
genes_all <- rownames(srt)
curated_keep <- unique(curated_Treg_cytokine[curated_Treg_cytokine %in% genes_all])
if (length(curated_keep) >= 6) {  # require a minimal footprint
  program_gene_sets$Treg_cytokine_IL10_TGFB_IL35 <- curated_keep
}

dir.create("program_gene_sets", showWarnings = FALSE)
iwalk(program_gene_sets, ~ write_csv(tibble(gene = .x),
                                     file = file.path("program_gene_sets", paste0(.y, ".csv"))))
cat("Programs built:\n"); print(names(program_gene_sets))

# ------------------------------
# 4) Score programs with UCell
# ------------------------------
stopifnot(length(program_gene_sets) > 0)
DefaultAssay(srt) <- "RNA"
program_gene_sets_f <- imap(program_gene_sets, function(gs, nm){
  keep <- unique(gs[gs %in% genes_all])
  message(sprintf("[CHECK] %s: %d / %d genes present", nm, length(keep), length(gs)))
  if (length(keep) == 0) NULL else keep
}) |> discard(is.null)
if (length(program_gene_sets_f) == 0) stop("No program had genes present.")

pre_cols <- colnames(srt@meta.data)
srt <- Seurat::AddModuleScore(object = srt, features = program_gene_sets_f,
                              method = "UCell", name = make.names(names(program_gene_sets_f)))
post_cols <- colnames(srt@meta.data)
added_cols <- setdiff(post_cols, pre_cols)
added_by_name <- unlist(mapply(function(nm){ grep(paste0("^", make.names(nm), "\\d+$"), added_cols, value = TRUE) },
                               names(program_gene_sets_f)))
if (length(added_by_name) != length(program_gene_sets_f)) added_by_name <- grep("\\d+$", added_cols, value=TRUE)
stopifnot(length(added_by_name) == length(program_gene_sets_f))
desired <- paste0(make.names(names(program_gene_sets_f)), "_UCell")
colnames(srt@meta.data)[match(added_by_name, colnames(srt@meta.data))] <- desired
score_cols <- desired
message("[INFO] UCell columns: ", paste(score_cols, collapse = ", "))

# Export patient summaries
cell_scores <- srt@meta.data |>
  dplyr::select(patient_id, mets, cell_class, atlas_label, all_of(score_cols)) |>
  as_tibble()
patient_scores <- cell_scores |>
  group_by(patient_id, mets, cell_class) |>
  summarize(across(all_of(score_cols), ~mean(.x, na.rm=TRUE)), .groups = "drop")
write_csv(patient_scores, "program_patient_scores.csv")
write_csv(markers_tbl, "atlas_label_markers_by_class.csv")

# -------------------------------------------------------------------------
# 4B) Tex↑ / Tex↓ without pseudotime & without lmer (replicate-aware)
# -------------------------------------------------------------------------
cd8 <- subset(
  srt,
  subset = cell_class %in% c("CD8+ T-cells","T cells") &
           !atlas_label %in% c("cTNI23 (NK CD16A+)","cTNI24 (NK GZMK+)","cTNI25 (NK XCL1+)")
)
if (ncol(cd8) < 50) stop("Too few CD8 cells after NK removal.")

E <- Seurat::GetAssayData(cd8, slot = "data")        # genes x cells
tex_score <- cd8$CD8_exhausted_UCell
if (is.null(tex_score)) stop("CD8_exhausted_UCell not found; ensure program scoring ran.")

keep <- rowSums(E > 0) >= max(20, floor(0.02 * ncol(E)))
E <- E[keep, , drop = FALSE]

cov_df <- cd8@meta.data[, c("nCount_RNA","percent.mt","S.Score","G2M.Score","patient_id")]
cov_df$nCount_RNA <- scale(cov_df$nCount_RNA)
cov_df$percent.mt <- scale(cov_df$percent.mt)
cov_df$S.Score    <- scale(cov_df$S.Score)
cov_df$G2M.Score  <- scale(cov_df$G2M.Score)

tex_res <- residuals(lm(tex_score ~ nCount_RNA + percent.mt + S.Score + G2M.Score, data = cov_df))

min_cells_per_patient <- 30
meta_one_gene <- function(g){
  z_list <- list(); w_list <- list()
  for (pid in unique(cov_df$patient_id)) {
    idx <- which(cov_df$patient_id == pid)
    if (length(idx) < min_cells_per_patient) next
    x <- as.numeric(E[g, idx]); y <- tex_res[idx]
    if (all(x == 0)) next
    r <- suppressWarnings(cor(x, y, method = "spearman"))
    if (is.na(r)) next
    r <- max(min(r, 0.999999), -0.999999)
    z <- atanh(r) * sqrt(length(idx) - 3)
    z_list[[length(z_list)+1]] <- z
    w_list[[length(w_list)+1]] <- sqrt(length(idx) - 3)
  }
  k <- length(z_list)
  if (k == 0) return(c(rho_meta = NA, p_meta = NA, k = 0))
  z_vec <- unlist(z_list); w_vec <- unlist(w_list)
  Z <- sum(z_vec) / sqrt(sum(w_vec^2))
  p <- 2*pnorm(-abs(Z))
  rho_hat <- tanh( Z / sqrt(mean(w_vec^2)) )
  c(rho_meta = rho_hat, p_meta = p, k = k)
}

message("[Tex] Computing replicate-aware meta-correlations (no pseudotime, no lmer)...")
res_mat <- t(vapply(rownames(E), meta_one_gene, c(rho_meta=NA_real_, p_meta=NA_real_, k=NA_real_)))
res_df  <- as_tibble(res_mat) |> mutate(gene = rownames(E),
                                        q_meta = p.adjust(p_meta, "BH"),
                                        k = as.integer(k)) |>
          relocate(gene)

min_patients_with_signal <- 3
tex_up   <- res_df %>% filter(!is.na(rho_meta), q_meta < 0.05, rho_meta > 0, k >= min_patients_with_signal) %>%
  arrange(desc(rho_meta)) %>% slice_head(n = 200) %>% pull(gene)
tex_down <- res_df %>% filter(!is.na(rho_meta), q_meta < 0.05, rho_meta < 0, k >= min_patients_with_signal) %>%
  arrange(rho_meta) %>% slice_head(n = 200) %>% pull(gene)

dir.create("program_gene_sets", showWarnings = FALSE)
write_csv(res_df, "program_gene_sets/Tex_meta_gene_table.csv")
write_csv(tibble(gene = tex_up),   "program_gene_sets/Tex_up.csv")
write_csv(tibble(gene = tex_down), "program_gene_sets/Tex_down.csv")
message(sprintf("[Tex] Selected %d Tex↑ and %d Tex↓ genes (FDR<0.05; ≥%d patients).",
                length(tex_up), length(tex_down), min_patients_with_signal))

tex_sets_present <- list(
  Tex_up   = tex_up[tex_up   %in% rownames(srt)],
  Tex_down = tex_down[tex_down %in% rownames(srt)]
)
tex_sets_present <- tex_sets_present[ lengths(tex_sets_present) > 0 ]
if (length(tex_sets_present) >= 1) {
  pre <- colnames(srt@meta.data)
  srt <- AddModuleScore(srt, features = tex_sets_present, method = "UCell", name = "Tex")
  post <- setdiff(colnames(srt@meta.data), pre)
  idx <- grep("^Tex\\d+$", post, value = TRUE)
  if (length(idx) >= 1) {
    new_names <- paste0(names(tex_sets_present), "_UCell")
    colnames(srt@meta.data)[match(idx[seq_along(new_names)], colnames(srt@meta.data))] <- new_names
    message("[Tex] Added: ", paste(new_names, collapse = ", "))
  }
} else {
  warning("[Tex] Tex gene lists empty or absent from object; skipping re-score.")
}

# --------------------------------------
# 5) (Optional) basic sanity correlation
# --------------------------------------
if (exists("patient_scores")) {
  cd8_ps <- dplyr::filter(patient_scores, grepl("CD8|T cell|T-cells", cell_class))
  if (all(c("CD8_cytotoxic_UCell","CD8_exhausted_UCell") %in% colnames(cd8_ps)) && nrow(cd8_ps) >= 5) {
    r <- suppressWarnings(cor(cd8_ps$CD8_cytotoxic_UCell, cd8_ps$CD8_exhausted_UCell, method = "spearman"))
    message(sprintf("[CHECK] Spearman(CD8_cytotoxic, CD8_exhausted) = %.2f (expect negative)", r))
  }
}

# ----------------
# Outputs summary
# ----------------
# program_gene_sets/:
#   - <PROGRAM>.csv           (atlas-driven; now includes Treg_suppressive)
#   - Treg_cytokine_IL10_TGFB_IL35.csv (if ≥6 genes detected)
#   - Tex_meta_gene_table.csv ; Tex_up.csv ; Tex_down.csv
# program_patient_scores.csv ; atlas_label_markers_by_class.csv
###############################################################################
```

Generate the plots and statistics for each program
```{r}
###############################################################################
# PURPOSE
# Batch screen of multiple gene programs at the **cell level**:
# for each specified program and its matching cell class, test score ~ metastasis.
# Adds Treg programs to program_specs.
###############################################################################
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(purrr); library(stringr)
  library(tibble); library(readr); library(ggplot2)
})

stopifnot(exists("srt"))
meta <- srt@meta.data

# (Optional safety) collapse any duplicated UCell columns by row-wise mean
dup_names <- names(meta)[duplicated(names(meta))]
if (length(dup_names) > 0) {
  message("[CLEANUP] Collapsing duplicated columns in srt@meta.data ...")
  for (nm in unique(dup_names)) {
    idx <- which(names(meta) == nm)
    if (length(idx) == 1) next
    if (all(vapply(meta[, idx, drop=FALSE], is.numeric, TRUE))) {
      combined <- rowMeans(meta[, idx, drop=FALSE], na.rm = TRUE)
    } else {
      m <- meta[, idx, drop=FALSE]
      combined <- apply(m, 1, function(x){
        i <- which(!is.na(x))
        if (length(i) == 0) NA else x[tail(i, 1)]
      })
    }
    meta <- meta[, -idx, drop=FALSE]
    meta[[nm]] <- combined
  }
  srt@meta.data <- meta
}

# ------------------------------------------------------------------------------
# Programs to test (restrict each program to the right class)
# ------------------------------------------------------------------------------
program_specs <- tibble::tibble(
  program      = c("CD8_exhausted_UCell",
                   "CD8_cytotoxic_UCell",
                   "CD8_naive_memory_UCell",
                   "CAF_inflammatory_UCell",
                   "CAF_myofibroblast_UCell",
                   "CAF_stem_niche_UCell",
                   "Tumor_epithelial_UCell",
                   # NEW: Treg programs (note: columns only appear if genes existed)
                   "Treg_suppressive_UCell",
                   "Treg_cytokine_IL10_TGFB_IL35_UCell"),
  class_filter = list(
    c("CD8+ T-cells"),    # CD8_exhausted
    c("CD8+ T-cells"),    # CD8_cytotoxic
    c("CD8+ T-cells"),    # CD8_naive_memory
    c("Fibroblasts"),               # CAF_inflammatory
    c("Fibroblasts"),               # CAF_myofibroblast
    c("Fibroblasts"),               # CAF_stem_niche
    c("Tumor cells"),               # Tumor_epithelial
    c("Treg"),                      # Treg_suppressive
    c("Treg")                       # Treg cytokine panel
  ),
  pretty       = c("CD8 exhausted", "CD8 cytotoxic", "CD8 naive/memory",
                   "CAF inflammatory", "CAF myofibroblast", "CAF stem/niche",
                   "Tumor epithelial",
                   "Treg suppressive (atlas)", "Treg IL10/TGFB/IL35 (curated)")
)

# Keep only programs that actually exist in metadata
ucols <- colnames(srt@meta.data)
program_specs <- program_specs %>% filter(program %in% ucols)
if (nrow(program_specs) == 0) stop("None of the specified program columns exist in srt@meta.data.")

# ------------------------------------------------------------------------------
# Helper: one program → plot + Wilcoxon + summary row
# ------------------------------------------------------------------------------
format_p <- function(p){
  if (is.na(p)) return("p = NA")
  if (p < 1e-4) sprintf("p = %.2e", p) else sprintf("p = %.3f", p)
}

plot_program_by_mets <- function(program, class_filter, pretty_name, outdir){
  df <- srt@meta.data %>%
    tibble::as_tibble(rownames = "cell_id") %>%
    dplyr::filter(.data$cell_class %in% class_filter) %>%
    dplyr::select(cell_id, mets, score = dplyr::all_of(program)) %>%
    tidyr::drop_na(mets, score)

  df <- df %>%
    mutate(mets = as.integer(mets),
           mets_fac = factor(mets, levels = c(0,1), labels = c("Non-mets","Mets")))

  n0 <- sum(df$mets == 0, na.rm = TRUE)
  n1 <- sum(df$mets == 1, na.rm = TRUE)
  message(sprintf("[BALANCE] %-30s | class=%-22s | n0=%d  n1=%d  n=%d",
                  program, paste(class_filter, collapse=","),
                  n0, n1, nrow(df)))

  if (length(unique(df$mets)) < 2 || nrow(df) < 2) {
    return(tibble(
      prog_id      = program,
      prog_pretty  = pretty_name,
      n_cells      = nrow(df),
      n_mets0      = n0,
      n_mets1      = n1,
      median0      = NA_real_,
      median1      = NA_real_,
      p            = NA_real_,
      plot_file    = NA_character_
    ))
  }

  wt   <- suppressWarnings(stats::wilcox.test(score ~ mets, data = df, exact = FALSE))
  pval <- wt$p.value
  med0 <- stats::median(df$score[df$mets == 0], na.rm = TRUE)
  med1 <- stats::median(df$score[df$mets == 1], na.rm = TRUE)
  p_lab <- format_p(pval)

  y_min <- min(df$score, na.rm = TRUE)
  y_max <- max(df$score, na.rm = TRUE)
  pad   <- 0.08 * (y_max - y_min + 1e-8)
  y_lab <- y_max + pad

  p <- ggplot(df, aes(x = mets_fac, y = score, fill = mets_fac)) +
    geom_violin(trim = TRUE, alpha = 0.85) +
    geom_boxplot(width = 0.18, outlier.shape = NA, color = "black") +
    scale_fill_manual(values = c("Non-mets" = "#E69F00", "Mets" = "#56B4E9"), name = "Group") +
    annotate("text", x = 1.5, y = y_lab, label = p_lab, size = 4.2, fontface = "bold") +
    expand_limits(y = y_lab) +
    labs(
      title = paste0(pretty_name, " (cell-level)"),
      x = NULL,
      y = paste0(pretty_name, " score")
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold"),
          plot.margin = unit(c(10, 12, 10, 10), "pt")) +
    coord_cartesian(clip = "off")

  # --- Save as PDF instead of PNG ---
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
  file_out <- file.path(outdir, paste0("violin_celllevel_", make.names(pretty_name), ".pdf"))
  ggsave(file_out, p, width = 5.6, height = 4.0, device = cairo_pdf)

  tibble(
    prog_id      = program,
    prog_pretty  = pretty_name,
    n_cells      = nrow(df),
    n_mets0      = n0,
    n_mets1      = n1,
    median0      = med0,
    median1      = med1,
    p            = pval,
    plot_file    = file_out
  )
}
# ------------------------------------------------------------------------------
# Run across programs
# ------------------------------------------------------------------------------
outdir <- "analysis_outputs/celllevel_program_programs"
res_list <- purrr::pmap(
  list(program_specs$program, program_specs$class_filter, program_specs$pretty),
  plot_program_by_mets,
  outdir = outdir
)

summary_tbl <- if (utils::packageVersion("purrr") >= "1.0.0") {
  purrr::list_rbind(res_list)
} else {
  dplyr::bind_rows(res_list)
} %>%
  dplyr::arrange(prog_pretty)

readr::write_csv(summary_tbl, file.path(outdir, "celllevel_wilcoxon_summary.csv"))

print(
  summary_tbl %>%
    transmute(Program = prog_pretty,
              `n cells` = n_cells,
              `n non-mets` = n_mets0,
              `n mets` = n_mets1,
              median_non_mets = signif(median0, 3),
              median_mets     = signif(median1, 3),
              p = signif(p, 3),
              plot = plot_file)
)
```


CellChat: Making the Cell chat objects
```{r}
###############################################################################
# PURPOSE
# -------
# Run CellChat v2 on Non-mets vs Mets cohorts from Seurat object `srt`.
# - Splits Seurat object by srt$mets (0 = Non-mets, 1 = Mets)
# - Groups by cell_class (CD8+ T-cells, Treg, Fibroblasts, Tumor cells, etc.)
# - Builds CellChat objects for each group
# - Saves results to "analysis_outputs/cellchat_non" and "analysis_outputs/cellchat_mets"
#
# REQUIREMENTS
# ------------
# - CellChat v2 installed
# - Seurat object `srt` with:
#     srt$mets        : 0 = Non-mets, 1 = Mets
#     srt$cell_class  : broad cell type labels
###############################################################################

suppressPackageStartupMessages({
  library(CellChat)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(ggplot2)
  library(patchwork)
  library(future)
  library(presto)   # fast Wilcoxon for identifyOverExpressed*
  library(Seurat)
})

# ---------------------------------------------------------------------------
# Global options
# ---------------------------------------------------------------------------
#options(future.globals.maxSize = 8000 * 1024^2)  # allow large Seurat/CellChat objects
#future::plan("multisession", workers = 4) # do parallel
# ---------------------------------------------------------------------------
# Helper: prep CellChat inputs from Seurat
# ---------------------------------------------------------------------------
prep_cellchat_inputs_from_seurat <- function(srt, cells_keep, group_col = "cell_class", min.cells = 30) {
  # subset Seurat object
  srt_sub <- subset(srt, cells = Cells(srt)[cells_keep])
  expr    <- GetAssayData(srt_sub, slot = "data", assay = "RNA")   # genes x cells
  meta    <- srt_sub@meta.data %>%
    dplyr::select(all_of(group_col)) %>%
    dplyr::rename(labels = !!sym(group_col)) %>%
    tibble::rownames_to_column("cell_id")
  # filter small groups
  keep_groups <- names(which(table(meta$labels) >= min.cells))
  meta <- meta %>% filter(labels %in% keep_groups)
  expr <- expr[, meta$cell_id, drop = FALSE]
  list(data = expr, meta = meta)
}

# ---------------------------------------------------------------------------
# Helper: run CellChat end-to-end for one cohort
# ---------------------------------------------------------------------------
run_cellchat <- function(data.input, meta.input, project_name, outdir) {
  message("[CellChat] Running for ", project_name)
  cellchat <- createCellChat(object = data.input, meta = meta.input, group.by = "labels")
  CellChatDB <- CellChatDB.human
  cellchat@DB <- CellChatDB
  cellchat <- subsetData(cellchat)                  # subset signaling genes
  cellchat <- identifyOverExpressedGenes(cellchat) # requires presto
  cellchat <- identifyOverExpressedInteractions(cellchat)
  cellchat <- computeCommunProb(cellchat)
  cellchat <- filterCommunication(cellchat, min.cells = 30)
  cellchat <- computeCommunProbPathway(cellchat)
  cellchat <- aggregateNet(cellchat)
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
  saveRDS(cellchat, file.path(outdir, paste0("cellchat_", project_name, ".rds")))
  return(cellchat)
}

# ---------------------------------------------------------------------------
# Build inputs from Seurat (Option B, since first run)
# ---------------------------------------------------------------------------
stopifnot(exists("srt"))
out_base <- "analysis_outputs"

nm_in <- prep_cellchat_inputs_from_seurat(
  srt, cells_keep = srt$mets == 0, group_col = "cell_class", min.cells = 30
)
m_in  <- prep_cellchat_inputs_from_seurat(
  srt, cells_keep = srt$mets == 1, group_col = "cell_class", min.cells = 30
)

# ---------------------------------------------------------------------------
# Run CellChat for Non-mets and Mets
# ---------------------------------------------------------------------------
cellchat_non <- run_cellchat(
  data.input   = nm_in$data,
  meta.input   = nm_in$meta,
  project_name = "NonMets",
  outdir       = file.path(out_base, "cellchat_non")
)

cellchat_mets <- run_cellchat(
  data.input   = m_in$data,
  meta.input   = m_in$meta,
  project_name = "Mets",
  outdir       = file.path(out_base, "cellchat_mets")
)

###############################################################################
# ✅ Outputs:
# - RDS files saved in analysis_outputs/cellchat_non and analysis_outputs/cellchat_mets
# - Each RDS can be reloaded and passed to CellChat plotting functions
###############################################################################
```

The following are for visualizing CellChat results:
Making Exhaustion groups
```{r}
###############################################################################
# PURPOSE
# Subset CellChat to exhaustion/suppressive pathways (based on your DB names),
# optionally restrict to Tumor / Fibroblast / CD8 / Treg, and export bar plots
# comparing NonMets vs Mets (interaction COUNT and WEIGHT).
#
# REQUIRES:
# - cellchat_non, cellchat_mets (already processed through computeCommunProb,
#   computeCommunProbPathway, aggregateNet, and netAnalysis_computeCentrality)
###############################################################################
suppressPackageStartupMessages({
  library(CellChat)
  library(dplyr)
  library(ggplot2)
})

stopifnot(inherits(cellchat_non, "CellChat"), inherits(cellchat_mets, "CellChat"))

outdir <- "analysis_outputs/cellchat_plots_exhaustion"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# 1) Exhaustion/suppressive pathway names present in YOUR DB
db_paths <- sort(unique(cellchat_non@DB$interaction$pathway_name))
exhaustion_paths_try <- c(
  # inhibitory checkpoints
  "PD-L1", "PDL2", "VISTA", "BTLA", "CD160", "LAIR1", "CD200", "SIRP",
  # TIGIT axis (via PVR/NECTIN ligands)
  "PVR", "NECTIN",
  # LAG3 axis proxies
  "MHC-II",
  # TIM-3 axis proxy (galectins)
  "GALECTIN",
  # immunosuppressive cytokines
  "TGFb", "IL10",
  # (optional co-stim to contextualize balance)
  "CD80", "CD86" # CTLA-4 competes with these; useful reference
)
exhaustion_paths <- intersect(exhaustion_paths_try, db_paths)

if (length(exhaustion_paths) == 0) {
  stop("No requested exhaustion/suppressive pathways found. Available examples:\n",
       paste(utils::head(db_paths, 40), collapse = ", "))
}
message("[CellChat] Using exhaustion/suppressive pathways: ",
        paste(exhaustion_paths, collapse = ", "))

# 2) (Optional) restrict to key sender/receiver groups
groups_keep     <- c("Tumor cells", "Fibroblasts", "CD8+ T-cells", "Treg")
restrict_groups <- TRUE

restrict_groups_fun <- function(cc, groups_keep, restrict = TRUE) {
  if (!restrict) return(cc)
  present <- intersect(levels(cc@idents), groups_keep)
  if (length(present) == 0)
    stop("None of groups_keep are present in this CellChat object. ",
         "Available: ", paste(levels(cc@idents), collapse = ", "))
  subsetCellChat(cc, idents.use = present)
}

# 3) Subset to exhaustion pathways and recompute centrality (v2 API)
subset_exhaustion <- function(cc, paths, groups_keep, restrict = TRUE) {
  cc2 <- restrict_groups_fun(cc, groups_keep, restrict)
  cc2 <- subsetData(cc2)
  # recompute centrality on pathway-level network (required for some plots)
  cc2 <- netAnalysis_computeCentrality(cc2, slot.name = "netP")
  cc2
}

cellchat_non_ex  <- subset_exhaustion(cellchat_non,  exhaustion_paths, groups_keep, restrict_groups)
cellchat_mets_ex <- subset_exhaustion(cellchat_mets, exhaustion_paths, groups_keep, restrict_groups)

# 4) Merge and compare (bar plots)
object.list      <- list(NonMets = cellchat_non_ex, Mets = cellchat_mets_ex)
cellchat_merged  <- mergeCellChat(object.list, add.names = names(object.list))

p_cnt <- compareInteractions(cellchat_merged, measure = "count")
p_wt  <- compareInteractions(cellchat_merged, measure = "weight")

ggsave(file.path(outdir, "compare_interactions_exhaustion_count.pdf"),
       plot = p_cnt, width = 6, height = 5, device = cairo_pdf)
ggsave(file.path(outdir, "compare_interactions_exhaustion_weight.pdf"),
       plot = p_wt,  width = 6, height = 5, device = cairo_pdf)

message("[CellChat] Saved exhaustion-focused bar plots → ", outdir)
```

Making Bar plots
```{r}
###############################################################################
# PURPOSE
# Make per–cell-type bar plots (NonMets vs Mets) for **exhaustion/suppressive**
# signaling only, using existing CellChat v2 objects:
#   - cellchat_non, cellchat_mets
#
# WHAT IT DOES
# 1) Selects exhaustion/suppressive pathways (PD-L1/PDL2, VISTA, BTLA, CD160,
#    LAIR1, CD200, SIRP, PVR/NECTIN (TIGIT axis), MHC-II (LAG3 proxy),
#    GALECTIN (TIM-3 proxy), IL10, TGFb).
# 2) Aggregates those pathway-specific adjacency matrices (cc@netP$prob)
#    into a single adjacency per dataset.
# 3) For each focal group (Tumor cells, Fibroblasts, CD8+ T-cells, Treg),
#    computes:
#       - COUNT = number of edges touching the group (incoming + outgoing)
#       - WEIGHT = sum of edge weights touching the group
# 4) Exports one PDF per group for COUNT and WEIGHT.
###############################################################################
suppressPackageStartupMessages({
  library(CellChat)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(stringr)
})

# --- Sanity: need v2 objects with pathway-level networks computed ---
stopifnot(inherits(cellchat_non, "CellChat"),
          inherits(cellchat_mets, "CellChat"))
# These must already have been run earlier on each object:
# computeCommunProb(); computeCommunProbPathway(); aggregateNet(); netAnalysis_computeCentrality()

outdir <- "analysis_outputs/cellchat_plots_exhaustion_bygroup"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# 1) Define the exhaustion/suppressive pathway names present in YOUR DB
db_paths <- sort(unique(cellchat_non@DB$interaction$pathway_name))
exhaustion_paths_try <- c(
  # inhibitory checkpoints
  "PD-L1","PDL2","VISTA","BTLA","CD160","LAIR1","CD200","SIRP",
  # TIGIT axis (via PVR/NECTIN ligands)
  "PVR","NECTIN",
  # LAG-3 axis proxy
  "MHC-II",
  # TIM-3 axis proxy
  "GALECTIN",
  # suppressive cytokines
  "TGFb","IL10"
)
exhaustion_paths <- intersect(exhaustion_paths_try, db_paths)
if (length(exhaustion_paths) == 0) {
  stop("No requested exhaustion/suppressive pathways found in this DB. Available examples:\n",
       paste(utils::head(db_paths, 40), collapse = ", "))
}
message("[CellChat] Using exhaustion/suppressive pathways: ",
        paste(exhaustion_paths, collapse = ", "))

# 2) Focal groups we want separate figures for
groups_focus <- c("Tumor cells","Fibroblasts","CD8+ T-cells","Treg")

# --- helpers ----------------------------------------------------------

# Sum pathway-level adjacency matrices (probability) into one aggregated matrix
sum_pathway_adj <- function(cc, pathways) {
  stopifnot(!is.null(cc@netP$prob))
  prob_arr <- cc@netP$prob  # 3D: senders x receivers x pathways
  pw_all   <- dimnames(prob_arr)[[3]]
  keep     <- intersect(pathways, pw_all)
  if (length(keep) == 0) {
    stop("Requested pathways are not in cc@netP$pathways. Available: ",
         paste(pw_all, collapse = ", "))
  }
  idx <- match(keep, pw_all)
  # sum over the 3rd dimension (selected pathways)
  apply(prob_arr[ , , idx, drop = FALSE], c(1, 2), sum)
}

# Compute group-level COUNT & WEIGHT from a (sender x receiver) adjacency
group_stats_from_adj <- function(adj, group, mode = c("both","incoming","outgoing")) {
  mode <- match.arg(mode)
  if (!group %in% rownames(adj) || !group %in% colnames(adj)) {
    stop("Group '", group, "' not found in adjacency.")
  }
  # ignore self-loops when counting
  diag(adj) <- 0
  out_vec <- adj[group, , drop = TRUE]   # group -> others
  in_vec  <- adj[ , group, drop = TRUE]  # others -> group

  if (mode == "both") {
    weight <- sum(out_vec, na.rm = TRUE) + sum(in_vec, na.rm = TRUE)
    count  <- sum(out_vec > 0, na.rm = TRUE) + sum(in_vec > 0, na.rm = TRUE)
  } else if (mode == "outgoing") {
    weight <- sum(out_vec, na.rm = TRUE)
    count  <- sum(out_vec > 0, na.rm = TRUE)
  } else { # incoming
    weight <- sum(in_vec, na.rm = TRUE)
    count  <- sum(in_vec > 0, na.rm = TRUE)
  }
  tibble::tibble(group = group, count = count, weight = weight)
}

# Build a tidy summary for one CellChat object
summarize_one_cc <- function(cc, label, pathways, mode = "both", groups_focus = NULL) {
  adj_sum <- sum_pathway_adj(cc, pathways)
  grps    <- if (is.null(groups_focus)) rownames(adj_sum) else intersect(groups_focus, rownames(adj_sum))
  if (length(grps) == 0) stop("No focus groups found in adjacency.")
  dplyr::bind_rows(lapply(grps, function(g) group_stats_from_adj(adj_sum, g, mode = mode))) |>
    dplyr::mutate(dataset = label, .before = 1)
}

# 3) Summaries for NonMets and Mets (mode = "both" by default)
sum_non <- summarize_one_cc(cellchat_non, "NonMets", exhaustion_paths, mode = "both")
sum_mets <- summarize_one_cc(cellchat_mets, "Mets", exhaustion_paths, mode = "both")
sum_all <- bind_rows(sum_non, sum_mets)

# 4) Plotting per group (NonMets vs Mets) for COUNT and WEIGHT
save_group_bar <- function(df, grp, measure = c("count","weight")) {
  measure <- match.arg(measure)
  dd <- df %>% filter(group == grp)
  if (nrow(dd) == 0) return(invisible(NULL))
  ylab <- if (measure == "count") "Number of inferred interactions"
          else "Interaction strength (probability sum)"
  p <- ggplot(dd, aes(x = dataset, y = .data[[measure]], fill = dataset)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = signif(.data[[measure]], 3)), vjust = -0.4, size = 3.5) +
    scale_fill_manual(values = c(NonMets = "#808080", Mets = "#cc0000")) +
    labs(title = paste0(grp, " — exhaustion/suppressive signaling"),
         x = NULL, y = ylab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold"))
  fn <- file.path(outdir, paste0("bar_", make.names(grp), "_", measure, ".pdf"))
  ggsave(fn, plot = p, width = 5.6, height = 4.0, device = cairo_pdf)
  message("Saved: ", fn)
}

invisible(lapply(groups_focus, function(g) save_group_bar(sum_all, g, "count")))
invisible(lapply(groups_focus, function(g) save_group_bar(sum_all, g, "weight")))

message("[CellChat] Saved per–cell-type exhaustion bar plots → ", outdir)
```

Making Scatter plots
```{r}
###############################################################################
# PURPOSE
# Plot signaling role scatter (sending vs receiving) restricted to exhaustion/
# suppressive pathways, separately for NonMets and Mets.
#
# REQUIRES:
# - cellchat_non_ex, cellchat_mets_ex (subsetted to exhaustion pathways)
# - netAnalysis_computeCentrality already run on them
###############################################################################
suppressPackageStartupMessages({
  library(CellChat)
  library(ggplot2)
  library(patchwork)
})

stopifnot(inherits(cellchat_non_ex, "CellChat"), inherits(cellchat_mets_ex, "CellChat"))

outdir <- "analysis_outputs/cellchat_plots_exhaustion"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# Compute centrality if not yet done
cellchat_non_ex  <- netAnalysis_computeCentrality(cellchat_non_ex,  slot.name = "netP")
cellchat_mets_ex <- netAnalysis_computeCentrality(cellchat_mets_ex, slot.name = "netP")

# Plot sending vs receiving roles (scatter) for exhaustion-only networks
p_non <- netAnalysis_signalingRole_scatter(cellchat_non_ex, title = "NonMets (Exhaustion)")
p_mets <- netAnalysis_signalingRole_scatter(cellchat_mets_ex, title = "Mets (Exhaustion)")

# Combine side by side
p_comb <- p_non + p_mets

# Save to PDF
pdf(file.path(outdir, "signalingRole_scatter_exhaustion.pdf"), width = 11, height = 5, useDingbats = FALSE)
print(p_comb)
dev.off()

message("[CellChat] Saved exhaustion-specific signaling role scatter → ", outdir)
```


```{r}
###############################################################################
# PURPOSE
# 1) Diagnose zeros in exhaustion circles for focus groups (Tumor/Fibro/CD8/Treg)
#    by printing per-pathway, per edge (sender→receiver) sums from netP.
# 2) If they’re truly all ~0, recompute a *rescue* CellChat object that is
#    restricted to exhaustion pathways and uses slightly more permissive options,
#    then plot circles again for the focus groups.
#
# INPUTS (already processed with CellChat v2 pipeline):
#   - cellchat_non, cellchat_mets (have @netP$prob etc.)
#
# OUTPUTS:
#   analysis_outputs/cellchat_plots_exhaustion/*
###############################################################################
suppressPackageStartupMessages({
  library(CellChat)
  library(dplyr)
  library(tidyr)
})

stopifnot(inherits(cellchat_non, "CellChat"), inherits(cellchat_mets, "CellChat"))
outdir <- "analysis_outputs/cellchat_plots_exhaustion"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# Exhaustion / suppressive pathways to test (must match your DB)
exhaustion_paths <- c(
  "PD-L1","PDL2","VISTA","BTLA","CD160","LAIR1","CD200","SIRP",
  "PVR","NECTIN",          # TIGIT axis
  "MHC-II","GALECTIN",     # LAG3 / TIM-3 proxies
  "TGFb","IL10"            # suppressive cytokines
)
focus_groups <- c("Tumor cells","Fibroblasts","CD8+ T-cells","Treg")
drop_self_loops <- TRUE

# ---------- helpers ----------
present_paths_netP <- function(cc) {
  if (is.null(cc@netP) || is.null(cc@netP$prob) || length(dim(cc@netP$prob)) < 3) character(0)
  else dimnames(cc@netP$prob)[[3]]
}

# Per-pathway edge matrix for selected groups (sums of netP probability)
extract_pathway_matrix <- function(cc, pathway, groups_keep = NULL, drop_diag = TRUE) {
  pw_names <- present_paths_netP(cc)
  if (!(pathway %in% pw_names)) return(NULL)
  M <- cc@netP$prob[, , pathway, drop = FALSE][,,1]
  if (!is.null(groups_keep)) {
    g <- intersect(rownames(M), groups_keep)
    if (length(g) >= 2) {
      M <- M[g, g, drop = FALSE]
    } else {
      return(matrix(0, 0, 0))
    }
  }
  if (drop_diag && length(diag(M))) diag(M) <- 0
  M
}

# Print top non-zero edges per pathway for the focus groups
diagnose_edges <- function(cc, label, paths, groups_keep) {
  cat(sprintf("\n[Diag] %s — per-pathway sender→receiver sums within focus groups\n", label))
  avail <- intersect(paths, present_paths_netP(cc))
  if (length(avail) == 0) {
    cat("  (no requested pathways present in netP)\n")
    return(invisible(NULL))
  }
  for (pw in avail) {
    M <- extract_pathway_matrix(cc, pw, groups_keep = groups_keep, drop_diag = TRUE)
    if (is.null(M) || length(M) == 0) next
    nz <- which(M > 0, arr.ind = TRUE)
    if (nrow(nz) == 0) {
      cat(sprintf("  %s: all zero within focus groups\n", pw))
    } else {
      df <- data.frame(
        sender = rownames(M)[nz[,1]],
        receiver = colnames(M)[nz[,2]],
        prob = M[nz]
      ) |>
        arrange(desc(prob))
      cat(sprintf("  %s: top edges (n=%d)\n", pw, nrow(df)))
      print(head(df, 12), row.names = FALSE)
    }
  }
}

# Build exhaustion-only adjacency from netP for plotting
make_adj_from_netP <- function(cc, keep_paths, groups_keep = NULL, drop_diag = TRUE) {
  pw_names <- present_paths_netP(cc)
  keep_pw  <- intersect(keep_paths, pw_names)
  rn <- dimnames(cc@netP$prob)[[1]]; cn <- dimnames(cc@netP$prob)[[2]]
  zeros <- list(
    count  = matrix(0, length(rn), length(cn), dimnames = list(rn, cn)),
    weight = matrix(0, length(rn), length(cn), dimnames = list(rn, cn))
  )
  if (length(keep_pw) == 0) return(zeros)
  arr <- cc@netP$prob[, , keep_pw, drop = FALSE]
  mat_count  <- apply(arr > 0, c(1, 2), sum)
  mat_weight <- apply(arr,    c(1, 2), sum)
  if (!is.null(groups_keep)) {
    g <- intersect(rownames(mat_count), groups_keep)
    if (length(g) >= 2) {
      mat_count  <- mat_count [g, g, drop = FALSE]
      mat_weight <- mat_weight[g, g, drop = FALSE]
    } else {
      mat_count[,]  <- 0; mat_weight[,] <- 0
    }
  }
  if (drop_diag) { diag(mat_count) <- 0; diag(mat_weight) <- 0 }
  list(count = mat_count, weight = mat_weight)
}

plot_circle <- function(mat, title, file, weight.scale = TRUE, edge.weight.max = NULL) {
  if (!is.matrix(mat) || all(mat == 0, na.rm = TRUE)) {
    message(sprintf("[WARN] %s has no non-zero edges — skipping plot.", title)); return(invisible(NULL))
  }
  vertex_wt <- rowSums(mat) + colSums(mat)
  if (all(vertex_wt == 0)) vertex_wt[] <- 1
  pdf(file, width = 6, height = 6, useDingbats = FALSE)
  netVisual_circle(
    mat,
    vertex.weight   = vertex_wt,
    weight.scale    = weight.scale,
    edge.weight.max = edge.weight.max,
    label.edge      = FALSE,
    title.name      = title
  )
  dev.off()
}

# ---------- (1) DIAGNOSE ----------
diag_non_paths  <- intersect(exhaustion_paths, present_paths_netP(cellchat_non))
diag_mets_paths <- intersect(exhaustion_paths, present_paths_netP(cellchat_mets))
cat("[Diag] NonMets has exhaustion pathways in netP:", paste(diag_non_paths, collapse = ", "), "\n")
cat("[Diag] Mets    has exhaustion pathways in netP:", paste(diag_mets_paths, collapse = ", "), "\n")

diagnose_edges(cellchat_non,  "NonMets", exhaustion_paths, focus_groups)
diagnose_edges(cellchat_mets, "Mets",    exhaustion_paths, focus_groups)

# Try to plot (focus groups) with current netP
adj_non_focus  <- make_adj_from_netP(cellchat_non,  exhaustion_paths, groups_keep = focus_groups, drop_diag = drop_self_loops)
adj_mets_focus <- make_adj_from_netP(cellchat_mets, exhaustion_paths, groups_keep = focus_groups, drop_diag = drop_self_loops)

edge_max_count  <- max(c(adj_non_focus$count,  adj_mets_focus$count),  na.rm = TRUE)
edge_max_weight <- max(c(adj_non_focus$weight, adj_mets_focus$weight), na.rm = TRUE)
if (!is.finite(edge_max_count))  edge_max_count  <- NULL
if (!is.finite(edge_max_weight)) edge_max_weight <- NULL

plot_circle(adj_non_focus$count,  "NonMets — Exhaustion (COUNT, Tumor/Fibro/CD8/Treg)",
            file.path(outdir, "circle_exhaustion_count_NonMets_focus.pdf"),
            weight.scale = FALSE, edge.weight.max = edge_max_count)
plot_circle(adj_mets_focus$count, "Mets — Exhaustion (COUNT, Tumor/Fibro/CD8/Treg)",
            file.path(outdir, "circle_exhaustion_count_Mets_focus.pdf"),
            weight.scale = FALSE, edge.weight.max = edge_max_count)

plot_circle(adj_non_focus$weight,  "NonMets — Exhaustion (WEIGHT, Tumor/Fibro/CD8/Treg)",
            file.path(outdir, "circle_exhaustion_weight_NonMets_focus.pdf"),
            weight.scale = TRUE, edge.weight.max = edge_max_weight)
plot_circle(adj_mets_focus$weight, "Mets — Exhaustion (WEIGHT, Tumor/Fibro/CD8/Treg)",
            file.path(outdir, "circle_exhaustion_weight_Mets_focus.pdf"),
            weight.scale = TRUE, edge.weight.max = edge_max_weight)

# ---------- (2) OPTIONAL RESCUE: recompute with only exhaustion DB -----------
# If the printed diagnostics show zeros across the board for the focus groups,
# run the block below (uncomment) to build “rescue” objects restricted to
# exhaustion LR and slightly relaxed parameters.

# ---- helper to restrict DB to specific pathways and recompute netP ----------
recompute_exhaustion_only <- function(cc, keep_paths,
                                      nboot = 100, trim = 0.1, Kh = 0.5,
                                      raw.use = TRUE, seed = 1) {
  cc2 <- cc
  # keep only interactions in keep_paths
  keep_int <- which(cc2@DB$interaction$pathway_name %in% keep_paths)
  cc2@DB$interaction <- cc2@DB$interaction[keep_int, , drop = FALSE]
  cc2@LR$LRsig       <- cc2@LR$LRsig      [cc2@LR$LRsig$pathway_name %in% keep_paths, , drop = FALSE]

  # refresh signaling genes and expression subset
  cc2 <- subsetData(cc2)

  # more permissive knobs
  cc2@options$parameter$raw.use <- raw.use
  cc2@options$parameter$nboot   <- nboot
  cc2@options$parameter$trim    <- trim
  cc2@options$parameter$Kh      <- Kh
  cc2@options$parameter$seed.use <- seed

  # recompute
  cc2 <- computeCommunProb(cc2)
  cc2 <- computeCommunProbPathway(cc2)
  cc2 <- aggregateNet(cc2)
  cc2 <- netAnalysis_computeCentrality(cc2, slot.name = "netP")
  cc2
}
message("[CellChat] Diagnostics complete. See console for per-pathway edges within focus groups.")
```
Making circle plots
```{r}
###############################################################################
# PURPOSE
# Plot clean, interpretable circle plots for a *subset* of cell types
# (e.g., Tumor/Fibro/CD8/Treg) showing either:
#   - number of interactions (COUNT), or
#   - aggregate interaction strength (WEIGHT),
# for each cohort (NonMets vs Mets), using the full (all-pathway) networks.
#
# WHY
# Exhaustion-only circles were ~0. Here we visualize the overall
# communication “traffic” but restrict to a few key players to avoid clutter.
#
# INPUTS: cellchat_non, cellchat_mets (already processed through pipeline)
# OUTPUTS: analysis_outputs/cellchat_plots_subset/*.pdf
###############################################################################
suppressPackageStartupMessages({
  library(CellChat)
  library(dplyr)
})

stopifnot(inherits(cellchat_non, "CellChat"), inherits(cellchat_mets, "CellChat"))

outdir <- "analysis_outputs/cellchat_plots_subset"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# ---- Choose the small set of cell types to display (edit as needed) ----------
focus_groups <- c("Tumor cells","Fibroblasts","CD8+ T-cells","Treg")
drop_self_loops <- TRUE   # set FALSE if you want autocrine edges visible

# ---- Helper to subset a CellChat adjacency (count or weight) to chosen groups -
subset_net_matrix <- function(cc, measure = c("count","weight"),
                              groups_keep = NULL, drop_diag = TRUE) {
  measure <- match.arg(measure)
  mat <- switch(measure,
                count  = cc@net$count,
                weight = cc@net$weight)
  # keep only existing groups in this object
  keep <- if (is.null(groups_keep)) rownames(mat) else intersect(groups_keep, rownames(mat))
  if (length(keep) < 2) {
    # return an all-zero matrix with intended dimnames so plotting can gracefully skip
    Z <- matrix(0, length(groups_keep), length(groups_keep),
                dimnames = list(groups_keep, groups_keep))
    return(Z)
  }
  M <- mat[keep, keep, drop = FALSE]
  if (drop_diag && length(diag(M))) diag(M) <- 0
  M
}

# ---- Build objects list for compact looping ----------------------------------
object.list <- list(NonMets = cellchat_non, Mets = cellchat_mets)

# ---- Extract focused matrices for both cohorts --------------------------------
mats_count  <- lapply(object.list, subset_net_matrix, measure = "count",
                      groups_keep = focus_groups, drop_diag = drop_self_loops)
mats_weight <- lapply(object.list, subset_net_matrix, measure = "weight",
                      groups_keep = focus_groups, drop_diag = drop_self_loops)

# ---- Shared edge scales across cohorts (fair visual comparison) ---------------
edge_max_count  <- max(unlist(mats_count),  na.rm = TRUE);  if (!is.finite(edge_max_count))  edge_max_count  <- NULL
edge_max_weight <- max(unlist(mats_weight), na.rm = TRUE);  if (!is.finite(edge_max_weight)) edge_max_weight <- NULL

# ---- Plot saver around netVisual_circle (works with plain matrices) ----------
plot_circle_mat <- function(mat, title, file, weight.scale = TRUE, edge.weight.max = NULL) {
  if (!is.matrix(mat) || all(mat == 0, na.rm = TRUE)) {
    message(sprintf("[WARN] %s has no non-zero edges — skipping plot.", title)); return(invisible(NULL))
  }
  vertex_wt <- rowSums(mat) + colSums(mat)
  if (all(vertex_wt == 0)) vertex_wt[] <- 1
  pdf(file, width = 6, height = 6, useDingbats = FALSE)
  netVisual_circle(
    mat,
    vertex.weight   = vertex_wt,
    weight.scale    = weight.scale,     # TRUE = use edge width scaling for weights
    edge.weight.max = edge.weight.max,  # keeps scale consistent across cohorts
    label.edge      = FALSE,
    title.name      = title
  )
  dev.off()
}

# ---- Plot COUNT circles (number of interactions) ------------------------------
for (nm in names(mats_count)) {
  plot_circle_mat(
    mats_count[[nm]],
    paste0(nm, " — Number of interactions (", paste(focus_groups, collapse="/"), ")"),
    file.path(outdir, paste0("circle_COUNT_", nm, "_focus.pdf")),
    weight.scale = FALSE,
    edge.weight.max = edge_max_count
  )
}

# ---- Plot WEIGHT circles (interaction strength) -------------------------------
for (nm in names(mats_weight)) {
  plot_circle_mat(
    mats_weight[[nm]],
    paste0(nm, " — Interaction strength (", paste(focus_groups, collapse="/"), ")"),
    file.path(outdir, paste0("circle_WEIGHT_", nm, "_focus.pdf")),
    weight.scale = TRUE,
    edge.weight.max = edge_max_weight
  )
}

message("[CellChat] Subset circle plots written → ", outdir)

# -------------------------------------------------------------------------------
# OPTIONAL: show only *selected* pairs (e.g., CD8→Tumor & Treg→CD8) as circles.
# Set all other senders/receivers to 0, leaving labels intact.
# -------------------------------------------------------------------------------
# example_pairs <- list(
#   c("CD8+ T-cells","Tumor cells"),
#   c("Treg","CD8+ T-cells")
# )
# mask_to_pairs <- function(M, pairs) {
#   keep <- matrix(FALSE, nrow(M), ncol(M), dimnames = dimnames(M))
#   for (pr in pairs) {
#     s <- pr[1]; r <- pr[2]
#     if (s %in% rownames(M) && r %in% colnames(M)) keep[s, r] <- TRUE
#   }
#   M[!keep] <- 0
#   M
# }
# for (nm in names(mats_weight)) {
#   Msel <- mask_to_pairs(mats_weight[[nm]], example_pairs)
#   plot_circle_mat(
#     Msel,
#     paste0(nm, " — Strength (selected pairs)"),
#     file.path(outdir, paste0("circle_WEIGHT_", nm, "_selected_pairs.pdf")),
#     weight.scale = TRUE,
#     edge.weight.max = edge_max_weight
#   )
# }
```

Circle plots for collagen and ADGRG
```{r}
###############################################################################
# PURPOSE
# Export circle/aggregate plots for specific exhaustion-relevant pathways
# (e.g., COLLAGEN and ADGRG) from your restricted CellChat objects:
#   - cellchat_non_ex
#   - cellchat_mets_ex
#
# NOTES (CellChat v2):
# - Do NOT pass `title.name` to netVisual_aggregate(); v2 sets titles internally.
# - If you want a shared edge scale across cohorts, set `use_shared_edge_scale <- TRUE`.
###############################################################################
suppressPackageStartupMessages({ library(CellChat) })

stopifnot(inherits(cellchat_non_ex, "CellChat"), inherits(cellchat_mets_ex, "CellChat"))

outdir <- "analysis_outputs/cellchat_plots_exhaustion_pathways"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# Pathways to plot explicitly
pathways_interest <- c("COLLAGEN", "ADGRG")

# If you want both cohorts to use the same max edge weight in the circle plots
use_shared_edge_scale <- FALSE  # set TRUE if desired

# Compute a shared edge cap (optional)
edge_cap <- NULL
if (use_shared_edge_scale) {
  # Pull pathway-level 3D arrays and compute a rough global max over selected pathways
  get_pw_max <- function(cc, pws) {
    if (is.null(cc@netP) || is.null(cc@netP$prob)) return(0)
    keep <- intersect(pws, dimnames(cc@netP$prob)[[3]])
    if (length(keep) == 0) return(0)
    max(cc@netP$prob[, , keep, drop = FALSE], na.rm = TRUE)
  }
  edge_cap <- max(get_pw_max(cellchat_non_ex, pathways_interest),
                  get_pw_max(cellchat_mets_ex, pathways_interest), na.rm = TRUE)
  if (!is.finite(edge_cap) || edge_cap <= 0) edge_cap <- NULL
}

# Helper: plot one pathway for one CellChat object
plot_pathway_pdf <- function(cc, pathway, label, edge.weight.max = NULL) {
  if (is.null(cc@netP) || is.null(cc@netP$prob)) {
    message(sprintf("[SKIP] %s: netP not available for %s", pathway, label)); return(invisible(NULL))
  }
  if (!(pathway %in% cc@netP$pathways)) {
    message(sprintf("[SKIP] %s not in pathways for %s", pathway, label)); return(invisible(NULL))
  }

  file_out <- file.path(outdir, paste0("circle_", pathway, "_", label, ".pdf"))
  pdf(file_out, width = 6, height = 6, useDingbats = FALSE)
  # IMPORTANT: no `title.name` here (CellChat v2 handles titles internally)
  netVisual_aggregate(
    cc,
    signaling       = pathway,
    layout          = "circle",
    vertex.label.cex = 0.9,
    label.edge      = TRUE,
    edge.width.max  = 12,
    edge.weight.max = edge.weight.max
  )
  dev.off()
  message("[OK] Wrote: ", file_out)
}

# Run for each cohort (NonMets vs Mets) on each pathway
for (pw in pathways_interest) {
  plot_pathway_pdf(cellchat_non_ex,  pw, "NonMets_ex", edge.weight.max = edge_cap)
  plot_pathway_pdf(cellchat_mets_ex, pw, "Mets_ex",    edge.weight.max = edge_cap)
}
```

Making Heatmaps
```{r}
suppressPackageStartupMessages({
  library(CellChat)
  library(ComplexHeatmap)
  library(grid)   # for unit()
})

outdir <- "analysis_outputs/cellchat_plots_exhaustion"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# Helper: save heatmap with padding + transparent background
save_heatmap_pdf <- function(cellchat_obj, file, measure = "weight",
                             width = 7, height = 6,
                             padding_mm = c(25, 25, 25, 25)) {
  pdf(file, width = width, height = height, useDingbats = FALSE, bg = "transparent")
  ht <- netVisual_heatmap(cellchat_obj, measure = measure)
  draw(ht, padding = unit(padding_mm, "mm"))  # add whitespace around the heatmap
  dev.off()
  message("[OK] Wrote heatmap: ", file)
}

# Mets
save_heatmap_pdf(cellchat_mets_ex,
                 file = file.path(outdir, "Heatmap_interactions_exhaustion_mets.pdf"),
                 measure = "weight",
                 width = 7, height = 6,
                 padding_mm = c(30, 30, 30, 30))  # increase these for more white space

# Non-mets
save_heatmap_pdf(cellchat_non_ex,
                 file = file.path(outdir, "Heatmap_interactions_exhaustion_nonmets.pdf"),
                 measure = "weight",
                 width = 7, height = 6,
                 padding_mm = c(30, 30, 30, 30))
```

Differential expression analysis
```{r}
###############################################################################
# PURPOSE
# One-model (NOT per-patient meta):
# Test gene ~ exhaustion (UCell) with MAST (hurdle GLM), blocking by patient
# and adjusting for QC covariates. Extract the continuous-part beta for
# "exhaustion" and combine with lrTest p-values.
###############################################################################
suppressPackageStartupMessages({
  library(Seurat)
  library(MAST)
  library(dplyr)
  library(tibble)
  library(readr)
})

stopifnot(exists("srt"))
stopifnot(all(c("patient_id","cell_class") %in% colnames(srt@meta.data)))

# -------------------- knobs --------------------
cell_classes_focus <- c("CD8+ T-cells","T cells")
EXH_COL <- if ("CD8_exhausted_UCell" %in% colnames(srt@meta.data)) "CD8_exhausted_UCell" else
           if (all(c("Tex_up_UCell","Tex_down_UCell") %in% colnames(srt@meta.data))) {
             srt$CD8_exhausted_UCell <- srt$Tex_up_UCell - srt$Tex_down_UCell; "CD8_exhausted_UCell"
           } else stop("No exhaustion score found.")
USE_RANDOM_EFFECT <- FALSE
min_detect_frac   <- 0.05
min_detect_abs    <- 50

# -------------------- data prep --------------------
cells_focus <- WhichCells(srt, expression = cell_class %in% cell_classes_focus)
if (length(cells_focus) < 100) stop("Too few focus cells for analysis.")
DefaultAssay(srt) <- "RNA"

expr <- Seurat::GetAssayData(srt, slot = "data")[, cells_focus, drop = FALSE]
meta <- srt@meta.data[cells_focus, , drop = FALSE] %>%
  as_tibble(rownames = "cell") %>%
  transmute(
    cell,
    patient    = factor(patient_id),
    exhaustion = .data[[EXH_COL]],
    nCount_RNA, percent.mt, S.Score, G2M.Score
  ) %>%
  tidyr::drop_na(patient, exhaustion, nCount_RNA, percent.mt, S.Score, G2M.Score)

expr <- expr[, meta$cell, drop = FALSE]

detect_counts <- Matrix::rowSums(expr > 0)
min_detect <- max(min_detect_abs, ceiling(min_detect_frac * ncol(expr)))
keep_genes <- names(which(detect_counts >= min_detect))
expr <- expr[keep_genes, , drop = FALSE]

sca <- MAST::FromMatrix(
  exprsArray = as.matrix(expr),
  cData      = data.frame(row.names = meta$cell, meta)
)
colData(sca)$cngeneson <- scale(colSums(assay(sca) > 0))
colData(sca)$nCount_RNA <- scale(colData(sca)$nCount_RNA)
colData(sca)$percent.mt <- scale(colData(sca)$percent.mt)
colData(sca)$S.Score    <- scale(colData(sca)$S.Score)
colData(sca)$G2M.Score  <- scale(colData(sca)$G2M.Score)

# -------------------- fit model --------------------
if (USE_RANDOM_EFFECT) {
  message("[MAST] Fitting with random intercept for patient …")
  fml <- ~ exhaustion + cngeneson + nCount_RNA + percent.mt + S.Score + G2M.Score + (1|patient)
  fit <- zlm(formula = fml, sca = sca, method = "glmer", ebayes = FALSE,
             fitArgsD = list(nAGQ = 0))
  lr  <- lrTest(fit, "exhaustion")
} else {
  message("[MAST] Fitting with patient as fixed blocking factor …")
  contrasts(colData(sca)$patient) <- contr.sum(nlevels(colData(sca)$patient))
  fml <- ~ patient + exhaustion + cngeneson + nCount_RNA + percent.mt + S.Score + G2M.Score
  fit <- zlm(formula = fml, sca = sca, method = "bayesglm", ebayes = TRUE)
  lr  <- lrTest(fit, "exhaustion")
}

# -------------------- extract results (correct orientation) -------------------
# coef(fit, "C") returns a matrix with ROWS = genes and COLS = predictors.
coefC <- suppressWarnings(coef(fit, "C"))  # genes x predictors

# find the exhaustion COLUMN robustly
find_exh_col <- function(nms) {
  if (is.null(nms)) return(NA_integer_)
  hit <- which(nms == "exhaustion")
  if (!length(hit)) hit <- grep("^`?\\s*exhaustion\\s*`?$", nms)
  if (!length(hit)) hit <- grep("exhaustion", nms, ignore.case = TRUE)
  if (length(hit)) hit[1] else NA_integer_
}
exh_col_idx <- find_exh_col(colnames(coefC))
if (is.na(exh_col_idx)) {
  stop(
    "Couldn't find an 'exhaustion' coefficient in the continuous part (columns).\n",
    "Example predictor colnames: ", paste(head(colnames(coefC), 16), collapse = ", ")
  )
}

# named vector (by gene) of exhaustion betas from the continuous part
beta_vec <- coefC[, exh_col_idx]
names(beta_vec) <- rownames(coefC)

# lrTest gives the combined hurdle p-value per gene
lr_df <- as.data.frame(lr)
lr_df$gene <- rownames(lr_df)

# align and finalize
p_col <- intersect(c("hurdle.Pr(>Chisq)", "Pr(>Chisq)"), colnames(lr_df))
if (length(p_col) == 0) stop("Couldn't find a p-value column in lrTest output.")
p_col <- p_col[1]

res <- tibble(
  gene        = lr_df$gene,
  p_combined  = lr_df[[p_col]],     # <-- use hurdle.Pr(>Chisq)
  coef_cont   = beta_vec[lr_df$gene]
) %>%
  mutate(
    q_combined = p.adjust(p_combined, "BH"),
    direction  = dplyr::case_when(
      is.na(coef_cont) ~ "unknown",
      coef_cont > 0    ~ "up_with_exhaustion",
      coef_cont < 0    ~ "down_with_exhaustion",
      TRUE             ~ "flat"
    )
  ) %>%
  arrange(q_combined, dplyr::desc(abs(coef_cont)))

# -------------------- save --------------------
dir.create("program_DE_exhaustion_mast", showWarnings = FALSE)
write_csv(res, file.path("program_DE_exhaustion_mast", "mast_exhaustion_results_full.csv"))

top_up   <- res %>% filter(q_combined < 0.05, direction == "up_with_exhaustion")
top_down <- res %>% filter(q_combined < 0.05, direction == "down_with_exhaustion")
write_csv(top_up,   file.path("program_DE_exhaustion_mast", "mast_exhaustion_top_up.csv"))
write_csv(top_down, file.path("program_DE_exhaustion_mast", "mast_exhaustion_top_down.csv"))

message(sprintf("[MAST] Genes tested: %d | q<0.05 up=%d down=%d → program_DE_exhaustion_mast/",
                nrow(res), nrow(top_up), nrow(top_down)))
```

UMAP
```{r}
## ------------------------------------------------------------
## Panel B — Global UMAP (labels on) + separate legend PDF
## ------------------------------------------------------------
suppressPackageStartupMessages({
  library(Seurat); library(ggplot2); library(dplyr); library(ggrepel); library(cowplot)
})

stopifnot(exists("srt"), "umap" %in% names(srt@reductions), "cell_class" %in% colnames(srt@meta.data))

# --- palette (same as before) ---
cell_levels <- sort(unique(srt$cell_class))
pal <- c(
  "CD8+ T-cells"="dodgerblue3","T cells"="steelblue3","Treg"="mediumorchid3",
  "CD4+ T-cells"="royalblue1","TFH cells"="lightskyblue3",
  "NK cells"="slateblue4","Monocytes"="darkorange2","Macrophages"="chocolate3",
  "Dendritic cells"="tan3","Plasma cells"="goldenrod2","B cells"="gold3",
  "Fibroblasts"="seagreen3","Endothelial cells"="darkseagreen","Pericytes"="olivedrab3",
  "Tumor cells"="firebrick2","Goblet cells"="sienna3","Smooth muscle cells"="darkolivegreen",
  "Schwann cells"="gray40","Innate lymphoid cells"="darkturquoise","Granulocytes"="thistle4",
  "Enteroendocrine cells"="plum3","MAIT cells"="darkmagenta","iNKT cells"="mediumslateblue",
  "γδ T-cells"="royalblue4","T cells (other)"="grey35","Unmapped"="gray70"
)
pal_use <- pal[intersect(names(pal), cell_levels)]

# --- data.frame for plotting + centroids ---
emb <- Embeddings(srt, "umap") %>% as.data.frame()
emb$cell_class <- srt$cell_class
centroids <- emb %>%
  group_by(cell_class) %>%
  summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")

# Manually nudge a few crowded labels to reduce overlap (edit as needed)
nudge_tbl <- tibble::tribble(
  ~cell_class,        ~dx,   ~dy,
  "CD8+ T-cells",      1.2,  -0.8,
  "γδ T-cells",       -1.2,   0.6,
  "iNKT cells",        0.9,   0.6,
  "MAIT cells",       -0.8,  -0.6
)
centroids <- centroids %>%
  left_join(nudge_tbl, by = "cell_class") %>%
  mutate(
    dx = dplyr::coalesce(dx, 0),
    dy = dplyr::coalesce(dy, 0),
    label_x = UMAP_1 + dx,
    label_y = UMAP_2 + dy
  )

# base UMAP (points)
p_points <- ggplot(emb, aes(UMAP_1, UMAP_2, color = cell_class)) +
  geom_point(size = 0.2, alpha = 0.85) +
  theme_classic(base_size = 11) +
  ggtitle("Global cell classes") +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none",                       # <-- legend OFF in the main figure
    plot.margin = margin(15, 70, 15, 15)           # extra right margin for labels
  )
if (length(pal_use)) p_points <- p_points + scale_color_manual(values = pal_use)

# add non-overlapping labels (repel) at nudged centroid positions
set.seed(12348934)
p_classes_main <- p_points +
  geom_text_repel(
    data = centroids,
    aes(x = label_x, y = label_y, label = cell_class),  # no color mapping here
    inherit.aes = FALSE,
    size = 3.3, fontface = "bold",
    color = "black",               # <-- force labels to black
    box.padding = 0.35, point.padding = 0.6, max.overlaps = Inf,
    segment.size = 0.2, min.segment.length = 0
  ) +
  # ensure the canvas includes any nudged labels (prevents clipping)
  expand_limits(x = centroids$label_x, y = centroids$label_y) +
  coord_cartesian(clip = "off")

# ----- make a standalone legend figure -----
# Build a copy with legend on, then extract it
p_with_legend <- p_points + theme(legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1), ncol = 1))
leg_g <- cowplot::get_legend(p_with_legend)

legend_plot <- cowplot::ggdraw(leg_g) +
  theme(plot.margin = margin(10, 10, 10, 10))

# ----- save PDFs -----
dir.create("analysis_outputs/umap", showWarnings = FALSE, recursive = TRUE)

# main UMAP (labels ON, legend OFF)
ggsave("analysis_outputs/umap/PanelB1_umap_cell_classes.pdf",
       p_classes_main, width = 6.8, height = 5.8, device = cairo_pdf, limitsize = FALSE)

# legend-only PDF (for Illustrator)
ggsave("analysis_outputs/umap/PanelB1_umap_cell_classes_LEGEND.pdf",
       legend_plot, width = 3.2, height = 7.5, device = cairo_pdf, limitsize = FALSE)
```


