---
title: "Untitled"
author: "Alos Diallo"
date: "2024-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages required
```{r message=FALSE, warning=FALSE, include=FALSE}
library(Seurat)
library(futile.logger)
library(ggplot2)
library(stringr)
library(patchwork)
library(harmony)
library(readr)
library(viridis)
library(RCurl)
library(cowplot)
library(phateR)
library(compositions)
library(tidyverse)
library(dplyr)
library(hdf5r)
library(Matrix)
library(data.table)
library(ComplexHeatmap)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
library(Matrix)
library(reticulate)
#library(loomR)
library(anndata)
library(DESeq2)
library(viridis)
library(textshaping)
library(EnhancedVolcano)
library(DataExplorer)


```


```{r}


readPath.patient_11 <- 'F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/11-C9/count/sample_filtered_feature_bc_matrix/'

patient_11.Data <- Read10X(readPath.patient_11)
```




```{r}
patient_11 <- CreateSeuratObject(counts = patient_11.Data, project = "colon_work", min.cells = 3, min.features = 200)
patient_11[["percent.mt"]] <- PercentageFeatureSet(patient_11, pattern = "^MT-")
```

```{r}
# Quality control - adjust parameters based on your data and QC plots
patient_11 <- subset(patient_11, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)


# Visualize QC metrics as a violin plot
VlnPlot(patient_11, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

plot1 <- FeatureScatter(patient_11, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(patient_11, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

# Normalization
patient_11 <- NormalizeData(patient_11, normalization.method = "LogNormalize", scale.factor = 10000)

# Find variable features
patient_11 <- FindVariableFeatures(patient_11, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(patient_11), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(patient_11)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

# Scale the data
all.genes <- rownames(patient_11)
patient_11 <- ScaleData(patient_11, features = all.genes)
patient_11 <- ScaleData(patient_11, vars.to.regress = "percent.mt")

# Perform PCA
patient_11 <- RunPCA(patient_11,features = VariableFeatures(object = patient_11))
VizDimLoadings(patient_11, dims = 1:2, reduction = "pca")
DimPlot(patient_11, reduction = "pca") + NoLegend()
DimHeatmap(patient_11, dims = 1, cells = 500, balanced = TRUE)


DimHeatmap(patient_11, dims = 1:15, cells = 500, balanced = TRUE)

```
```{r}
ElbowPlot(patient_11)
```


```{r}
patient_11 <- FindNeighbors(patient_11, dims = 1:10)
patient_11 <- FindClusters(patient_11, resolution = 0.5)
```
```{r}
patient_11 <- RunUMAP(patient_11, dims = 1:10)
# note that you can set `label = TRUE` or use the LabelClusters function to help label
# individual clusters
DimPlot(patient_11, reduction = "umap")
saveRDS(patient_11, file = "F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/patient_11.rds")

```


```{r}
DimPlot(patient_11, reduction = "umap")
```


```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(patient_11, ident.1 = 2)
head(cluster2.markers, n = 5)

# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(patient_11, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)

# find markers for every cluster compared to all remaining cells, report only the positive
# ones
patient_11.markers <- FindAllMarkers(patient_11, only.pos = TRUE)
patient_11.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
```

```{r}
cluster0.markers <- FindMarkers(patient_11, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
```
mait_markers = ["TRAV1-2", "KLRB1", "CD3D", "CD3E", "CD3G", "MR1"]  

```{r}
VlnPlot(patient_11, features = c("TRAV1-2", "KLRB1", "CD3D", "CD3E", "CD3G", "MR1"))
```
caf_markers = ["ACTA2", "FAP", "PDGFRB"]


```{r}
VlnPlot(patient_11, features = c("ACTA2", "FAP", "PDGFRB"))
```

```{r}
FeaturePlot(patient_11, features = c("TRAV1-2", "KLRB1", "CD3D", "CD3E", "CD3G", "MR1"))
```


caf_markers = ["ACTA2", "FAP", "PDGFRB"]
```{r}
FeaturePlot(patient_11, features = c("ACTA2", "FAP", "PDGFRB"))
```

tumor_markers = ["MKI67", "EPCAM"]
```{r}
FeaturePlot(patient_11, features = c("MKI67", "EPCAM"))
```
```{r}
patient_11.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(patient_11, features = top10$gene) + NoLegend()
```

A Single Cell Atlas of MMRd and MMRp Colorectal Cancer
```{r}
h5_file <- 'F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/GSE178341_crc10x_full_c295v4_submit.h5'

# Load the data into a Seurat object
seurat_object <- Read10X_h5(h5_file)

# Create a Seurat object
seurat_object <- CreateSeuratObject(counts = seurat_object, project = "CRC-Single-Cell")


# Set the path to the cluster file
cluster_file <- "F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/GSE178341_crc10x_full_c295v4_submit_cluster.csv.gz"

# Read the cluster data
cluster_data <- read.csv(gzfile(cluster_file), row.names = 1)

# Add the cluster information to the Seurat object's metadata
seurat_object <- AddMetaData(seurat_object, metadata = cluster_data)

# Set the path to the metadata file
metadata_file <- "F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/GSE178341_crc10x_full_c295v4_submit_metatables.csv.gz"

# Read the metadata
metadata <- read.csv(gzfile(metadata_file), row.names = 1)

# Add the metadata to the Seurat object's metadata
seurat_object <- AddMetaData(seurat_object, metadata = metadata)
```

```{r}
# Find markers in your dataset
markers.your_data <- FindAllMarkers(seurat_object, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# View the markers
head(markers.your_data)

```
```{r}
colnames(seurat_object@misc)
v = seurat_object@meta.data
head(v)
```


```{r}
# Replace "clMidwayPr" with the exact name of the column from your metadata if different
seurat_object <- AddMetaData(seurat_object, metadata = seurat_object@meta.data$cl295v11SubFull, col.name = "cl295v11SubFull")
Idents(seurat_object) <- seurat_object@meta.data$cl295v11SubFull
```

```{r}
# Run PCA
seurat_object <- NormalizeData(seurat_object)
seurat_object <- FindVariableFeatures(seurat_object)
seurat_object <- ScaleData(seurat_object)
seurat_object <- RunPCA(seurat_object)
seurat_object <- RunUMAP(seurat_object, dims = 1:30)

saveRDS(seurat_object, file = "F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/atlas_data.rds")
```



```{r}
DimPlot(seurat_object, label = TRUE, group.by = "cl295v11SubFull", raster = FALSE) +
  theme(legend.text = element_text(size = 3), legend.title = element_text(size = 3)) +  # Adjust size as needed
  NoLegend()  # Use this if you want to remove the legend entirely


```
```{r}
library(ggplot2)

# Generate the UMAP plot and adjust legend text size
p <- DimPlot(seurat_object, label = TRUE, group.by = "cl295v11SubFull", raster = FALSE) +
  theme(legend.text = element_text(size = 6),  # Adjust text size
        legend.title = element_text(size = 7)) +  # Adjust title size
  guides(colour = guide_legend(override.aes = list(size = 2)))  # Adjust symbol size

# Print the plot
print(p)
```


```{r}
options(future.globals.maxSize = 600 * 1024^2)
# Find transfer anchors
anchors <- FindTransferAnchors(reference = seurat_object, query = patient_11, dims = 1:30)


saveRDS(anchors, file = "F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/patient_11_anchors.rds")

# Transfer cell type information
patient_11 <- TransferData(anchorset = anchors, reference = seurat_object, query = patient_11, refdata = seurat_object@meta.data$cl295v11SubFull)
saveRDS(patient_11, file = "F:/GS/PhD/Christensen_Lab/Colon_Project/single_cell/patient_11_processed.rds")
```




```{r}
table(patient_11@meta.data$predicted.id)
```
```{r}
DimPlot(patient_11, reduction = "umap", group.by = "predicted.id", label = TRUE)
```
```{r}
# Create the UMAP plot without the labels on the plot points
umap_plot <- DimPlot(patient_11, reduction = "umap", group.by = "predicted.id", label = FALSE)


# Save the plot to a file
ggsave("umap_plot.png", plot = umap_plot, width = 20, height = 10, dpi = 600)

```

```{r}
patient_11$predicted.id
```


```{r}
levels(as.factor(patient_11@meta.data$predicted.id))
```
Here I want to only include cell types I am interested in.
```{r}
# Define the consolidated labels for each group
caf_labels <- c("cS27 (CXCL14+ CAF)", "cS28 (GREM1+ CAF)", "cS29 (MMP3+ CAF)", "cS30 (CAF CCL8 Fibro-like)", "cS31 (CAF stem niche Fibro-like)")
t_cell_labels <- c("cTNI01 (CD4+ IL7R+)", "cTNI02 (CD4+ IL7R+SELL+)", "cTNI10 (CD8+ IL7R+)")
tumor_labels <- c("cE01 (Stem/TA-like)", "cE02 (Stem/TA-like/Immature Goblet)", "cE03 (Stem/TA-like prolif)", "cE04 (Enterocyte 1)", "cE05 (Enterocyte 2)", "cE06 (Immature Goblet)", "cE07 (Goblet/Enterocyte)", "cE08 (Goblet)")

# Create a new column for consolidated labels
patient_11$consolidated_labels <- patient_11$predicted.id

# Consolidate labels
patient_11$consolidated_labels[patient_11$predicted.id %in% caf_labels] <- "CAF"
patient_11$consolidated_labels[patient_11$predicted.id %in% t_cell_labels] <- "T-Cell"
patient_11$consolidated_labels[patient_11$predicted.id %in% tumor_labels] <- "Tumor"

# Now specify the selected cell types for the plot
selected_cell_types <- c("Tumor", "CAF", "T-Cell")

# Set labels to NA for cells not in the selected cell types
patient_11$consolidated_labels[!patient_11$consolidated_labels %in% selected_cell_types] <- NA

# Plot the UMAP with only selected labels
DimPlot(patient_11, reduction = "umap", group.by = "consolidated_labels", label = FALSE) 


```





To consolidate labels
"cE01 (Stem/TA-like)"                
 [5] "cE02 (Stem/TA-like/Immature Goblet)" "cE03 (Stem/TA-like prolif)"          "cE04 (Enterocyte 1)"                 "cE05 (Enterocyte 2)"                
 [9] "cE06 (Immature Goblet)"              "cE07 (Goblet/Enterocyte)"            "cE08 (Goblet)"                       "cE10 (Tuft)" 

```{r}
# Define a common label for similar cell types, e.g., "Tumor"
common_label <- "Tumor"
tumor_labels <- c("cE01 (Stem/TA-like)" , "cE02 (Stem/TA-like/Immature Goblet)", "cE03 (Stem/TA-like prolif)", "cE04 (Enterocyte 1)","cE05 (Enterocyte 2)","cE06 (Immature Goblet)","cE07 (Goblet/Enterocyte)","cE08 (Goblet)")  # Replace with the actual labels you want to group

selected_cell_types <- c("Tumor")
# Replace the detailed labels with the common label
patient_11$consolidated_labels <- patient_11$predicted.id
patient_11$consolidated_labels[patient_11$predicted.id %in% tumor_labels] <- common_label

# Set labels to NA for cells not in the selected cell types
patient_11$consolidated_labels[!patient_11$consolidated_labels %in% selected_cell_types] <- NA

# Check if there are any non-NA labels before plotting
if(any(!is.na(patient_11$consolidated_labels))) {
  # Use the updated labels for plotting
  DimPlot(patient_11, reduction = "umap", group.by = "consolidated_labels", label = FALSE)
} else {
  warning("All selected_labels are NA. Cannot create a DimPlot with no labels.")
}
```




