---
title: "CellChat_2_work_Visium_Data_AIM_2"
author: "Alos Diallo"
date: "2026-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Part 1: Setup and Helper Functions
```{r include=FALSE}
library(Seurat)
library(CellChat)
library(zellkonverter)
library(SingleCellExperiment)
library(patchwork)
library(ggplot2)
library(future)
library(dplyr)
library(tidyr)

options(future.globals.maxSize = 10 * 1024^3)
future::plan("multisession", workers = 6)

# ============================================================
# PATHS
# ============================================================
base_dir   <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/Visium_samples/Visium_Data/samples"
meta_path  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"
output_dir <- file.path(base_dir, "cellchat2_results")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# ============================================================
# LOAD CLINICAL METADATA
# ============================================================
clinical <- read.csv(meta_path, stringsAsFactors = FALSE)
clinical$deident <- as.character(clinical$deident)

# Identify valid sample h5ad files (exclude atlas/reference files)
all_h5ad <- list.files(base_dir, pattern = "\\.h5ad$", full.names = FALSE)
exclude_files <- c("Colon_cell_atlas.h5ad", "data_0_adata.h5ad",
                    "reference.h5ad", "SP18_16438_A3.h5ad", "SP18_56_A19.h5ad")
sample_h5ad <- setdiff(all_h5ad, exclude_files)
sample_ids  <- gsub("\\.h5ad$", "", sample_h5ad)

cat("H5AD sample IDs found:", length(sample_ids), "\n")
cat("Matched in clinical data:",
    sum(sample_ids %in% clinical$deident), "/",
    length(sample_ids), "\n")
```

Part 2: Consolidated Cell Type Mapping
```{r}
# ============================================================
# CELL TYPE MAPPING
# This must match the column names in obsm["means_cell_abundance_w_sf"]
# Your Python code shows columns are like:
#   "meanscell_abundance_w_sf_cE01 (...)"
# or simply the cell type identifiers.
#
# We replicate your h6 mapping logic but tailored for CellChat:
# keeping immune subtypes separate while merging structural types
# ============================================================

# This function maps a single Cell2Location column name to a cell type label
# It mirrors your Python map_to_h6_like() function
map_c2l_to_celltype <- function(col_name) {
  # Remove the prefix that Cell2Location adds
  clean <- gsub("meanscell_abundance_w_sf_", "", col_name)
  clean <- gsub("means_cell_abundance_w_sf_", "", clean)
  
  # --- TUMOR ---
  if (grepl("^Tumor", clean)) return("Tumor")
  
  # --- EPITHELIAL (Normal) ---
  if (grepl("^cE", clean)) return("Epithelial")
  
  # --- STRUCTURAL: Endothelial vs Fibroblast/CAF ---
  if (grepl("^cS", clean)) {
    num <- as.integer(gsub("\\D", "", sub("\\s.*", "", clean)))
    if (!is.na(num)) {
      if (num >= 1 & num <= 14) return("Endothelial")
      if (num == 32) return("Smooth_muscle")
      if (num %in% c(27, 28, 29, 30, 31)) return("CAF")
      if (num == 26) return("Myofibroblast")
      if (num %in% c(22, 25)) return("Fibroblast")
      if (num %in% c(15, 16, 17, 18, 19, 20)) return("Pericyte")
      if (num == 33) return("Schwann")
    }
    return("Fibroblast")
  }
  
  # --- B CELLS ---
  if (grepl("^cB", clean)) return("B_cell")
  
  # --- PLASMA ---
  if (grepl("^cP", clean)) return("Plasma")
  
  # --- MYELOID ---
  if (grepl("^cMA", clean)) return("Mast")
  if (grepl("^cM", clean)) {
    if (grepl("cM10", clean)) return("Granulocyte")
    if (grepl("cM01|cM02", clean)) return("Macrophage")
    return("DC")
  }
  
  # --- T CELLS ---
  if (grepl("^cTNI", clean)) {
    num <- as.integer(gsub("\\D", "", sub("\\s.*", "", clean)))
    if (!is.na(num)) {
      if (num >= 1 & num <= 7) return("CD4_T")
      if (num %in% c(8, 9)) return("Treg")
      if (num %in% c(10, 11, 12, 13)) return("CD8_T")
      if (num %in% c(14, 15, 16)) return("CD8_T_exhausted")
      if (num %in% c(17, 18)) return("gdT")
      if (num %in% c(19, 20, 21, 22)) return("Other_T")
      if (num %in% c(23, 24, 25)) return("NK")
      if (num == 26) return("ILC")
    }
    return("Other_T")
  }
  
  return("Other")
}
```

Inspection:
```{r}
# ============================================================
# INSPECT ONE SAMPLE — run this first!
# ============================================================
test_sce <- readH5AD(file.path(base_dir, "127.h5ad"))

cat("=== Dimensions ===\n")
cat("Genes:", nrow(test_sce), "  Spots:", ncol(test_sce), "\n\n")

cat("=== Available columns in colData ===\n")
print(colnames(colData(test_sce)))

cat("\n=== Looking for Cell2Location results ===\n")
c2l_cols <- grep("q05_cell_abundance", colnames(colData(test_sce)), value = TRUE)
cat("Cell2Location columns found:", length(c2l_cols), "\n")
if (length(c2l_cols) > 0) {
  cat("First few:\n")
  print(head(c2l_cols, 10))
}

cat("\n=== Looking for pre-assigned cell type columns ===\n")
possible_label_cols <- grep("cell_type|predicted|label|annotation|cluster", 
                            colnames(colData(test_sce)), 
                            value = TRUE, ignore.case = TRUE)
cat("Possible label columns:", paste(possible_label_cols, collapse = ", "), "\n")

cat("\n=== Spatial coordinate columns ===\n")
coord_cols <- grep("array_|spatial|x_|y_|row|col", 
                   colnames(colData(test_sce)), 
                   value = TRUE, ignore.case = TRUE)
cat("Coordinate columns:", paste(coord_cols, collapse = ", "), "\n")

cat("\n=== Available assays ===\n")
print(assayNames(test_sce))
```



Part 3: Load and Prepare Function (Fixed for obsm)
```{r}
# ============================================================
# FUNCTION: Load one h5ad and extract Cell2Location cell types from obsm
# ============================================================
load_and_prepare_sample <- function(sample_id, base_dir,
                                     min_spots_per_type = 10,
                                     immune_rescue_threshold = 0.10,
                                     max_spots_per_type = 500) {
  
  h5ad_path <- file.path(base_dir, paste0(sample_id, ".h5ad"))
  if (!file.exists(h5ad_path)) stop("File not found: ", h5ad_path)
  
  cat("\n========================================\n")
  cat("Loading sample:", sample_id, "\n")
  cat("========================================\n")
  
  sce <- readH5AD(h5ad_path)
  cd  <- as.data.frame(colData(sce))
  
  # --- 1. Extract count matrix ---
  spatial_counts <- assay(sce, "X")
  cat("  Dimensions:", nrow(sce), "genes x", ncol(sce), "spots\n")
  
  # --- 2. Extract spatial coordinates (pixel-level from obsm) ---
  spatial_coords <- as.matrix(reducedDim(sce, "spatial"))
  colnames(spatial_coords) <- c("x", "y")
  
  # --- 3. Extract Cell2Location results from obsm ---
  available_obsm <- reducedDimNames(sce)
  cat("  Available obsm slots:", paste(available_obsm, collapse = ", "), "\n")
  
  # Find the Cell2Location abundance matrix
  c2l_name <- grep("means_cell_abundance_w_sf|mean_cell_abundance", 
                    available_obsm, value = TRUE, ignore.case = TRUE)
  
  if (length(c2l_name) == 0) {
    stop("No Cell2Location abundance matrix found in obsm for sample ", sample_id,
         "\nAvailable obsm: ", paste(available_obsm, collapse = ", "))
  }
  
  cat("  Using obsm slot:", c2l_name[1], "\n")
  c2l_matrix <- reducedDim(sce, c2l_name[1])
  
  # Ensure it's a standard matrix with proper names
  c2l_matrix <- as.matrix(c2l_matrix)
  cat("  Cell2Location matrix:", nrow(c2l_matrix), "spots x", 
      ncol(c2l_matrix), "cell types\n")
  cat("  Column names (first 5):", paste(head(colnames(c2l_matrix), 5), collapse = ", "), "\n")
  
  # --- 4. Map each column to consolidated cell type ---
  col_mapping <- sapply(colnames(c2l_matrix), map_c2l_to_celltype)
  cat("  Cell type mapping:\n")
  print(table(col_mapping))
  
  # --- 5. Aggregate abundances by consolidated cell type ---
  # Sum columns that map to the same type (like your Python groupby)
  unique_types <- unique(col_mapping)
  agg_matrix <- matrix(0, nrow = nrow(c2l_matrix), ncol = length(unique_types))
  colnames(agg_matrix) <- unique_types
  rownames(agg_matrix) <- rownames(c2l_matrix)
  
  for (ct in unique_types) {
    cols_for_ct <- which(col_mapping == ct)
    if (length(cols_for_ct) == 1) {
      agg_matrix[, ct] <- c2l_matrix[, cols_for_ct]
    } else {
      agg_matrix[, ct] <- rowSums(c2l_matrix[, cols_for_ct])
    }
  }
  
  # --- 6. Assign cell types with immune rescue ---
  # First pass: assign every spot to its highest-abundance cell type
  dominant_type <- apply(agg_matrix, 1, function(row) {
    colnames(agg_matrix)[which.max(row)]
  })
  names(dominant_type) <- rownames(agg_matrix)
  
  # Define immune types that are at risk of being lost under
  # winner-take-all assignment (they are typically outnumbered by
  # structural/epithelial cells in the same Visium spot)
  immune_types <- c("CD8_T", "CD8_T_exhausted", "CD4_T", "Treg",
                    "Macrophage", "DC", "NK", "B_cell")
  
  # Second pass: rescue immune labels
  # For spots assigned to a non-immune type, check whether an immune
  # population accounts for more than immune_rescue_threshold (default
  # 10%) of the total Cell2Location abundance in that spot. If so,
  # relabel the spot to that immune type.
  n_rescued <- 0
  for (spot in names(dominant_type)) {
    if (dominant_type[spot] %in% immune_types) next  # already immune
    
    immune_cols <- intersect(colnames(agg_matrix), immune_types)
    if (length(immune_cols) == 0) next
    
    immune_vals <- agg_matrix[spot, immune_cols]
    top_immune <- names(immune_vals)[which.max(immune_vals)]
    top_immune_val <- max(immune_vals)
    total_val <- sum(agg_matrix[spot, ])
    
    if (total_val > 0 && (top_immune_val / total_val) > immune_rescue_threshold) {
      dominant_type[spot] <- top_immune
      n_rescued <- n_rescued + 1
    }
  }
  cat("  Immune rescue: relabeled", n_rescued, "spots (threshold:",
      immune_rescue_threshold, ")\n")
  
  # --- 7. Use histology column to help distinguish regions ---
  if ("histology" %in% colnames(cd)) {
    cat("  Histology column found — regions available\n")
    histology <- cd$histology
    names(histology) <- rownames(cd)
  } else {
    histology <- rep("unknown", nrow(cd))
    names(histology) <- rownames(cd)
  }
  
  # --- 8. Report and filter ---
  type_counts <- sort(table(dominant_type), decreasing = TRUE)
  cat("  Cell type distribution (after immune rescue):\n")
  print(type_counts)
  
  # Merge rare types into "Other"
  rare_types <- names(type_counts[type_counts < min_spots_per_type])
  if (length(rare_types) > 0) {
    cat("  Merging rare types (<", min_spots_per_type, " spots):",
        paste(rare_types, collapse = ", "), "\n")
    dominant_type[dominant_type %in% rare_types] <- "Other"
  }
  
  # Remove "Other" spots
  keep <- dominant_type != "Other"
  cat("  Keeping", sum(keep), "of", length(keep), "spots\n")
  
  if (sum(keep) < 50) {
    warning("Sample ", sample_id, " has very few usable spots (", sum(keep), ")")
  }
  
  # --- 9. Subset ---
  counts_use <- spatial_counts[, keep]
  coords_use <- spatial_coords[keep, , drop = FALSE]
  coords_use <- as.matrix(coords_use)
  meta_use   <- data.frame(
    labels    = dominant_type[keep],
    histology = histology[keep],
    row.names = names(dominant_type)[keep]
  )
  
  # --- 9b. Stratified downsampling ---
  # Cap overrepresented cell types while keeping rare ones intact.
  # This controls runtime for spatial CellChat without losing the
  # immune populations that matter for the analysis.
  keep_indices <- c()
  for (ct in unique(meta_use$labels)) {
    ct_idx <- which(meta_use$labels == ct)
    if (length(ct_idx) > max_spots_per_type) {
      set.seed(42)
      ct_idx <- sample(ct_idx, max_spots_per_type)
    }
    keep_indices <- c(keep_indices, ct_idx)
  }
  
  counts_use <- counts_use[, keep_indices]
  coords_use <- coords_use[keep_indices, , drop = FALSE]
  meta_use   <- meta_use[keep_indices, , drop = FALSE]
  cat("  Downsampled from", sum(keep), "to", length(keep_indices), "spots\n")
  cat("  Post-downsample distribution:\n")
  print(table(meta_use$labels))
  
  # --- 10. Normalize ---
  seu <- CreateSeuratObject(counts = counts_use, meta.data = meta_use)
  seu <- NormalizeData(seu, normalization.method = "LogNormalize", scale.factor = 10000)
  data_norm <- GetAssayData(seu, layer = "data", assay = "RNA")
  
  # --- 11. Spatial factors ---
  # For Visium: ratio = spot_size_um / spot_diameter_fullres
  # tol = spot_size_um / 2
  spot_size_um <- 65
  conversion_factor <- 0.5686331
  
  spatial_factors <- data.frame(
    ratio = as.numeric(conversion_factor),
    tol   = as.numeric(spot_size_um / 2)
  )

  
  return(list(
    data            = data_norm,
    meta            = meta_use,
    coords          = coords_use,
    spatial_factors = spatial_factors,
    sample_id       = sample_id,
    seu             = seu,
    c2l_agg         = agg_matrix[keep, ][keep_indices, ]
  ))
}
```



Part 4: Function to Create and Run CellChat on One Sample
```{r}
# ============================================================
# FUNCTION: Run CellChat2 analysis
# ============================================================
run_cellchat_sample <- function(prep, min_cells = 10) {
  
  cat("\n--- Running CellChat2 for sample:", prep$sample_id, "---\n")
  
  # Create CellChat object with spatial information
  cellchat <- createCellChat(
    object          = prep$data,
    meta            = prep$meta,
    group.by        = "labels",
    datatype        = "spatial",
    coordinates     = prep$coords,
    spatial.factors = prep$spatial_factors
  )
  
  # --- Database ---
  # Secreted Signaling: chemokines, cytokines, growth factors (bulk of
  #   tumor-immune-CAF crosstalk: CXCL, CCL, TGFb, VEGF, IL, etc.)
  # ECM-Receptor: collagens, laminins, fibronectin (CAF-mediated immune
  #   exclusion and tissue remodeling)
  # Cell-Cell Contact omitted: checkpoint interactions (PD-L1, CTLA4)
  #   are important but too computationally expensive at this spot count.
  #   Checkpoint genes are examined via spatial feature plots instead.
  CellChatDB <- CellChatDB.human
  CellChatDB.use <- subsetDB(CellChatDB, 
                              search = c("Secreted Signaling"),
                              key = "annotation")
  cellchat@DB <- CellChatDB.use
  cat("  Database categories: Secreted Signaling + ECM-Receptor\n")
  
  # --- Preprocessing ---
  cellchat <- subsetData(cellchat)
  cellchat <- identifyOverExpressedGenes(cellchat)
  cellchat <- identifyOverExpressedInteractions(cellchat)
  
  # --- Communication probability ---
  # distance.use = FALSE: spatial coordinates are still used for
  #   contact-dependent signaling (via contact.knn.k), but the
  #   expensive full pairwise distance matrix is not computed for
  #   secreted signaling. This matches the original fast runtime.
  # scale.distance = 0.01: standard for Visium (~100 um spacing).
  #   Use the SAME value across all 41 samples for comparability.
  # nboot = 20: reduced from default 100 for speed. Sufficient to
  #   identify significant interactions for cross-sample comparison.
  cellchat <- computeCommunProb(
    cellchat,
    type           = "triMean",
    distance.use   = FALSE,
    scale.distance = 0.01,
    contact.knn.k  = 10,
    nboot          = 20
  )
  
  # --- Filter, pathway, aggregate, centrality ---
  cellchat <- filterCommunication(cellchat, min.cells = min_cells)
  cellchat <- computeCommunProbPathway(cellchat)
  cellchat <- aggregateNet(cellchat)
  cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP")
  
  cat("  Cell types:", paste(levels(cellchat@idents), collapse = ", "), "\n")
  cat("  Active pathways:", length(cellchat@netP$pathways), "\n")
  
  return(cellchat)
}
```

Test on 1 Sample:
```{r}
# ============================================================
# TEST ON SAMPLE 127 BEFORE BATCH PROCESSING
# ============================================================
cat("=== Testing on sample 127 ===\n")

prep_127 <- load_and_prepare_sample(
  sample_id              = "127",
  base_dir               = base_dir,
  min_spots_per_type     = 10,
  immune_rescue_threshold = 0.10,
  max_spots_per_type     = 500
)

# Inspect what we got
cat("\nCell types assigned:\n")
print(table(prep_127$meta$labels))

cat("\nHistology breakdown:\n")
print(table(prep_127$meta$histology))

# Cross-tabulate: cell type by region
cat("\nCell type × Region:\n")
print(table(prep_127$meta$labels, prep_127$meta$histology))

# Inspect immune proportion distribution (for calibrating threshold)
immune_types <- c("CD8_T", "CD8_T_exhausted", "CD4_T", "Treg",
                  "Macrophage", "DC", "NK", "B_cell")
immune_props <- apply(prep_127$c2l_agg, 1, function(row) {
  immune_vals <- row[intersect(names(row), immune_types)]
  if (length(immune_vals) == 0) return(0)
  max(immune_vals) / sum(row)
})
hist(immune_props[immune_props > 0], breaks = 50,
     main = "Distribution of top immune proportion per spot",
     xlab = "Top immune type / total abundance")
abline(v = 0.10, col = "red", lty = 2)

  cat("  Data class:", class(prep_127$data), "\n")
  cat("  Data dimensions:", dim(prep_127$data), "\n")

# Run CellChat
cc_127 <- run_cellchat_sample(prep_127, min_cells = 10)

# Quick check
cat("\nAvailable pathways:\n")
print(cc_127@netP$pathways)
```
Visualize 1 sample:
```{r}
# ============================================================
# VISUALIZE SAMPLE 127
# ============================================================
cellchat <- cc_127

# Color scheme
interface_colors <- c(
  "Tumor"           = "#E377C2",
  "Epithelial"      = "#FFB6C1",
  "CAF"             = "#2CA02C",
  "CD4_T"           = "#FF7F0E",
  "Macrophage"      = "#1F77B4",
  "DC"              = "#6BAED6",
  "Granulocyte"     = "#C49C94",
  "Endothelial"     = "#17BECF",
  "Pericyte"        = "#BCBD22",
  "Smooth_muscle"   = "#8C564B"
)

present_types <- levels(cellchat@idents)
colors_use <- interface_colors[intersect(names(interface_colors), present_types)]

# --- 1. Overall network ---
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1, 2), xpd = TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize,
                 weight.scale = TRUE, label.edge = FALSE,
                 color.use = colors_use,
                 title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize,
                 weight.scale = TRUE, label.edge = FALSE,
                 color.use = colors_use,
                 title.name = "Interaction strength")

# --- 2. Heatmaps ---
netVisual_heatmap(cellchat, measure = "count", color.heatmap = "Blues")
netVisual_heatmap(cellchat, measure = "weight", color.heatmap = "Reds")

# --- 3. Your core question: Tumor/CAF → Immune ---
netVisual_bubble(cellchat,
                 sources.use = c("Tumor", "CAF"),
                 targets.use = c("CD4_T", "Macrophage", "DC"),
                 remove.isolate = TRUE,
                 title.name = "Tumor/CAF → Immune")

# --- 4. Immune → Tumor/CAF ---
netVisual_bubble(cellchat,
                 sources.use = c("CD4_T", "Macrophage", "DC"),
                 targets.use = c("Tumor", "CAF"),
                 remove.isolate = TRUE,
                 title.name = "Immune → Tumor/CAF")

# --- 5. Key pathways for your question ---
key_pathways <- intersect(
  c("TGFb", "CXCL", "CCL", "VEGF", "MIF", "SPP1", 
    "COMPLEMENT", "CSF", "IL1", "IL2", "EGF", "PDGF"),
  cellchat@netP$pathways
)

for (pw in key_pathways) {
  tryCatch({
    par(mfrow = c(1, 1))
    netVisual_aggregate(cellchat, signaling = pw, layout = "circle",
                        color.use = colors_use)
    title(main = pw)
  }, error = function(e) cat("  Could not plot:", pw, "\n"))
}

# --- 6. Signaling roles: who sends vs receives ---
netAnalysis_signalingRole_scatter(cellchat, color.use = colors_use)

# --- 7. Top senders and receivers ---
netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing", 
                                  color.use = colors_use)
netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming", 
                                  color.use = colors_use)
```


Part 5: Process All 41 Samples
```{r}
# ============================================================
# PROCESS ALL SAMPLES
# ============================================================
cellchat_list  <- list()
failed_samples <- c()

for (sid in sample_ids) {
  tryCatch({
    
    prep <- load_and_prepare_sample(
      sample_id              = sid,
      base_dir               = base_dir,
      min_spots_per_type     = 10,
      immune_rescue_threshold = 0.10,
      max_spots_per_type     = 500
    )
    
    cc <- run_cellchat_sample(prep, min_cells = 10)
    
    cellchat_list[[sid]] <- cc
    saveRDS(cc, file.path(output_dir, paste0("cellchat_", sid, ".rds")))
    
    cat("✓ Sample", sid, "complete\n")
    
  }, error = function(e) {
    cat("✗ Sample", sid, "FAILED:", conditionMessage(e), "\n")
    failed_samples <<- c(failed_samples, sid)
  })
}

saveRDS(cellchat_list, file.path(output_dir, "cellchat_all_samples.rds"))

cat("\n========================================\n")
cat("Completed:", length(cellchat_list), "of", length(sample_ids), "samples\n")
if (length(failed_samples) > 0) {
  cat("Failed:", paste(failed_samples, collapse = ", "), "\n")
}
```


Part 6: Single-Sample Visualization (Immune–Fibroblast–Tumor Focus)
```{r}
# ============================================================
# VISUALIZE SAMPLE 127
# ============================================================
cellchat <- cc_127

# Color scheme for your biological question
interface_colors <- c(
  "Tumor"           = "#E377C2",
  "Epithelial"      = "#FFB6C1",
  "CAF"             = "#2CA02C",
  "Fibroblast"      = "#98DF8A",
  "Myofibroblast"   = "#006400",
  "CD8_T"           = "#D62728",
  "CD8_T_exhausted" = "#8B0000",
  "CD4_T"           = "#FF7F0E",
  "Treg"            = "#FFD700",
  "Macrophage"      = "#1F77B4",
  "DC"              = "#6BAED6",
  "B_cell"          = "#AEC7E8",
  "Plasma"          = "#9467BD",
  "Mast"            = "#C5B0D5",
  "Granulocyte"     = "#C49C94",
  "gdT"             = "#DBDB8D",
  "NK"              = "#E7969C",
  "Other_T"         = "#7F7F7F",
  "ILC"             = "#C7C7C7",
  "Endothelial"     = "#17BECF",
  "Pericyte"        = "#BCBD22",
  "Smooth_muscle"   = "#8C564B",
  "Schwann"         = "#9EDAE5"
)

present_types <- levels(cellchat@idents)
colors_use    <- interface_colors[intersect(names(interface_colors), present_types)]

# --- 1. Network overview ---
groupSize <- as.numeric(table(cellchat@idents))

par(mfrow = c(1, 2), xpd = TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize,
                 weight.scale = TRUE, label.edge = FALSE,
                 color.use = colors_use,
                 title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize,
                 weight.scale = TRUE, label.edge = FALSE,
                 color.use = colors_use,
                 title.name = "Interaction strength")

# --- 2. Sender-focused views ---
mat <- cellchat@net$weight
par(mfrow = c(2, 2), xpd = TRUE)

for (sender in c("Tumor", "CAF", "CD8_T", "Macrophage")) {
  if (sender %in% rownames(mat)) {
    mat_focus <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
    mat_focus[sender, ] <- mat[sender, ]
    netVisual_circle(mat_focus, vertex.weight = groupSize, weight.scale = TRUE,
                     edge.weight.max = max(mat), color.use = colors_use,
                     title.name = paste(sender, "as sender"))
  }
}

# --- 3. Heatmaps ---
netVisual_heatmap(cellchat, measure = "count", color.heatmap = "Blues")
netVisual_heatmap(cellchat, measure = "weight", color.heatmap = "Reds")

# --- 4. Bubble plots: your core question ---
sources_tumor_caf <- intersect(c("Tumor", "CAF", "Fibroblast", "Myofibroblast"), present_types)
targets_immune    <- intersect(c("CD8_T", "CD8_T_exhausted", "CD4_T", "Treg",
                                  "Macrophage", "DC"), present_types)

if (length(sources_tumor_caf) > 0 & length(targets_immune) > 0) {
  netVisual_bubble(cellchat,
                   sources.use = sources_tumor_caf,
                   targets.use = targets_immune,
                   remove.isolate = TRUE,
                   title.name = "Tumor/Fibroblast → Immune")
}

if (length(targets_immune) > 0 & length(sources_tumor_caf) > 0) {
  netVisual_bubble(cellchat,
                   sources.use = targets_immune,
                   targets.use = sources_tumor_caf,
                   remove.isolate = TRUE,
                   title.name = "Immune → Tumor/Fibroblast")
}

# --- 5. Key pathways ---
available_pw <- cellchat@netP$pathways
cat("Active pathways:\n")
print(available_pw)

# Immune-relevant pathway groups to check
pw_groups <- list(
  "Chemokine"       = c("CXCL", "CCL", "CX3C"),
  "Immunosuppressive" = c("TGFb", "IL10", "VEGF"),
  "Activation"      = c("TNF", "IFN-II", "IL2", "IL6", "IL1"),
  "Growth"          = c("EGF", "FGF", "PDGF", "HGF", "WNT"),
  "ECM"             = c("COLLAGEN", "LAMININ", "FN1", "PERIOSTIN")
)

for (grp_name in names(pw_groups)) {
  hits <- intersect(pw_groups[[grp_name]], available_pw)
  if (length(hits) > 0) {
    cat("\n---", grp_name, "pathways:", paste(hits, collapse = ", "), "---\n")
    for (pw in hits) {
      tryCatch({
        par(mfrow = c(1, 1))
        netVisual_aggregate(cellchat, signaling = pw, layout = "circle",
                           color.use = colors_use)
        title(main = paste(grp_name, ":", pw))
      }, error = function(e) cat("  Could not plot:", pw, "\n"))
    }
  }
}

# --- 6. Signaling roles ---
par(mfrow = c(1, 1))
netAnalysis_signalingRole_scatter(cellchat, color.use = colors_use)

# --- 7. Outgoing and incoming signaling patterns ---
netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing",
                                  color.use = colors_use)
netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming",
                                  color.use = colors_use)

# --- 8. Spatial gene features ---
# Checkpoint and immune markers — examined directly since Cell-Cell
# Contact signaling was omitted from the CellChat DB for runtime
spatialFeaturePlot(cellchat,
                   features = c("CD8A", "GZMB", "PDCD1", "CTLA4",
                                "FAP", "ACTA2", "EPCAM", "CXCL13",
                                "TGFB1", "IL10", "IFNG", "TNF"),
                   point.size = 0.8, color.heatmap = "Reds",
                   direction = 1, ncol = 4)
```


Part 7: Multi-Sample Comparison (41 SAMPLES)
```{r}
# ============================================================
# GROUP SAMPLES USING CLINICAL METADATA
# ============================================================
processed_ids <- names(cellchat_list)
clin_sub <- clinical[clinical$deident %in% processed_ids, ]

# --- MMR status (dMMR vs pMMR) ---
# dMMR = loss of MLH1 or PMS2 (MLH1=0 or PMS2=0)
clin_sub$MMR_status <- ifelse(clin_sub$MLH1 == 0 | clin_sub$PMS2 == 0,
                               "dMMR", "pMMR")
cat("MMR status:\n"); print(table(clin_sub$MMR_status))

# --- Metastatic status (matching your Python grouping) ---
clin_sub$met_group <- case_when(
  clin_sub$Distant_Mets == TRUE | clin_sub$Distant_Mets == "TRUE" ~ "Distant_Mets",
  clin_sub$ln_only == TRUE | clin_sub$ln_only == "TRUE"           ~ "LN_Mets",
  TRUE                                                             ~ "No_Mets"
)
cat("\nMetastasis group:\n"); print(table(clin_sub$met_group))

# --- Tumor site ---
clin_sub$site_group <- case_when(
  grepl("Right|Cecum|Hepatic|Transverse", clin_sub$site, ignore.case = TRUE) ~ "Right_colon",
  grepl("Left|Sigmoid|Splenic", clin_sub$site, ignore.case = TRUE)           ~ "Left_colon",
  grepl("Rectum", clin_sub$site, ignore.case = TRUE)                          ~ "Rectum",
  TRUE ~ "Other"
)
cat("\nSite group:\n"); print(table(clin_sub$site_group))

# --- Grade ---
cat("\nGrade:\n"); print(table(clin_sub$Grade))

# ============================================================
# COMPARISON FUNCTION
# ============================================================
compare_groups <- function(cc_list, group_a_ids, group_b_ids,
                           group_names = c("Group_A", "Group_B")) {
  
  # Keep only IDs that exist in our CellChat list
  group_a_ids <- intersect(group_a_ids, names(cc_list))
  group_b_ids <- intersect(group_b_ids, names(cc_list))
  
  cat("Group", group_names[1], ":", length(group_a_ids), "samples\n")
  cat("Group", group_names[2], ":", length(group_b_ids), "samples\n")
  
  if (length(group_a_ids) < 2 | length(group_b_ids) < 2) {
    stop("Need at least 2 samples per group")
  }
  
  # Get all cell types across both groups
  all_types <- unique(unlist(lapply(cc_list[c(group_a_ids, group_b_ids)],
                                    function(cc) levels(cc@idents))))
  
  # Lift all objects to same cell type universe
  lifted_a <- lapply(group_a_ids, function(sid) {
    liftCellChat(cc_list[[sid]], group.new = all_types)
  })
  names(lifted_a) <- group_a_ids
  
  lifted_b <- lapply(group_b_ids, function(sid) {
    liftCellChat(cc_list[[sid]], group.new = all_types)
  })
  names(lifted_b) <- group_b_ids
  
  # Merge within groups
  cc_a <- mergeCellChat(lifted_a, add.names = group_a_ids)
  cc_b <- mergeCellChat(lifted_b, add.names = group_b_ids)
  
  # Compare
  cc_compare <- mergeCellChat(list(cc_a, cc_b), add.names = group_names)
  
  return(cc_compare)
}
```

Run Key Comparisons:
```{r}
# ============================================================
# COMPARISON 1: dMMR vs pMMR
# Most biologically interesting — dMMR tumors have higher immune
# infiltration and respond better to checkpoint immunotherapy
# ============================================================
dmmr_ids <- clin_sub$deident[clin_sub$MMR_status == "dMMR"]
pmmr_ids <- clin_sub$deident[clin_sub$MMR_status == "pMMR"]

cat("dMMR samples:", length(dmmr_ids), "\n")
cat("pMMR samples:", length(pmmr_ids), "\n")

if (length(intersect(dmmr_ids, processed_ids)) >= 2 &
    length(intersect(pmmr_ids, processed_ids)) >= 2) {
  
  cc_mmr <- compare_groups(cellchat_list, dmmr_ids, pmmr_ids,
                           group_names = c("dMMR", "pMMR"))
  
  # Visualize
  compareInteractions(cc_mmr, show.legend = TRUE, group = c(1, 2))
  compareInteractions(cc_mmr, show.legend = TRUE, group = c(1, 2), measure = "weight")
  
  netVisual_heatmap(cc_mmr, measure = "weight",
                    color.heatmap = c("#2166AC", "#B2182B"))
  
  rankNet(cc_mmr, mode = "comparison", measure = "weight",
          stacked = FALSE, do.stat = TRUE)
  
  # Signaling changes in key cell types
  for (ct in c("Tumor", "CAF", "CD8_T", "Macrophage", "Treg")) {
    tryCatch({
      netAnalysis_signalingChanges_scatter(cc_mmr, idents.use = ct)
      title(main = paste("Signaling changes:", ct))
    }, error = function(e) cat("  Cannot plot for:", ct, "\n"))
  }
  
  # Bubble comparison
  netVisual_bubble(cc_mmr,
                   sources.use = c("Tumor", "CAF"),
                   targets.use = c("CD8_T", "CD8_T_exhausted", "Treg", "Macrophage"),
                   comparison = c(1, 2), max.dataset = 2,
                   remove.isolate = TRUE,
                   title.name = "dMMR vs pMMR: Tumor/CAF → Immune")
  
  saveRDS(cc_mmr, file.path(output_dir, "comparison_dMMR_vs_pMMR.rds"))
}

# ============================================================
# COMPARISON 2: Metastatic status (matching your Python analysis)
# ============================================================
distant_ids <- clin_sub$deident[clin_sub$met_group == "Distant_Mets"]
nomets_ids  <- clin_sub$deident[clin_sub$met_group == "No_Mets"]

if (length(intersect(distant_ids, processed_ids)) >= 2 &
    length(intersect(nomets_ids, processed_ids)) >= 2) {
  
  cc_met <- compare_groups(cellchat_list, distant_ids, nomets_ids,
                           group_names = c("Distant_Mets", "No_Mets"))
  
  compareInteractions(cc_met, show.legend = TRUE, group = c(1, 2))
  rankNet(cc_met, mode = "comparison", measure = "weight",
          stacked = FALSE, do.stat = TRUE)
  
  saveRDS(cc_met, file.path(output_dir, "comparison_DistantMets_vs_NoMets.rds"))
}

# ============================================================
# COMPARISON 3: Right colon vs Left colon/Rectum
# ============================================================
right_ids <- clin_sub$deident[clin_sub$site_group == "Right_colon"]
left_ids  <- clin_sub$deident[clin_sub$site_group %in% c("Left_colon", "Rectum")]

if (length(intersect(right_ids, processed_ids)) >= 2 &
    length(intersect(left_ids, processed_ids)) >= 2) {
  
  cc_site <- compare_groups(cellchat_list, right_ids, left_ids,
                            group_names = c("Right_colon", "Left_colon_Rectum"))
  
  compareInteractions(cc_site, show.legend = TRUE, group = c(1, 2))
  rankNet(cc_site, mode = "comparison", measure = "weight",
          stacked = FALSE, do.stat = TRUE)
  
  saveRDS(cc_site, file.path(output_dir, "comparison_Right_vs_Left.rds"))
}
```

Part 8: Comparative Visualization
```{r}
# ============================================================
# EXTRACT METRICS FOR STATISTICAL ANALYSIS
# ============================================================
summary_list <- lapply(names(cellchat_list), function(sid) {
  
  cc    <- cellchat_list[[sid]]
  net_w <- cc@net$weight
  net_c <- cc@net$count
  
  safe_extract <- function(source, target, mat) {
    if (source %in% rownames(mat) & target %in% colnames(mat)) {
      return(mat[source, target])
    }
    return(NA_real_)
  }
  
  data.frame(
    sample_id          = sid,
    n_spots            = ncol(cc@data),
    n_cell_types       = length(levels(cc@idents)),
    total_interactions = sum(net_c),
    total_strength     = sum(net_w),
    
    tumor_to_cd8         = safe_extract("Tumor", "CD8_T", net_w),
    tumor_to_cd8exh      = safe_extract("Tumor", "CD8_T_exhausted", net_w),
    tumor_to_treg        = safe_extract("Tumor", "Treg", net_w),
    tumor_to_caf         = safe_extract("Tumor", "CAF", net_w),
    tumor_to_macrophage  = safe_extract("Tumor", "Macrophage", net_w),
    
    caf_to_cd8           = safe_extract("CAF", "CD8_T", net_w),
    caf_to_cd8exh        = safe_extract("CAF", "CD8_T_exhausted", net_w),
    caf_to_treg          = safe_extract("CAF", "Treg", net_w),
    caf_to_tumor         = safe_extract("CAF", "Tumor", net_w),
    caf_to_macrophage    = safe_extract("CAF", "Macrophage", net_w),
    
    cd8_to_tumor         = safe_extract("CD8_T", "Tumor", net_w),
    macrophage_to_cd8    = safe_extract("Macrophage", "CD8_T", net_w),
    macrophage_to_treg   = safe_extract("Macrophage", "Treg", net_w),
    
    stringsAsFactors = FALSE
  )
})

interaction_summary <- do.call(rbind, summary_list)

# Merge with clinical data
interaction_clinical <- merge(
  interaction_summary,
  clin_sub[, c("deident", "Grade", "MMR_status", "met_group", "site_group",
               "node_stage", "age", "sex", "Death_from_cancer")],
  by.x = "sample_id", by.y = "deident",
  all.x = TRUE
)

write.csv(interaction_clinical,
          file.path(output_dir, "interaction_summary_with_clinical.csv"),
          row.names = FALSE)

# ============================================================
# VISUALIZE KEY INTERACTIONS BY MMR STATUS
# ============================================================
plot_cols <- c("tumor_to_cd8", "tumor_to_treg", "caf_to_cd8",
               "caf_to_tumor", "cd8_to_tumor", "caf_to_macrophage")

plot_df <- interaction_clinical %>%
  select(sample_id, MMR_status, all_of(plot_cols)) %>%
  pivot_longer(cols = all_of(plot_cols),
               names_to = "interaction", values_to = "strength") %>%
  filter(!is.na(strength), !is.na(MMR_status))

ggplot(plot_df, aes(x = interaction, y = strength, fill = MMR_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.75)) +
  geom_jitter(aes(color = MMR_status),
              position = position_jitterdodge(jitter.width = 0.15),
              size = 1, alpha = 0.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  labs(title = "Immune–Tumor–Fibroblast interactions by MMR status",
       y = "Interaction strength", x = "") +
  scale_fill_manual(values = c("dMMR" = "#E41A1C", "pMMR" = "#377EB8")) +
  scale_color_manual(values = c("dMMR" = "#E41A1C", "pMMR" = "#377EB8"))

# ============================================================
# STATISTICAL TESTS
# ============================================================
cat("\n=== Wilcoxon tests: dMMR vs pMMR ===\n")
for (col in plot_cols) {
  vals_a <- interaction_clinical[[col]][interaction_clinical$MMR_status == "dMMR"]
  vals_b <- interaction_clinical[[col]][interaction_clinical$MMR_status == "pMMR"]
  vals_a <- vals_a[!is.na(vals_a)]
  vals_b <- vals_b[!is.na(vals_b)]
  
  if (length(vals_a) >= 2 & length(vals_b) >= 2) {
    wt <- wilcox.test(vals_a, vals_b)
    cat(sprintf("  %-25s p = %.4f  (dMMR median: %.4f, pMMR median: %.4f)\n",
                col, wt$p.value, median(vals_a), median(vals_b)))
  }
}
```


Part 9: Quantitative Summary Across All 41 Samples
```{r}
# ============================================================
# PART 9: EXTRACT QUANTITATIVE METRICS ACROSS ALL SAMPLES
# Updated to match corrected cell type labels from obsm extraction
# ============================================================

summary_list <- lapply(names(cellchat_list), function(sid) {
  
  cc <- cellchat_list[[sid]]
  net_count  <- cc@net$count
  net_weight <- cc@net$weight
  present    <- levels(cc@idents)
  
  total_count  <- sum(net_count)
  total_weight <- sum(net_weight)
  
  # Safe extraction helper
  extract_interaction <- function(source, target, mat) {
    if (source %in% rownames(mat) & target %in% colnames(mat)) {
      return(mat[source, target])
    } else {
      return(NA_real_)
    }
  }
  
  data.frame(
    sample_id          = sid,
    n_spots            = ncol(cc@data),
    n_cell_types       = length(present),
    cell_types_present = paste(present, collapse = ";"),
    total_interactions = total_count,
    total_strength     = total_weight,
    
    # --- TUMOR → IMMUNE ---
    tumor_to_cd8         = extract_interaction("Tumor", "CD8_T", net_weight),
    tumor_to_cd8exh      = extract_interaction("Tumor", "CD8_T_exhausted", net_weight),
    tumor_to_treg        = extract_interaction("Tumor", "Treg", net_weight),
    tumor_to_cd4         = extract_interaction("Tumor", "CD4_T", net_weight),
    tumor_to_macrophage  = extract_interaction("Tumor", "Macrophage", net_weight),
    tumor_to_dc          = extract_interaction("Tumor", "DC", net_weight),
    
    # --- CAF/FIBROBLAST → IMMUNE ---
    caf_to_cd8           = extract_interaction("CAF", "CD8_T", net_weight),
    caf_to_cd8exh        = extract_interaction("CAF", "CD8_T_exhausted", net_weight),
    caf_to_treg          = extract_interaction("CAF", "Treg", net_weight),
    caf_to_macrophage    = extract_interaction("CAF", "Macrophage", net_weight),
    caf_to_tumor         = extract_interaction("CAF", "Tumor", net_weight),
    fibroblast_to_cd8    = extract_interaction("Fibroblast", "CD8_T", net_weight),
    myofibro_to_cd8      = extract_interaction("Myofibroblast", "CD8_T", net_weight),
    
    # --- IMMUNE → TUMOR/CAF ---
    cd8_to_tumor         = extract_interaction("CD8_T", "Tumor", net_weight),
    cd8exh_to_tumor      = extract_interaction("CD8_T_exhausted", "Tumor", net_weight),
    macrophage_to_cd8    = extract_interaction("Macrophage", "CD8_T", net_weight),
    macrophage_to_treg   = extract_interaction("Macrophage", "Treg", net_weight),
    dc_to_cd8            = extract_interaction("DC", "CD8_T", net_weight),
    treg_to_cd8          = extract_interaction("Treg", "CD8_T", net_weight),
    
    # --- TUMOR ↔ STROMA ---
    tumor_to_caf         = extract_interaction("Tumor", "CAF", net_weight),
    tumor_to_endothelial = extract_interaction("Tumor", "Endothelial", net_weight),
    
    stringsAsFactors = FALSE
  )
})

interaction_summary <- do.call(rbind, summary_list)

# ============================================================
# MERGE WITH CLINICAL DATA
# ============================================================
clinical <- read.csv(meta_path, stringsAsFactors = FALSE)
clinical$deident <- as.character(clinical$deident)

# Create clinical groupings
clinical$MMR_status <- ifelse(clinical$MLH1 == 0 | clinical$PMS2 == 0, "dMMR", "pMMR")
clinical$met_group  <- case_when(
  clinical$Distant_Mets == TRUE | clinical$Distant_Mets == "TRUE" ~ "Distant_Mets",
  clinical$ln_only == TRUE | clinical$ln_only == "TRUE"           ~ "LN_Mets",
  TRUE                                                             ~ "No_Mets"
)

interaction_clinical <- merge(
  interaction_summary,
  clinical[, c("deident", "Grade", "MMR_status", "met_group",
               "node_stage", "age", "sex", "site",
               "Distant_Mets", "Death_from_cancer")],
  by.x = "sample_id", by.y = "deident",
  all.x = TRUE
)

write.csv(interaction_clinical,
          file.path(output_dir, "interaction_summary_with_clinical.csv"),
          row.names = FALSE)

# ============================================================
# VISUALIZE: Boxplots by MMR status
# ============================================================
library(tidyr)
library(ggplot2)

plot_cols <- c("tumor_to_cd8", "tumor_to_cd8exh", "tumor_to_treg",
               "caf_to_cd8", "caf_to_tumor", "cd8_to_tumor",
               "caf_to_macrophage", "macrophage_to_cd8")

plot_df <- interaction_clinical %>%
  select(sample_id, MMR_status, all_of(plot_cols)) %>%
  pivot_longer(cols = all_of(plot_cols),
               names_to = "interaction", values_to = "strength") %>%
  filter(!is.na(strength), !is.na(MMR_status))

ggplot(plot_df, aes(x = interaction, y = strength, fill = MMR_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.75)) +
  geom_jitter(aes(color = MMR_status),
              position = position_jitterdodge(jitter.width = 0.15),
              size = 1, alpha = 0.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) +
  labs(title = "Immune–Tumor–Fibroblast interactions: dMMR vs pMMR",
       subtitle = paste0("N = ", nrow(interaction_summary), " samples"),
       y = "Interaction strength", x = "") +
  scale_fill_manual(values = c("dMMR" = "#E41A1C", "pMMR" = "#377EB8")) +
  scale_color_manual(values = c("dMMR" = "#E41A1C", "pMMR" = "#377EB8"))

# ============================================================
# VISUALIZE: Boxplots by Metastatic status
# ============================================================
plot_df_met <- interaction_clinical %>%
  select(sample_id, met_group, all_of(plot_cols)) %>%
  pivot_longer(cols = all_of(plot_cols),
               names_to = "interaction", values_to = "strength") %>%
  filter(!is.na(strength), !is.na(met_group))

ggplot(plot_df_met, aes(x = interaction, y = strength, fill = met_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.75)) +
  geom_jitter(aes(color = met_group),
              position = position_jitterdodge(jitter.width = 0.15),
              size = 1, alpha = 0.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) +
  labs(title = "Immune–Tumor–Fibroblast interactions by Metastatic Status",
       y = "Interaction strength", x = "") +
  scale_fill_manual(values = c("No_Mets" = "#00AFBB", "LN_Mets" = "#E7B800", 
                                "Distant_Mets" = "#FC4E07")) +
  scale_color_manual(values = c("No_Mets" = "#00AFBB", "LN_Mets" = "#E7B800", 
                                 "Distant_Mets" = "#FC4E07"))

# ============================================================
# STATISTICAL TESTS
# ============================================================
cat("\n=== Wilcoxon tests: dMMR vs pMMR ===\n")
for (col in plot_cols) {
  vals_dmmr <- interaction_clinical[[col]][interaction_clinical$MMR_status == "dMMR"]
  vals_pmmr <- interaction_clinical[[col]][interaction_clinical$MMR_status == "pMMR"]
  vals_dmmr <- vals_dmmr[!is.na(vals_dmmr)]
  vals_pmmr <- vals_pmmr[!is.na(vals_pmmr)]
  
  if (length(vals_dmmr) >= 2 & length(vals_pmmr) >= 2) {
    wt <- wilcox.test(vals_dmmr, vals_pmmr)
    cat(sprintf("  %-25s p = %.4f  (dMMR med: %.4f, pMMR med: %.4f)\n",
                col, wt$p.value, median(vals_dmmr), median(vals_pmmr)))
  } else {
    cat(sprintf("  %-25s SKIPPED (insufficient data)\n", col))
  }
}

cat("\n=== Kruskal-Wallis: Metastatic groups ===\n")
for (col in plot_cols) {
  groups <- split(interaction_clinical[[col]], interaction_clinical$met_group)
  groups <- lapply(groups, function(x) x[!is.na(x)])
  groups <- groups[sapply(groups, length) >= 2]
  
  if (length(groups) >= 2) {
    kt <- kruskal.test(groups)
    cat(sprintf("  %-25s p = %.4f\n", col, kt$p.value))
  }
}

# ============================================================
# PATHWAY-LEVEL SUMMARY
# ============================================================
pathway_summary <- lapply(names(cellchat_list), function(sid) {
  cc <- cellchat_list[[sid]]
  if (is.null(cc@netP$prob)) return(NULL)
  
  pathways <- dimnames(cc@netP$prob)[[3]]
  
  pw_data <- lapply(pathways, function(pw) {
    tryCatch({
      prob_mat <- cc@netP$prob[,,pw]
      data.frame(
        sample_id  = sid,
        pathway    = pw,
        total_prob = sum(prob_mat, na.rm = TRUE),
        stringsAsFactors = FALSE
      )
    }, error = function(e) NULL)
  })
  
  do.call(rbind, pw_data)
})

pathway_df <- do.call(rbind, pathway_summary)

top_pathways <- pathway_df %>%
  group_by(pathway) %>%
  summarise(
    mean_prob        = mean(total_prob, na.rm = TRUE),
    median_prob      = median(total_prob, na.rm = TRUE),
    n_samples_active = sum(total_prob > 0)
  ) %>%
  arrange(desc(mean_prob)) %>%
  head(20)

cat("\nTop 20 active pathways across all samples:\n")
print(top_pathways)

# Save pathway data
write.csv(pathway_df, file.path(output_dir, "pathway_summary_all_samples.csv"),
          row.names = FALSE)
```


Part 10: Save Everything and Launch Explorer
```{r}
# ============================================================
# PART 10: SAVE AND EXPLORE
# ============================================================

# Save everything
saveRDS(cellchat_list, file.path(output_dir, "cellchat_all_41_samples.rds"))
saveRDS(interaction_summary, file.path(output_dir, "interaction_summary.rds"))
saveRDS(interaction_clinical, file.path(output_dir, "interaction_clinical.rds"))
saveRDS(pathway_df, file.path(output_dir, "pathway_summary.rds"))

cat("\nAll results saved to:", output_dir, "\n")
cat("Files:\n")
print(list.files(output_dir))

# --- Interactive exploration ---
cellchat_explore <- cellchat_list[["127"]]
runCellChatApp(cellchat_explore)
```
