---
title: "Cell_Type_Deconvolution_Analysis_DNA_Methylation"
author: "Alos Diallo"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# =========================================================================
# DNA methylation Cell Type Deconvolution using HiTIMED - COAD DH Dataset
# Code by: Alos Diallo
# 2023,2024
# =========================================================================


```{r message=FALSE, warning=FALSE, include=FALSE}
library(minfi)
library(sesame)
library(pheatmap)
library(minfiData)
library(FlowSorted.Blood.EPIC)
library(HiTIMED)
library(ggplot2)
library(dplyr)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICv2manifest)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICmanifest)
library(limma)
library(qvalue)
library(sva)
library(ENmix)
library(DataExplorer)
library(tidyr)
library(EpiDISH)
library(tibble)
```

Loading Data
```{r}
targets <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/dm_57_samples.csv")
targets$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets) <- targets$patient
targets$SampleID<- targets$patient


# Convert relevant columns in `targets` to factors
targets <- targets %>%
  mutate(
    MLH1 = factor(MLH1, levels = c(0, 1), labels = c("MLH1 Negative", "MLH1 Positive")),
    MSH2 = factor(MSH2),
    sex = factor(sex, levels = c("M", "F")),
    Distant_Mets = factor(Distant_Mets, levels = c(FALSE, TRUE), labels = c("No Distant Mets", "Distant Mets")),
    any_mets = factor(any_mets, levels = c(FALSE, TRUE), labels = c("No Mets", "Mets")),
    ln_only = factor(ln_only, levels = c(FALSE, TRUE), labels = c("No LN Mets", "LN Mets"))
  )
```

```{r}
idat = "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/"
RGset_32 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/",recursive = TRUE) 
#RGset_25 = openSesame(idat , func = getBetas) 
RGset_25 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/",recursive = TRUE) 
```


Normalize my Data
```{r}
Noob_25_m = preprocessNoob(RGset_25)
Noob_32_m = preprocessNoob(RGset_32)
```

Getting the beta values
```{r}
Betas_25<-getBeta(Noob_25_m)
Betas_32<-getBeta(Noob_32_m)
```

```{r}
Betas_25<- sesame::betasCollapseToPfx(Betas_25)
#Betas_32<- sesame::betasCollapseToPfx(Betas_32)
colnames(Betas_25) = colnames(Noob_25_m)
```

```{r}
DH_CRC_Betas <- merge(Betas_25, Betas_32, by = "row.names")
rownames(DH_CRC_Betas) <- DH_CRC_Betas$Row.names
DH_CRC_Betas <- DH_CRC_Betas[,-1] 
```


Loading the TCGA Data:
```{r warning=FALSE, include=FALSE}
tcga.data <- readRDS("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/TCGA_data.rds")
tcga.dnam <- tcga.data[["dnam"]]
tcga.pheno <- tcga.data[["pheno"]]
tcga.cell <- tcga.data[["cell_types"]]
msi_dat<-readRDS("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/msi_tcga.rds")
tcga.cell$COAD_Deconv1#$MSS
tcga.cell$COAD_Deconv2#MASS

tcga_T = t(tcga.dnam)

```

Running EpiDish to obtain Fibroblast information
```{r}
# Define sources
sources <- list(
  DH = DH_CRC_Betas,
  TCGA = tcga_T
  # GTEx = GTEx_T
)

# Define the reference matrix (e.g., centEpiFibIC.m from EpiDISH)
ref_matrix <- centEpiFibIC.m

# Function to run EpiDISH on a source
run_epidish <- function(beta_matrix, source_name) {
  # Run EpiDISH
  results <- epidish(beta.m = beta_matrix, ref.m = ref_matrix, method = "RPC")
  
  # Extract estimated cell type proportions
  cell_type_proportions <- results$estF
  
  # Convert matrix to a data frame and scale proportions to percentages
  cell_type_proportions_df <- as.data.frame(cell_type_proportions) * 100
  
  # Add metadata columns
  cell_type_proportions_df$Source <- source_name
  cell_type_proportions_df$SampleID <- rownames(cell_type_proportions)
  
  return(cell_type_proportions_df)
}

# Apply EpiDISH to all sources
epidish_results <- lapply(names(sources), function(source_name) {
  run_epidish(sources[[source_name]], source_name)
})

# Combine results from all sources into a single data frame
epidish_combined_results <- do.call(rbind, epidish_results)

# Inspect the combined results
head(epidish_combined_results)
```


Running HiTIMED
```{r include=FALSE}
sources <- list(
  DH = DH_CRC_Betas,
  TCGA = tcga_T
  #GTEx = GTEx_T
)

all_results <- lapply(names(sources), function(source_name) {
  lapply(1:6, function(i) {
    result <- HiTIMED_deconvolution(sources[[source_name]], "COAD", i, "tumor")
    result$Source <- source_name
    result
  })
})

names(all_results) <- names(sources)  # Label results by source
```

```{r}
data_levels <- lapply(1:6, function(level) {
  bind_rows(
    all_results$DH[[level]],
    all_results$TCGA[[level]],
    all_results$GTEx[[level]]
  )
})
names(data_levels) <- paste0("Level_", 1:6)

data_levels <- lapply(1:6, function(level) {
  data <- data_levels[[level]]
  data$ID <- paste0("Level_", level)
  data
})

# Combine all levels into one dataset
deconvolution_results <- bind_rows(data_levels)
deconvolution_results$SampleID <- rownames(deconvolution_results)

```

Rearranging the data so that I can plot it properly.
```{r}
# Reshape data to long format
long_data <- deconvolution_results %>%
  pivot_longer(
    -c(SampleID, ID, Source),  # Exclude SampleID, ID, and Source from reshaping
    names_to = "CellType",
    values_to = "Proportion"
  )


# Define HiTIMED levels and their corresponding cell types
hitimed_levels <- list(
  Level_1 = c("Tumor", "Other"),
  Level_2 = c("Tumor", "Immune", "Angiogenic"),
  Level_3 = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal"),
  Level_4 = c("Tumor", "Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK", 
              "Endothelial", "Epithelial", "Stromal"),
  Level_5 = c("Tumor", "Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", 
              "CD4T", "CD8T", "NK", "Endothelial", "Epithelial", "Stromal"),
  Level_6 = c("Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv", "Bmem", 
              "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC", "NK", 
              "Bas", "Eos", "Neu")
)

```

Printing out the box plot for the DH data and TCGA data
```{r}
# Function to plot cell types for a specific HiTIMED level and source
plot_hitimed_level <- function(data, level, source = NULL) {
  # Get the cell types for the specified level
  selected_cell_types <- hitimed_levels[[level]]
  
  if (is.null(selected_cell_types)) {
    warning(paste("Level", level, "is not defined in hitimed_levels. Skipping this level."))
    return(NULL)
  }
  
  # Filter the data for the specific level and cell types
  filtered_data <- data %>%
    filter(CellType %in% selected_cell_types)
  
  # Further filter by Source if specified
  if (!is.null(source)) {
    filtered_data <- filtered_data %>%
      filter(Source == source)
  }
  
  if (nrow(filtered_data) == 0) {
    warning(paste("No data available for Level", level, "after filtering."))
    return(NULL)
  }
  
  # Plot the data without grouping by any variable
  plot <- ggplot(filtered_data, aes(x = CellType, y = Proportion)) +
    geom_boxplot(fill = "skyblue") +  # Use a fixed fill color
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"  # Remove legend
    ) +
    labs(
      x = "Cell Type",
      y = "Cell Type Proportion",
      title = paste("Distribution of Cell Types for HiTIMED", level, "(", source, ")")
    )
  
  return(plot)
}

# Example usage with all levels, for DH or TCGA data only
all_levels <- names(hitimed_levels)

# Generate plots for DH data
for (level in all_levels) {
  plot <- plot_hitimed_level(data = long_data, level = level, source = "DH")
  
  # Save only if the plot is successfully created
  if (!is.null(plot)) {
    ggsave(paste0("HiTIMED_", level, "_Boxplot_DH.png"), plot, width = 10, height = 8)
    print(plot)  # Display the plot in R
  }
}

# Generate plots for TCGA data
for (level in all_levels) {
  plot <- plot_hitimed_level(data = long_data, level = level, source = "TCGA")
  
  # Save only if the plot is successfully created
  if (!is.null(plot)) {
    ggsave(paste0("HiTIMED_", level, "_Boxplot_TCGA.png"), plot, width = 10, height = 8)
    print(plot)  # Display the plot in R
  }
}
```


```{r}
# Extract SampleID (everything before the ellipsis)
deconvolution_results$SampleID <- gsub("\\.\\.\\..*", "", rownames(deconvolution_results))

# Extract HiTIMED_Level (everything after the last underscore and before the ellipsis)
deconvolution_results$HiTIMED_Level <- gsub(".*\\.\\.\\.(.*)", "\\1", rownames(deconvolution_results))

# Verify the results
head(deconvolution_results[, c("SampleID", "HiTIMED_Level")])

# Ensure targets has a SampleID column
targets$SampleID <- rownames(targets)

# Merge deconvolution_results and targets based on the new SampleID column
combined_results <- deconvolution_results %>%
  inner_join(targets, by = "SampleID")

# Check the structure of the combined dataframe
str(combined_results)

# Filter for DH data
dh_results <- combined_results %>%
  filter(Source == "DH")

# Reshape DH data into long format
reshaped_dh_data <- dh_results %>%
  pivot_longer(
    cols = where(is.numeric),  # Only numeric columns
    names_to = "CellType",
    values_to = "Proportion"
  )

# Define HiTIMED levels and their corresponding cell types
hitimed_levels <- list(
  Level_1 = c("Tumor", "Other"),
  Level_2 = c("Tumor", "Immune", "Angiogenic"),
  Level_3 = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal"),
  Level_4 = c("Tumor", "Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK", 
              "Endothelial", "Epithelial", "Stromal"),
  Level_5 = c("Tumor", "Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", 
              "CD4T", "CD8T", "NK", "Endothelial", "Epithelial", "Stromal"),
  Level_6 = c("Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv", "Bmem", 
              "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC", "NK", 
              "Bas", "Eos", "Neu")
)
```



```{r}
# Function to plot HiTIMED levels with stratification
plot_stratified_hitimed <- function(data, level, stratify_by = NULL) {
  # Get cell types for the specified HiTIMED level
  selected_cell_types <- hitimed_levels[[level]]
  
  # Filter data for specified cell types
  level_data <- data %>%
    filter(CellType %in% selected_cell_types)
  
  if (nrow(level_data) == 0) {
    warning(paste("No data available for Level", level, "after filtering."))
    return(NULL)
  }
  
  # Generate plot
  plot <- ggplot(level_data, aes(x = CellType, y = Proportion)) +
    geom_boxplot(aes(fill = .data[[stratify_by]])) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    ) +
    labs(
      fill = stratify_by,
      x = "Cell Type",
      y = "Cell Type Proportion",
      title = paste("HiTIMED Level", level, "-", stratify_by, "Stratification")
    )
  
  return(plot)
}

# Example usage: Plot for Level_1, stratified by 'any_mets'
plot <- plot_stratified_hitimed(data = reshaped_dh_data, level = "Level_1", stratify_by = "any_mets")
if (!is.null(plot)) {
  ggsave("HiTIMED_Level_1_Stratified_by_any_mets.png", plot, width = 10, height = 8)
  print(plot)
}
```
Here I am combining what I have from HiTIMED and EpiDISH to plot cell type proportions 
```{r}
# Step 1: Extract fibroblast information from EpiDISH results
fibroblast_data <- epidish_combined_results %>%
  select(SampleID, Source, Fibroblasts = Fib)  # Rename 'Fib' to 'Fibroblasts' for clarity

# Step 2: Extract HiTIMED Level 6 results
hitimed_level_6 <- lapply(all_results, function(source_results) {
  source_result <- source_results[[6]]  # Extract Level 6
  
  # Convert row names into a SampleID column
  source_result <- source_result %>%
    rownames_to_column(var = "SampleID")  # Move row names into "SampleID"
  
  # Remove 'Stromal' column if it exists and keep all other columns
  source_result %>%
    select(-Stromal, everything())  # Exclude Stromal cells, keep all others
})

# Combine HiTIMED Level 6 results into a single data frame
hitimed_level_6 <- bind_rows(hitimed_level_6)

# Step 3: Merge HiTIMED Level 6 with EpiDISH fibroblast data
combined_data <- hitimed_level_6 %>%
  left_join(fibroblast_data, by = c("SampleID", "Source"))  # Match by SampleID and Source

# Step 4: Merge with DH metadata (from `targets`) for stratification
combined_data <- combined_data %>%
  inner_join(targets, by = "SampleID")  # Add metadata like 'any_mets'

# Step 5: Filter for DH data only
dh_combined_data <- combined_data %>%
  filter(Source == "DH")  # Keep only DH samples

# Step 6: Reshape data for visualization (include only specific cell types)
cell_types_to_include <- c("Tumor", "Endothelial", "Epithelial", "Fibroblasts", "Bnv", "Bmem", 
                           "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC", "NK", 
                           "Bas", "Eos", "Neu")

long_dh_data <- dh_combined_data %>%
  pivot_longer(
    cols = where(is.numeric),  # Include only numeric columns for reshaping
    names_to = "CellType",
    values_to = "Proportion"
  ) %>%
  filter(CellType %in% cell_types_to_include)  # Include only selected cell types

# Ensure CellType is a factor with a meaningful order
long_dh_data <- long_dh_data %>%
  mutate(CellType = factor(CellType, levels = cell_types_to_include))  # Set order of CellType

# Generate boxplots for selected cell types stratified by 'any_mets'
all_celltypes_boxplot <- ggplot(long_dh_data, aes(x = CellType, y = Proportion, fill = any_mets)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers for clarity
  theme_minimal() +
  labs(
    title = "Selected Cell Type Proportions Stratified by Any Metastasis",
    x = "Cell Type",
    y = "Proportion",
    fill = "Metastasis Status"
  ) +
  scale_fill_brewer(palette = "Set2") +  # Use a color palette
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

# Display the plot
print(all_celltypes_boxplot)

# Save the plot
ggsave("Selected_CellTypes_Boxplot_DH_AnyMets.png", all_celltypes_boxplot, width = 12, height = 8)

# OPTIONAL: Faceted version
facet_celltypes_boxplot <- ggplot(long_dh_data, aes(x = any_mets, y = Proportion, fill = any_mets)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~ CellType, scales = "free_y") +  # Create separate panels for each cell type
  theme_minimal() +
  labs(
    title = "Selected Cell Type Proportions by Any Metastasis (Faceted)",
    x = "Metastasis Status",
    y = "Proportion",
    fill = "Metastasis Status"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")  # Format facet labels
  )

# Display the faceted plot
print(facet_celltypes_boxplot)

# Save the faceted plot
ggsave("Facet_Selected_CellTypes_Boxplot_DH_AnyMets.png", facet_celltypes_boxplot, width = 16, height = 10)

```



