---
title: "Cell_Type_Deconvolution_Analysis_DNA_Methylation"
author: "Alos Diallo"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# =========================================================================
# DNA methylation Cell Type Deconvolution using HiTIMED - COAD DH Dataset
# Code by: Alos Diallo
# 2023,2024
# =========================================================================


```{r message=FALSE, warning=FALSE, include=FALSE}
library(minfi)
library(sesame)
library(pheatmap)
library(minfiData)
library(FlowSorted.Blood.EPIC)
library(HiTIMED)
library(ggplot2)
library(dplyr)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICv2manifest)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICmanifest)
library(limma)
library(qvalue)
library(sva)
library(ENmix)
library(DataExplorer)
library(tidyr)
library(EpiDISH)
library(tibble)
library(vegan)
library(rstatix)
```

Loading Data
```{r}
targets <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/dm_57_samples.csv")
targets$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets) <- targets$patient
targets$SampleID<- targets$patient


# Convert relevant columns in `targets` to factors
targets <- targets %>%
  mutate(
    MLH1 = factor(MLH1, levels = c(0, 1), labels = c("MLH1 Negative", "MLH1 Positive")),
    MSH2 = factor(MSH2),
    sex = factor(sex, levels = c("M", "F")),
    Distant_Mets = factor(Distant_Mets, levels = c(FALSE, TRUE), labels = c("No Distant Mets", "Distant Mets")),
    any_mets = factor(any_mets, levels = c(FALSE, TRUE), labels = c("No Mets", "Mets")),
    ln_only = factor(ln_only, levels = c(FALSE, TRUE), labels = c("No LN Mets", "LN Mets"))
  )
```

```{r}
idat = "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/"
RGset_32 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/",recursive = TRUE) 
#RGset_25 = openSesame(idat , func = getBetas) 
RGset_25 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/",recursive = TRUE) 
```


Normalize my Data
```{r}
Noob_25_m = preprocessNoob(RGset_25)
Noob_32_m = preprocessNoob(RGset_32)
```

Getting the beta values
```{r}
Betas_25<-getBeta(Noob_25_m)
Betas_32<-getBeta(Noob_32_m)
```

```{r}
Betas_25<- sesame::betasCollapseToPfx(Betas_25)
#Betas_32<- sesame::betasCollapseToPfx(Betas_32)
colnames(Betas_25) = colnames(Noob_25_m)
```

```{r}
DH_CRC_Betas <- merge(Betas_25, Betas_32, by = "row.names")
rownames(DH_CRC_Betas) <- DH_CRC_Betas$Row.names
DH_CRC_Betas <- DH_CRC_Betas[,-1] 
```


Loading the TCGA Data:
```{r warning=FALSE, include=FALSE}
tcga.data <- readRDS("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/TCGA_data.rds")
tcga.dnam <- tcga.data[["dnam"]]
tcga.pheno <- tcga.data[["pheno"]]
tcga.cell <- tcga.data[["cell_types"]]
msi_dat<-readRDS("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/msi_tcga.rds")
tcga.cell$COAD_Deconv1#$MSS
tcga.cell$COAD_Deconv2#MASS

tcga_T = t(tcga.dnam)

```

Running EpiDish to obtain Fibroblast information
```{r}
# Define sources
sources <- list(
  DH = DH_CRC_Betas,
  TCGA = tcga_T
  # GTEx = GTEx_T
)

# Define the reference matrix (e.g., centEpiFibIC.m from EpiDISH)
ref_matrix <- centEpiFibIC.m

# Function to run EpiDISH on a source
run_epidish <- function(beta_matrix, source_name) {
  # Run EpiDISH
  results <- epidish(beta.m = beta_matrix, ref.m = ref_matrix, method = "RPC")
  
  # Extract estimated cell type proportions
  cell_type_proportions <- results$estF
  
  # Convert matrix to a data frame and scale proportions to percentages
  cell_type_proportions_df <- as.data.frame(cell_type_proportions) * 100
  
  # Add metadata columns
  cell_type_proportions_df$Source <- source_name
  cell_type_proportions_df$SampleID <- rownames(cell_type_proportions)
  
  return(cell_type_proportions_df)
}

# Apply EpiDISH to all sources
epidish_results <- lapply(names(sources), function(source_name) {
  run_epidish(sources[[source_name]], source_name)
})

# Combine results from all sources into a single data frame
epidish_combined_results <- do.call(rbind, epidish_results)

# Inspect the combined results
head(epidish_combined_results)
```


Running HiTIMED
```{r include=FALSE}
sources <- list(
  DH = DH_CRC_Betas,
  TCGA = tcga_T
  #GTEx = GTEx_T
)

all_results <- lapply(names(sources), function(source_name) {
  lapply(1:6, function(i) {
    result <- HiTIMED_deconvolution(sources[[source_name]], "COAD", i, "tumor")
    result$Source <- source_name
    result
  })
})

names(all_results) <- names(sources)  # Label results by source
```

```{r}
data_levels <- lapply(1:6, function(level) {
  bind_rows(
    all_results$DH[[level]],
    all_results$TCGA[[level]],
    all_results$GTEx[[level]]
  )
})
names(data_levels) <- paste0("Level_", 1:6)

data_levels <- lapply(1:6, function(level) {
  data <- data_levels[[level]]
  data$ID <- paste0("Level_", level)
  data
})

# Combine all levels into one dataset
deconvolution_results <- bind_rows(data_levels)
deconvolution_results$SampleID <- rownames(deconvolution_results)

```

Rearranging the data so that I can plot it properly.
```{r}
# Reshape data to long format
long_data <- deconvolution_results %>%
  pivot_longer(
    -c(SampleID, ID, Source),  # Exclude SampleID, ID, and Source from reshaping
    names_to = "CellType",
    values_to = "Proportion"
  )


# Define HiTIMED levels and their corresponding cell types
hitimed_levels <- list(
  Level_1 = c("Tumor", "Other"),
  Level_2 = c("Tumor", "Immune", "Angiogenic"),
  Level_3 = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal"),
  Level_4 = c("Tumor", "Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK", 
              "Endothelial", "Epithelial", "Stromal"),
  Level_5 = c("Tumor", "Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", 
              "CD4T", "CD8T", "NK", "Endothelial", "Epithelial", "Stromal"),
  Level_6 = c("Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv", "Bmem", 
              "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC", "NK", 
              "Bas", "Eos", "Neu")
)

```

Printing out the box plot for the DH data and TCGA data
```{r}
# Function to plot cell types for a specific HiTIMED level and source
plot_hitimed_level <- function(data, level, source = NULL) {
  # Get the cell types for the specified level
  selected_cell_types <- hitimed_levels[[level]]
  
  if (is.null(selected_cell_types)) {
    warning(paste("Level", level, "is not defined in hitimed_levels. Skipping this level."))
    return(NULL)
  }
  
  # Filter the data for the specific level and cell types
  filtered_data <- data %>%
    filter(CellType %in% selected_cell_types)
  
  # Further filter by Source if specified
  if (!is.null(source)) {
    filtered_data <- filtered_data %>%
      filter(Source == source)
  }
  
  if (nrow(filtered_data) == 0) {
    warning(paste("No data available for Level", level, "after filtering."))
    return(NULL)
  }
  
  # Plot the data without grouping by any variable
  plot <- ggplot(filtered_data, aes(x = CellType, y = Proportion)) +
    geom_boxplot(fill = "skyblue") +  # Use a fixed fill color
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"  # Remove legend
    ) +
    labs(
      x = "Cell Type",
      y = "Cell Type Proportion",
      title = paste("Distribution of Cell Types for HiTIMED", level, "(", source, ")")
    )
  
  return(plot)
}

# Example usage with all levels, for DH or TCGA data only
all_levels <- names(hitimed_levels)

# Generate plots for DH data
for (level in all_levels) {
  plot <- plot_hitimed_level(data = long_data, level = level, source = "DH")
  
  # Save only if the plot is successfully created
  if (!is.null(plot)) {
    ggsave(paste0("HiTIMED_", level, "_Boxplot_DH.png"), plot, width = 10, height = 8)
    print(plot)  # Display the plot in R
  }
}

# Generate plots for TCGA data
for (level in all_levels) {
  plot <- plot_hitimed_level(data = long_data, level = level, source = "TCGA")
  
  # Save only if the plot is successfully created
  if (!is.null(plot)) {
    ggsave(paste0("HiTIMED_", level, "_Boxplot_TCGA.png"), plot, width = 10, height = 8)
    print(plot)  # Display the plot in R
  }
}
```


```{r}
# Extract SampleID (everything before the ellipsis)
deconvolution_results$SampleID <- gsub("\\.\\.\\..*", "", rownames(deconvolution_results))

# Extract HiTIMED_Level (everything after the last underscore and before the ellipsis)
deconvolution_results$HiTIMED_Level <- gsub(".*\\.\\.\\.(.*)", "\\1", rownames(deconvolution_results))

# Verify the results
head(deconvolution_results[, c("SampleID", "HiTIMED_Level")])

# Ensure targets has a SampleID column
targets$SampleID <- rownames(targets)

# Merge deconvolution_results and targets based on the new SampleID column
combined_results <- deconvolution_results %>%
  inner_join(targets, by = "SampleID")

# Check the structure of the combined dataframe
str(combined_results)

# Filter for DH data
dh_results <- combined_results %>%
  filter(Source == "DH")

# Reshape DH data into long format
reshaped_dh_data <- dh_results %>%
  pivot_longer(
    cols = where(is.numeric),  # Only numeric columns
    names_to = "CellType",
    values_to = "Proportion"
  )

# Define HiTIMED levels and their corresponding cell types
hitimed_levels <- list(
  Level_1 = c("Tumor", "Other"),
  Level_2 = c("Tumor", "Immune", "Angiogenic"),
  Level_3 = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal"),
  Level_4 = c("Tumor", "Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK", 
              "Endothelial", "Epithelial", "Stromal"),
  Level_5 = c("Tumor", "Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", 
              "CD4T", "CD8T", "NK", "Endothelial", "Epithelial", "Stromal"),
  Level_6 = c("Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv", "Bmem", 
              "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC", "NK", 
              "Bas", "Eos", "Neu")
)
```



```{r}
# Function to plot HiTIMED levels with stratification
plot_stratified_hitimed <- function(data, level, stratify_by = NULL) {
  # Get cell types for the specified HiTIMED level
  selected_cell_types <- hitimed_levels[[level]]
  
  # Filter data for specified cell types
  level_data <- data %>%
    filter(CellType %in% selected_cell_types)
  
  if (nrow(level_data) == 0) {
    warning(paste("No data available for Level", level, "after filtering."))
    return(NULL)
  }
  
  # Generate plot
  plot <- ggplot(level_data, aes(x = CellType, y = Proportion)) +
    geom_boxplot(aes(fill = .data[[stratify_by]])) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    ) +
    labs(
      fill = stratify_by,
      x = "Cell Type",
      y = "Cell Type Proportion",
      title = paste("HiTIMED Level", level, "-", stratify_by, "Stratification")
    )
  
  return(plot)
}

# Example usage: Plot for Level_1, stratified by 'any_mets'
plot <- plot_stratified_hitimed(data = reshaped_dh_data, level = "Level_5", stratify_by = "any_mets")
if (!is.null(plot)) {
  ggsave("HiTIMED_Level_1_Stratified_by_any_mets.png", plot, width = 10, height = 8)
  print(plot)
}
```
Here I am combining what I have from HiTIMED and EpiDISH to plot cell type proportions 
```{r}
# Step 1: Extract fibroblast information from EpiDISH results
fibroblast_data <- epidish_combined_results %>%
  select(SampleID, Source, Fibroblasts = Fib)  # Rename 'Fib' to 'Fibroblasts' for clarity

# Step 2: Extract HiTIMED Level 6 results
hitimed_level_6 <- lapply(all_results, function(source_results) {
  source_result <- source_results[[6]]  # Extract Level 6
  
  # Convert row names into a SampleID column
  source_result <- source_result %>%
    rownames_to_column(var = "SampleID")  # Move row names into "SampleID"
  
  # Remove 'Stromal' column if it exists and keep all other columns
  source_result %>%
    select(-Stromal, everything())  # Exclude Stromal cells, keep all others
})

# Combine HiTIMED Level 6 results into a single data frame
hitimed_level_6 <- bind_rows(hitimed_level_6)

# Step 3: Merge HiTIMED Level 6 with EpiDISH fibroblast data
combined_data <- hitimed_level_6 %>%
  left_join(fibroblast_data, by = c("SampleID", "Source"))  # Match by SampleID and Source

# Step 4: Merge with DH metadata (from `targets`) for stratification
combined_data <- combined_data %>%
  inner_join(targets, by = "SampleID")  # Add metadata like 'any_mets'

# Step 5: Filter for DH data only
dh_combined_data <- combined_data %>%
  filter(Source == "DH")  # Keep only DH samples

# Step 6: Reshape data for visualization (include only specific cell types)
cell_types_to_include <- c("Tumor", "Endothelial", "Epithelial", "Fibroblasts", "Bnv", "Bmem", 
                           "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC", "NK", 
                           "Bas", "Eos", "Neu")

long_dh_data <- dh_combined_data %>%
  pivot_longer(
    cols = where(is.numeric),  # Include only numeric columns for reshaping
    names_to = "CellType",
    values_to = "Proportion"
  ) %>%
  filter(CellType %in% cell_types_to_include)  # Include only selected cell types

# Ensure CellType is a factor with a meaningful order
long_dh_data <- long_dh_data %>%
  mutate(CellType = factor(CellType, levels = cell_types_to_include))  # Set order of CellType

# Generate boxplots for selected cell types stratified by 'any_mets'
all_celltypes_boxplot <- ggplot(long_dh_data, aes(x = CellType, y = Proportion, fill = any_mets)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers for clarity
  theme_minimal() +
  labs(
    title = "Selected Cell Type Proportions Stratified by Any Metastasis",
    x = "Cell Type",
    y = "Proportion",
    fill = "Metastasis Status"
  ) +
  scale_fill_brewer(palette = "Set2") +  # Use a color palette
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

# Display the plot
print(all_celltypes_boxplot)

# Save the plot
ggsave("Selected_CellTypes_Boxplot_DH_AnyMets.png", all_celltypes_boxplot, width = 12, height = 8)

# OPTIONAL: Faceted version
facet_celltypes_boxplot <- ggplot(long_dh_data, aes(x = any_mets, y = Proportion, fill = any_mets)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~ CellType, scales = "free_y") +  # Create separate panels for each cell type
  theme_minimal() +
  labs(
    title = "Selected Cell Type Proportions by Any Metastasis (Faceted)",
    x = "Metastasis Status",
    y = "Proportion",
    fill = "Metastasis Status"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")  # Format facet labels
  )

# Display the faceted plot
print(facet_celltypes_boxplot)

# Save the faceted plot
ggsave("Facet_Selected_CellTypes_Boxplot_DH_AnyMets.png", facet_celltypes_boxplot, width = 16, height = 10)

```
Here we are looking at Cell Type distrobutions but my mets status, but not stratified by MLH1.
```{r}
# Step 1: Filter and categorize data for "No Mets," "LN Mets," and "Distant Mets"
stratified_data <- dh_combined_data %>%
  mutate(
    Metastasis_Group = case_when(
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Mets",
      Distant_Mets == "Distant Mets" ~ "Distant Mets",
      any_mets == "No Mets" ~ "No Mets"
    )
  ) %>%
  filter(!is.na(Metastasis_Group))  # Ensure no missing values in grouping

# Step 2: Reshape data for visualization (HiTIMED Level 6 results, including specific cell types)
long_stratified_data <- stratified_data %>%
  pivot_longer(
    cols = where(is.numeric),  # Include only numeric columns
    names_to = "CellType",
    values_to = "Proportion"
  ) %>%
  filter(CellType %in% cell_types_to_include) %>%  # Keep only selected cell types
  mutate(
    CellType = factor(CellType, levels = cell_types_to_include),  # Order CellType
    Metastasis_Group = factor(Metastasis_Group, levels = c("No Mets", "LN Mets", "Distant Mets"))  # Order groups
  )

# Step 3: Generate boxplots for each cell type with three metastasis groups
stratified_boxplot <- ggplot(long_stratified_data, aes(x = Metastasis_Group, y = Proportion, fill = Metastasis_Group)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers for clarity
  facet_wrap(~ CellType, scales = "free_y", ncol = 3) +  # Create facets for each cell type
  theme_minimal() +
  labs(
    title = "Cell Type Proportions Stratified by Metastasis Status",
    x = "Metastasis Group",
    y = "Proportion",
    fill = "Metastasis Group"
  ) +
  scale_fill_brewer(palette = "Set2") +  # Use a color palette
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")  # Format facet labels
  )

# Step 4: Display and save the plot
print(stratified_boxplot)
ggsave("Stratified_CellType_Boxplot.png", stratified_boxplot, width = 16, height = 12)
```

Plots for each cell type individually saved.  I did this because it was hard to see the plots when they were all together.  This works, but there are no stats and or jiggers.
```{r}
# Step 1: Filter and categorize data for MLH1 status and metastasis groups
stratified_ml_data <- dh_combined_data %>%
  mutate(
    Metastasis_Group = case_when(
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Mets",
      Distant_Mets == "Distant Mets" ~ "Distant Mets",
      any_mets == "No Mets" ~ "No Mets"
    )
  ) %>%
  filter(!is.na(Metastasis_Group), !is.na(MLH1))  # Ensure no missing values

# Step 2: Reshape data for visualization
long_ml_data <- stratified_ml_data %>%
  pivot_longer(
    cols = where(is.numeric),  # Include only numeric columns
    names_to = "CellType",
    values_to = "Proportion"
  ) %>%
  filter(CellType %in% cell_types_to_include) %>%  # Include specific cell types
  mutate(
    CellType = factor(CellType, levels = cell_types_to_include),  # Order cell types
    Metastasis_Group = factor(Metastasis_Group, levels = c("No Mets", "LN Mets", "Distant Mets")),  # Order groups
    MLH1 = factor(MLH1, levels = c("MLH1 Negative", "MLH1 Positive"))  # Order MLH1
  )

# Create a directory to save the plots (optional)
output_dir <- "MLH1_CellType_Plots"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Loop through each cell type and create a separate plot
unique_cell_types <- unique(long_ml_data$CellType)

for (cell_type in unique_cell_types) {
  # Filter data for the current cell type
  cell_type_data <- long_ml_data %>% filter(CellType == cell_type)
  
  # Create the plot
  cell_type_plot <- ggplot(cell_type_data, aes(x = Metastasis_Group, y = Proportion, fill = Metastasis_Group)) +
    geom_boxplot(outlier.shape = NA) +  # Remove outliers for clarity
    facet_wrap(~ MLH1, scales = "free_y") +  # Separate facets by MLH1 status
    theme_minimal(base_size = 14) +  # Increase base font size for readability
    labs(
      title = paste("Proportions of", cell_type, "Stratified by Metastasis and MLH1 Status"),
      x = "Metastasis Group",
      y = "Proportion",
      fill = "Metastasis Group"
    ) +
    scale_fill_brewer(palette = "Set2") +  # Use a color palette
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Larger x-axis text
      legend.position = "bottom",
      legend.title = element_text(size = 14),  # Larger legend title
      legend.text = element_text(size = 12),  # Larger legend text
      strip.text = element_text(size = 14, face = "bold")  # Larger facet labels
    )
  
  # Save each plot with a unique file name
  file_name <- paste0(output_dir, "/", cell_type, "_MLH1_Boxplot.png")
  ggsave(file_name, cell_type_plot, width = 12, height = 8, dpi = 300)
}
```



Setting up date for Permanova analysis
```{r}
# Define the cell types to include based on my analysis plan
cell_types_to_include <- c(
  "Tumor", "Endothelial", "Epithelial", "Fibroblasts",
  "Bnv", "Bmem", "CD4nv", "CD4mem", "Treg",
  "CD8nv", "CD8mem", "Mono", "DC", "NK",
  "Bas", "Eos", "Neu"
)

# Verify that all specified cell types are present in the data
missing_cols <- setdiff(cell_types_to_include, colnames(dh_combined_data))
if(length(missing_cols) > 0){
  stop(paste("Missing cell type columns:", paste(missing_cols, collapse = ", ")))
}

# Extract the cell type proportion data
cell_type_matrix <- dh_combined_data %>%
  select(all_of(cell_types_to_include)) %>%
  as.matrix()

# Ensure all cell type columns are numeric
if(!all(sapply(cell_type_matrix, is.numeric))){
  stop("Not all cell type proportion columns are numeric.")
}

# Set row names to SampleID for reference (optional)
rownames(cell_type_matrix) <- dh_combined_data$SampleID

# Ensure 'any_mets' is a factor with appropriate levels
dh_combined_data$any_mets <- factor(dh_combined_data$any_mets, levels = c("No Mets", "Mets"))
```

Perform PCoA
```{r}
# Compute Bray-Curtis distance matrix
bray_dist <- vegdist(cell_type_matrix, method = "bray")

# Perform PCoA
pcoa_results <- cmdscale(bray_dist, eig = TRUE, k = 2)

# Create a dataframe for plotting
pcoa_df <- data.frame(
  SampleID = rownames(cell_type_matrix),
  PC1 = pcoa_results$points[,1],
  PC2 = pcoa_results$points[,2],
  any_mets = dh_combined_data$any_mets
)

# Plot PCoA
ggplot(pcoa_df, aes(x = PC1, y = PC2, color = any_mets)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "PCoA of Cell Type Proportions",
    x = "PC1",
    y = "PC2",
    color = "Metastasis Status"
  ) +
  scale_color_brewer(palette = "Set1")
```

```{r}
# Assuming 'bray_dist' is your distance matrix and 'cell_type_matrix' is your data
# Perform PCoA using vegan's cmdscale or ape's pcoa
pcoa_results <- cmdscale(bray_dist, eig = TRUE, k = 2)

# Extract eigenvectors (coordinates)
pc1 <- pcoa_results$points[,1]
pc2 <- pcoa_results$points[,2]

# Calculate correlations between cell types and principal coordinates
correlations <- cor(cell_type_matrix, cbind(pc1, pc2))
print(correlations)
```
```{r}
# Calculate species scores (variable loadings)
species_scores <- envfit(pcoa_results, cell_type_matrix, permutations = 999)
species_vectors <- as.data.frame(scores(species_scores, display = "vectors"))
species_vectors$CellType <- rownames(species_vectors)
# Using dplyr
library(dplyr)

species_vectors <- species_vectors %>%
  rename(
    PC1 = Dim1,
    PC2 = Dim2
  )


# Create PCoA dataframe
pcoa_df <- data.frame(
  SampleID = rownames(cell_type_matrix),
  PC1 = pcoa_results$points[,1],
  PC2 = pcoa_results$points[,2],
  any_mets = dh_combined_data$any_mets
)

ggplot(pcoa_df, aes(x = PC1, y = PC2, color = any_mets)) +
  geom_point(alpha = 0.7) +
  
  # geom_segment with inherit.aes = FALSE
  geom_segment(
    data = species_vectors,
    aes(x = 0, y = 0, xend = PC1 * max(pcoa_df$PC1), yend = PC2 * max(pcoa_df$PC2)),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black",
    inherit.aes = FALSE  # Prevent inheritance of main aes
  ) +
  
  # geom_text with inherit.aes = FALSE
  geom_text(
    data = species_vectors,
    aes(x = PC1 * max(pcoa_df$PC1) * 1.1, y = PC2 * max(pcoa_df$PC2) * 1.1, label = CellType),
    size = 3,
    inherit.aes = FALSE  # Prevent inheritance of main aes
  ) +
  
  theme_minimal() +
  labs(
    title = "PCoA of Cell Type Proportions with Biplot Vectors",
    x = "PCoA1",
    y = "PCoA2",
    color = "Metastasis Status"
  )
```
Here I am generating a PCoA plot but for MLH1+ and MLH1- seperately. 
```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(vegan)
library(ggpubr)  # Optional, for advanced annotations

# Step 1: Compute Bray-Curtis distance matrix
bray_dist <- vegdist(cell_type_matrix, method = "bray")

# Step 2: Perform PCoA
pcoa_results <- cmdscale(bray_dist, eig = TRUE, k = 2)

# Step 3: Create a dataframe for plotting
pcoa_df <- data.frame(
  SampleID = rownames(cell_type_matrix),
  PC1 = pcoa_results$points[,1],
  PC2 = pcoa_results$points[,2],
  any_mets = dh_combined_data$any_mets,
  MLH1_status = dh_combined_data$MLH1  # Include MLH1 status
)

# Correct Factor Recoding for MLH1_status
pcoa_df <- pcoa_df %>%
  mutate(
    MLH1_status = factor(
      MLH1_status, 
      levels = c("MLH1 Positive", "MLH1 Negative"),  # Match actual levels
      labels = c("MLH1+", "MLH1-")  # Desired labels
    )
  )

# Split the PCoA dataframe into MLH1+ and MLH1- subsets
pcoa_df_MLH1_positive <- pcoa_df %>%
  filter(MLH1_status == "MLH1+")

pcoa_df_MLH1_negative <- pcoa_df %>%
  filter(MLH1_status == "MLH1-")

# Step 4: Calculate Species Vectors
species_scores <- envfit(pcoa_results, cell_type_matrix, permutations = 999)
species_vectors <- as.data.frame(scores(species_scores, display = "vectors"))
species_vectors$CellType <- rownames(species_vectors)
colnames(species_vectors)[1:2] <- c("PC1", "PC2")  # Rename columns

# Scale vectors based on PCoA axes
scale_factor_pc1 <- max(pcoa_df$PC1) * 0.8
scale_factor_pc2 <- max(pcoa_df$PC2) * 0.8

species_vectors <- species_vectors %>%
  mutate(
    PC1_end = PC1 * scale_factor_pc1,
    PC2_end = PC2 * scale_factor_pc2
  )

# Step 5: Generate PCoA Plot for MLH1+ Patients
pcoa_plot_MLH1_positive <- ggplot(pcoa_df_MLH1_positive, aes(x = PC1, y = PC2, color = any_mets)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_segment(data = species_vectors,
               aes(x = 0, y = 0, xend = PC1_end, yend = PC2_end),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "black") +
  geom_text_repel(data = species_vectors,
                  aes(x = PC1_end * 1.05, y = PC2_end * 1.05, label = CellType),
                  size = 3,
                  inherit.aes = FALSE) +
  theme_minimal() +
  labs(
    title = "PCoA of Cell Type Proportions - MLH1+ Patients",
    x = "PCoA1",
    y = "PCoA2",
    color = "Metastasis Status"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

# Display and save MLH1+ PCoA plot
print(pcoa_plot_MLH1_positive)
ggsave("PCoA_MLH1_Positive_Biplot.png", pcoa_plot_MLH1_positive, width = 10, height = 8, dpi = 300)

# Step 6: Generate PCoA Plot for MLH1- Patients
pcoa_plot_MLH1_negative <- ggplot(pcoa_df_MLH1_negative, aes(x = PC1, y = PC2, color = any_mets)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_segment(data = species_vectors,
               aes(x = 0, y = 0, xend = PC1_end, yend = PC2_end),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "black") +
  geom_text_repel(data = species_vectors,
                  aes(x = PC1_end * 1.05, y = PC2_end * 1.05, label = CellType),
                  size = 3,
                  inherit.aes = FALSE) +
  theme_minimal() +
  labs(
    title = "PCoA of Cell Type Proportions - MLH1- Patients",
    x = "PCoA1",
    y = "PCoA2",
    color = "Metastasis Status"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

# Display and save MLH1- PCoA plot
print(pcoa_plot_MLH1_negative)
ggsave("PCoA_MLH1_Negative_Biplot.png", pcoa_plot_MLH1_negative, width = 10, height = 8, dpi = 300)

# Step 7: (Optional) Create Combined Faceted PCoA Plot with Specific Vectors

# Function to compute species vectors for a given subset
compute_species_vectors <- function(cell_type_matrix_subset, pcoa_results_subset) {
  species_fit <- envfit(pcoa_results_subset, cell_type_matrix_subset, permutations = 999)
  species_vec <- as.data.frame(scores(species_fit, display = "vectors"))
  species_vec$CellType <- rownames(species_vec)
  colnames(species_vec)[1:2] <- c("PC1", "PC2")
  
  # Scale vectors
  scale_factor_pc1 <- max(pcoa_results_subset$points[,1]) * 0.8
  scale_factor_pc2 <- max(pcoa_results_subset$points[,2]) * 0.8
  
  species_vec <- species_vec %>%
    mutate(
      PC1_end = PC1 * scale_factor_pc1,
      PC2_end = PC2 * scale_factor_pc2
    )
  
  return(species_vec)
}

# Compute species vectors for MLH1+ patients
species_vectors_MLH1_positive <- compute_species_vectors(
  cell_type_matrix_subset = cell_type_matrix[rownames(pcoa_df_MLH1_positive), ],
  pcoa_results_subset = cmdscale(vegdist(cell_type_matrix[rownames(pcoa_df_MLH1_positive), ], method = "bray"), eig = TRUE, k = 2)
)

# Compute species vectors for MLH1- patients
species_vectors_MLH1_negative <- compute_species_vectors(
  cell_type_matrix_subset = cell_type_matrix[rownames(pcoa_df_MLH1_negative), ],
  pcoa_results_subset = cmdscale(vegdist(cell_type_matrix[rownames(pcoa_df_MLH1_negative), ], method = "bray"), eig = TRUE, k = 2)
)

# Combine species vectors with MLH1 status
species_vectors_combined <- bind_rows(
  species_vectors_MLH1_positive %>% mutate(MLH1_status = "MLH1+"),
  species_vectors_MLH1_negative %>% mutate(MLH1_status = "MLH1-")
)

# Combined PCoA dataframe
pcoa_df_combined <- pcoa_df

# Combined PCoA plot with faceting by MLH1_status and specific species vectors
pcoa_plot_combined_specific <- ggplot(pcoa_df_combined, aes(x = PC1, y = PC2, color = any_mets)) +
  geom_point(alpha = 0.7, size = 3) +
  
  # Add biplot vectors specific to each facet
  geom_segment(data = species_vectors_combined,
               aes(x = 0, y = 0, xend = PC1_end, yend = PC2_end),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "black") +
  
  # Add labels for cell types using ggrepel
  geom_text_repel(data = species_vectors_combined,
                  aes(x = PC1_end * 1.05, y = PC2_end * 1.05, label = CellType),
                  size = 3,
                  inherit.aes = FALSE) +
  
  facet_wrap(~ MLH1_status) +  # Facet by MLH1_status
  theme_minimal() +
  labs(
    title = "PCoA of Cell Type Proportions by MLH1 Status",
    x = "PCoA1",
    y = "PCoA2",
    color = "Metastasis Status"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 16, face = "bold")  # Enhance facet labels
  )

# Display the combined faceted PCoA plot with specific vectors
print(pcoa_plot_combined_specific)

# Save the combined faceted plot
ggsave("PCoA_Combined_Faceted_MLH1_Status_SpecificVectors.png", 
       plot = pcoa_plot_combined_specific, 
       width = 14, height = 8, dpi = 300)
```



Perform PERMANOVA
```{r}
# Perform PERMANOVA using adonis2
permanova_results <- adonis2(
  cell_type_matrix ~ any_mets,
  data = dh_combined_data,
  method = "bray",
  permutations = 999
)

# View the results
print(permanova_results)
```

```{r}
# Assess homogeneity of dispersion
dispersion <- betadisper(bray_dist, dh_combined_data$any_mets)

# View dispersion results
print(dispersion)

# Perform permutation test for homogeneity
dispersion_test <- permutest(dispersion, permutations = 999)

# View dispersion test results
print(dispersion_test)
```
Performing the Wilcoxon rank-sum test (also known as the Mann-Whitney U test) to compare the proportions of CD8+ T-cells and Fibroblasts between metastatic (Mets) and non-metastatic (No Mets) colorectal cancer samples.

```{r}
# Define the cell types of interest
cell_types_of_interest <- c("CD8mem", "Fibroblasts")
```

```{r}
# Perform Wilcoxon rank-sum tests for the specified cell types
wilcox_results <- long_dh_data %>%
  filter(CellType %in% cell_types_of_interest) %>%  # Select relevant cell types
  group_by(CellType) %>%                             # Group by cell type
  wilcox_test(Proportion ~ any_mets) %>%            # Perform the test
  adjust_pvalue(method = "BH") %>%                  # Adjust p-values for multiple testing
  add_significance()                                 # Add significance labels

# View the results
print(wilcox_results)
```
```{r}
# Filter data for the cell types of interest
plot_data <- long_dh_data %>%
  filter(CellType %in% cell_types_of_interest)

# Create boxplots
ggplot(plot_data, aes(x = any_mets, y = Proportion, fill = any_mets)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~ CellType, scales = "free_y") +  # Create separate panels for each cell type
  theme_minimal() +
  labs(
    title = "Cell Type Proportions by Metastasis Status",
    x = "Metastasis Status",
    y = "Proportion",
    fill = "Metastasis Status"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )
```
```{r}
# Calculate effect size
wilcox_results <- wilcox_results %>%
  mutate(effect_size = statistic / (n1 * n2))

# View the updated results
print(wilcox_results)
```


```{r}
# Load necessary packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(vegan)

# Step 1: Extract fibroblast information from EpiDISH results
fibroblast_data <- epidish_combined_results %>%
  select(SampleID, Source, Fibroblasts = Fib)  # Rename 'Fib' to 'Fibroblasts' for clarity

# Step 2: Extract HiTIMED Levels 1-6 with conditional 'Stromal' removal
hitimed_all_levels <- lapply(all_results, function(source_results) {
  # Assuming levels 1 to 6 are stored as elements 1 to 6
  levels_data <- lapply(1:6, function(level) {
    level_result <- source_results[[level]]  # Extract specific level
    
    level_result <- level_result %>%
      rownames_to_column(var = "SampleID") %>%  # Move row names into "SampleID"
      select(-any_of("Stromal"), everything()) %>%  # Conditionally remove 'Stromal' if it exists
      mutate(Level = paste0("Level", level))    # Add 'Level' column
    
    return(level_result)
  })
  
  bind_rows(levels_data)  # Combine all levels for the current source
})


# Combine all sources and levels into a single dataframe
hitimed_all_levels <- bind_rows(hitimed_all_levels)

# Step 3: Merge HiTIMED Levels with EpiDISH Fibroblast Data
combined_data <- hitimed_all_levels %>%
  left_join(fibroblast_data, by = c("SampleID", "Source"))

# Step 4: Merge with Metadata
combined_data <- combined_data %>%
  inner_join(targets, by = "SampleID")  # Add metadata like 'any_mets'

# Step 5: Filter for DH Data Only
dh_combined_data <- combined_data %>%
  filter(Source == "DH")  # Keep only DH samples

# Step 6: Reshape Data for Visualization (include only specific cell types)
cell_types_to_include <- c("Tumor", "Endothelial", "Epithelial", "Fibroblasts", 
                           "Bnv", "Bmem", "CD4nv", "CD4mem", "Treg", 
                           "CD8nv", "CD8mem", "Mono", "DC", "NK", 
                           "Bas", "Eos", "Neu")

long_dh_data <- dh_combined_data %>%
  pivot_longer(
    cols = all_of(cell_types_to_include),  # Include only specified cell types
    names_to = "CellType",
    values_to = "Proportion"
  ) %>%
  mutate(
    CellType = factor(CellType, levels = cell_types_to_include),  # Set factor levels
    Level = factor(Level, levels = paste0("Level", 1:6))         # Ensure 'Level' is a factor
  )

# Step 7: Generate Boxplots with Faceting and Data Points

# Generate faceted boxplots with individual data points
facet_celltypes_boxplot <- ggplot(long_dh_data, aes(x = any_mets, y = Proportion, fill = any_mets)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Boxplots without outliers
  geom_jitter(aes(color = any_mets), width = 0.2, alpha = 0.5, size = 1.5) +  # Overlay points
  facet_grid(Level ~ CellType, scales = "free_y") +  # Facet by 'Level' and 'CellType'
  theme_minimal() +
  labs(
    title = "Cell Type Proportions by Metastasis Status Across HiTIMED Levels",
    x = "Metastasis Status",
    y = "Proportion",
    fill = "Metastasis Status",
    color = "Metastasis Status"
  ) +
  scale_fill_brewer(palette = "Set2") +  # Aesthetic color palette
  scale_color_brewer(palette = "Set2") +  # Match point colors to fill
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")  # Enhance facet labels
  )

# Display the faceted plot
print(facet_celltypes_boxplot)

# Save the faceted plot
ggsave("Facet_Selected_CellTypes_Boxplot_DH_AnyMets.png", 
       plot = facet_celltypes_boxplot, 
       width = 16, height = 10, dpi = 300)

# OPTIONAL: Generate Separate Plots for Each HiTIMED Level

# List of levels
levels_list <- levels(long_dh_data$Level)

# Create a directory to save individual plots
dir.create("HiTIMED_Level_Plots", showWarnings = FALSE)

# Loop through each level and generate plots
for (lvl in levels_list) {
  # Filter data for the current level
  level_data <- long_dh_data %>%
    filter(Level == lvl)
  
  # Generate boxplot with points
  level_plot <- ggplot(level_data, aes(x = any_mets, y = Proportion, fill = any_mets)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(aes(color = any_mets), width = 0.2, alpha = 0.5, size = 1.5) +  # Add points
    facet_wrap(~ CellType, scales = "free_y") +  # Facet by cell type within the level
    theme_minimal() +
    labs(
      title = paste("HiTIMED", lvl, "- Cell Type Proportions by Metastasis Status"),
      x = "Metastasis Status",
      y = "Proportion",
      fill = "Metastasis Status",
      color = "Metastasis Status"
    ) +
    scale_fill_brewer(palette = "Set2") +
    scale_color_brewer(palette = "Set2") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      strip.text = element_text(size = 8, face = "bold")  # Format facet labels
    )
  
  # Display the plot
  print(level_plot)
  
  # Save the plot
  ggsave(filename = paste0("HiTIMED_Level_Plots/Boxplot_", lvl, ".png"), 
         plot = level_plot, 
         width = 12, height = 8, dpi = 300)
}
```



