---
title: "Cell_Type_Deconvolution_Analysis_DNA_Methylation"
author: "Alos Diallo"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# =========================================================================
# DNA methylation Cell Type Deconvolution using HiTIMED - COAD DH Dataset
# Code by: Alos Diallo
# 2023,2024
# =========================================================================


```{r message=FALSE, warning=FALSE, include=FALSE}
library(minfi)
library(sesame)
library(pheatmap)
library(minfiData)
library(FlowSorted.Blood.EPIC)
library(HiTIMED)
library(ggplot2)
library(dplyr)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICv2manifest)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICmanifest)
library(limma)
library(qvalue)
library(sva)
library(ENmix)
library(DataExplorer)
library(tidyr)
library(EpiDISH)
library(tibble)
library(vegan)
library(rstatix)
```

Loading Data
```{r}
targets <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/dm_57_samples.csv")
targets$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets) <- targets$patient
targets$SampleID<- targets$patient


# Convert relevant columns in `targets` to factors
targets <- targets %>%
  mutate(
    MLH1 = factor(MLH1, levels = c(0, 1), labels = c("MLH1 Negative", "MLH1 Positive")),
    MSH2 = factor(MSH2),
    sex = factor(sex, levels = c("M", "F")),
    Distant_Mets = factor(Distant_Mets, levels = c(FALSE, TRUE), labels = c("No Distant Mets", "Distant Mets")),
    any_mets = factor(any_mets, levels = c(FALSE, TRUE), labels = c("No Mets", "Mets")),
    ln_only = factor(ln_only, levels = c(FALSE, TRUE), labels = c("No LN Mets", "LN Mets"))
  )
```

```{r}
idat = "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/"
RGset_32 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/",recursive = TRUE) 
#RGset_25 = openSesame(idat , func = getBetas) 
RGset_25 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/",recursive = TRUE) 
```


Normalize my Data
```{r}
Noob_25_m = preprocessNoob(RGset_25)
Noob_32_m = preprocessNoob(RGset_32)
```

Getting the beta values
```{r}
Betas_25<-getBeta(Noob_25_m)
Betas_32<-getBeta(Noob_32_m)
```

```{r}
Betas_25<- sesame::betasCollapseToPfx(Betas_25)
#Betas_32<- sesame::betasCollapseToPfx(Betas_32)
colnames(Betas_25) = colnames(Noob_25_m)
```

```{r}
DH_CRC_Betas <- merge(Betas_25, Betas_32, by = "row.names")
rownames(DH_CRC_Betas) <- DH_CRC_Betas$Row.names
DH_CRC_Betas <- DH_CRC_Betas[,-1] 
```


Running EpiDish to obtain Fibroblast information
```{r}
# Define the beta matrix and reference matrix
beta_matrix <- DH_CRC_Betas  # Your dataset
ref_matrix <- centEpiFibIC.m  # Reference matrix for EpiDISH (e.g., centEpiFibIC.m)

# Run EpiDISH
epidish_results <- epidish(beta.m = beta_matrix, ref.m = ref_matrix, method = "RPC")

# Extract the estimated cell type proportions
cell_type_proportions <- epidish_results$estF

# Convert the matrix to a data frame and scale proportions to percentages
cell_type_proportions_df <- as.data.frame(cell_type_proportions) * 100

# Add SampleID as a column from the row names
cell_type_proportions_df$SampleID <- rownames(cell_type_proportions)

# Inspect the results
head(cell_type_proportions_df)
```



Running HiTIMED to generate cell type proportions
```{r}
HiTIMED_result<-HiTIMED_deconvolution(DH_CRC_Betas,"COAD",6,"tumor")

HiTIMED_result$SampleID <- rownames(HiTIMED_result)
```


```{r}
# Merge HiTIMED results, targets, and EpiDISH fibroblast data
combined_data <- HiTIMED_result %>%
  left_join(targets, by = "SampleID") %>%
  left_join(cell_type_proportions_df %>% select(SampleID, Fib), by = "SampleID")  # Add 'Fib' column

combined_data$SampleID = combined_data$patient
```

	•	No Metastasis (if any_mets == "No Mets"),
	•	Lymph Node Metastasis (if ln_only == "LN Mets" and Distant_Mets == "No Distant Mets"),
	•	Distant Metastasis (if Distant_Mets == "Distant Mets").
	
```{r}
# ----- 1. Create a metastasis category factor -----
combined_data <- combined_data %>%
  mutate(
    mets_cat = case_when(
      any_mets == "No Mets" ~ "No Metastasis",
      # Ensure that LN-only does not include Distant Mets
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Metastasis",
      Distant_Mets == "Distant Mets" ~ "Distant Metastasis",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    mets_cat = factor(
      mets_cat,
      levels = c("No Metastasis", "LN Metastasis", "Distant Metastasis")
    )
  )
```


```{r}
# ----- 2. Reshape your data for plotting -----
# Define my relevant cell-type columns based on the HiTIMED paper:
cell_types <- c(
  "Tumor","Endothelial","Epithelial","Stromal","Bnv","Bmem","CD4nv","CD4mem",
  "Treg","CD8nv","CD8mem","Mono","DC","NK","Bas","Eos","Neu","Fib"
)

long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )
```
	
	
```{r}
# ----- 3. Plot using ggplot2 -----
# A single faceted figure for all cell types:
ggplot(long_data, aes(x = mets_cat, y = proportion, fill = MLH1)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Cell Type Proportions by Metastasis Category and MLH1 Status",
    x = "Metastasis Category",
    y = "Proportion",
    fill = "MLH1 Status"
  ) +
  theme_bw()
```

Here I am looking at Distant Mets colored by MLH1
```{r}

# Specify the cell type columns to plot
cell_types <- c(
  "Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv",
  "Bmem", "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem",
  "Mono", "DC", "NK", "Bas", "Eos", "Neu", "Fib"
)

# Pivot data to "long" format: one row per (sample x cell_type)
long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )

# Plot: compare No Distant Mets vs. Distant Mets, coloring by MLH1 status
ggplot(long_data, aes(x = Distant_Mets, y = proportion, fill = MLH1)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Cell Type Proportions by Distant Metastasis Status and MLH1",
    x = "Distant Metastasis Status",
    y = "Proportion",
    fill = "MLH1 Status"
  ) +
  theme_bw()
```

```{r}
combined_data <- combined_data %>%
  mutate(
    mets_cat = case_when(
      # no metastases at all
      any_mets == "No Mets" ~ "No Mets",  

      # LN-only metastasis: any_mets == "Mets" BUT distant_mets == "No Distant Mets"
      # i.e., LN mets == "LN Mets"
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Mets",
      
      # distant metastasis
      Distant_Mets == "Distant Mets" ~ "Distant Mets",
      
      TRUE ~ NA_character_
    ),
    # factor with 3 distinct levels
    mets_cat = factor(mets_cat, levels = c("No Mets", "LN Mets", "Distant Mets"))
  )

```

```{r}
# Define the function
plot_cell_types <- function(data,
                            cell_type_cols,
                            group_var   = "Distant_Mets",  # This can be "any_mets", "ln_only", etc.
                            fill_var    = "MLH1",
                            scales      = "free_y",        # free_y, free_x, or fixed
                            plot_title  = "Cell Type Proportions",
                            x_label     = NULL,
                            y_label     = "Proportion") {
  
  # If x_label isn't specified, use group_var by default
  if (is.null(x_label)) {
    x_label <- group_var
  }
  
  # Pivot my data to long format
  long_data <- data %>%
    pivot_longer(
      cols = all_of(cell_type_cols),
      names_to = "cell_type",
      values_to = "proportion"
    )
  
  # Create the box plot
  ggplot(long_data, aes_string(x = group_var, y = "proportion", fill = fill_var)) +
    geom_boxplot() +
    facet_wrap(~ cell_type, scales = scales) +
    labs(
      title = plot_title,
      x     = x_label,
      y     = y_label,
      fill  = fill_var
    ) +
    theme_bw()
}

# Suppose I have a vector of all your cell type columns:
cell_types <- c(
  "Tumor","Endothelial","Epithelial","Stromal","Bnv","Bmem",
  "CD4nv","CD4mem","Treg","CD8nv","CD8mem","Mono","DC",
  "NK","Bas","Eos","Neu","Fib"
)

# Now, just call the function with my desired grouping variable.

## 1. By Distant_Mets
plot_cell_types(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "Distant_Mets",      # Compare No Distant Mets vs Distant Mets
  fill_var       = "MLH1",
  scales         = "free_y",            # each facet has its own y-scale
  plot_title     = "Cell Type Proportions by Distant Metastasis Status and MLH1"
)

## 2. By any_mets
plot_cell_types(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "any_mets",          # Compare No Mets vs Mets
  fill_var       = "MLH1",
  plot_title     = "Cell Type Proportions by Any Mets Status and MLH1"
)

## 3. By ln_only
plot_cell_types(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "ln_only",           # Compare No LN Mets vs LN Mets
  fill_var       = "MLH1",
  plot_title     = "Cell Type Proportions by LN Metastasis Status and MLH1"
)
```

Here I am looking at No mets at all vs Distant mets, slightly different but more distinct.  The above analysis as done before includes samples that are LN mets. 
```{r}

combined_data <- combined_data %>%
  mutate(
    mets_binary = case_when(
      any_mets == "No Mets"                  ~ "No Mets",
      Distant_Mets == "Distant Mets"         ~ "Distant Mets",
      TRUE                                   ~ NA_character_
    ),
    mets_binary = factor(
      mets_binary, 
      levels = c("No Mets", "Distant Mets")
    )
  )


cell_types <- c(
  "Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv", "Bmem",
  "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC",
  "NK", "Bas", "Eos", "Neu", "Fib"
)

long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )

library(ggplot2)

plot_data <- long_data %>%
  filter(!is.na(mets_binary))  # keep only No Mets & Distant Mets

ggplot(plot_data, aes(x = mets_binary, y = proportion, fill = MLH1)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Cell Type Proportions: Distant Mets vs. No Mets",
    x = "Metastasis Status",
    y = "Proportion",
    fill = "MLH1 Status"
  ) +
  theme_bw()
```

Making a function to carry out the plotting
```{r}
combined_data <- combined_data %>%
  mutate(
    # 1) No Mets vs. LN Mets only
    no_mets_vs_ln = case_when(
      any_mets == "No Mets"        ~ "No Mets",
      ln_only == "LN Mets"         ~ "LN Mets",
      TRUE                         ~ NA_character_
    ),
    no_mets_vs_ln = factor(no_mets_vs_ln, levels = c("No Mets", "LN Mets")),
    
    # 2) No Mets vs. Distant Mets
    no_mets_vs_distant = case_when(
      any_mets == "No Mets"          ~ "No Mets",
      Distant_Mets == "Distant Mets" ~ "Distant Mets",
      TRUE                           ~ NA_character_
    ),
    no_mets_vs_distant = factor(no_mets_vs_distant, 
                                levels = c("No Mets", "Distant Mets")),
    
    # 3) No Mets vs. Any Mets
    no_mets_vs_any = case_when(
      any_mets == "No Mets" ~ "No Mets",
      any_mets == "Mets"    ~ "Any Mets",
      TRUE                  ~ NA_character_
    ),
    no_mets_vs_any = factor(no_mets_vs_any, levels = c("No Mets", "Any Mets"))
  )

plot_cell_types <- function(data,
                            cell_type_cols,
                            group_var   = "any_mets",  # or "no_mets_vs_ln", etc.
                            fill_var    = "MLH1",
                            scales      = "free_y",
                            plot_title  = "Cell Type Proportions") {
  # Pivot wide -> long, storing cell type in "cell_type", proportion in "proportion"
  data_long <- data %>%
    pivot_longer(
      cols = all_of(cell_type_cols),
      names_to = "cell_type",
      values_to = "proportion"
    )
  
  # Plot
  ggplot(data_long, aes_string(x = group_var, y = "proportion", fill = fill_var)) +
    geom_boxplot() +
    facet_wrap(~ cell_type, scales = scales) +
    labs(
      title = plot_title,
      x     = group_var,
      y     = "Proportion",
      fill  = fill_var
    ) +
    theme_bw()
}

# 1) No Mets vs. LN Mets
plot_cell_types(
  data = combined_data %>% filter(!is.na(no_mets_vs_ln)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_ln",
  fill_var       = "MLH1",
  scales         = "free_y",
  plot_title     = "No Mets vs. LN Mets"
)

# 2) No Mets vs. Distant Mets
plot_cell_types(
  data = combined_data %>% filter(!is.na(no_mets_vs_distant)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_distant",
  fill_var       = "MLH1",
  scales         = "free_y",
  plot_title     = "No Mets vs. Distant Mets"
)

# 3) No Mets vs. Any Mets
plot_cell_types(
  data = combined_data %>% filter(!is.na(no_mets_vs_any)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_any",
  fill_var       = "MLH1",
  scales         = "free_y",
  plot_title     = "No Mets vs. Any Mets"
)




```


