---
title: "Cell_Type_Deconvolution_Analysis_DNA_Methylation"
author: "Alos Diallo"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# =========================================================================
# DNA methylation Cell Type Deconvolution using HiTIMED - COAD DH Dataset
# Code by: Alos Diallo
# 2023,2024
# =========================================================================


```{r message=FALSE, warning=FALSE, include=FALSE}
library(minfi)
library(sesame)
library(pheatmap)
library(minfiData)
library(FlowSorted.Blood.EPIC)
library(HiTIMED)
library(ggplot2)
library(dplyr)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICv2manifest)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICmanifest)
library(limma)
library(qvalue)
library(sva)
library(ENmix)
library(DataExplorer)
library(tidyr)
library(EpiDISH)
library(tibble)
library(vegan)
library(rstatix)
```

Loading Data
```{r}
targets <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/dm_57_samples.csv")
targets$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets) <- targets$patient
targets$SampleID<- targets$patient

targets_expanded <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv")
targets_expanded$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets_expanded) <- targets_expanded$patient
targets_expanded$SampleID<- targets_expanded$patient
targets = targets_expanded


# Convert relevant columns in `targets` to factors
targets <- targets %>%
  mutate(
    MLH1 = factor(MLH1, levels = c(0, 1), labels = c("MLH1 Negative", "MLH1 Positive")),
    MSH2 = factor(MSH2),
    sex = factor(sex, levels = c("M", "F")),
    Distant_Mets = factor(Distant_Mets, levels = c(FALSE, TRUE), labels = c("No Distant Mets", "Distant Mets")),
    any_mets = factor(any_mets, levels = c(FALSE, TRUE), labels = c("No Mets", "Mets")),
    ln_only = factor(ln_only, levels = c(FALSE, TRUE), labels = c("No LN Mets", "LN Mets"))
  )
```

```{r}
idat = "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/"
RGset_32 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/",recursive = TRUE) 
#RGset_25 = openSesame(idat , func = getBetas) 
RGset_25 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/",recursive = TRUE) 
```


Normalize my Data
```{r}
Noob_25_m = preprocessNoob(RGset_25)
Noob_32_m = preprocessNoob(RGset_32)
```

Getting the beta values
```{r}
Betas_25<-getBeta(Noob_25_m)
Betas_32<-getBeta(Noob_32_m)
```

```{r}
Betas_25<- sesame::betasCollapseToPfx(Betas_25)
#Betas_32<- sesame::betasCollapseToPfx(Betas_32)
colnames(Betas_25) = colnames(Noob_25_m)
```

```{r}
DH_CRC_Betas <- merge(Betas_25, Betas_32, by = "row.names")
rownames(DH_CRC_Betas) <- DH_CRC_Betas$Row.names
DH_CRC_Betas <- DH_CRC_Betas[,-1] 
```


Running EpiDish to obtain Fibroblast information
```{r}
# Define the beta matrix and reference matrix
beta_matrix <- DH_CRC_Betas  # Your dataset
ref_matrix <- centEpiFibIC.m  # Reference matrix for EpiDISH (e.g., centEpiFibIC.m)

# Run EpiDISH
epidish_results <- epidish(beta.m = beta_matrix, ref.m = ref_matrix, method = "RPC")

# Extract the estimated cell type proportions
cell_type_proportions <- epidish_results$estF

# Convert the matrix to a data frame and scale proportions to percentages
cell_type_proportions_df <- as.data.frame(cell_type_proportions) * 100

# Add SampleID as a column from the row names
cell_type_proportions_df$SampleID <- rownames(cell_type_proportions)

# Inspect the results
head(cell_type_proportions_df)
```



Running HiTIMED to generate cell type proportions
```{r}
HiTIMED_result<-HiTIMED_deconvolution(DH_CRC_Betas,"COAD",5,"tumor")

HiTIMED_result$SampleID <- rownames(HiTIMED_result)

HiTIMED_result_immune<-HiTIMED_deconvolution(DH_CRC_Betas,"COAD",2,"tumor")

HiTIMED_result_immune$SampleID <- rownames(HiTIMED_result)

HiTIMED_result_immune_sub<-HiTIMED_deconvolution(DH_CRC_Betas,"COAD",3,"tumor")

HiTIMED_result_immune_sub$SampleID <- rownames(HiTIMED_result)

write.csv(HiTIMED_result,"/Users/adiallo/Desktop/HiTIMED_result.csv")
write.csv(HiTIMED_result_immune,"/Users/adiallo/Desktop/HiTIMED_Level_2.csv")

```


```{r}
# Merge HiTIMED results, targets, and EpiDISH fibroblast data
combined_data <- HiTIMED_result %>%
  left_join(targets, by = "SampleID") %>%
  left_join(cell_type_proportions_df %>% select(SampleID, Fib), by = "SampleID")  # Add 'Fib' column

combined_data$SampleID = combined_data$patient
```

	•	No Metastasis (if any_mets == "No Mets"),
	•	Lymph Node Metastasis (if ln_only == "LN Mets" and Distant_Mets == "No Distant Mets"),
	•	Distant Metastasis (if Distant_Mets == "Distant Mets").
	
```{r}
# ----- 1. Create a metastasis category factor -----
combined_data <- combined_data %>%
  mutate(
    mets_cat = case_when(
      any_mets == "No Mets" ~ "No Metastasis",
      # Ensure that LN-only does not include Distant Mets
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Metastasis",
      Distant_Mets == "Distant Mets" ~ "Distant Metastasis",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    mets_cat = factor(
      mets_cat,
      levels = c("No Metastasis", "LN Metastasis", "Distant Metastasis")
    )
  )
```

Stacked Barplots:

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

# Define cell-type variables
cell_type_vars <- c("Tumor", "Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", "CD4T", 
                    "CD8T", "NK", "Endothelial", "Epithelial", "Stromal")

# Ensure mets_cat is a factor with the correct order
combined_data$mets_cat <- factor(combined_data$mets_cat, 
                                 levels = c("No Metastasis", "LN Metastasis", "Distant Metastasis"))

# Reshape data to long format
df_long <- combined_data %>%
  select(SampleID, mets_cat, all_of(cell_type_vars)) %>%
  pivot_longer(cols = cell_type_vars, names_to = "CellType", values_to = "Proportion")

# 1️⃣ Order samples within each metastasis category by "Tumor" proportion
tumor_order <- df_long %>%
  filter(CellType == "Tumor") %>%
  arrange(mets_cat, desc(Proportion)) %>%  # Sort within each disease state
  pull(SampleID)  # Extract ordered SampleID

# Convert SampleID to factor to enforce order in plots
df_long$SampleID <- factor(df_long$SampleID, levels = tumor_order)

# 2️⃣ Order cell types from most to least abundant (across all samples)
cell_type_order <- df_long %>%
  group_by(CellType) %>%
  summarize(MeanProportion = mean(Proportion, na.rm = TRUE)) %>%
  arrange(desc(MeanProportion)) %>%  # Most abundant first
  pull(CellType)  # Extract ordered cell type names

# Convert CellType to factor with ordered levels (most abundant at bottom of stacked bar)
df_long$CellType <- factor(df_long$CellType, levels = rev(cell_type_order))  # Reverse to put most abundant at bottom

# Define custom colors
my_14_colors <- c(
  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3",
  "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3",
  "#1F78B4", "#33A02C", "#FB9A99", "#FDBF6F",
  "#CAB2D6", "#6A3D9A", "#E31A1C", "#B15928", "#C51B8A"
)

# Function to generate plots for each metastasis category
plot_cell_proportions <- function(data, metastasis_label) {
  ggplot(data, aes(x = SampleID, y = Proportion, fill = CellType)) +
    geom_bar(stat = "identity", position = "fill") + 
    scale_fill_manual(values = my_14_colors) +
    scale_y_continuous(labels = percent_format()) +  
    labs(x = "Sample ID", 
         y = "Cell Type Proportion (%)", 
         title = paste("Cell Type Proportions:", metastasis_label)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme_minimal()
}

# Generate plots for each metastasis category with the ordered SampleID and CellType
df_no_mets <- df_long %>% filter(mets_cat == "No Metastasis")
df_ln_mets <- df_long %>% filter(mets_cat == "LN Metastasis")
df_dist_mets <- df_long %>% filter(mets_cat == "Distant Metastasis")

plot_cell_proportions(df_no_mets, "No Metastasis")
plot_cell_proportions(df_ln_mets, "LN Metastasis")
plot_cell_proportions(df_dist_mets, "Distant Metastasis")
```



```{r}
# Load necessary libraries
library(dplyr)

# Define cell-type variables
cell_type_vars <- c("Tumor", "Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", 
                    "CD4T", "CD8T", "NK", "Endothelial", "Epithelial", "Stromal", "Fib")

# Compute summary statistics for each cell type across metastasis categories
summary_stats <- combined_data %>%
  group_by(mets_cat) %>%
  summarise(across(all_of(cell_type_vars), 
                   list(mean = mean, median = median, sd = sd, min = min, 
                        q1 = ~quantile(.x, 0.25), q3 = ~quantile(.x, 0.75), max = max), 
                   .names = "{.col}_{.fn}"))

# Print summary statistics
print(summary_stats)

# If you want to save the results to a CSV file:
# write.csv(summary_stats, "cell_type_summary_stats.csv", row.names = FALSE)
```


"Tumor", "Bas", "Eos", 
                      "Neu", "Mono", "DC", "Bnv", "Bmem", "CD4T", 
                      "CD8T", "NK", "Endothelial", "Epithelial", 
                      "Stromal","Fib"


	
```{r}
Bmem = combined_data$Bmem/HiTIMED_result_immune_sub$Lymphocyte
combined_data$Bmem_Norm = Bmem


# ----- 2. Reshape your data for plotting -----
# Define my relevant cell-type columns based on the HiTIMED paper:
cell_types <- c("Bmem_Norm"
)

long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(ggpubr)  # for stat_compare_means()
library(viridis)

# Define the pairwise comparisons between metastasis categories
comparisons <- list(
  c("No Metastasis", "LN Metastasis"),
  c("No Metastasis", "Distant Metastasis"),
  c("LN Metastasis", "Distant Metastasis")
)

# Create the plot with statistics overlaid
ggplot(long_data, aes(x = mets_cat, y = proportion, fill = mets_cat)) +
  geom_boxplot() +
  geom_jitter(color = "black", size = 0.4, alpha = 0.9) + 
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_viridis(discrete = TRUE, option = "plasma", alpha = 0.6) +
  labs(
    title = "Cell Type Proportions by Metastasis Category",
    x = "Metastasis Category",
    y = "Proportion"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}

###############################################################################
# FULL EXAMPLE CODE BLOCK
###############################################################################

# Load Required Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)       # for stat_compare_means()
library(viridis)

#------------------------------------------------------------------------------
# 1) Compute Normalized Columns in combined_data
#    (Adjust if your column names differ)
#------------------------------------------------------------------------------
combined_data$CD8T_Norm <- combined_data$CD8T / HiTIMED_result_immune$Immune
combined_data$CD4_Norm  <- combined_data$CD4T / HiTIMED_result_immune$Immune
combined_data$Mono_Norm <- combined_data$Mono / HiTIMED_result_immune$Immune
combined_data$NK_Norm   <- combined_data$NK   / HiTIMED_result_immune$Immune

#------------------------------------------------------------------------------
# 2) Create a Column with Three Metastasis Categories
#    (Adapt this to match however your metastasis info is stored)
#------------------------------------------------------------------------------
# ----- 1. Create a metastasis category factor -----
combined_data <- combined_data %>%
  mutate(
    mets_cat = case_when(
      any_mets == "No Mets" ~ "No Metastasis",
      # Ensure that LN-only does not include Distant Mets
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Metastasis",
      Distant_Mets == "Distant Mets" ~ "Distant Metastasis",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    mets_cat = factor(
      mets_cat,
      levels = c("No Metastasis", "LN Metastasis", "Distant Metastasis")
    )
  )

#------------------------------------------------------------------------------
# 3) Pivot to Long Format
#------------------------------------------------------------------------------
cell_types <- c("CD8T_Norm", "CD4_Norm", "Mono_Norm", "NK_Norm")

long_data <- combined_data %>%
  pivot_longer(
    cols      = all_of(cell_types),
    names_to  = "cell_type",
    values_to = "proportion"
  )

#------------------------------------------------------------------------------
# 4) Define Pairwise Comparisons for the Three Groups
#------------------------------------------------------------------------------
comparisons <- list(
  c("No Metastasis", "LN Metastasis"),
  c("No Metastasis", "Distant Metastasis"),
  c("LN Metastasis", "Distant Metastasis")
)


# Ensure mets_cat is a factor with the correct order
combined_data$mets_cat <- factor(combined_data$mets_cat, 
                                 levels = c("No Metastasis", "LN Metastasis", "Distant Metastasis"))
#------------------------------------------------------------------------------
# 5) Generate the Faceted Plot with T-Test Results Overlaid
#------------------------------------------------------------------------------
ggplot(long_data, aes(x = mets_cat, y = proportion, fill = mets_cat)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.2,   # Horizontal jitter so points don't overlap
    color = "black",
    size  = 0.5,
    alpha = 0.8
  ) +
  facet_wrap(~ cell_type, scales = "free_y") + # One facet per cell type
  stat_compare_means(
    comparisons      = comparisons,  # Pairwise t-tests among the 3 categories
    method           = "t.test",
    label            = "p.signif",   # "p.signif" = stars; "p.format" = numeric p-values
    p.adjust.method  = "BH"          # e.g., Benjamini-Hochberg correction for multiple comparisons
  ) +
  scale_fill_viridis(
    discrete = TRUE,
    option   = "plasma",
    alpha    = 0.6
  ) +
  labs(
    title = "Cell Type Proportions by Metastasis Category",
    x     = "Metastasis Category",
    y     = "Proportion"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
```{r}

cell_types <- c("Fib", "CD8T", "Tumor")  # ✅ Now includes all three key cell types
long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )


# Define the key cell types to visualize
selected_cells <- c("Fib", "CD8T", "Tumor")

# Compute mean proportions for these cell types across metastasis categories
trend_data <- long_data %>%
  filter(cell_type %in% selected_cells) %>%
  group_by(cell_type, mets_cat) %>%
  summarise(mean_proportion = mean(proportion, na.rm = TRUE), .groups = "drop")

# Convert data to ensure correct factor ordering for metastasis categories
trend_data$mets_cat <- factor(trend_data$mets_cat, 
                              levels = c("No Metastasis", "LN Metastasis", "Distant Metastasis"))

# Create the line plot
ggplot(trend_data, aes(x = mets_cat, y = mean_proportion, group = cell_type, color = cell_type)) +
  geom_line(size = 1.5) +  # Line for each cell type
  geom_point(size = 3) +   # Points for each mean value
  scale_color_manual(values = c("Fib" = "red", "CD8T" = "blue", "Tumor" = "black")) + # Custom colors
  labs(
    title = "Trends of Fibroblasts, CD8 T-cells, and Tumors Across Metastasis Categories",
    x = "Metastasis Category",
    y = "Mean Cell Type Proportion",
    color = "Cell Type"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```
Cell Type Plots stratified by MLH1:
```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
# combined_data <- combined_data %>%
#   mutate(
#     cd8_total   =  (CD8nv + CD8mem)
#   )

# Assuming 'CD8mem' is your column for CD8 proportions
# and 'mets_cat' has levels: "No Metastasis", "LN Metastasis", "Distant Metastasis"
# Also assuming 'MLH1' has exactly two levels: "MLH1 Negative", "MLH1 Positive"

p_MLH1_pos <- combined_data %>%
  filter(MLH1 == "MLH1 Positive") %>%
  ggplot(aes(x = mets_cat, y = CD8_Norm)) +
  geom_boxplot(fill = "#66C2A5") +
  labs(
    title = "CD8 Proportions: MLH1 Positive",
    x = "Metastasis Category",
    y = "CD8 Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_MLH1_neg <- combined_data %>%
  filter(MLH1 == "MLH1 Negative") %>%
  ggplot(aes(x = mets_cat, y = CD8_Norm)) +
  geom_boxplot(fill = "#FC8D62") +
  labs(
    title = "CD8 Proportions: MLH1 Negative",
    x = "Metastasis Category",
    y = "CD8 Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display both plots side by side
p_MLH1_pos + p_MLH1_neg
```
Looking at Tumor Grade
```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
# combined_data <- combined_data %>%
#   mutate(
#     cd8_total   =  (CD8nv + CD8mem)
#   )

# Assuming 'CD8mem' is your column for CD8 proportions
# and 'mets_cat' has levels: "No Metastasis", "LN Metastasis", "Distant Metastasis"
# Also assuming 'MLH1' has exactly two levels: "MLH1 Negative", "MLH1 Positive"

p_Grade_one <- combined_data %>%
  filter(Grade == "1") %>%
  ggplot(aes(x = mets_cat, y = Fib)) +
  geom_boxplot(fill = "#66C2A5") +
  labs(
    title = "Fibroblast Proportions: Tumor Grade 1",
    x = "Metastasis Category",
    y = "Fibroblast Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_Grade_two <- combined_data %>%
  filter(Grade == "2") %>%
  ggplot(aes(x = mets_cat, y = Fib)) +
  geom_boxplot(fill = "#FC8D62") +
  labs(
    title = "Fibroblast Proportions: Tumor Grade 2",
    x = "Metastasis Category",
    y = "Fibroblast Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_Grade_three <- combined_data %>%
  filter(Grade == "3") %>%
  ggplot(aes(x = mets_cat, y = Fib)) +
  geom_boxplot(fill = "#FC8D62") +
  labs(
    title = "Fibroblast Proportions: Tumor Grade 3",
    x = "Metastasis Category",
    y = "Fibroblast Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display both plots side by side
p_Grade_one + p_Grade_two + p_Grade_three
```

Here I am looking at Distant Mets colored by MLH1
```{r}

# Specify the cell type columns to plot
cell_types <- c(
  "Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv",
  "Bmem", "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem",
  "Mono", "DC", "NK", "Bas", "Eos", "Neu", "Fib"
)

# Pivot data to "long" format: one row per (sample x cell_type)
long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )

# Plot: compare No Distant Mets vs. Distant Mets, coloring by MLH1 status
ggplot(long_data, aes(x = Distant_Mets, y = proportion, fill = MLH1)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Cell Type Proportions by Distant Metastasis Status and MLH1",
    x = "Distant Metastasis Status",
    y = "Proportion",
    fill = "MLH1 Status"
  ) +
  theme_bw()
```

```{r}
combined_data <- combined_data %>%
  mutate(
    mets_cat = case_when(
      # no metastases at all
      any_mets == "No Mets" ~ "No Mets",  

      # LN-only metastasis: any_mets == "Mets" BUT distant_mets == "No Distant Mets"
      # i.e., LN mets == "LN Mets"
      ln_only == "LN Mets" & Distant_Mets == "No Distant Mets" ~ "LN Mets",
      
      # distant metastasis
      Distant_Mets == "Distant Mets" ~ "Distant Mets",
      
      TRUE ~ NA_character_
    ),
    # factor with 3 distinct levels
    mets_cat = factor(mets_cat, levels = c("No Mets", "LN Mets", "Distant Mets"))
  )

```

```{r}
# Define the function
plot_cell_types <- function(data,
                            cell_type_cols,
                            group_var   = "Distant_Mets",  # This can be "any_mets", "ln_only", etc.
                            fill_var    = "MLH1",
                            scales      = "free_y",        # free_y, free_x, or fixed
                            plot_title  = "Cell Type Proportions",
                            x_label     = NULL,
                            y_label     = "Proportion") {
  
  # If x_label isn't specified, use group_var by default
  if (is.null(x_label)) {
    x_label <- group_var
  }
  
  # Pivot my data to long format
  long_data <- data %>%
    pivot_longer(
      cols = all_of(cell_type_cols),
      names_to = "cell_type",
      values_to = "proportion"
    )
  
  # Create the box plot
  ggplot(long_data, aes_string(x = group_var, y = "proportion", fill = fill_var)) +
    geom_boxplot() +
    facet_wrap(~ cell_type, scales = scales) +
    labs(
      title = plot_title,
      x     = x_label,
      y     = y_label,
      fill  = fill_var
    ) +
    theme_bw()
}

# Suppose I have a vector of all your cell type columns:
cell_types <- c(
  "Tumor","Endothelial","Epithelial","Stromal","Bnv","Bmem",
  "CD4nv","CD4mem","Treg","CD8nv","CD8mem","Mono","DC",
  "NK","Bas","Eos","Neu","Fib"
)

# Now, just call the function with my desired grouping variable.

## 1. By Distant_Mets
plot_cell_types(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "Distant_Mets",      # Compare No Distant Mets vs Distant Mets
  fill_var       = "MLH1",
  scales         = "free_y",            # each facet has its own y-scale
  plot_title     = "Cell Type Proportions by Distant Metastasis Status and MLH1"
)

## 2. By any_mets
plot_cell_types(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "any_mets",          # Compare No Mets vs Mets
  fill_var       = "MLH1",
  plot_title     = "Cell Type Proportions by Any Mets Status and MLH1"
)

## 3. By ln_only
plot_cell_types(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "ln_only",           # Compare No LN Mets vs LN Mets
  fill_var       = "MLH1",
  plot_title     = "Cell Type Proportions by LN Metastasis Status and MLH1"
)
```

Re-creating the old plots
```{r}
plot_cell_types_facetMLH1 <- function(data,
                                      cell_type_cols,
                                      group_var   = "Distant_Mets",
                                      facet_var   = "MLH1",   # new argument
                                      scales      = "free_y",
                                      plot_title  = "Cell Type Proportions",
                                      x_label     = NULL,
                                      y_label     = "Proportion") {
  if (is.null(x_label)) x_label <- group_var
  
  data_long <- data %>%
    pivot_longer(
      cols = all_of(cell_type_cols),
      names_to = "cell_type",
      values_to = "proportion"
    )
  
  ggplot(data_long, aes_string(x = group_var, y = "proportion", fill = group_var)) +
    geom_boxplot() +
    facet_grid(as.formula(paste(facet_var, "~ cell_type")), scales = scales) +
    labs(
      title = plot_title,
      x     = x_label,
      y     = y_label,
      fill  = group_var
    ) +
    theme_bw()
}

plot_cell_types_facetMLH1(
  data           = combined_data,
  cell_type_cols = cell_types,
  group_var      = "Distant_Mets",
  facet_var      = "MLH1",
  scales         = "free_y",
  plot_title     = "Cell Type Proportions by Distant_Mets and MLH1"
)
```


```{r}
# Create two subsets of your combined_data:
combined_data_MLH1pos <- combined_data %>%
  filter(MLH1 == "MLH1 Positive")

combined_data_MLH1neg <- combined_data %>%
  filter(MLH1 == "MLH1 Negative")

plot_cell_types <- function(data,
                            cell_type_cols,
                            group_var   = "Distant_Mets",
                            fill_var    = "MLH1",  # originally coloring by MLH1
                            scales      = "free_y",
                            plot_title  = "Cell Type Proportions",
                            x_label     = NULL,
                            y_label     = "Proportion") {
  
  if (is.null(x_label)) x_label <- group_var
  
  data_long <- data %>%
    pivot_longer(
      cols = all_of(cell_type_cols),
      names_to = "cell_type",
      values_to = "proportion"
    )
  
  ggplot(data_long, aes_string(x = group_var, y = "proportion", fill = fill_var)) +
    geom_boxplot() +
    facet_wrap(~ cell_type, scales = scales) +
    labs(
      title = plot_title,
      x     = x_label,
      y     = y_label,
      fill  = fill_var
    ) +
    theme_bw()
}

plot_cell_types(
  data = combined_data_MLH1pos,
  cell_type_cols = cell_types,
  group_var      = "Distant_Mets",
  fill_var       = "Distant_Mets",   # <-- changed here
  scales         = "free_y",
  plot_title     = "MLH1 Positive: Cell Type Proportions by Distant_Mets"
)

# A) Distant_Mets
plot_cell_types(
  data           = combined_data_MLH1pos,
  cell_type_cols = cell_types,
  group_var      = "Distant_Mets",      # No Distant Mets vs Distant Mets
  fill_var       = "Distant_Mets",      # fill color by the same grouping
  scales         = "free_y",
  plot_title     = "MLH1+ : Distant_Mets"
)

# B) any_mets
plot_cell_types(
  data           = combined_data_MLH1pos,
  cell_type_cols = cell_types,
  group_var      = "any_mets",          # No Mets vs Mets
  fill_var       = "any_mets",
  scales         = "free_y",
  plot_title     = "MLH1+ : any_mets"
)

# C) ln_only
plot_cell_types(
  data           = combined_data_MLH1pos,
  cell_type_cols = cell_types,
  group_var      = "ln_only",           # No LN Mets vs LN Mets
  fill_var       = "ln_only",
  scales         = "free_y",
  plot_title     = "MLH1+ : ln_only"
)

# A) Distant_Mets
plot_cell_types(
  data           = combined_data_MLH1neg,
  cell_type_cols = cell_types,
  group_var      = "Distant_Mets",
  fill_var       = "Distant_Mets",
  scales         = "free_y",
  plot_title     = "MLH1- : Distant_Mets"
)

# B) any_mets
plot_cell_types(
  data           = combined_data_MLH1neg,
  cell_type_cols = cell_types,
  group_var      = "any_mets",
  fill_var       = "any_mets",
  scales         = "free_y",
  plot_title     = "MLH1- : any_mets"
)

# C) ln_only
plot_cell_types(
  data           = combined_data_MLH1neg,
  cell_type_cols = cell_types,
  group_var      = "ln_only",
  fill_var       = "ln_only",
  scales         = "free_y",
  plot_title     = "MLH1- : ln_only"
)


```




Here I am looking at No mets at all vs Distant mets, slightly different but more distinct.  The above analysis as done before includes samples that are LN mets. 
```{r}

combined_data <- combined_data %>%
  mutate(
    mets_binary = case_when(
      any_mets == "No Mets"                  ~ "No Mets",
      Distant_Mets == "Distant Mets"         ~ "Distant Mets",
      TRUE                                   ~ NA_character_
    ),
    mets_binary = factor(
      mets_binary, 
      levels = c("No Mets", "Distant Mets")
    )
  )


cell_types <- c(
  "Tumor", "Endothelial", "Epithelial", "Stromal", "Bnv", "Bmem",
  "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", "Mono", "DC",
  "NK", "Bas", "Eos", "Neu", "Fib"
)

long_data <- combined_data %>%
  pivot_longer(
    cols = all_of(cell_types),
    names_to = "cell_type",
    values_to = "proportion"
  )

library(ggplot2)

plot_data <- long_data %>%
  filter(!is.na(mets_binary))  # keep only No Mets & Distant Mets

ggplot(plot_data, aes(x = mets_binary, y = proportion, fill = MLH1)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Cell Type Proportions: Distant Mets vs. No Mets",
    x = "Metastasis Status",
    y = "Proportion",
    fill = "MLH1 Status"
  ) +
  theme_bw()
```

Making a function to carry out the plotting
```{r}
combined_data <- combined_data %>%
  mutate(
    # 1) No Mets vs. LN Mets only
    no_mets_vs_ln = case_when(
      any_mets == "No Mets"        ~ "No Mets",
      ln_only == "LN Mets"         ~ "LN Mets",
      TRUE                         ~ NA_character_
    ),
    no_mets_vs_ln = factor(no_mets_vs_ln, levels = c("No Mets", "LN Mets")),
    
    # 2) No Mets vs. Distant Mets
    no_mets_vs_distant = case_when(
      any_mets == "No Mets"          ~ "No Mets",
      Distant_Mets == "Distant Mets" ~ "Distant Mets",
      TRUE                           ~ NA_character_
    ),
    no_mets_vs_distant = factor(no_mets_vs_distant, 
                                levels = c("No Mets", "Distant Mets")),
    
    # 3) No Mets vs. Any Mets
    no_mets_vs_any = case_when(
      any_mets == "No Mets" ~ "No Mets",
      any_mets == "Mets"    ~ "Any Mets",
      TRUE                  ~ NA_character_
    ),
    no_mets_vs_any = factor(no_mets_vs_any, levels = c("No Mets", "Any Mets"))
  )

plot_cell_types <- function(data,
                            cell_type_cols,
                            group_var   = "any_mets",  # or "no_mets_vs_ln", etc.
                            fill_var    = "MLH1",
                            scales      = "free_y",
                            plot_title  = "Cell Type Proportions") {
  # Pivot wide -> long, storing cell type in "cell_type", proportion in "proportion"
  data_long <- data %>%
    pivot_longer(
      cols = all_of(cell_type_cols),
      names_to = "cell_type",
      values_to = "proportion"
    )
  
  # Plot
  ggplot(data_long, aes_string(x = group_var, y = "proportion", fill = fill_var)) +
    geom_boxplot() +
    facet_wrap(~ cell_type, scales = scales) +
    labs(
      title = plot_title,
      x     = group_var,
      y     = "Proportion",
      fill  = fill_var
    ) +
    theme_bw()
}

# 1) No Mets vs. LN Mets
plot_cell_types(
  data = combined_data %>% filter(!is.na(no_mets_vs_ln)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_ln",
  fill_var       = "MLH1",
  scales         = "free_y",
  plot_title     = "No Mets vs. LN Mets"
)

# 2) No Mets vs. Distant Mets
plot_cell_types(
  data = combined_data %>% filter(!is.na(no_mets_vs_distant)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_distant",
  fill_var       = "MLH1",
  scales         = "free_y",
  plot_title     = "No Mets vs. Distant Mets"
)

# 3) No Mets vs. Any Mets
plot_cell_types(
  data = combined_data %>% filter(!is.na(no_mets_vs_any)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_any",
  fill_var       = "MLH1",
  scales         = "free_y",
  plot_title     = "No Mets vs. Any Mets"
)




```
```{r}
# Split data by MLH1
combined_data_MLH1pos <- combined_data %>%
  filter(MLH1 == "MLH1 Positive")

combined_data_MLH1neg <- combined_data %>%
  filter(MLH1 == "MLH1 Negative")

# 1) No Mets vs. LN Mets, MLH1+
plot_cell_types(
  data = combined_data_MLH1pos %>% filter(!is.na(no_mets_vs_ln)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_ln",  # x-axis
  fill_var       = "no_mets_vs_ln",  # ensures two colors (one for "No Mets", one for "LN Mets")
  scales         = "free_y",
  plot_title     = "MLH1+ : No Mets vs. LN Mets"
)

# 2) No Mets vs. Distant Mets, MLH1+
plot_cell_types(
  data = combined_data_MLH1pos %>% filter(!is.na(no_mets_vs_distant)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_distant",   # x-axis
  fill_var       = "no_mets_vs_distant",   # ensures two colors (one for "No Mets", one for "Distant Mets")
  scales         = "free_y",
  plot_title     = "MLH1+ : No Mets vs. Distant Mets"
)

# 3) No Mets vs. Any Mets, MLH1+
plot_cell_types(
  data = combined_data_MLH1pos %>% filter(!is.na(no_mets_vs_any)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_any",       # x-axis
  fill_var       = "no_mets_vs_any",       # ensures two colors (one for "No Mets", one for "Any Mets")
  scales         = "free_y",
  plot_title     = "MLH1+ : No Mets vs. Any Mets"
)

# 1) No Mets vs. LN Mets, MLH1-
plot_cell_types(
  data = combined_data_MLH1neg %>% filter(!is.na(no_mets_vs_ln)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_ln",
  fill_var       = "no_mets_vs_ln",
  scales         = "free_y",
  plot_title     = "MLH1- : No Mets vs. LN Mets"
)

# 2) No Mets vs. Distant Mets, MLH1-
plot_cell_types(
  data = combined_data_MLH1neg %>% filter(!is.na(no_mets_vs_distant)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_distant",
  fill_var       = "no_mets_vs_distant",
  scales         = "free_y",
  plot_title     = "MLH1- : No Mets vs. Distant Mets"
)

# 3) No Mets vs. Any Mets, MLH1-
plot_cell_types(
  data = combined_data_MLH1neg %>% filter(!is.na(no_mets_vs_any)),
  cell_type_cols = cell_types,
  group_var      = "no_mets_vs_any",
  fill_var       = "no_mets_vs_any",
  scales         = "free_y",
  plot_title     = "MLH1- : No Mets vs. Any Mets"
)


```
Examinine the ratio of Fib to CD8 Cells
```{r}
# For example, suppose you created this ratio column before:
combined_data <- combined_data %>%
  mutate(
    ratio_fib_cd8   = Fib   / (CD8nv + CD8mem),
    ratio_tumor_cd8 = Tumor / (CD8nv + CD8mem),
    ratio_treg_cd8  = Treg  / (CD8nv + CD8mem)
  )


combined_data %>%
  # (Optionally) filter out NA if any exist
  filter(!is.na(mets_cat)) %>%
  ggplot(aes(x = mets_cat, y = ratio_fib_cd8, fill = mets_cat)) +
  geom_boxplot() +
  labs(
    title = "Fib / (CD8nv + CD8mem) for No Mets, LN Mets, Distant Mets",
    x     = "Metastasis Category",
    y     = "Fib / CD8 ratio",
    fill  = "Metastasis Category"
  ) +
  theme_bw()

# Tumor / CD8 ratio across No Mets, LN Mets, Distant Mets
combined_data %>%
  filter(!is.na(mets_cat)) %>%
  ggplot(aes(x = mets_cat, y = ratio_tumor_cd8, fill = mets_cat)) +
  geom_boxplot() +
  labs(
    title = "Tumor / (CD8nv + CD8mem) for No Mets, LN Mets, Distant Mets",
    x     = "Metastasis Category",
    y     = "Tumor / CD8 ratio"
  ) +
  theme_bw()

combined_data %>%
  filter(!is.na(mets_cat)) %>%
  ggplot(aes(x = mets_cat, y = ratio_treg_cd8, fill = mets_cat)) +
  geom_boxplot() +
  labs(
    title = "Treg / (CD8nv + CD8mem) for No Mets, LN Mets, Distant Mets",
    x     = "Metastasis Category",
    y     = "Treg / CD8 ratio"
  ) +
  theme_bw()


# Suppose we already have these columns in combined_data:
# ratio_fib_cd8, ratio_tumor_cd8, ratio_treg_cd8

long_ratios <- combined_data %>%
  select(ratio_fib_cd8, ratio_tumor_cd8, ratio_treg_cd8,
         no_mets_vs_ln, mets_cat, MLH1) %>%   # plus any relevant grouping columns
  pivot_longer(
    cols      = starts_with("ratio_"),
    names_to  = "ratio_type",
    values_to = "ratio_value"
  )

# Now create a boxplot faceted by ratio type
ggplot(long_ratios, aes(x = no_mets_vs_ln, y = ratio_value, fill = no_mets_vs_ln)) +
  geom_boxplot() +
  facet_wrap(~ ratio_type, scales = "free_y") +  # each ratio has its own y-axis scale
  labs(
    title = "Tumor/CD8, Treg/CD8, and Fib/CD8 Ratios: No Mets vs. LN Mets",
    x     = "Metastasis Category",
    y     = "Ratio Value"
  ) +
  theme_bw()
```
```{r}
library(ggplot2)
library(dplyr)
library(patchwork)

# Assuming 'CD8mem' is your column for CD8 proportions
# and 'mets_cat' has levels: "No Metastasis", "LN Metastasis", "Distant Metastasis"
# Also assuming 'MLH1' has exactly two levels: "MLH1 Negative", "MLH1 Positive"

p_MLH1_pos <- combined_data %>%
  filter(MLH1 == "MLH1 Positive") %>%
  ggplot(aes(x = mets_cat, y = Epithelial)) +
  geom_boxplot(fill = "#66C2A5") +
  labs(
    title = "Epithelial Proportions: MLH1 Positive",
    x = "Metastasis Category",
    y = "Epithelial Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_MLH1_neg <- combined_data %>%
  filter(MLH1 == "MLH1 Negative") %>%
  ggplot(aes(x = mets_cat, y = Epithelial)) +
  geom_boxplot(fill = "#FC8D62") +
  labs(
    title = "Epithelial Proportions: MLH1 Negative",
    x = "Metastasis Category",
    y = "Epithelial Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display both plots side by side
p_MLH1_pos + p_MLH1_neg
```
```{r}
library(ggplot2)
library(dplyr)

# One box plot showing Fib proportions for all MLH1 statuses
# across your three metastasis categories.
p_fib_all <- combined_data %>%
  ggplot(aes(x = mets_cat, y = Tumor)) +
  geom_boxplot(fill = "#66C2A5") +
  labs(
    title = "Tumor Proportions (All MLH1 Statuses)",
    x = "Metastasis Category",
    y = "Tumor Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_fib_all
```

