---
title: "Deconvolution_DH_and_Healthy"
author: "Alos Diallo"
date: "2025-07-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Good don't touch
Suplemental Figure 1: Validation box plots
```{r}
###############################################################################
#  deconv_DH_GTEx_TCGA_boxplots.R  ·  Figure 2 draft (v1.6)
#  DH tumors vs GTEx healthy vs TCGA adjacent normals
#  HiTIMED (h = 1:6) + EpiDISH(Fib)  → boxplots (PDF, no titles)
#  Author: Alos Diallo  ·  Date: 2025-07-22
###############################################################################

## ── USER PATHS ──────────────────────────────────────────────────────────────
path_dh_epic_v1    <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
path_dh_epic_v2    <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
path_dh_meta_csv   <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

path_gtex_beta_rds <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/GTEx/GTEx_samples.rds"

path_tcga_normal_idats <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/normal_idats"
path_tcga_normal_meta  <- file.path(path_tcga_normal_idats, "tcga_normal_metadata.csv")

out_root <- "/Users/adiallo/Desktop/Thesis/Results/Fig2_allNormals"
ih_vec   <- 1:6

redo_preprocess_dh <- TRUE
cache_beta_dh_rds  <- file.path(out_root, "beta_DH_combined_noXY.rds")
remove_sex_chr     <- TRUE

reuse_if_exists <- TRUE
set.seed(42)

## ── LIBS ────────────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(minfi);  library(sesame);   library(ENmix)
  library(HiTIMED); library(EpiDISH)
  library(tidyverse); library(ggpubr); library(rstatix)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
})

dir.create(out_root, showWarnings = FALSE, recursive = TRUE)
msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

## ── REUSE HELPER ────────────────────────────────────────────────────────────
use_if_exists <- function(name, fun){
  if (reuse_if_exists && exists(name, envir = .GlobalEnv)) {
    msg("Using", name, "from RAM.")
    get(name, envir = .GlobalEnv)
  } else {
    val <- fun()
    assign(name, val, envir = .GlobalEnv)
    val
  }
}

## ── LOADERS & HELPERS ───────────────────────────────────────────────────────
read_beta_DH <- function(path_v1, path_v2, cache_rds = NULL, redo = FALSE, drop_sex = TRUE){
  if (!is.null(cache_rds) && file.exists(cache_rds) && !redo){
    msg("Reading DH β from cache: ", cache_rds)
    return(readRDS(cache_rds))
  }
  msg("Reading DH IDATs…")
  rg1 <- read.metharray.exp(path_v1, recursive = TRUE, extended = TRUE, force = TRUE)
  rg2 <- read.metharray.exp(path_v2, recursive = TRUE, extended = TRUE, force = TRUE)

  msg("NOOB normalisation (DH)…")
  beta1 <- getBeta(preprocessNoob(rg1, dyeMethod = "single"))
  beta2 <- getBeta(preprocessNoob(rg2, dyeMethod = "single")) %>% sesame::betasCollapseToPfx()

  common <- intersect(rownames(beta1), rownames(beta2))
  beta   <- cbind(beta1[common, ], beta2[common, ])

  bad <- union(ENmix::QCinfo(rg1)$badCpG, ENmix::QCinfo(rg2)$badCpG)
  if (drop_sex) {
    anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
    bad  <- union(bad, rownames(anno)[anno$chr %in% c("chrX","chrY")])
  }
  beta <- beta[setdiff(rownames(beta), bad), ]

  if (!is.null(cache_rds)) saveRDS(beta, cache_rds)
  beta
}

read_beta_TCGA <- function(path_idats, drop_sex = TRUE){
  msg("Reading TCGA normal IDATs…")
  rg <- read.metharray(
    basenames = unique(sub("_Grn.idat|_Red.idat", "",
                           list.files(path_idats, full.names = TRUE, pattern = "idat$"))),
    force = TRUE)
  colnames(rg) <- sub("_.*", "", colnames(rg))

  msg("NOOB normalisation (TCGA)…")
  beta <- getBeta(preprocessNoob(rg))
  if (drop_sex) {
    anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
    beta <- beta[!(rownames(beta) %in% rownames(anno)[anno$chr %in% c("chrX","chrY")]), ]
  }
  sesame::betasCollapseToPfx(beta)
}

read_beta_GTEx <- function(path_rds, drop_sex = TRUE){
  msg("Reading GTEx β-matrix RDS…")
  beta <- readRDS(path_rds) %>% as.matrix()
  if (drop_sex) {
    anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
    beta <- beta[!(rownames(beta) %in% rownames(anno)[anno$chr %in% c("chrX","chrY")]), ]
  }
  sesame::betasCollapseToPfx(beta)
}

safe_hitimed <- function(beta_mat, tumor_type = "COAD", h = 1, tissue_type = "tumor"){
  out <- HiTIMED_deconvolution(beta_mat,
                               tumor_type = tumor_type,
                               h          = h,
                               tissue_type = tissue_type)
  if (is.null(out) || length(out) == 0)
    stop(sprintf("HiTIMED returned empty output (h=%s). Likely lost signature CpGs.", h))
  as.data.frame(out) %>% rownames_to_column("SampleID")
}

run_hitimed_all <- function(beta_mat, tumor_type = "COAD", tissue_type = "tumor", h_vec = 1:6){
  res <- setNames(vector("list", length(h_vec)), paste0("h", h_vec))
  for (h in h_vec) {
    msg("    HiTIMED: h =", h)
    res[[paste0("h", h)]] <- safe_hitimed(beta_mat, tumor_type, h, tissue_type)
  }
  res
}

run_epidish_fib <- function(beta_mat){
  data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
  common <- intersect(rownames(beta_mat), rownames(centEpiFibIC.m))
  epidish(beta_mat[common, ], centEpiFibIC.m[common, ], "RPC")$estF %>%
    as.data.frame() %>% dplyr::select(Fib) %>% rownames_to_column("SampleID")
}

normalise_fractions <- function(df){
  master <- c("Bas","Eos","Neu","Mono","DC","Bnv","Bmem",
              "CD4nv","CD4mem","Treg","CD8nv","CD8mem","NK","Immune")
  immune_cols <- intersect(master, names(df))
  if (length(immune_cols) < 2) return(df)
  df %>%
    mutate(Immune_total = rowSums(across(all_of(immune_cols)), na.rm = TRUE)) %>%
    mutate(across(all_of(immune_cols),
                  ~ if_else(Immune_total > 0, .x / Immune_total, NA_real_),
                  .names = "{.col}_norm"))
}

make_group_var <- function(meta_df){
  meta_df %>%
    mutate(group4 = case_when(
      dataset == "GTEx"                 ~ "GTEx healthy",
      dataset == "TCGA_N"               ~ "TCGA adjacent normal",
      dataset == "DH" & any_mets == "No Mets" ~ "DH non-mets",
      dataset == "DH" & any_mets != "No Mets" ~ "DH mets",
      TRUE ~ dataset
    )) %>%
    mutate(group4 = factor(group4,
                           levels = c("GTEx healthy","TCGA adjacent normal",
                                      "DH non-mets","DH mets")))
}

# UPDATED plotting function with new jitter parameters
# UPDATED plotting function with "thin box" settings
draw_plots_level <- function(combined_df, cell_cols, level_tag, out_dir){
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  long_df <- combined_df %>%
    pivot_longer(cols = all_of(cell_cols), names_to = "Cell", values_to = "Prop") %>%
    filter(!is.na(group4))

  stats_tbl <- long_df %>%
    group_by(Cell) %>%
    kruskal_test(Prop ~ group4) %>%
    left_join(
      long_df %>% group_by(Cell) %>%
        dunn_test(Prop ~ group4, p.adjust.method = "BH"),
      by = "Cell", suffix = c(".kw", ".dunn")
    )
  write_csv(stats_tbl, file.path(out_dir, paste0("stats_KW_Dunn_", level_tag, ".csv")))

  pal_groups <- c("GTEx healthy"         = "#1b9e77",
                  "TCGA adjacent normal" = "#7570b3",
                  "DH non-mets"          = "#d95f02",
                  "DH mets"              = "#e7298a")

  msg("    Saving individual cell plots (PDF)...")
  purrr::walk(unique(long_df$Cell), function(cl){
    dfc   <- filter(long_df, Cell == cl)
    comps <- combn(levels(dfc$group4), 2, simplify = FALSE)

    p <- ggplot(dfc, aes(group4, Prop, fill = group4)) +
      geom_boxplot(
        outlier.shape = NA,
        width = 0.25, # <--- EDIT: Was 0.65
        alpha = 0.85, # <--- EDIT: Added alpha for consistency
        color = "gray25" # <--- EDIT: Added color for consistency
      ) +
      geom_jitter(
        width = 0.05, # <--- EDIT: Was 0.2
        size = 0.8, # <--- EDIT: Was 1.2, reduced to match tighter jitter
        alpha = 0.6 # <--- EDIT: Was 0.8
      ) +
      stat_compare_means(comparisons = comps, method = "wilcox.test",
                         label = "p.signif", hide.ns = TRUE, size = 3) +
      scale_fill_manual(values = pal_groups, guide = "none") +
      # <--- EDIT: Added y-axis expansion to give stats more space
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
      theme_bw(base_size = 11) +
      theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8), # <--- EDIT: 90 degrees
        plot.title  = element_blank(),
        axis.title.x = element_blank(), # <--- EDIT: Added to remove x-axis label
        axis.ticks = element_blank(), # <--- EDIT: Added to match example
        panel.grid = element_blank(), # <--- EDIT: Was panel.grid.major/minor
        panel.border = element_blank() # <--- EDIT: Added to match example
      ) +
      labs(x = NULL, y = "Proportion")

    ggsave(file.path(out_dir,
                     paste0("BOX_", gsub("[^A-Za-z0-9]+","_", cl), "_", level_tag, ".pdf")),
           p,
           width = 2.5, # <--- EDIT: Was 4.2. Reduced width to compress groups.
           height = 4.2,
           device = cairo_pdf)
  })

  msg("    Saving faceted plot (PDF)...")
  comps_all <- combn(levels(long_df$group4), 2, simplify = FALSE)
  p_all <- ggplot(long_df, aes(group4, Prop, fill = group4)) +
    geom_boxplot(
      outlier.shape = NA,
      width = 0.25, # <--- EDIT: Was 0.65
      alpha = 0.85, # <--- EDIT: Was 0.7, matched style
      color = "gray25" # <--- EDIT: Added color
    ) +
    geom_jitter(
      width = 0.05, # <--- EDIT: Was 0.15
      size = 0.4, # <--- EDIT: Was 0.8, matched style
      alpha = 0.6 # <--- EDIT: Was 0.7, matched style
    ) +
    facet_wrap(~ Cell, scales = "free_y", ncol = 5) +
    stat_compare_means(comparisons = comps_all, method = "wilcox.test",
                       label = "p.signif", hide.ns = TRUE, size = 2.1) + # <--- EDIT: Was 2.8
    scale_fill_manual(values = pal_groups, guide = "none") +
    # <--- EDIT: Added y-axis expansion
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7), # <--- EDIT: 90 degrees
      plot.title  = element_blank(),
      axis.title.x = element_blank(), # <--- EDIT: Added
      axis.ticks = element_blank(), # <--- EDIT: Added
      panel.grid = element_blank(), # <--- EDIT: Was panel.grid.major/minor
      panel.border = element_blank(), # <--- EDIT: Added
      strip.background = element_blank(), # <--- EDIT: Added
      strip.text = element_text(face = "bold", size = 8.5) # <--- EDIT: Added
    ) +
    labs(x = NULL, y = "Proportion")

  ggsave(file.path(out_dir, paste0("FACET_allCells_", level_tag, ".pdf")),
         p_all,
         width = 8.0, # <--- EDIT: Was 11. Reduced to compress facets.
         height = 8.8,
         device = cairo_pdf)
}

## ── MAIN ────────────────────────────────────────────────────────────────────
msg("======= Starting deconvolution pipeline =======")

# 1) β-matrices
DH_beta   <- use_if_exists("DH_beta",   function() read_beta_DH(path_dh_epic_v1, path_dh_epic_v2, cache_beta_dh_rds, redo_preprocess_dh, remove_sex_chr))
GTEx_beta <- use_if_exists("GTEx_beta", function() read_beta_GTEx(path_gtex_beta_rds, remove_sex_chr))
TCGA_beta <- use_if_exists("TCGA_beta", function() read_beta_TCGA(path_tcga_normal_idats, remove_sex_chr))

# 2) Metadata

meta_raw <- read_csv(path_dh_meta_csv, show_col_types = FALSE)           # read once
has_node <- "node_stage" %in% names(meta_raw)                         # flag

meta_DH <- meta_raw %>%
 
  mutate(
    # 1. Create a unique SampleID
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),

    # 2. Create a clean factor for MLH1 status
    MLH1_status = factor(ifelse(MLH1 == 1, "MLH1 Positive", "MLH1 Negative")),

    # 3. Create the main 3-level metastasis category (renamed to 'mets_cat')
     mets_cat = case_when(
       Distant_Mets == TRUE               ~ "Distant",
       has_node & node_stage > 0          ~ "LN-only",
       !has_node & ln_only == TRUE        ~ "LN-only",
       TRUE                               ~ "No Mets"
     ),

    # 4. Convert 'mets_cat' to an ordered factor for plotting
    mets_cat = factor(mets_cat,
                      # <--- THIS IS THE FIX: Levels now match the values from Step 3
                      levels = c("No Mets", "LN-only", "Distant")),

    # 5. Create the simplified binary factor for any metastasis
     any_mets = factor(ifelse(mets_cat == "No Mets",
                                    "No Mets", "Mets")),
     ln_only  = factor(ifelse(mets_cat == "LN-only",
                                    "LN Mets", "No LN Mets")),
     dist     = factor(ifelse(mets_cat == "Distant",
                                    "Distant Mets", "No Distant Mets")),

    # 6. Add the dataset label
    dataset = "DH"
  )

meta_TCGA <- read_csv(path_tcga_normal_meta, show_col_types = FALSE) %>%
  rename(SampleID = geo_accession) %>%
  mutate(dataset = "TCGA_N") %>%
  dplyr::select(SampleID, dataset)

meta_GTEx <- tibble(SampleID = colnames(GTEx_beta), dataset = "GTEx")

# 3) HiTIMED + Fibro
msg("Running HiTIMED for DH (tumor)…")
HIT_DH   <- use_if_exists("HIT_DH",   function() run_hitimed_all(DH_beta,   tumor_type = "COAD", tissue_type = "tumor",  h_vec = ih_vec))
msg("Running HiTIMED for GTEx (healthy)…")
HIT_GTEx <- use_if_exists("HIT_GTEx", function() run_hitimed_all(GTEx_beta, tumor_type = "COAD", tissue_type = "normal", h_vec = ih_vec))
msg("Running HiTIMED for TCGA normals…")
HIT_TCGA <- use_if_exists("HIT_TCGA", function() run_hitimed_all(TCGA_beta, tumor_type = "COAD", tissue_type = "normal", h_vec = ih_vec))

msg("Running EpiDISH Fibroblast…")
fib_DH   <- use_if_exists("fib_DH",   function() run_epidish_fib(DH_beta))
fib_GTEx <- use_if_exists("fib_GTEx", function() run_epidish_fib(GTEx_beta))
fib_TCGA <- use_if_exists("fib_TCGA", function() run_epidish_fib(TCGA_beta))

library(dplyr)
## 4) Loop per h-level
for (h in ih_vec) {
  level_tag <- paste0("h", h)
  msg("\n==============================")
  msg(">>> Level:", level_tag)

  frac_DH   <- HIT_DH  [[level_tag]]
  frac_GTEx <- HIT_GTEx[[level_tag]]
  frac_TCGA <- HIT_TCGA[[level_tag]]

  make_combined <- function(frac_df, meta_df, fib_df, label){
    meta_df %>%
      mutate(dataset = label) %>%
      left_join(frac_df, by = "SampleID") %>%
      left_join(fib_df , by = "SampleID") %>%
      normalise_fractions()
  }

  comb_DH   <- make_combined(frac_DH,   meta_DH,   fib_DH,   "DH")
  comb_GTEx <- make_combined(frac_GTEx, meta_GTEx, fib_GTEx, "GTEx")
  comb_TCGA <- make_combined(frac_TCGA, meta_TCGA, fib_TCGA, "TCGA_N")

  combined  <- bind_rows(comb_GTEx, comb_TCGA, comb_DH) %>% make_group_var()

  cell_cols <- c(setdiff(names(frac_DH), "SampleID"),
                 "Fib",
                 grep("_norm$", names(combined), value = TRUE))

  out_lvl <- file.path(out_root, level_tag)
  draw_plots_level(combined, cell_cols, level_tag, out_lvl)
  msg("✓  Level", level_tag, "done →", out_lvl)
}

msg("\n✅  All levels complete. Outputs in: ", out_root)
```

Good don't touch
Figure 2a: Stacked Bar plots
```{r}
###############################################################################
#  stackbars_DH_after_deconv.R  ·  run AFTER the main deconv script
#  Stacked bars for DH only (No/LN/Distant mets)
#    • No sample IDs, no y-axis labels/ticks
#    • No facet strip labels
#    • No grid lines
#    • Legend written as a separate PDF
#    • Width/height scale with #samples
#  Author: Alos Diallo  ·  Date: 2025-07-22
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(scales)
  library(cowplot); library(grid)
})

## ── USER SETTINGS ───────────────────────────────────────────────────────────
level_tag        <- "h6"   # pick a level present in HIT_DH
drop_norm_cols   <- TRUE   # remove *_norm cols from bars
cell_type_vars   <- NULL   # NULL → auto-detect
hide_sample_ids  <- TRUE
hide_y_axis      <- TRUE
hide_facet_strips<- TRUE

# sizing
per_sample_w   <- 0.25     # inches per sample (tweak as needed)
min_width_each <- 12
min_width_all  <- 20
height_each    <- 6
height_all     <- 7

# legend PDF size
legend_width   <- 3
legend_height  <- 6

# base palette (auto-extended if needed)
base_cols <- c(
  "#66C2A5","#FC8D62","#8DA0CB","#E78AC3",
  "#A6D854","#FFD92F","#E5C494","#B3B3B3",
  "#1F78B4","#33A02C","#FB9A99","#FDBF6F",
  "#CAB2D6","#6A3D9A","#E31A1C","#B15928","#C51B8A"
)

## ── PATHS ───────────────────────────────────────────────────────────────────
if (!exists("out_root")) out_root <- getwd()
out_dir <- file.path(out_root, level_tag, "stackbars_DH")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

## ── CHECKS ──────────────────────────────────────────────────────────────────
stopifnot(exists("HIT_DH"), exists("fib_DH"), exists("meta_DH"))
if (!level_tag %in% names(HIT_DH)) stop("level_tag not found in HIT_DH.")
if (!"mets_cat" %in% names(meta_DH)) stop("meta_DH must contain 'mets_cat'.")

## ── BUILD DH DATA ───────────────────────────────────────────────────────────
frac_DH <- HIT_DH[[level_tag]]
comb_DH <- meta_DH %>%
  left_join(frac_DH, by = "SampleID") %>%
  left_join(fib_DH , by = "SampleID")


comb_DH$mets_cat <- factor(comb_DH$mets_cat,
                           levels = c("No Metastasis","LN Metastasis","Distant Metastasis"))

## ── CHOOSE CELL COLUMNS ─────────────────────────────────────────────────────
if (is.null(cell_type_vars)) {
  meta_cols <- c("SampleID","mets_cat","any_mets_fct","dataset","Immune_total")
  ct_df <- comb_DH %>% dplyr::select(-any_of(meta_cols)) %>% select(where(is.numeric))
  if (drop_norm_cols) ct_df <- ct_df %>% dplyr::select(-matches("_norm$"))
  cell_type_vars <- names(ct_df)
}
if (!length(cell_type_vars)) stop("No numeric cell-type columns found.")

## ── LONG FORMAT & ORDERINGS ─────────────────────────────────────────────────
df_long <- comb_DH %>%
  dplyr::select(SampleID, mets_cat, all_of(cell_type_vars)) %>%
  pivot_longer(all_of(cell_type_vars), names_to = "CellType", values_to = "Proportion")

tumor_col <- if ("Tumor" %in% cell_type_vars) "Tumor" else cell_type_vars[1]

tumor_order <- df_long %>%
  filter(CellType == tumor_col) %>%
  arrange(mets_cat, desc(Proportion)) %>%
  pull(SampleID)

df_long$SampleID <- factor(df_long$SampleID, levels = tumor_order)

cell_type_order <- df_long %>%
  group_by(CellType) %>%
  summarise(mean_prop = mean(Proportion, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_prop)) %>%
  pull(CellType)

df_long$CellType <- factor(df_long$CellType, levels = rev(cell_type_order))

## ── PALETTE (auto-extend) ───────────────────────────────────────────────────
n_ct   <- nlevels(df_long$CellType)
pal_ct <- setNames(c(base_cols, scales::hue_pal()(max(0, n_ct - length(base_cols))))[1:n_ct],
                   levels(df_long$CellType))

## ── THEMING ─────────────────────────────────────────────────────────────────
x_text  <- if (hide_sample_ids) element_blank() else element_text(angle = 90, vjust = 0.5, hjust = 1)
x_ticks <- if (hide_sample_ids) element_blank() else element_line()

base_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = x_text,
    axis.ticks.x    = x_ticks,
    axis.ticks.y    = element_blank(),
    plot.title      = element_blank(),
    legend.position = "right"
  )

if (hide_facet_strips) {
  base_theme <- base_theme +
    theme(strip.text.x   = element_blank(),
          strip.background = element_blank())
}

if (hide_y_axis) {
  base_theme <- base_theme +
    theme(axis.title.y = element_blank(),
          axis.text.y  = element_blank(),
          axis.ticks.y = element_blank())
}

## ── PLOTTERS ────────────────────────────────────────────────────────────────
plot_cell_props <- function(data, pal, show_legend = FALSE){
  p <- ggplot(data, aes(SampleID, Proportion, fill = CellType)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(labels = percent_format()) +
    labs(x = NULL, y = "Cell type proportion") +
    base_theme
  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}

save_legend_pdf <- function(p_with_leg, file, w = legend_width, h = legend_height){
  leg_grob <- cowplot::get_legend(p_with_leg)
  leg_plot <- cowplot::ggdraw(leg_grob)
  ggsave(filename = file, plot = leg_plot,
         width = w, height = h, device = cairo_pdf)
}

## ── LEGEND EXPORT ───────────────────────────────────────────────────────────
dummy_for_leg <- plot_cell_props(df_long, pal_ct, show_legend = TRUE)
legend_file   <- file.path(out_dir, paste0("LEGEND_CellTypes_", level_tag, ".pdf"))
save_legend_pdf(dummy_for_leg, legend_file)
message("Saved legend: ", legend_file)

## ── SAVE PER-GROUP PDFs ─────────────────────────────────────────────────────
for (cat in levels(df_long$mets_cat)) {
  dcat <- df_long %>% filter(mets_cat == cat)
  p    <- plot_cell_props(dcat, pal_ct, show_legend = FALSE)

  n_samp <- length(unique(dcat$SampleID))
  w <- max(min_width_each, n_samp * per_sample_w)

  fn <- file.path(out_dir,
                  paste0("STACKBAR_", gsub("[^A-Za-z0-9]+","_", cat),
                         "_", level_tag, ".pdf"))
  ggsave(fn, p, width = w, height = height_each, device = cairo_pdf)
  message("Saved: ", fn)
}

## ── FACET (ALL GROUPS) PDF ──────────────────────────────────────────────────
p_all <- ggplot(df_long, aes(SampleID, Proportion, fill = CellType)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_grid(~ mets_cat, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = pal_ct) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Cell type proportion") +
  base_theme +
  theme(legend.position = "none")

n_total <- length(unique(df_long$SampleID))
w_all   <- max(min_width_all, n_total * per_sample_w)

fn_all <- file.path(out_dir, paste0("STACKBAR_allGroups_", level_tag, ".pdf"))
ggsave(fn_all, p_all, width = w_all, height = height_all, device = cairo_pdf)
message("Saved: ", fn_all)

message("✓ Done.")
```

```{r}
###############################################################################
# stackbars_DH_COMPLETE_DEBUG.R
# Purpose: Full deconvolution and visualization for DH tumor progression.
###############################################################################

## ── 1. LIBS ─────────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(minfi)
  library(sesame)
  library(ENmix)
  library(HiTIMED)
  library(EpiDISH)
  library(tidyverse)
  library(scales)
  library(cowplot)
  library(grid)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
})

## ── 2. PATHS & SETTINGS ─────────────────────────────────────────────────────
# Paths based on your Thesis data structure
path_dh_epic_v1  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
path_dh_epic_v2  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
path_dh_meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

# Output directory for results
out_root  <- "/Users/adiallo/Desktop/Thesis/Results/Fig2_allNormals"
level_tag <- "h6" 

# Create the folder immediately
out_dir <- file.path(out_root, level_tag, "stackbars_DH")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

## ── 3. DATA PROCESSING ──────────────────────────────────────────────────────
msg("Loading IDATs and performing NOOB normalization...")

# Read and normalize DH datasets
rg1 <- read.metharray.exp(path_dh_epic_v1, recursive = TRUE, extended = TRUE, force = TRUE)
rg2 <- read.metharray.exp(path_dh_epic_v2, recursive = TRUE, extended = TRUE, force = TRUE)

beta1 <- getBeta(preprocessNoob(rg1, dyeMethod = "single"))
beta2 <- getBeta(preprocessNoob(rg2, dyeMethod = "single")) %>% sesame::betasCollapseToPfx()

# Combine matrices
common <- intersect(rownames(beta1), rownames(beta2))
DH_beta <- cbind(beta1[common, ], beta2[common, ])

# Run Deconvolution (h6 + Fibroblasts)
msg("Running HiTIMED (Level 6) and EpiDISH...")
frac_DH <- HiTIMED_deconvolution(DH_beta, tumor_type = "COAD", h = 6, tissue_type = "tumor") %>%
  as.data.frame() %>% 
  rownames_to_column("SampleID")

# EpiDISH for Fibroblasts
data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
common_fib <- intersect(rownames(DH_beta), rownames(centEpiFibIC.m))
fib_DH <- epidish(DH_beta[common_fib, ], centEpiFibIC.m[common_fib, ], "RPC")$estF %>%
  as.data.frame() %>% 
  dplyr::select(Fib) %>% 
  rownames_to_column("SampleID")

# Load and process Metadata
meta_DH <- read_csv(path_dh_meta_csv, show_col_types = FALSE) %>%
  mutate(
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    mets_cat = case_when(
      Distant_Mets == TRUE ~ "Distant Metastasis",
      ("node_stage" %in% names(.) & node_stage > 0) ~ "LN Metastasis",
      TRUE ~ "No Metastasis"
    ),
    mets_cat = factor(mets_cat, levels = c("No Metastasis","LN Metastasis","Distant Metastasis"))
  )

# Combine for plotting
comb_DH <- meta_DH %>% 
  left_join(frac_DH, by = "SampleID") %>% 
  left_join(fib_DH, by = "SampleID")

## ── 4. VISUALIZATION PREP ───────────────────────────────────────────────────
# Clean up numeric columns for plotting
ct_vars <- comb_DH %>% 
  dplyr::select(-any_of(c("SampleID", "mets_cat", "Immune_total", "any_mets_fct", "dataset"))) %>% 
  select(where(is.numeric))
cell_type_vars <- names(ct_vars)

df_long <- comb_DH %>%
  dplyr::select(SampleID, mets_cat, all_of(cell_type_vars)) %>%
  pivot_longer(all_of(cell_type_vars), names_to = "CellType", values_to = "Proportion")

# Sort samples by Tumor proportion (descending) within categories
tumor_col <- if ("Tumor" %in% cell_type_vars) "Tumor" else cell_type_vars[1]
tumor_order <- df_long %>% 
  filter(CellType == tumor_col) %>% 
  arrange(mets_cat, desc(Proportion)) %>% 
  pull(SampleID)

df_long$SampleID <- factor(df_long$SampleID, levels = unique(tumor_order))

# Order CellTypes by average abundance
cell_type_order <- df_long %>% 
  group_by(CellType) %>% 
  summarise(m = mean(Proportion, na.rm = TRUE)) %>% 
  arrange(desc(m)) %>% 
  pull(CellType)

df_long$CellType <- factor(df_long$CellType, levels = rev(cell_type_order))

# Set colors
base_cols <- c("#66C2A5","#FC8D62","#8DA0CB","#E78AC3","#A6D854","#FFD92F",
               "#E5C494","#B3B3B3","#1F78B4","#33A02C","#FB9A99","#FDBF6F",
               "#CAB2D6","#6A3D9A","#E31A1C","#B15928","#C51B8A")
pal_ct <- setNames(base_cols[1:nlevels(df_long$CellType)], levels(df_long$CellType))

## ── 5. SAVING OUTPUTS ───────────────────────────────────────────────────────
msg("Starting file export to:", out_dir)

# 1. Legend PDF
p_leg <- ggplot(df_long, aes(SampleID, Proportion, fill = CellType)) + 
  geom_bar(stat="identity", position="fill") + 
  scale_fill_manual(values=pal_ct) + 
  theme_minimal()
leg_plot <- cowplot::ggdraw(cowplot::get_legend(p_leg))
ggsave(file.path(out_dir, "LEGEND_h6.pdf"), leg_plot, width=3, height=6, device=cairo_pdf)

# 2. All Groups Faceted PDF
p_all <- ggplot(df_long, aes(SampleID, Proportion, fill = CellType)) +
  geom_bar(stat="identity", position="fill") +
  facet_grid(~ mets_cat, scales="free_x", space="free_x") +
  scale_fill_manual(values=pal_ct) +
  theme_minimal() + 
  theme(legend.position="none", axis.text.x=element_blank(), panel.grid=element_blank()) +
  scale_y_continuous(labels = percent_format())
ggsave(file.path(out_dir, "STACKBAR_allGroups_h6.pdf"), p_all, width=20, height=7, device=cairo_pdf)

# 3. Individual Group PDFs
for (cat in levels(df_long$mets_cat)) {
  dcat <- df_long %>% filter(mets_cat == cat)
  if(nrow(dcat) == 0) next
  
  p_ind <- ggplot(dcat, aes(SampleID, Proportion, fill = CellType)) +
    geom_bar(stat="identity", position="fill") +
    scale_fill_manual(values=pal_ct) +
    theme_minimal() + 
    theme(legend.position="none", axis.text.x=element_blank(), panel.grid=element_blank()) +
    scale_y_continuous(labels = percent_format())
  
  ind_path <- file.path(out_dir, paste0("STACKBAR_", gsub(" ", "_", cat), "_h6.pdf"))
  ggsave(ind_path, p_ind, width=12, height=6, device=cairo_pdf)
  msg("Saved:", ind_path)
}

msg("✅ PROCESS COMPLETE. Files are located in:", out_dir)
```


Old
Figure 2C: Disease progression box plots
```{r}
###############################################################################
#  deconv_DH_progression_boxplots.R · Figure 2 draft (v2.8 - P-Value Position)
#  DH tumors: No Mets vs LN-only Mets vs Distant Mets
#  HiTIMED (h = 1:6) + EpiDISH(Fib) → faceted + individual boxplots (PDF)
#  Author: Alos Diallo · Date: 2025-08-08
#  Adapted from v2.7 to fix p-value label positioning.
###############################################################################

## ── USER PATHS ──────────────────────────────────────────────────────────────
path_dh_epic_v1   <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
path_dh_epic_v2   <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
path_dh_meta_csv  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

out_root <- "/Users/adiallo/Desktop/Thesis/Results/Fig2_DH_progression"
ih_vec   <- 1:6

redo_preprocess_dh <- TRUE
cache_beta_dh_rds  <- file.path(out_root, "beta_DH_combined_noXY.rds")
remove_sex_chr     <- TRUE

reuse_if_exists <- TRUE
set.seed(42)

## ── LIBS ────────────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(minfi);  library(sesame);   library(ENmix)
  library(HiTIMED); library(EpiDISH)
  library(tidyverse); library(ggpubr); library(rstatix)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
})

dir.create(out_root, showWarnings = FALSE, recursive = TRUE)
msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

## ── REUSE HELPER & LOADERS (Unchanged) ──────────────────────────────────────
use_if_exists <- function(name, fun){
  if (reuse_if_exists && exists(name, envir = .GlobalEnv)) {
    msg("Using", name, "from RAM.")
    get(name, envir = .GlobalEnv)
  } else {
    val <- fun()
    assign(name, val, envir = .GlobalEnv)
    val
  }
}

read_beta_DH <- function(path_v1, path_v2, cache_rds = NULL, redo = FALSE, drop_sex = TRUE){
  if (!is.null(cache_rds) && file.exists(cache_rds) && !redo){
    msg("Reading DH β from cache: ", cache_rds)
    return(readRDS(cache_rds))
  }
  msg("Reading DH IDATs…")
  rg1 <- read.metharray.exp(path_v1, recursive = TRUE, extended = TRUE, force = TRUE)
  rg2 <- read.metharray.exp(path_v2, recursive = TRUE, extended = TRUE, force = TRUE)
  msg("NOOB normalisation (DH)…")
  beta1 <- getBeta(preprocessNoob(rg1, dyeMethod = "single"))
  beta2 <- getBeta(preprocessNoob(rg2, dyeMethod = "single")) %>% sesame::betasCollapseToPfx()
  common <- intersect(rownames(beta1), rownames(beta2))
  beta   <- cbind(beta1[common, ], beta2[common, ])
  bad <- union(ENmix::QCinfo(rg1)$badCpG, ENmix::QCinfo(rg2)$badCpG)
  if (drop_sex) {
    anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
    bad  <- union(bad, rownames(anno)[anno$chr %in% c("chrX","chrY")])
  }
  beta <- beta[setdiff(rownames(beta), bad), ]
  if (!is.null(cache_rds)) saveRDS(beta, cache_rds)
  beta
}

run_hitimed_all <- function(beta_mat, tumor_type = "COAD", tissue_type = "tumor", h_vec = 1:6){
  res <- setNames(vector("list", length(h_vec)), paste0("h", h_vec))
  for (h in h_vec) {
    msg("    HiTIMED: h =", h)
    res[[paste0("h", h)]] <- HiTIMED_deconvolution(beta_mat, tumor_type, h, tissue_type) %>%
      as.data.frame() %>% rownames_to_column("SampleID")
  }
  res
}

run_epidish_fib <- function(beta_mat){
  data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
  common <- intersect(rownames(beta_mat), rownames(centEpiFibIC.m))
  epidish(beta_mat[common, ], centEpiFibIC.m[common, ], "RPC")$estF %>%
    as.data.frame() %>% dplyr::select(Fib) %>% rownames_to_column("SampleID")
}

normalise_fractions <- function(df){
  master <- c("Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem",
              "CD4nv", "CD4mem", "Treg", "CD4T",
              "CD8nv", "CD8mem", "CD8T",
              "NK", "Immune")
  immune_cols <- intersect(master, names(df))
  if (length(immune_cols) < 2) return(df)
  df %>%
    mutate(Immune_total = rowSums(across(all_of(immune_cols)), na.rm = TRUE)) %>%
    mutate(across(all_of(immune_cols),
                  ~ if_else(Immune_total > 0, .x / Immune_total, NA_real_),
                  .names = "{.col}_norm"))
}

make_progression_group <- function(meta_df){
  has_node_stage_col <- "node_stage" %in% names(meta_df)
  meta_df %>%
    mutate(
      mets_progression = case_when(
        Distant_Mets == TRUE ~ "Distant Mets",
        (has_node_stage_col & !is.na(node_stage) & node_stage > 0) | ln_only == TRUE ~ "LN-only Mets",
        TRUE ~ "No Mets"
      ),
      mets_progression = factor(mets_progression,
                                levels = c("No Mets", "LN-only Mets", "Distant Mets"))
    )
}

## ── PLOTTING FUNCTION (Updated) ─────────────────────────────────────────────
draw_progression_plots <- function(combined_df, cell_cols, level_tag, out_dir){
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  long_df <- combined_df %>%
    pivot_longer(cols = all_of(cell_cols), names_to = "Cell", values_to = "Prop") %>%
    filter(!is.na(mets_progression))
  pal_progression <- c("No Mets" = "#00AFBB", "LN-only Mets" = "#E7B800", "Distant Mets" = "#FC4E07")
  comps <- combn(levels(long_df$mets_progression), 2, simplify = FALSE)

  msg("    Saving individual cell plots (PDF)…")
  purrr::walk(unique(long_df$Cell), function(cl) {
    df_cell <- filter(long_df, Cell == cl)
    
    # <<< Determine y-position for the KW p-value dynamically
    # This places it at 95% of the range, above the highest data point.
    kw_y_pos <- max(df_cell$Prop, na.rm = TRUE) * 0.95

    p_indiv <- ggplot(df_cell, aes(mets_progression, Prop, fill = mets_progression)) +
      geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.7, color = "gray20") +
      geom_jitter(width = 0.15, size = 0.8, alpha = 0.6) +
      # <<< Let ggpubr handle pairwise bracket positions automatically
      stat_compare_means(comparisons = comps, method = "wilcox.test", label = "p.signif", hide.ns = TRUE, size = 3.5, tip.length = 0.01) +
      # <<< Use the calculated y-position for the KW p-value
      stat_compare_means(method = "kruskal.test", label.y = kw_y_pos, size = 4) +
      scale_fill_manual(values = pal_progression, guide = "none") +
      # <<< Expand the top of the plot by 20% to make ample room
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
      theme_bw(base_size = 12) +
      theme(
        axis.text.x = element_text(angle = 35, hjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
      ) +
      labs(x = NULL, y = "Estimated Proportion")
    safe_cell_name <- gsub("[^A-Za-z0-9]+", "_", cl)
    ggsave(file.path(out_dir, paste0("BOX_", safe_cell_name, "_", level_tag, ".pdf")), p_indiv, width = 4.5, height = 5, device = cairo_pdf)
  })

  msg("    Saving faceted plot for progression (PDF)…")
  p_all <- ggplot(long_df, aes(mets_progression, Prop, fill = mets_progression)) +
    geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.7, color = "gray20") +
    geom_jitter(width = 0.15, size = 0.5, alpha = 0.6) +
    facet_wrap(~ Cell, scales = "free_y", ncol = 5) +
    # <<< Let ggpubr handle positioning automatically, using step.increase for stacking
    stat_compare_means(comparisons = comps, method = "wilcox.test", label = "p.signif", hide.ns = TRUE, size = 2.8, step.increase = 0.07, tip.length = 0.01) +
    # <<< Place KW p-value at the top 98% of each facet
    stat_compare_means(method = "kruskal.test", label.y.npc = 0.98, size = 3) +
    scale_fill_manual(values = pal_progression, name = "Progression Status") +
    theme_bw(base_size = 11) +
    theme(
        axis.text.x = element_text(angle = 35, hjust = 1),
        plot.title  = element_blank(),
        legend.position = "bottom",
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "gray90", color = "gray50")
    ) +
    labs(x = NULL, y = "Estimated Proportion")
  ggsave(file.path(out_dir, paste0("FACET_allCells_progression_", level_tag, ".pdf")), p_all, width = 11, height = 9.5, device = cairo_pdf)
}

## ── MAIN (Unchanged) ────────────────────────────────────────────────────────
msg("======= Starting DH Progression deconvolution pipeline =======")

# 1) β-matrix
DH_beta  <- use_if_exists("DH_beta", function() read_beta_DH(path_dh_epic_v1, path_dh_epic_v2, cache_beta_dh_rds, redo_preprocess_dh, remove_sex_chr))

# 2) Metadata
meta_DH <- read_csv(path_dh_meta_csv, show_col_types = FALSE) %>%
  mutate(SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_")) %>%
  make_progression_group() %>%
  dplyr::select(SampleID, mets_progression, everything())

# 3) HiTIMED + Fibroblast Deconvolution
msg("Running HiTIMED for DH (tumor)…")
HIT_DH  <- use_if_exists("HIT_DH", function() run_hitimed_all(DH_beta, tumor_type = "COAD", h_vec = ih_vec))

msg("Running EpiDISH Fibroblast for DH…")
fib_DH  <- use_if_exists("fib_DH", function() run_epidish_fib(DH_beta))

## 4) Loop per h-level, Combine Data, and Plot
for (h in ih_vec) {
  level_tag <- paste0("h", h)
  msg("\n==============================")
  msg(">>> Level:", level_tag)

  frac_DH  <- HIT_DH[[level_tag]]

  combined  <- meta_DH %>%
    left_join(frac_DH, by = "SampleID") %>%
    left_join(fib_DH , by = "SampleID") %>%
    normalise_fractions()

  cell_cols <- c(setdiff(names(frac_DH), "SampleID"),
                 "Fib",
                 grep("_norm$", names(combined), value = TRUE))

  cell_cols <- unique(cell_cols)
  cell_cols <- cell_cols[sapply(cell_cols, function(c) c %in% names(combined) && !all(is.na(combined[[c]])))]

  out_lvl <- file.path(out_root, level_tag)
  draw_progression_plots(combined, cell_cols, level_tag, out_lvl)
  msg("✓  Level", level_tag, "done →", out_lvl)
}

```

Figure 2B
```{r}
#New Code 10/20/25
###############################################################################
#  Compact normalized facet plot (HiTIMED level 5 only)
#  Automatically combines Bmem/Bnv -> Bcell_norm and always retains facet
#  Output: FACET_normCells_progression_h5_compact_BcellsCombined_fixed.pdf
#  VERSION: Reverted numeric x-axis. Using ggsave width to control spacing.
###############################################################################

if ("HIT_DH" %in% ls() && "fib_DH" %in% ls() && "meta_DH" %in% ls()) {

  msg("\n==============================")
  msg(">>> Generating compact normalized facet plot for HiTIMED level 5 (B cells combined)")

  level_tag <- "h5"
  frac_DH   <- HIT_DH[[level_tag]]

  combined <- meta_DH %>%
    left_join(frac_DH, by = "SampleID") %>%
    left_join(fib_DH, by = "SampleID") %>%
    normalise_fractions()

  # --- Collapse B-cell subtypes into a single SUMMED column ---
  b_all_cols <- intersect(names(combined), c("Bmem", "Bnv", "Bmem_norm", "Bnv_norm", "Bcell"))
  b_candidates_for_sum <- intersect(names(combined), c("Bmem_norm", "Bnv_norm"))

  if (length(b_candidates_for_sum) > 0) {
    msg("    Collapsing B-cell subtypes (using _norm) -> single 'Bcell_norm' column.")
    combined$Bcell_norm <- rowSums(combined[, b_candidates_for_sum], na.rm = TRUE)
  } else {
    msg("    No explicit B-cell _norm subtypes found -- checking for existing Bcell columns.")
    if ("Bcell" %in% names(combined)) {
      combined$Bcell_norm <- combined$Bcell
    } else {
      b_candidates_base <- intersect(names(combined), c("Bmem", "Bnv"))
      if(length(b_candidates_base) > 0) {
          msg("    Using non-normalized B-cell subtypes (summing).")
          combined$Bcell_norm <- rowSums(combined[, b_candidates_base], na.rm = TRUE)
      } else {
          msg("    No Bcell columns found. Creating placeholder NA column.")
          combined$Bcell_norm <- NA_real_
      }
    }
  }

  if (!"Bcell_norm" %in% names(combined)) {
    combined$Bcell_norm <- NA_real_
  }

  b_cols_to_drop <- setdiff(b_all_cols, "Bcell_norm")
  combined <- combined %>% dplyr::select(-any_of(b_cols_to_drop))
  combined <- combined %>% dplyr::select(-any_of(b_candidates_for_sum))
  # --- End of B-cell fix ---


  # --- Prefer *_norm versions when they exist ---
  cell_cols <- setdiff(names(combined), c("SampleID", "mets_progression"))
  cell_cols_base <- gsub("_norm$", "", cell_cols)
  keep_norm <- sapply(unique(cell_cols_base), function(base) {
    norm_col <- paste0(base, "_norm")
    if (norm_col %in% cell_cols) norm_col else base
  })
  cell_cols <- intersect(keep_norm, names(combined))

  # --- Facet order (3x5 grid) grouped by biology ---
  facet_order <- c(
    "Bas_norm", "Bcell_norm", "CD4T_norm", "CD8T_norm", "Treg_norm",    # Adaptive
    "DC_norm", "Endothelial", "Eos_norm", "Epithelial", "Fib",         # Stromal/interface
    "NK_norm", "Mono_norm", "Neu_norm", "Tumor"                        # Innate/tumor
  )
  facet_order <- intersect(facet_order, cell_cols)

  # --- Prepare long-form data (numeric only) ---
  numeric_cell_cols <- cell_cols[sapply(cell_cols, function(c) is.numeric(combined[[c]]))]

  # --- REVERT: No more x_pos column ---
  long_df <- combined %>%
    pivot_longer(cols = all_of(numeric_cell_cols), names_to = "Cell", values_to = "Prop") %>%
    filter(!is.na(mets_progression)) %>%
    mutate(Cell = factor(Cell, levels = facet_order)) %>%
    filter(!is.na(Cell))

  # --- Plot settings ---
  pal_progression <- c("No Mets" = "#00AFBB",
                       "LN-only Mets" = "#E7B800",
                       "Distant Mets" = "#FC4E07")
  comps <- combn(levels(long_df$mets_progression), 2, simplify = FALSE)

  # --- PLOT CODE BLOCK (WITH P-VALUES) ---
  msg("    Creating compact facet plot (Level 5, p-value-only labels) ...")
  
  # --- REVERT: Back to discrete x-axis ---
  p_compact <- ggplot(long_df, aes(x = mets_progression, y = Prop, 
                                   fill = mets_progression)) +
    geom_boxplot(
      outlier.shape = NA,
      width = 0.25,        # Thin boxes
      alpha = 0.85,
      color = "gray25"
    ) +
    geom_jitter(
      width = 0.05,        # Tight jitter
      size = 0.4,
      alpha = 0.6
    ) +
    facet_wrap(~ Cell, scales = "free_y", ncol = 5) +
    stat_compare_means(
      comparisons = comps,
      method = "wilcox.test",
      label = "p.format",
      hide.ns = TRUE,
      size = 2.1,
      step.increase = 0.05,
      tip.length = 0.006
    ) +
    stat_compare_means(
      method = "kruskal.test",
      label = "p.format",
      label.y.npc = 0.98,
      size = 2.4
    ) +
    scale_fill_manual(
      values = pal_progression,
      name = "Progression Status"
    ) +
    # --- REVERT: Remove scale_x_continuous ---
    theme_bw(base_size = 10.5) +
    theme(
      axis.text.x = element_text(
        angle = 90, vjust = 0.5, hjust = 1, size = 7
      ),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = 8.5),
      panel.grid = element_blank(),
      panel.border = element_blank(),
      panel.spacing = unit(0.15, "lines"),
      plot.margin = margin(5, 5, 5, 5)
    ) +
    labs(y = "Estimated Proportion")

  # --- Output (WITH P-VALUES) ---
  out_dir <- file.path(out_root, level_tag)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  out_path <- file.path(out_dir, "FACET_normCells_progression_h5_compact_BcellsCombined_fixed.pdf")

  # --- FIX: Reduce width to 7.5 to force groups closer ---
  ggsave(out_path, p_compact, width = 7.5, height = 6.5, device = cairo_pdf)
  msg("V  Compact normalized plot (B cells combined, fixed) saved -> ", out_path)
  
  
  # --- PLOT CODE BLOCK (NO P-VALUES) ---
  msg("    Creating additional plot (Level 5, NO p-values) ...")
  
  # --- REVERT: Back to discrete x-axis ---
  p_compact_no_p <- ggplot(long_df, aes(x = mets_progression, y = Prop, 
                                         fill = mets_progression)) +
    geom_boxplot(
      outlier.shape = NA,
      width = 0.25,      # Thin boxes
      alpha = 0.85,
      color = "gray25"
    ) +
    geom_jitter(
      width = 0.05,      # Tight jitter
      size = 0.4,
      alpha = 0.6
    ) +
    facet_wrap(~ Cell, scales = "free_y", ncol = 5) +
    scale_fill_manual(
      values = pal_progression,
      name = "Progression Status"
    ) +
    # --- REVERT: Remove scale_x_continuous ---
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    theme_bw(base_size = 10.5) +
    theme(
      axis.text.x = element_text(
        angle = 90, vjust = 0.5, hjust = 1, size = 7
      ),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = 8.5),
      panel.grid = element_blank(),
      panel.border = element_blank(),
      panel.spacing = unit(0.15, "lines"),
      plot.margin = margin(5, 5, 5, 5)
    ) +
    labs(y = "Estimated Proportion")

  # --- Output (NO P-VALUES) ---
  out_path_no_p <- file.path(out_dir, "FACET_normCells_progression_h5_compact_NO_PVAL.pdf")
  
  # --- FIX: Reduce width to 7.5 to force groups closer ---
  ggsave(out_path_no_p, p_compact_no_p, width = 7.5, height = 6.5, device = cairo_pdf)
  msg("V  Additional plot (no p-values) saved -> ", out_path_no_p)

}


###############################################################################
#  Individual cell type plots (HiTIMED level 3) - "Plot B" and "C"
#  FIX: This single block finds ALL cell types in h3 (granular or broad)
#       and plots them with the final "thin box" theme.
#       REMOVED all normalization and summing logic.
#  Output: BOX_[CellName]_h3.pdf for each cell type found.
###############################################################################

if ("HIT_DH" %in% ls() && "fib_DH" %in% ls() && "meta_DH" %in% ls()) {

  msg("\n==============================")
  msg(">>> Generating ALL individual cell plots for HiTIMED level 3")

  level_tag <- "h3" 
  frac_DH   <- HIT_DH[[level_tag]] # This has "Myeloid", "Lymphocyte", etc.

  # --- FIX: NO normalization, NO B-cell summing ---
  # Just combine the metadata with the raw h3 fractions and Fib
  combined <- meta_DH %>%
    left_join(frac_DH, by = "SampleID") %>%
    left_join(fib_DH, by = "SampleID")
    # --- normalise_fractions() REMOVED ---
    # --- B-cell logic REMOVED ---

  # --- FIX: Define cell_cols based *only* on deconvolution output ---
  # Get all HiTIMED h3 cols (minus SampleID)
  cell_cols <- setdiff(names(frac_DH), "SampleID") 
  
  # Add Fibroblast if it exists
  if ("Fib" %in% names(combined)) {
    cell_cols <- c(cell_cols, "Fib")
  }
  
  # Ensure they all actually exist in the 'combined' df and are numeric
  cell_cols <- intersect(cell_cols, names(combined))
  cell_cols <- cell_cols[sapply(cell_cols, function(c) {
    is.numeric(combined[[c]]) && !all(is.na(combined[[c]]))
  })]
  
  msg(paste("    Found", length(cell_cols), "cell types to plot:", paste(cell_cols, collapse=", ")))
  
  # --- Plot settings ---
  pal_progression <- c("No Mets" = "#00AFBB",
                       "LN-only Mets" = "#E7B800",
                       "Distant Mets" = "#FC4E07")
  comps <- combn(levels(combined$mets_progression), 2, simplify = FALSE)
  out_dir <- file.path(out_root, level_tag)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  # --- Loop and create individual plots ---
  msg(paste("    Looping and saving individual plots for", level_tag, "..."))

  for (cell_name in cell_cols) {
    
    df_cell <- combined %>%
      dplyr::select(SampleID, mets_progression, Prop = all_of(cell_name)) %>%
      filter(!is.na(mets_progression) & !is.na(Prop))

    if (nrow(df_cell) < 3) {
      msg(paste("    - Skipping", cell_name, "(no/insufficient data)"))
      next
    }

    # --- Use the "thin box" h5 theme ---
    p_indiv <- ggplot(df_cell, aes(x = mets_progression, y = Prop, 
                                   fill = mets_progression)) +
      geom_boxplot(
        outlier.shape = NA,
        width = 0.25, # Thin box (from h5 code)
        alpha = 0.85,
        color = "gray25"
      ) +
      geom_jitter(
        width = 0.05, # Tight jitter (from h5 code)
        size = 0.8,
        alpha = 0.6
      ) +
      stat_compare_means(
        comparisons = comps,
        method = "wilcox.test",
        label = "p.format",
        hide.ns = TRUE,
        size = 3.5, 
        step.increase = 0.05,
        tip.length = 0.01
      ) +
      stat_compare_means(
        method = "kruskal.test",
        label = "p.format",
        label.y.npc = 0.98,
        size = 4 
      ) +
      scale_fill_manual(
        values = pal_progression,
        guide = "none"
      ) +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
      theme_bw(base_size = 12) + # Base 12 for single plots
      theme(
        # --- 90-degree (vertical) text, size 7 (matches h5 code) ---
        axis.text.x = element_text(
          angle = 90, vjust = 0.5, hjust = 1, size = 7
        ),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        strip.background = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5)
      ) +
      labs(
        y = "Estimated Proportion",
        title = cell_name 
      )

    safe_cell_name <- gsub("[^A-Za-z0-9]+", "_", cell_name)
    out_path_indiv <- file.path(out_dir, paste0("BOX_", safe_cell_name, "_", level_tag, ".pdf"))
    
    # --- Save with a NARROW width (2.0") to force groups together ---
    ggsave(out_path_indiv, p_indiv, width = 2.0, height = 5, device = cairo_pdf)
  }

  msg("V  Individual", level_tag, "plots saved -> ", out_dir)
}

```
New 
Comparisons for all: This should replace everything

This will produce plots with the p-vales labeled and tables showing correctedd p-values
```{r}
###############################################################################
#  Fig1_DH_Final_ROBUST.R
#  Features:
#  1. ROBUST: Skips rare cells with insufficient data to prevent crashes.
#  2. SAFETY: Uses tryCatch() to ignore stat errors for specific cells.
#  3. PLOTS: Numeric P-values (p=0.05).
#  4. TABLES: Exports CSVs to 'hX' folders.
#  Author: Alos Diallo (Assisted by Gemini)
###############################################################################

## ── PATHS & SETUP ───────────────────────────────────────────────────────────
path_dh_epic_v1  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
path_dh_epic_v2  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
path_dh_meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"
out_root         <- "/Users/adiallo/Desktop/Thesis/Results/Figure1"

dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

suppressPackageStartupMessages({
  library(minfi); library(sesame); library(ENmix)
  library(HiTIMED); library(EpiDISH); library(tidyverse); 
  library(ggpubr); library(rstatix); library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  library(RColorBrewer)
})

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

## ── LEVEL DEFINITIONS ───────────────────────────────────────────────────────
get_level_defs <- function(h) {
  if (h == 1) {
    return(list(norm = c(), raw  = c("Tumor", "Other")))
  } else if (h == 2) {
    return(list(norm = c(), raw  = c("Tumor", "Immune", "Angiogenic")))
  } else if (h == 3) {
    return(list(norm = c(), raw  = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal")))
  } else if (h == 4) {
    return(list(
      norm = c("Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  } else if (h == 5) {
    return(list(
      norm = c("Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", "CD4T", "CD8T", "NK"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  } else if (h == 6) {
    return(list(
      norm = c("Bnv", "Bmem", "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", 
               "Mono", "DC", "NK", "Bas", "Eos", "Neu"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  }
}

## ── DATA LOADING ────────────────────────────────────────────────────────────
read_beta_DH <- function() {
  msg("Reading DH IDATs...")
  rg1 <- read.metharray.exp(path_dh_epic_v1, recursive = TRUE, force = TRUE, extended = TRUE)
  rg2 <- read.metharray.exp(path_dh_epic_v2, recursive = TRUE, force = TRUE, extended = TRUE)
  msg("NOOB Normalization...")
  beta1 <- getBeta(preprocessNoob(rg1, dyeMethod = "single"))
  beta2 <- getBeta(preprocessNoob(rg2, dyeMethod = "single")) %>% sesame::betasCollapseToPfx()
  common <- intersect(rownames(beta1), rownames(beta2))
  beta   <- cbind(beta1[common, ], beta2[common, ])
  msg("Running ENmix QC...")
  bad <- union(ENmix::QCinfo(rg1)$badCpG, ENmix::QCinfo(rg2)$badCpG)
  anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  sex_cpgs <- rownames(anno)[anno$chr %in% c("chrX","chrY")]
  beta <- beta[setdiff(rownames(beta), union(bad, sex_cpgs)), ]
  msg(paste("Final Beta Matrix:", nrow(beta), "CpGs x", ncol(beta), "Samples"))
  return(beta)
}

if (!exists("DH_beta")) DH_beta <- read_beta_DH()

## ── METADATA CLEANING ───────────────────────────────────────────────────────
msg("Processing Metadata...")
meta_raw <- read_csv(path_dh_meta_csv, show_col_types = FALSE)

meta_DH <- meta_raw %>%
  mutate(
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    
    site_clean = str_to_title(site),
    side = case_when(
      site_clean %in% c("Cecum", "Ascending Colon", "Hepatic Flexure", "Right Colon", "Transverse Colon") ~ "Right Side",
      site_clean %in% c("Splenic Flexure", "Descending Colon", "Left Colon", "Sigmoid", "Sigmoid Colon") ~ "Left Side",
      site_clean == "Rectum" ~ "Rectum",
      TRUE ~ NA_character_
    ),
    side = factor(side, levels = c("Right Side", "Left Side", "Rectum")),
    node_stage_clean = case_when(
      as.character(node_stage) == "0" ~ "N0",
      str_detect(as.character(node_stage), "^1") ~ "N1",
      str_detect(as.character(node_stage), "^2") ~ "N2",
      TRUE ~ NA_character_
    ),
    node_stage_clean = factor(node_stage_clean, levels = c("N0", "N1", "N2")),
    Grade = factor(as.character(Grade), levels = c("1", "2", "3")),

    mets_cat = case_when(
      Distant_Mets == TRUE ~ "Distant Mets",
      (!is.na(node_stage) & node_stage > 0) | ln_only == TRUE ~ "LN Mets",
      TRUE ~ "No Mets"
    ),
    mets_cat = factor(mets_cat, levels = c("No Mets", "LN Mets", "Distant Mets")),
    
    M_Stage = if_else(Distant_Mets == TRUE, "M1 (Distant)", "M0 (Locoregional)"),
    M_Stage = factor(M_Stage, levels = c("M0 (Locoregional)", "M1 (Distant)")),
    
    any_mets_bin = if_else(mets_cat == "No Mets", "No Mets", "Mets (LN+Dist)"),
    any_mets_bin = factor(any_mets_bin, levels = c("No Mets", "Mets (LN+Dist)")),
    
    MLH1_status = case_when(
      MLH1 == 1 ~ "MLH1 Intact",
      MLH1 == 0 ~ "MLH1 Deficient",
      TRUE ~ NA_character_
    ),
    MLH1_status = factor(MLH1_status, levels = c("MLH1 Intact", "MLH1 Deficient"))
  ) %>%
  filter(SampleID %in% colnames(DH_beta)) %>%
  dplyr::select(SampleID, mets_cat, any_mets_bin, M_Stage, MLH1_status, 
                site_clean, side, node_stage_clean, Grade)

## ── REFERENCE GENERATION ────────────────────────────────────────────────────
msg("Generating References...")
# HiTIMED returns Percentages (0-100)
h2_res <- HiTIMED_deconvolution(DH_beta, "COAD", 2, "tumor")
h2_df  <- as.data.frame(h2_res) %>% 
  rownames_to_column("SampleID") %>% 
  dplyr::rename(Total_Immune_Ref = Immune)

data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
common_cpg <- intersect(rownames(DH_beta), rownames(centEpiFibIC.m))
# EpiDISH returns Proportions (0-1), so we multiply by 100
fib_res <- epidish(DH_beta[common_cpg, ], centEpiFibIC.m[common_cpg, ], "RPC")$estF
fib_df  <- as.data.frame(fib_res * 100) %>%  
  rownames_to_column("SampleID") %>% dplyr::select(SampleID, Fib)

meta_refs <- meta_DH %>%
  inner_join(h2_df[, c("SampleID", "Total_Immune_Ref")], by = "SampleID") %>%
  inner_join(fib_df, by = "SampleID")

## ── ROBUST PLOTTING ENGINE ──────────────────────────────────────────────────

process_and_plot <- function(df, cell_cols, group_var, title_suffix, filename_suffix, out_path) {
  
  if(!group_var %in% names(df)) return()
  avail_cols <- intersect(cell_cols, names(df))
  if(length(avail_cols) == 0) return()
  
  long_df <- df %>%
    pivot_longer(cols = all_of(avail_cols), names_to = "Cell", values_to = "Prop") %>%
    filter(!is.na(.data[[group_var]]))
  
  # 1. STRICTER FILTERING
  # Only plot cells where AT LEAST 2 GROUPS have >= 3 samples.
  valid_stats <- long_df %>%
    filter(!is.na(Prop)) %>%
    group_by(Cell, .data[[group_var]]) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(Cell) %>%
    summarise(n_groups_pass = sum(n >= 3), total_n = sum(n), .groups = "drop")
  
  # FIX: Require 2+ viable groups to allow pairwise testing
  cells_to_plot <- valid_stats %>% filter(n_groups_pass >= 2) %>% pull(Cell)
  
  if(length(cells_to_plot) == 0) return()

  plot_df <- long_df %>% filter(Cell %in% cells_to_plot, !is.na(Prop))
  
  # 2. GENERATE STATISTICS TABLE (WITH SAFETY CATCH)
  # We wrap this in tryCatch because Wilcox test crashes on zero-variance/constant data
  final_stats_table <- tryCatch({
    
    stat_test <- plot_df %>%
      group_by(Cell) %>%
      wilcox_test(as.formula(paste("Prop ~", group_var))) %>%
      adjust_pvalue(method = "BH") %>%
      add_significance("p.adj")
    
    kruskal_test <- plot_df %>%
      group_by(Cell) %>%
      kruskal_test(as.formula(paste("Prop ~", group_var)))
    
    stat_test %>%
      dplyr::select(Cell, group1, group2, p, p.adj, p.adj.signif) %>%
      rename(Raw_P = p, FDR_P = p.adj, Significance = p.adj.signif) %>%
      left_join(kruskal_test %>% dplyr::select(Cell, p) %>% rename(Global_Kruskal_P = p), by = "Cell")
      
  }, error = function(e) {
    msg(paste("  [WARN] Stats calc failed for", filename_suffix, "(likely sparse data). Skipping table. Error:", e$message))
    return(NULL)
  })
  
  if(!is.null(final_stats_table)) {
    stats_file <- file.path(out_path, paste0("STATS_", filename_suffix, ".csv"))
    write_csv(final_stats_table, stats_file)
  }
  # 2b. GENERATE DESCRIPTIVE SUMMARY TABLE (Mean +/- SD)
  # ----------------------------------------------------
  summary_table <- plot_df %>%
    group_by(Cell, Group = .data[[group_var]]) %>%
    summarise(
      Mean   = mean(Prop, na.rm = TRUE),
      SD     = sd(Prop, na.rm = TRUE),
      Median = median(Prop, na.rm = TRUE),
      IQR    = IQR(Prop, na.rm = TRUE),
      Min    = min(Prop, na.rm = TRUE),
      Max    = max(Prop, na.rm = TRUE),
      N      = n(),
      .groups = "drop"
    ) %>%
    mutate(across(where(is.numeric), \(x) round(x, 3))) # Round to 3 decimals

  summary_file <- file.path(out_path, paste0("SUMMARY_TABLE_", filename_suffix, ".csv"))
  write_csv(summary_table, summary_file)
  # 3. PLOTTING
  pal <- switch(group_var,
    "mets_cat"         = c("No Mets"="#00AFBB", "LN Mets"="#E7B800", "Distant Mets"="#FC4E07"),
    "any_mets_bin"     = c("No Mets"="#00AFBB", "Mets (LN+Dist)"="#FC4E07"),
    "M_Stage"          = c("M0 (Locoregional)"="#377EB8", "M1 (Distant)"="#E41A1C"),
    "MLH1_status"      = c("MLH1 Intact"="#1F78B4", "MLH1 Deficient"="#E31A1C"),
    "side"             = c("Right Side"="#D95F02", "Left Side"="#7570B3", "Rectum"="#1B9E77"),
    "node_stage_clean" = c("N0"="#66A61E", "N1"="#E6AB02", "N2"="#A6761D"),
    "Grade"            = c("1"="#FBB4AE", "2"="#F0027F", "3"="#984EA3"),
    setNames(colorRampPalette(brewer.pal(8, "Dark2"))(length(unique(plot_df[[group_var]]))), 
             unique(plot_df[[group_var]]))
  )
  
  groups_present <- unique(as.character(plot_df[[group_var]]))
  if(length(groups_present) < 2) return()
  comps <- combn(groups_present, 2, simplify = FALSE)
  
  n_cells  <- length(unique(plot_df$Cell))
  n_cols   <- min(5, n_cells)
  n_rows   <- ceiling(n_cells / n_cols)
  p_height <- max(4, n_rows * 2.5)
  p_width  <- max(5, n_cols * 2.2)

  # Facet Plot
  p_facet <- ggplot(plot_df, aes(x = .data[[group_var]], y = Prop, fill = .data[[group_var]])) +
    geom_boxplot(outlier.shape = NA, width = 0.4, alpha = 0.8, color = "gray25") +
    geom_jitter(width = 0.1, size = 0.8, alpha = 0.5) +
    facet_wrap(~ Cell, scales = "free_y", ncol = n_cols) +
    stat_compare_means(comparisons = comps, method = "wilcox.test", 
                       label = "p.format", hide.ns = FALSE, size = 3) +
    stat_compare_means(method = "kruskal.test", label = "p.format", 
                       label.y.npc = 0.98, size = 3) +
    scale_fill_manual(values = pal, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) + 
    theme_bw(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
      axis.title.x = element_blank(),
      strip.background = element_rect(fill = "gray95"),
      strip.text = element_text(face = "bold"),
      panel.grid = element_blank()
    ) +
    labs(y = "Percentage", title = NULL)

  ggsave(file.path(out_path, paste0("FACET_", filename_suffix, ".pdf")), 
         p_facet, width = p_width, height = p_height, device = cairo_pdf)

  # Individual Plots
  indiv_path <- file.path(out_path, "Individual")
  dir.create(indiv_path, showWarnings = FALSE)
  
  for(cell in unique(plot_df$Cell)) {
    cell_data <- plot_df %>% filter(Cell == cell)
    p_indiv <- ggplot(cell_data, aes(x = .data[[group_var]], y = Prop, fill = .data[[group_var]])) +
      geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.8, color = "gray25") +
      geom_jitter(width = 0.1, size = 1.0, alpha = 0.6) +
      stat_compare_means(comparisons = comps, method = "wilcox.test", 
                         label = "p.format", hide.ns = FALSE, size = 4) +
      stat_compare_means(method = "kruskal.test", label = "p.format", 
                         label.y.npc = 0.95, size = 4) +
      scale_fill_manual(values = pal, guide = "none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) + 
      theme_bw(base_size = 14) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")
      ) +
      labs(y = "Percentage", title = cell)
    
    safe_name <- gsub("[^A-Za-z0-9]+", "_", cell)
    ggsave(file.path(indiv_path, paste0("BOX_", safe_name, "_", filename_suffix, ".pdf")),
           p_indiv, width = 3.5, height = 4.5, device = cairo_pdf)
  }
}

## ── MAIN LOOP ───────────────────────────────────────────────────────────────
msg("======= Starting Final ROBUST Analysis =======")

for (h in 1:6) {
  level_name <- paste0("h", h)
  msg("\n>>> Processing Level:", level_name)
  
  defs <- get_level_defs(h)
  res  <- HiTIMED_deconvolution(DH_beta, "COAD", h, "tumor")
  
  if (is.vector(res)) {
    frac_df <- data.frame(SampleID = names(res), res = res)
    names(frac_df)[2] <- paste0("Cell_h", h)
  } else {
    frac_df <- as.data.frame(res) %>% rownames_to_column("SampleID")
  }
  
  combined <- meta_refs %>% inner_join(frac_df, by = "SampleID")
  present  <- setdiff(names(frac_df), "SampleID")
  
  cols_norm <- intersect(present, defs$norm)
  cols_raw  <- intersect(present, defs$raw)
  if (h >= 3) cols_raw <- unique(c(cols_raw, "Fib"))

  # Normalization & SCALING
  if (length(cols_norm) > 0) {
    combined_norm <- combined %>%
      mutate(across(all_of(cols_norm), 
                    ~ if_else(Total_Immune_Ref > 0 & !is.na(Total_Immune_Ref), 
                              (.x / Total_Immune_Ref) * 100, NA_real_),
                    .names = "{.col}_norm"))
    immune_cols <- paste0(cols_norm, "_norm")
  } else {
    combined_norm <- combined
    immune_cols   <- NULL
  }
  
  lvl_dir <- file.path(out_root, level_name)
  dir.create(lvl_dir, showWarnings = FALSE)
  
  run_all_vars <- function(df, cols, prefix) {
    if(length(cols) == 0) return()
    process_and_plot(df, cols, "mets_cat", "Mets", paste0(prefix, "_Mets"), lvl_dir)
    process_and_plot(df, cols, "MLH1_status", "MLH1", paste0(prefix, "_MLH1"), lvl_dir)
    process_and_plot(filter(df, MLH1_status == "MLH1 Intact"), cols, "mets_cat", "MSS", paste0(prefix, "_Mets_IntactOnly"), lvl_dir)
    process_and_plot(filter(df, MLH1_status == "MLH1 Deficient"), cols, "mets_cat", "MSI", paste0(prefix, "_Mets_DeficientOnly"), lvl_dir)
    process_and_plot(df, cols, "side", "Side", paste0(prefix, "_Side_LR"), lvl_dir)
    process_and_plot(df, cols, "site_clean", "Site", paste0(prefix, "_Site_Detailed"), lvl_dir)
    process_and_plot(df, cols, "node_stage_clean", "NodeStage", paste0(prefix, "_NodeStage"), lvl_dir)
    process_and_plot(df, cols, "Grade", "Grade", paste0(prefix, "_Grade"), lvl_dir)
    process_and_plot(df, cols, "M_Stage", "MStage", paste0(prefix, "_M0_vs_M1"), lvl_dir)
    process_and_plot(df, cols, "any_mets_bin", "AnyMets", paste0(prefix, "_AnyMets_Binary"), lvl_dir)
  }

  if(!is.null(immune_cols)) {
    msg("  -> Generating Immune Plots (Normalized %)...")
    run_all_vars(combined_norm, immune_cols, "IMMUNE")
  }
  if(length(cols_raw) > 0) {
    msg("  -> Generating Structural Plots (Raw %)...")
    run_all_vars(combined_norm, cols_raw, "STRUCTURAL")
  }
}

msg("\n✅ Analysis Complete. Check 'hX' folders for PDFs (Raw P) and CSVs (Raw+FDR P).")
```

```{r}
# 2b. GENERATE DESCRIPTIVE SUMMARY TABLE (Mean +/- SD)
  # ----------------------------------------------------
  summary_table <- plot_df %>%
    group_by(Cell, Group = .data[[group_var]]) %>%
    summarise(
      Mean   = mean(Prop, na.rm = TRUE),
      SD     = sd(Prop, na.rm = TRUE),
      Median = median(Prop, na.rm = TRUE),
      IQR    = IQR(Prop, na.rm = TRUE),
      Min    = min(Prop, na.rm = TRUE),
      Max    = max(Prop, na.rm = TRUE),
      N      = n(),
      .groups = "drop"
    ) %>%
    mutate(across(where(is.numeric), \(x) round(x, 3))) # Round to 3 decimals

  summary_file <- file.path(out_path, paste0("SUMMARY_TABLE_", filename_suffix, ".csv"))
  write_csv(summary_table, summary_file)
```


Expanded set, will remove the above if this works:
```{r}
###############################################################################
#  Fig1_DH_MASTER_ALL_COMPARISONS.R
#  The "Kitchen Sink" Script: 14 distinct comparison sets per cell/level.
#  Includes: 
#    - Standard Mets/MLH1
#    - M0 vs M1
#    - MLH1 Stratification (Mets within Intact/Deficient)
#    - Grade/Node/Side Stratification (Within Mets/No Mets)
#  Author: Alos Diallo (Assisted by Gemini)
###############################################################################

## ── PATHS & SETUP ───────────────────────────────────────────────────────────
path_dh_epic_v1  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
path_dh_epic_v2  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
path_dh_meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"
out_root         <- "/Users/adiallo/Desktop/Thesis/Results/Figure1"

dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

suppressPackageStartupMessages({
  library(minfi); library(sesame); library(ENmix)
  library(HiTIMED); library(EpiDISH); library(tidyverse); 
  library(ggpubr); library(rstatix); library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  library(RColorBrewer)
})

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

## ── LEVEL DEFINITIONS ───────────────────────────────────────────────────────
get_level_defs <- function(h) {
  if (h == 1) {
    return(list(norm = c(), raw  = c("Tumor", "Other")))
  } else if (h == 2) {
    return(list(norm = c(), raw  = c("Tumor", "Immune", "Angiogenic")))
  } else if (h == 3) {
    return(list(norm = c(), raw  = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal")))
  } else if (h == 4) {
    return(list(
      norm = c("Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  } else if (h == 5) {
    return(list(
      norm = c("Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", "CD4T", "CD8T", "NK"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  } else if (h == 6) {
    return(list(
      norm = c("Bnv", "Bmem", "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", 
               "Mono", "DC", "NK", "Bas", "Eos", "Neu"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  }
}

## ── DATA LOADING ────────────────────────────────────────────────────────────
read_beta_DH <- function() {
  msg("Reading DH IDATs...")
  rg1 <- read.metharray.exp(path_dh_epic_v1, recursive = TRUE, force = TRUE, extended = TRUE)
  rg2 <- read.metharray.exp(path_dh_epic_v2, recursive = TRUE, force = TRUE, extended = TRUE)
  msg("NOOB Normalization...")
  beta1 <- getBeta(preprocessNoob(rg1, dyeMethod = "single"))
  beta2 <- getBeta(preprocessNoob(rg2, dyeMethod = "single")) %>% sesame::betasCollapseToPfx()
  common <- intersect(rownames(beta1), rownames(beta2))
  beta   <- cbind(beta1[common, ], beta2[common, ])
  msg("Running ENmix QC...")
  bad <- union(ENmix::QCinfo(rg1)$badCpG, ENmix::QCinfo(rg2)$badCpG)
  anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  sex_cpgs <- rownames(anno)[anno$chr %in% c("chrX","chrY")]
  beta <- beta[setdiff(rownames(beta), union(bad, sex_cpgs)), ]
  msg(paste("Final Beta Matrix:", nrow(beta), "CpGs x", ncol(beta), "Samples"))
  return(beta)
}

if (!exists("DH_beta")) DH_beta <- read_beta_DH()

## ── METADATA CLEANING ───────────────────────────────────────────────────────
msg("Processing Metadata...")
meta_raw <- read_csv(path_dh_meta_csv, show_col_types = FALSE)

meta_DH <- meta_raw %>%
  mutate(
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    
    # Clean Site/Side
    site_clean = str_to_title(site),
    side = case_when(
      site_clean %in% c("Cecum", "Ascending Colon", "Hepatic Flexure", "Right Colon", "Transverse Colon") ~ "Right Side",
      site_clean %in% c("Splenic Flexure", "Descending Colon", "Left Colon", "Sigmoid", "Sigmoid Colon") ~ "Left Side",
      site_clean == "Rectum" ~ "Rectum",
      TRUE ~ NA_character_
    ),
    side = factor(side, levels = c("Right Side", "Left Side", "Rectum")),
    
    # Clean Node
    node_stage_clean = case_when(
      as.character(node_stage) == "0" ~ "N0",
      str_detect(as.character(node_stage), "^1") ~ "N1",
      str_detect(as.character(node_stage), "^2") ~ "N2",
      TRUE ~ NA_character_
    ),
    node_stage_clean = factor(node_stage_clean, levels = c("N0", "N1", "N2")),
    
    # Clean Grade
    Grade = factor(as.character(Grade), levels = c("1", "2", "3")),

    # Mets Variables
    mets_cat = case_when(
      Distant_Mets == TRUE ~ "Distant Mets",
      (!is.na(node_stage) & node_stage > 0) | ln_only == TRUE ~ "LN Mets",
      TRUE ~ "No Mets"
    ),
    mets_cat = factor(mets_cat, levels = c("No Mets", "LN Mets", "Distant Mets")),
    
    M_Stage = if_else(Distant_Mets == TRUE, "M1 (Distant)", "M0 (Locoregional)"),
    M_Stage = factor(M_Stage, levels = c("M0 (Locoregional)", "M1 (Distant)")),
    
    any_mets_bin = if_else(mets_cat == "No Mets", "No Mets", "Mets (LN+Dist)"),
    any_mets_bin = factor(any_mets_bin, levels = c("No Mets", "Mets (LN+Dist)")),
    
    # MLH1
    MLH1_status = case_when(
      MLH1 == 1 ~ "MLH1 Intact",
      MLH1 == 0 ~ "MLH1 Deficient",
      TRUE ~ NA_character_
    ),
    MLH1_status = factor(MLH1_status, levels = c("MLH1 Intact", "MLH1 Deficient"))
  ) %>%
  filter(SampleID %in% colnames(DH_beta)) %>%
  dplyr::select(SampleID, mets_cat, any_mets_bin, M_Stage, MLH1_status, 
                site_clean, side, node_stage_clean, Grade)

## ── REFERENCE GENERATION ────────────────────────────────────────────────────
msg("Generating References...")
h2_res <- HiTIMED_deconvolution(DH_beta, "COAD", 2, "tumor")
h2_df  <- as.data.frame(h2_res) %>% 
  rownames_to_column("SampleID") %>% 
  dplyr::rename(Total_Immune_Ref = Immune)

data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
common_cpg <- intersect(rownames(DH_beta), rownames(centEpiFibIC.m))
fib_res <- epidish(DH_beta[common_cpg, ], centEpiFibIC.m[common_cpg, ], "RPC")$estF
fib_df  <- as.data.frame(fib_res) %>% 
  rownames_to_column("SampleID") %>% dplyr::select(SampleID, Fib)

meta_refs <- meta_DH %>%
  inner_join(h2_df[, c("SampleID", "Total_Immune_Ref")], by = "SampleID") %>%
  inner_join(fib_df, by = "SampleID")

## ── PLOTTING ENGINE ─────────────────────────────────────────────────────────

plot_panel_safe <- function(df, cell_cols, group_var, title_suffix, filename_suffix, out_path) {
  
  if(!group_var %in% names(df)) return()
  avail_cols <- intersect(cell_cols, names(df))
  if(length(avail_cols) == 0) return()
  
  long_df <- df %>%
    pivot_longer(cols = all_of(avail_cols), names_to = "Cell", values_to = "Prop") %>%
    filter(!is.na(.data[[group_var]]))
  
  # Min N=3 Check
  valid_stats <- long_df %>%
    filter(!is.na(Prop)) %>%
    group_by(Cell, .data[[group_var]]) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(Cell) %>%
    summarise(n_groups_pass = sum(n >= 3), total_n = sum(n), .groups = "drop")
  
  cells_to_plot <- valid_stats %>% filter(total_n > 0) %>% pull(Cell)
  if(length(cells_to_plot) == 0) return()

  plot_df <- long_df %>% filter(Cell %in% cells_to_plot, !is.na(Prop))
  
  # Colors
  pal <- switch(group_var,
    "mets_cat"         = c("No Mets"="#00AFBB", "LN Mets"="#E7B800", "Distant Mets"="#FC4E07"),
    "any_mets_bin"     = c("No Mets"="#00AFBB", "Mets (LN+Dist)"="#FC4E07"),
    "M_Stage"          = c("M0 (Locoregional)"="#377EB8", "M1 (Distant)"="#E41A1C"),
    "MLH1_status"      = c("MLH1 Intact"="#1F78B4", "MLH1 Deficient"="#E31A1C"),
    "side"             = c("Right Side"="#D95F02", "Left Side"="#7570B3", "Rectum"="#1B9E77"),
    "node_stage_clean" = c("N0"="#66A61E", "N1"="#E6AB02", "N2"="#A6761D"),
    "Grade"            = c("1"="#FBB4AE", "2"="#F0027F", "3"="#984EA3"),
    setNames(colorRampPalette(brewer.pal(8, "Dark2"))(length(unique(plot_df[[group_var]]))), 
             unique(plot_df[[group_var]]))
  )
  
  groups_present <- unique(as.character(plot_df[[group_var]]))
  if(length(groups_present) < 2) return()
  comps <- combn(groups_present, 2, simplify = FALSE)
  
  # Facet Dimensions
  n_cells  <- length(unique(plot_df$Cell))
  n_cols   <- min(5, n_cells)
  n_rows   <- ceiling(n_cells / n_cols)
  p_height <- max(4, n_rows * 2.5)
  p_width  <- max(5, n_cols * 2.2)

  # 1. FACET PLOT
  p_facet <- ggplot(plot_df, aes(x = .data[[group_var]], y = Prop, fill = .data[[group_var]])) +
    geom_boxplot(outlier.shape = NA, width = 0.4, alpha = 0.8, color = "gray25") +
    geom_jitter(width = 0.1, size = 0.8, alpha = 0.5) +
    facet_wrap(~ Cell, scales = "free_y", ncol = n_cols) +
    stat_compare_means(comparisons = comps, method = "wilcox.test", 
                       label = "p.signif", hide.ns = TRUE, size = 3) +
    stat_compare_means(method = "kruskal.test", label.y.npc = 0.98, size = 3) +
    scale_fill_manual(values = pal, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) + 
    theme_bw(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
      axis.title.x = element_blank(),
      strip.background = element_rect(fill = "gray95"),
      strip.text = element_text(face = "bold"),
      panel.grid = element_blank()
    ) +
    labs(y = "Proportion", title = NULL)

  ggsave(file.path(out_path, paste0("FACET_", filename_suffix, ".pdf")), 
         p_facet, width = p_width, height = p_height, device = cairo_pdf)

  # 2. INDIVIDUAL PLOTS
  indiv_path <- file.path(out_path, "Individual")
  dir.create(indiv_path, showWarnings = FALSE)
  
  for(cell in unique(plot_df$Cell)) {
    cell_data <- plot_df %>% filter(Cell == cell)
    p_indiv <- ggplot(cell_data, aes(x = .data[[group_var]], y = Prop, fill = .data[[group_var]])) +
      geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.8, color = "gray25") +
      geom_jitter(width = 0.1, size = 1.0, alpha = 0.6) +
      stat_compare_means(comparisons = comps, method = "wilcox.test", 
                         label = "p.signif", hide.ns = TRUE, size = 4) +
      stat_compare_means(method = "kruskal.test", label.y.npc = 0.95, size = 4) +
      scale_fill_manual(values = pal, guide = "none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) + 
      theme_bw(base_size = 14) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")
      ) +
      labs(y = "Proportion", title = cell)
    
    safe_name <- gsub("[^A-Za-z0-9]+", "_", cell)
    ggsave(file.path(indiv_path, paste0("BOX_", safe_name, "_", filename_suffix, ".pdf")),
           p_indiv, width = 3.5, height = 4.5, device = cairo_pdf)
  }
}

## ── MAIN LOOP ───────────────────────────────────────────────────────────────
msg("======= Starting Master Analysis (Levels 1-6) =======")

for (h in 1:6) {
  level_name <- paste0("h", h)
  msg("\n>>> Processing Level:", level_name)
  
  defs <- get_level_defs(h)
  res  <- HiTIMED_deconvolution(DH_beta, "COAD", h, "tumor")
  
  if (is.vector(res)) {
    frac_df <- data.frame(SampleID = names(res), res = res)
    names(frac_df)[2] <- paste0("Cell_h", h)
  } else {
    frac_df <- as.data.frame(res) %>% rownames_to_column("SampleID")
  }
  
  combined <- meta_refs %>% inner_join(frac_df, by = "SampleID")
  present  <- setdiff(names(frac_df), "SampleID")
  
  cols_norm <- intersect(present, defs$norm)
  cols_raw  <- intersect(present, defs$raw)
  if (h >= 3) cols_raw <- unique(c(cols_raw, "Fib"))

  # Normalization
  if (length(cols_norm) > 0) {
    combined_norm <- combined %>%
      mutate(across(all_of(cols_norm), 
                    ~ if_else(Total_Immune_Ref > 0 & !is.na(Total_Immune_Ref), 
                              .x / Total_Immune_Ref, NA_real_),
                    .names = "{.col}_norm"))
    immune_cols <- paste0(cols_norm, "_norm")
  } else {
    combined_norm <- combined
    immune_cols   <- NULL
  }
  
  lvl_dir <- file.path(out_root, level_name)
  dir.create(lvl_dir, showWarnings = FALSE)
  
  # ── MASTER RUNNER ──
  run_master_vars <- function(df, cols, prefix) {
    if(length(cols) == 0) return()
    
    # A. GLOBAL METS / STAGE
    plot_panel_safe(df, cols, "mets_cat", "Mets_3Level", paste0(prefix, "_Mets_3Level"), lvl_dir)
    plot_panel_safe(df, cols, "any_mets_bin", "AnyMets", paste0(prefix, "_AnyMets_Binary"), lvl_dir)
    plot_panel_safe(df, cols, "M_Stage", "MStage", paste0(prefix, "_M0_vs_M1"), lvl_dir)
    plot_panel_safe(df, cols, "MLH1_status", "MLH1", paste0(prefix, "_MLH1"), lvl_dir)
    
    # B. MLH1 STRATIFIED
    plot_panel_safe(filter(df, MLH1_status == "MLH1 Intact"), cols, "mets_cat", "MSS", paste0(prefix, "_Mets_IntactOnly"), lvl_dir)
    plot_panel_safe(filter(df, MLH1_status == "MLH1 Deficient"), cols, "mets_cat", "MSI", paste0(prefix, "_Mets_DeficientOnly"), lvl_dir)
    
    # C. GRADE
    plot_panel_safe(df, cols, "Grade", "Grade_All", paste0(prefix, "_Grade_All"), lvl_dir)
    plot_panel_safe(filter(df, mets_cat=="No Mets"), cols, "Grade", "Grade_NoMets", paste0(prefix, "_Grade_in_NoMets"), lvl_dir)
    plot_panel_safe(filter(df, any_mets_bin!="No Mets"), cols, "Grade", "Grade_Mets", paste0(prefix, "_Grade_in_Mets"), lvl_dir)
    
    # D. NODE STAGE
    plot_panel_safe(filter(df, any_mets_bin!="No Mets"), cols, "node_stage_clean", "Node_Mets", paste0(prefix, "_Node_in_Mets"), lvl_dir)
    plot_panel_safe(filter(df, mets_cat=="Distant Mets"), cols, "node_stage_clean", "Node_DistMets", paste0(prefix, "_Node_in_DistantMets"), lvl_dir)
    
    # E. SIDE
    plot_panel_safe(df, cols, "side", "Side_All", paste0(prefix, "_Side_All"), lvl_dir)
    plot_panel_safe(filter(df, mets_cat=="No Mets"), cols, "side", "Side_NoMets", paste0(prefix, "_Side_in_NoMets"), lvl_dir)
    plot_panel_safe(filter(df, any_mets_bin!="No Mets"), cols, "side", "Side_Mets", paste0(prefix, "_Side_in_Mets"), lvl_dir)
  }

  if(!is.null(immune_cols)) {
    msg("  -> Generating Immune Plots (Normalized)...")
    run_master_vars(combined_norm, immune_cols, "IMMUNE")
  }
  if(length(cols_raw) > 0) {
    msg("  -> Generating Structural Plots (Raw)...")
    run_master_vars(combined_norm, cols_raw, "STRUCTURAL")
  }
}

msg("\n✅ Analysis Complete. All 14 comparison sets generated per level.")
```


Comparisons for site that were the most significant
```{r}
###############################################################################
#  Fig1_DH_AllCells_AllPossibleSitePairs_FIXED.R
#  "Brute Force" Discovery Mode
#  FIXED: Prevents 'Tumor.x/Tumor.y' collision by cleaning reference dataframes.
#  Author: Alos Diallo (Assisted by Gemini)
###############################################################################

## ── PATHS & SETUP ───────────────────────────────────────────────────────────
path_dh_epic_v1  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
path_dh_epic_v2  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
path_dh_meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"
out_root         <- "/Users/adiallo/Desktop/Thesis/Results/Figure1/Site_Specific_Comparisons"

dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

suppressPackageStartupMessages({
  library(minfi); library(sesame); library(ENmix)
  library(HiTIMED); library(EpiDISH); library(tidyverse); 
  library(ggpubr); library(rstatix); library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
})

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

## ── 1. LEVEL DEFINITIONS (Strict Mapping) ───────────────────────────────────
get_level_defs <- function(h) {
  if (h == 1) {
    return(list(norm = c(), raw  = c("Tumor", "Other")))
  } else if (h == 2) {
    return(list(norm = c(), raw  = c("Tumor", "Immune", "Angiogenic")))
  } else if (h == 3) {
    return(list(norm = c(), raw  = c("Tumor", "Lymphocyte", "Myeloid", "Endothelial", "Epithelial", "Stromal")))
  } else if (h == 4) {
    return(list(
      norm = c("Granulocyte", "Mononuclear", "Tcell", "Bcell", "NK"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  } else if (h == 5) {
    return(list(
      norm = c("Bas", "Eos", "Neu", "Mono", "DC", "Bnv", "Bmem", "CD4T", "CD8T", "NK"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  } else if (h == 6) {
    return(list(
      norm = c("Bnv", "Bmem", "CD4nv", "CD4mem", "Treg", "CD8nv", "CD8mem", 
               "Mono", "DC", "NK", "Bas", "Eos", "Neu"),
      raw  = c("Tumor", "Endothelial", "Epithelial", "Stromal")
    ))
  }
}

## ── 2. DATA LOADING & METADATA ──────────────────────────────────────────────
read_beta_DH <- function() {
  msg("Reading DH IDATs...")
  rg1 <- read.metharray.exp(path_dh_epic_v1, recursive = TRUE, force = TRUE, extended = TRUE)
  rg2 <- read.metharray.exp(path_dh_epic_v2, recursive = TRUE, force = TRUE, extended = TRUE)
  msg("NOOB Normalization...")
  beta1 <- getBeta(preprocessNoob(rg1, dyeMethod = "single"))
  beta2 <- getBeta(preprocessNoob(rg2, dyeMethod = "single")) %>% sesame::betasCollapseToPfx()
  common <- intersect(rownames(beta1), rownames(beta2))
  beta   <- cbind(beta1[common, ], beta2[common, ])
  msg("Running ENmix QC...")
  bad <- union(ENmix::QCinfo(rg1)$badCpG, ENmix::QCinfo(rg2)$badCpG)
  anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  sex_cpgs <- rownames(anno)[anno$chr %in% c("chrX","chrY")]
  beta <- beta[setdiff(rownames(beta), union(bad, sex_cpgs)), ]
  msg(paste("Final Beta Matrix:", nrow(beta), "CpGs x", ncol(beta), "Samples"))
  return(beta)
}

if (!exists("DH_beta")) DH_beta <- read_beta_DH()

msg("Processing Metadata...")
meta_raw <- read_csv(path_dh_meta_csv, show_col_types = FALSE)

meta_DH <- meta_raw %>%
  mutate(
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    site_clean = str_to_title(site),
    site_clean = case_when(
        site_clean == "Sigmoid Colon" ~ "Sigmoid",
        TRUE ~ site_clean
    )
  ) %>%
  filter(SampleID %in% colnames(DH_beta)) %>%
  dplyr::select(SampleID, site_clean)

# ── AUTO-DETECT PAIRS ──
valid_sites <- meta_DH %>% 
  count(site_clean) %>% 
  filter(n >= 2) %>% 
  pull(site_clean)

msg(paste("Found", length(valid_sites), "sites with N>=2:", paste(valid_sites, collapse=", ")))

all_pairs <- t(combn(valid_sites, 2))
colnames(all_pairs) <- c("Site_A", "Site_B")
site_pairs <- as_tibble(all_pairs)

msg(paste("Generated", nrow(site_pairs), "unique site pairwise comparisons."))


## ── 3. REFERENCE GENERATION ─────────────────────────────────────────────────
msg("Generating References...")
h2_res <- HiTIMED_deconvolution(DH_beta, "COAD", 2, "tumor")

# FIX: Select ONLY the reference column we need to avoid collisions with "Tumor" later
h2_df  <- as.data.frame(h2_res) %>% 
          rownames_to_column("SampleID") %>% 
          dplyr::rename(Total_Immune_Ref = Immune) %>%
          dplyr::select(SampleID, Total_Immune_Ref) # <--- STRICT SELECTION

data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
common_cpg <- intersect(rownames(DH_beta), rownames(centEpiFibIC.m))
fib_res <- epidish(DH_beta[common_cpg, ], centEpiFibIC.m[common_cpg, ], "RPC")$estF
fib_df  <- as.data.frame(fib_res) %>% rownames_to_column("SampleID") %>% 
          dplyr::select(SampleID, Fib)

## ── 4. MAIN LOOP (LEVELS 1-6) ───────────────────────────────────────────────
msg("======= Starting Comprehensive Pairwise Analysis =======")

for (h in 1:6) {
  level_name <- paste0("h", h)
  msg("\n>>> Processing Level:", level_name)
  
  # A. Deconvolution
  res <- HiTIMED_deconvolution(DH_beta, "COAD", h, "tumor")
  if (is.vector(res)) {
    frac_df <- data.frame(SampleID = names(res), res = res)
    names(frac_df)[2] <- paste0("Cell_h", h)
  } else {
    frac_df <- as.data.frame(res) %>% rownames_to_column("SampleID")
  }

  # B. Merge
  # Join structure ensures clean names
  combined <- meta_DH %>%
    inner_join(h2_df, by="SampleID") %>%
    inner_join(fib_df, by="SampleID") %>%
    inner_join(frac_df, by="SampleID")
  
  # C. Identify Cells & Definitions
  present_cols <- setdiff(names(frac_df), "SampleID")
  defs         <- get_level_defs(h)
  cols_norm    <- intersect(present_cols, defs$norm)
  cols_raw     <- intersect(present_cols, defs$raw)
  if (h >= 3) cols_raw <- unique(c(cols_raw, "Fib")) # Always add Fib from h3+
  
  all_target_cells <- c(cols_norm, cols_raw)
  
  # Create Output Directory
  lvl_dir <- file.path(out_root, level_name)
  dir.create(lvl_dir, showWarnings = FALSE)
  
  msg(paste("   Plotting", length(all_target_cells), "cell types x", nrow(site_pairs), "pairs..."))

  # D. Loop through ALL Cells
  for(cell in all_target_cells) {
    
    is_immune <- cell %in% cols_norm
    
    # E. Loop through ALL Site Pairs
    for(i in 1:nrow(site_pairs)) {
      site_a <- site_pairs$Site_A[i]
      site_b <- site_pairs$Site_B[i]
      
      # Prepare Data (Safe if() for normalization)
      plot_data <- combined %>%
        filter(site_clean %in% c(site_a, site_b)) %>%
        dplyr::rename(Raw_Prop = all_of(cell)) %>%
        mutate(site_clean = factor(site_clean, levels = c(site_a, site_b)))
      
      if(is_immune) {
        plot_data$Final_Prop <- ifelse(plot_data$Total_Immune_Ref > 0, 
                                       plot_data$Raw_Prop / plot_data$Total_Immune_Ref, 
                                       NA)
      } else {
        plot_data$Final_Prop <- plot_data$Raw_Prop
      }
      
      plot_data <- plot_data %>% filter(!is.na(Final_Prop))
      
      # N Check (Min 2)
      n_a <- sum(plot_data$site_clean == site_a)
      n_b <- sum(plot_data$site_clean == site_b)
      
      if(n_a < 2 || n_b < 2) next # Skip insufficient data
      
      # Plot
      file_suffix <- if(is_immune) "_norm" else ""
      plot_title  <- paste0(cell, file_suffix)
      y_label     <- if(is_immune) "Prop (Norm)" else "Prop (Raw)"
      
      p <- ggplot(plot_data, aes(x = site_clean, y = Final_Prop, fill = site_clean)) +
        geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.8, color = "gray30") +
        geom_jitter(width = 0.1, size = 1.5, alpha = 0.6) +
        stat_compare_means(method = "wilcox.test", 
                           label = "p.format", size = 5, 
                           comparisons = list(c(site_a, site_b))) + 
        scale_fill_manual(values = c("#377EB8", "#E41A1C")) +
        theme_bw(base_size = 14) +
        theme(
          axis.title.x = element_blank(),
          legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"),
          panel.grid.major.x = element_blank()
        ) +
        labs(y = y_label, title = plot_title)
      
      safe_a <- gsub(" ", "", site_a)
      safe_b <- gsub(" ", "", site_b)
      fname  <- paste0("BOX_", cell, file_suffix, "_", safe_a, "_vs_", safe_b, ".pdf")
      
      ggsave(file.path(lvl_dir, fname), p, width = 3.5, height = 5, device = cairo_pdf)
    }
  }
}

msg("\n✅ Discovery Complete. Every valid site comparison has been plotted.")
```




