---
title: "MMR_Analysis_Colon_DM_Data"
author: "Alos Diallo"
date: "2024-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# =========================================================================
# Examination of DNA Missmatch Repair genes - COAD DH Dataset
# Code by: Alos Diallo
# 2023,2024
# =========================================================================


Loading libraries:
```{r message=FALSE, warning=FALSE, include=FALSE}
library(minfi)
library(sesame)
library(pheatmap)
library(minfiData)
library(FlowSorted.Blood.EPIC)
library(HiTIMED)
library(ggplot2)
library(dplyr)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICv2manifest)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICmanifest)
library(limma)
library(qvalue)
library(sva)
library(ENmix)
library(DataExplorer)
library(ggplot2)
library(ggrepel)
library(matrixStats)
library(EpiDISH)
library(tibble)
library(tidyr)
```

Loading the annotation file:
```{r}
data("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
```

Loading the data and metadata information
```{r}
targets <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/dm_57_samples.csv")
targets$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets) <- targets$patient
targets$SampleID<- targets$patient

targets_expanded <- read.csv("/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv")
targets_expanded$patient <- paste(targets$Sentrix_ID,targets$Sentrix_Position,sep="_")
rownames(targets_expanded) <- targets_expanded$patient
targets_expanded$SampleID<- targets_expanded$patient
```

```{r}
idat = "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/"
RGset_32 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/",recursive = TRUE) 
#RGset_25 = openSesame(idat , func = getBetas) 
RGset_25 = read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/",recursive = TRUE) 
```


Looking for bad CpG's using QCinfo
```{r}
sva32<-ctrlsva(RGset_32)
sva25<-ctrlsva(RGset_25)
SVA_combined <- rbind(sva25[,1:8], sva32)
SVA_combined_df <- as.data.frame(SVA_combined)

surrogateVariables <- sva
# Convert RGChannelSet to RGChannelSetExtended
RGset_32_extended <- read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/",recursive = TRUE,extended = TRUE)
RGset_25_extended <- read.metharray.exp("/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/",recursive = TRUE,extended = TRUE) 

# Run QCinfo function
qc_info32 <- QCinfo(RGset_32_extended)
qc_info25 <- QCinfo(RGset_25_extended)

# Display QC information
#print(qc_info32)
#print(qc_info25)
badCpG = c(qc_info32$badCpG,qc_info25$badCpG)
# Remove any suffix after an underscore
cleaned_badCpG <- sub("_.*", "", badCpG)
```


Normalizing 
```{r}
Noob_25_m = preprocessNoob(RGset_25)
Noob_32_m = preprocessNoob(RGset_32)

# Run functional normalization
#funnorm_25 <- preprocessFunnorm(RGset_25, bgCorr = TRUE,dyeCorr = TRUE)
#funnorm_32 <- preprocessFunnorm(RGset_32, bgCorr = TRUE,dyeCorr = TRUE)
```

Extracting the beta values from the normalized samples.  I am now running functional normalization as it contains a Noob option
```{r}
Betas_25<-getBeta(Noob_25_m)
Betas_32<-getBeta(Noob_32_m)

#Betas_25<-getBeta(funnorm_25)
#Betas_32<-getBeta(funnorm_32)
```

Colapsing the betas so that the data can be merged
```{r}
Betas_25<- sesame::betasCollapseToPfx(Betas_25)
#Betas_32<- sesame::betasCollapseToPfx(Betas_32)
colnames(Betas_25) = colnames(Noob_25_m)
```

Filter out sex linked probes
```{r}
# Extract probes located on the Y chromosome
cpg_y_probes <- rownames(anno[anno$chr == "chrY", ])  # Filter probes on Y chromosome

# Filter Y chromosome probes from Betas_25
cpg_y_probes_in_betas_25 <- cpg_y_probes[cpg_y_probes %in% rownames(Betas_25)]
Betas_25_filtered <- Betas_25[!rownames(Betas_25) %in% cpg_y_probes_in_betas_25, ]

# Filter Y chromosome probes from Betas_32
cpg_y_probes_in_betas_32 <- cpg_y_probes[cpg_y_probes %in% rownames(Betas_32)]
Betas_32_filtered <- Betas_32[!rownames(Betas_32) %in% cpg_y_probes_in_betas_32, ]

# Extract probes located on the X chromosome
cpg_x_probes <- rownames(anno[anno$chr == "chrX", ])

# Filter X chromosome probes from Betas_25
cpg_x_probes_in_betas_25 <- cpg_x_probes[cpg_x_probes %in% rownames(Betas_25_filtered)]
Betas_25_filtered <- Betas_25_filtered[!rownames(Betas_25_filtered) %in% cpg_x_probes_in_betas_25, ]

# Filter X chromosome probes from Betas_32
cpg_x_probes_in_betas_32 <- cpg_x_probes[cpg_x_probes %in% rownames(Betas_32_filtered)]
Betas_32_filtered <- Betas_32_filtered[!rownames(Betas_32_filtered) %in% cpg_x_probes_in_betas_32, ]
```


Merging the samples 
```{r}
# Ensure the same probes exist in both datasets
common_probes <- intersect(rownames(Betas_25_filtered), rownames(Betas_32_filtered))

# Subset both datasets to include only common probes
Betas_25_filtered <- Betas_25_filtered[common_probes, ]
Betas_32_filtered <- Betas_32_filtered[common_probes, ]

# Combine the beta values from both datasets
DH_CRC_Betas <- cbind(Betas_25_filtered, Betas_32_filtered) 

```


Running HiTIMED to generate cell type proportions
```{r}
HiTIMED_result<-HiTIMED_deconvolution(DH_CRC_Betas,"COAD",6,"tumor")
```

Running EpiDISH to obtain fibroblast proportions
```{r}
# Define the beta matrix and reference matrix
beta_matrix <- DH_CRC_Betas  # Your dataset
ref_matrix <- centEpiFibIC.m  # Reference matrix for EpiDISH (e.g., centEpiFibIC.m)

# Run EpiDISH
epidish_results <- epidish(beta.m = beta_matrix, ref.m = ref_matrix, method = "RPC")

# Extract the estimated cell type proportions
cell_type_proportions <- epidish_results$estF

# Convert the matrix to a data frame and scale proportions to percentages
cell_type_proportions_df <- as.data.frame(cell_type_proportions) * 100

# Add SampleID as a column from the row names
cell_type_proportions_df$SampleID <- rownames(cell_type_proportions)

# Inspect the results
head(cell_type_proportions_df)
```


Generate M values
```{r}
DH_CRC_M_value = BetaValueToMValue(DH_CRC_Betas)
```

Filter out bad CpG's
```{r}
# Identify good CpGs
all_cpgs <- rownames(DH_CRC_M_value)
good_cpgs <- setdiff(all_cpgs, cleaned_badCpG)
# Filter the M values to keep only good CpGs
filtered_DH_CRC_M_value <- DH_CRC_M_value[good_cpgs, ]
DH_CRC_M_value = filtered_DH_CRC_M_value
```





I am looking at the relationship between MLH1 status and metastasis to determine if it should be included in my linear models.  Here I am making a table and then running a fisher's exact test. 
```{r}
# Assuming your targets data is already loaded in R

# Create a contingency table for MLH1 and any_mets
contingency_table <- table(targets$MLH1, targets$any_mets)

# Add labels for clarity
dimnames(contingency_table) <- list(
  MLH1 = c("Wildtype", "Mutated"),
  Metastasis = c("No Mets", "Mets")
)

# View the contingency table
print(contingency_table)

# Perform Fisher's Exact Test
fisher_test <- fisher.test(contingency_table)

# View the results of the test
print(fisher_test)


# Step 1: Create a new column 'MMR'
targets_expanded$MMR <- ifelse(rowSums(targets_expanded[, c("MLH1", "MSH2", "PMS2", "MSH6")] == 1) > 0, 1, 0)

targets_expanded$MMR = as.factor(targets_expanded$MMR)
targets_expanded$any_mets = as.factor(targets_expanded$any_mets)
# Step 2: Verify the new column
table(targets_expanded$MMR)

# Step 1: Generate a contingency table for MMR and any_mets
contingency_table <- table(targets_expanded$MMR, targets_expanded$any_mets)
contingency_table

# Create a summary table for mutations in individual MMR genes
gene_summary <- colSums(targets_expanded[, c("MLH1", "MSH2", "PMS2", "MSH6")] == 1)

# Print the table
print(gene_summary)

# Fisher's test for a specific gene (e.g., MLH1 vs. any_mets)
contingency_table <- table(targets_expanded$MLH1, targets_expanded$any_mets)
fisher_test <- fisher.test(contingency_table)
print(fisher_test)
```

No noticable difference was found.  

Here I am running the same analysis on all of the categorical data:
	•	Higher values indicate smaller (more significant) p-values.
	•	Lower values indicate larger (less significant) p-values.
	
	•	\(-\log_{10}(\text{p-value}) = 1: Corresponds to a p-value of \(0.1\).
	•	\(-\log_{10}(\text{p-value}) = 2: Corresponds to a p-value of \(0.01\).
	•	\(-\log_{10}(\text{p-value}) = 3: Corresponds to a p-value of \(0.001\).	

Generating a Venn Diagram showing overplap of MMR gene mutations with Metastasis status
```{r}
# Load necessary library
library(VennDiagram)

# Step 1: Create logical vectors for each MMR gene mutation
mlh1_mutated <- targets_expanded$MLH1 == 1
msh2_mutated <- targets_expanded$MSH2 == 1
pms2_mutated <- targets_expanded$PMS2 == 1
msh6_mutated <- targets_expanded$MSH6 == 1

# Step 2: Generate a Venn diagram to visualize the overlap
venn_plot <- venn.diagram(
  x = list(
    MLH1 = which(mlh1_mutated),
    MSH2 = which(msh2_mutated),
    PMS2 = which(pms2_mutated),
    MSH6 = which(msh6_mutated)
  ),
  filename = NULL,  # Set to NULL to return the plot as an object
  category.names = c("MLH1", "MSH2", "PMS2", "MSH6"),
  fill = c("red", "blue", "green", "yellow"),
  alpha = 0.5,
  cat.cex = 0.8,
  cex = 0.9,
  margin = 0.1
)

grid.draw(venn_plot)
# Step 3: Save the plot or display it
pdf("MMR_gene_mutation_venn.pdf")
grid.draw(venn_plot)
dev.off()
```

Generating a heatmap that shows which mutations are positive for a given sample. 
```{r}
# Install and load pheatmap if not already installed
if (!requireNamespace("pheatmap", quietly = TRUE)) {
  install.packages("pheatmap")
}
library(pheatmap)

# Step 1: Create a mutation matrix for MMR genes
# Extract relevant columns for MMR genes (MLH1, MSH2, PMS2, MSH6)
mmr_genes <- targets_expanded[, c("MLH1", "MSH2", "PMS2", "MSH6")]

# Transpose the matrix so rows are genes and columns are samples
mutation_matrix <- t(as.matrix(mmr_genes))

# Step 2: Set colors for the heatmap
# Blue for mutation (1), white for no mutation (0)
annotation_colors <- list(
  Mutation = c("0" = "white", "1" = "blue")
)

# Step 3: Generate the heatmap
pheatmap(
  mutation_matrix,
  color = c("white", "blue"), # White for 0, blue for 1
  cluster_rows = FALSE,       # Do not cluster rows (genes)
  cluster_cols = FALSE,       # Do not cluster columns (samples)
  main = "MMR Gene Mutation Heatmap",
  fontsize = 10,
  legend_labels = c("No Mutation", "Mutation")
)
```

```{r}
# Step 1: Convert appropriate columns to factors
categorical_vars <- c("MLH1", "MSH2", "diagnosis", "node_stage", "sex", "site", "Distant_Mets", "any_mets", "ln_only")
targets[categorical_vars] <- lapply(targets[categorical_vars], factor)

# Debug: Check if all columns are converted
str(targets[categorical_vars])

# Step 2: Initialize the p-value matrix
pval_matrix <- matrix(NA, nrow = length(categorical_vars), ncol = length(categorical_vars),
                      dimnames = list(categorical_vars, categorical_vars))

# Step 3: Compute Fisher's Exact Test for all pairs
for (i in seq_along(categorical_vars)) {
  for (j in seq_along(categorical_vars)) {
    if (i != j) {
      # Create contingency table
      contingency_table <- table(targets[[categorical_vars[i]]], targets[[categorical_vars[j]]])
      
      # Check dimensions of the contingency table
      if (nrow(contingency_table) > 1 && ncol(contingency_table) > 1) {
        # Perform Fisher's Exact Test
        fisher_test <- fisher.test(contingency_table,simulate.p.value = TRUE, B = 1e6)
        
        # Store the p-value
        pval_matrix[i, j] <- fisher_test$p.value
      }
    }
  }
}

# Debug: Check the p-value matrix
print(pval_matrix)

# Replace NA values to avoid errors in log transformation
pval_matrix[is.na(pval_matrix)] <- 1

# Step 4: Visualize as a heatmap
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
library(pheatmap)

# Plot the heatmap
pheatmap(-log10(pval_matrix), 
         display_numbers = TRUE, 
         number_format = "%.2f",
         main = "Heatmap of Fisher's Exact Test (-log10 p-values)",
         cluster_rows = FALSE, 
         cluster_cols = FALSE)


```