---
title: "Production_Validation"
author: "Alos Diallo"
date: "2025-08-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Main EWAS
```{r}
# ewas_metastasis_h6_normImmune.R     · 2025-07-18
# Single HiTIMED h = 6 → immune-fraction normalisation + EpiScore
# Phenotypes: any_mets · ln_only · distant

## ── USER SETTINGS ───────────────────────────────────────────────────────────
idat_v1  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
idat_v2  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

phenotypes          <- c(any  = "any_mets",
                         ln   = "ln_only",
                         dist = "distant")
variance_percentile <- 0.90  # 0 ⇒ keep all CpGs

## NEW: centralize outputs/caches here
results_root <- "/Users/adiallo/Desktop/Thesis/Results"

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "]", ..., "\n")

suppressPackageStartupMessages({
  library(minfi)
  library(sesame)
  library(ENmix)
  library(matrixStats)
  library(limma)
  library(sva)
  library(HiTIMED)
  library(EpiDISH)
  library(tidyverse)
  library(glue)
})

read_csv_silent <- function(file, ...) {
  if (utils::packageVersion("readr") >= "2.0.0") {
    readr::read_csv(file, show_col_types = FALSE, ...)
  } else {
    suppressMessages(readr::read_csv(file, ...))
  }
}

## ── 1) β-matrix (cached) ────────────────────────────────────────────────────
## NEW: cache under Results
cache_dir <- file.path(results_root, "ewas_cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
beta_rds  <- file.path(cache_dir, "beta_comb_qc.rds")
## NEW: keep sample-origin manifests
manifest_csv <- file.path(cache_dir, "sample_manifest_v1_v2.csv")

read_pp <- function(path, tag) {
  rg  <- read.metharray.exp(path, recursive = TRUE, extended = TRUE, force = TRUE)
  bet <- getBeta(preprocessNoob(rg, dyeMethod = "single"))
  if (tag == "v2") bet <- betasCollapseToPfx(bet)
  bad <- ENmix::QCinfo(rg)$badCpG
  bet <- bet[ setdiff(rownames(bet), bad), , drop = FALSE]
  bet
}

## NEW: we’ll save sample IDs per platform so we can label Array_Version later
if (!file.exists(beta_rds)) {
  msg("Reading / normalising IDATs …")
  beta1 <- read_pp(idat_v1, "v1")
  beta2 <- read_pp(idat_v2, "v2")

  sample_ids_v1 <- colnames(beta1)
  sample_ids_v2 <- colnames(beta2)

  common    <- intersect(rownames(beta1), rownames(beta2))
  beta_comb <- cbind(beta1[common, , drop = FALSE],
                     beta2[common, , drop = FALSE])

  saveRDS(list(beta = beta_comb,
               sample_ids_v1 = sample_ids_v1,
               sample_ids_v2 = sample_ids_v2),
          beta_rds)

  ## write a small manifest for sanity/debug
  tibble(
    SampleID = c(sample_ids_v1, sample_ids_v2),
    Array_Version = c(rep("v1", length(sample_ids_v1)),
                      rep("v2", length(sample_ids_v2)))
  ) %>%
    distinct() %>%
    write_csv(manifest_csv)

  msg("β-matrix cached.")
} else {
  cache_obj <- readRDS(beta_rds)
  if (is.list(cache_obj) && !is.null(cache_obj$beta)) {
    beta_comb     <- cache_obj$beta
    sample_ids_v1 <- cache_obj$sample_ids_v1
    sample_ids_v2 <- cache_obj$sample_ids_v2
  } else {
    ## backward compat if the RDS used to be only the matrix
    beta_comb     <- cache_obj
    sample_ids_v1 <- character(0)
    sample_ids_v2 <- character(0)
  }
  msg("β-matrix loaded from cache.")
}

##--- Remove sex linked probes--##
get_annotation_for <- function(beta_mat) {
  # Prefer EPIC; fallback to 450k
  if (requireNamespace("IlluminaHumanMethylationEPICanno.ilm10b4.hg19", quietly = TRUE)) {
    suppressPackageStartupMessages(
      library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
    )
    a <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  } else if (requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    suppressPackageStartupMessages(
      library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
    )
    a <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  } else {
    stop("Install an Illumina methylation annotation package (EPIC or 450k).")
  }
  # Keep rows that exist in your beta matrix
  a[rownames(a) %in% rownames(beta_mat), , drop = FALSE]
}

anno <- get_annotation_for(beta_comb)

# Try multiple possible chromosome fields across annotation versions
chr_vec <- NULL
if ("chr" %in% colnames(anno))       chr_vec <- anno$chr
if (is.null(chr_vec) && "CHR" %in% colnames(anno)) chr_vec <- anno$CHR
if (is.null(chr_vec) && "seqnames" %in% colnames(anno)) chr_vec <- as.character(anno$seqnames)

if (is.null(chr_vec)) stop("Could not find chromosome column in annotation.")

is_sex_chr <- chr_vec %in% c("chrX","chrY","X","Y")
sex_probes <- rownames(anno)[is_sex_chr]

n_before <- nrow(beta_comb)
if (length(sex_probes) > 0) {
  beta_comb <- beta_comb[setdiff(rownames(beta_comb), sex_probes), , drop = FALSE]
}
msg(length(sex_probes), " chrX/chrY probes removed; ",
    n_before, " → ", nrow(beta_comb), " CpGs remain.")

## ── Variance filter ─────────────────────────────────────────────────────────
M.mat <- log2(beta_comb / (1 - beta_comb) + 1e-6)
if (variance_percentile > 0) {
  var_thr   <- quantile(rowVars(M.mat, TRUE), variance_percentile)
  keep      <- rowVars(M.mat, TRUE) > var_thr
  beta_comb <- beta_comb[keep, , drop = FALSE]
  M.mat     <- M.mat[keep, , drop = FALSE]
  msg(sum(keep), " CpGs kept (top ",
      100 * (1 - variance_percentile), "% var).")
}

## ── 2a) Define & run EpiScore for fibroblasts ────────────────────────────────
data(centEpiFibIC.m)  # built-in colon-fibroblast reference

run_episcore <- function(beta) {
  common <- intersect(rownames(beta), rownames(centEpiFibIC.m))
  est    <- epidish(beta[common, ],
                    centEpiFibIC.m[common, ],
                    method = "RPC")$estF
  as.data.frame(est) * 100    # columns = Epi, Fib, IC
}

msg("Running EpiScore (RPC) for fibroblast …")
fib_scores <- run_episcore(beta_comb) %>%
  rownames_to_column("SampleID")
# fib_scores has SampleID, Epi, Fib, IC

## ── 2b) HiTIMED deconvolution (h = 6) ────────────────────────────────────────
msg("Running HiTIMED  h = 6 …")
frac_h6 <- HiTIMED_deconvolution(beta_comb, "COAD", h = 6) %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(fib_scores, by = "SampleID")
# now frac_h6 includes a "Fib" column (from EpiScore)

## ── 3) Collapse / normalise fractions ───────────────────────────────────────
collapse_frac <- function(df) {
  get_ <- function(x) if (x %in% names(df)) df[[x]] else 0
  
  imm_cols <- c("CD8mem","CD8nv","CD4mem","CD4nv",
                "Treg","Bmem","Bnv","DC","Mono","NK")
  totalImm <- rowSums(df[imm_cols], na.rm = TRUE)
  totalImm[totalImm == 0] <- NA
  
  tibble(
    SampleID     = df$SampleID,
    Tumor        = get_("Tumor"),
    Fibroblast   = get_("Fib"),       # now uses EpiScore Fib
    StromalOther = get_("Endothelial") + get_("Epithelial"),
    CD8_T        = (get_("CD8mem") + get_("CD8nv")) / totalImm,
    LymphOther   = (get_("CD4mem") + get_("CD4nv") +
                    get_("Treg")   + get_("Bmem") + get_("Bnv")) / totalImm
  ) %>% mutate(across(where(is.numeric), as.numeric))
}

## ── 4) Tier templates ───────────────────────────────────────────────────────
tier_template <- list(
  Tier1 = c("age","sex_num","MLH1_num"),
  Tier2 = c("age","sex_num","MLH1_num","Tumor"),
  Tier3 = c("age","sex_num","MLH1_num","Tumor",
            "Fibroblast","StromalOther",
            "CD8_T","LymphOther")
)
trim_tiers <- function(tiers, avail) lapply(tiers, intersect, avail)

## ── 5) Loop over phenotypes ────────────────────────────────────────────────
for (pheno_tag in names(phenotypes)) {
  pheno_var <- phenotypes[[pheno_tag]]
  msg("\n===== ", pheno_var, " =====")
  
  meta <- read_csv_silent(meta_csv) %>%
    mutate(
      SampleID      = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
      any_mets      = as.integer(any_mets),
      distant       = as.integer(Distant_Mets),
      ln_only       = as.integer(ln_only),
      sex_num       = as.integer(sex == "M"),
      MLH1_num      = as.integer(MLH1 == 1)
    )

  ## NEW: assign Array_Version using the vectors we saved at read time
  meta <- meta %>%
    mutate(
      Array_Version = case_when(
        SampleID %in% sample_ids_v1 ~ "v1",
        SampleID %in% sample_ids_v2 ~ "v2",
        TRUE                        ~ NA_character_
      )
    )

  ## NEW: warn if any SampleID in meta not found in either list
  if (any(is.na(meta$Array_Version))) {
    miss <- meta$SampleID[is.na(meta$Array_Version)]
    msg("WARNING: ", length(miss), " meta SampleIDs not matched to v1/v2: ",
        paste(head(miss, 8), collapse = ", "),
        if (length(miss) > 8) " …" else "")
  }
  
  meta$pheno_bin <- dplyr::case_when(
    pheno_var == "any_mets" ~ meta$any_mets,
    pheno_var == "distant"  ~ meta$distant,
    pheno_var == "ln_only"  ~ (meta$ln_only & !meta$distant),
    TRUE                    ~ NA_integer_
  )
  stopifnot(all(meta$pheno_bin %in% 0:1, na.rm = TRUE))
  
  frac_df  <- collapse_frac(frac_h6)
  meta_all <- left_join(meta, frac_df, by = "SampleID")
  ok       <- complete.cases(meta_all[, tier_template$Tier3])
  meta_f   <- meta_all[ok, ]
  M_f      <- M.mat[, meta_f$SampleID, drop = FALSE]
  
  ## ComBat if both platforms are present after filtering
  if (length(unique(na.omit(meta_f$Array_Version))) > 1) {
    mod_cb <- model.matrix(~ age + sex_num + MLH1_num + pheno_bin, meta_f)
    M_cb   <- ComBat(M_f,
                     batch       = factor(meta_f$Array_Version),
                     mod         = mod_cb,
                     par.prior   = TRUE,
                     prior.plots = FALSE)
  } else {
    M_cb <- M_f
  }
  
  tier_list <- trim_tiers(tier_template, names(meta_f))
  fits <- purrr::map(tier_list, function(covs) {
    nz   <- covs[sapply(meta_f[covs], sd, TRUE) > 0]
    form <- if (length(nz) == 0) {
      ~ pheno_bin
    } else {
      as.formula(paste("~ pheno_bin +", paste(nz, collapse = " + ")))
    }
    des <- model.matrix(form, meta_f)
    eBayes(lmFit(M_cb, des), trend = TRUE, robust = TRUE)
  })
  
  ## NEW: write results under Results root
  out_dir <- file.path(results_root, glue("results_ewas_{pheno_tag}"), "H6")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(fits, file.path(out_dir, "ewas_fits.rds"))
  msg("✓  fits saved → ", out_dir)
}

msg("\n✅ Finished: Any, LN-only, Distant  (HiTIMED-h6, immune-normalised + EpiScore)")
```

Without Combat
```{r}
# ewas_metastasis_h6_normImmune.R     · 2025-07-18
# Single HiTIMED h = 6 → immune-fraction normalisation + EpiScore
# Phenotypes: any_mets · ln_only · distant
# NOTE: ComBat is DISABLED — same-cohort v1/v2; using intersected probes instead.

## ── USER SETTINGS ───────────────────────────────────────────────────────────
idat_v1  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/idats/"
idat_v2  <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/no_match/25_samples/"
meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

phenotypes          <- c(any  = "any_mets",
                         ln   = "ln_only",
                         dist = "distant")
variance_percentile <- 0.90  # 0 ⇒ keep all CpGs

## NEW: centralize outputs/caches here
results_root <- "/Users/adiallo/Desktop/Thesis/Results"

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "]", ..., "\n")

suppressPackageStartupMessages({
  library(minfi)
  library(sesame)
  library(ENmix)
  library(matrixStats)
  library(limma)
  # library(sva)   # <- not needed since ComBat is off
  library(HiTIMED)
  library(EpiDISH)
  library(tidyverse)
  library(glue)
})

read_csv_silent <- function(file, ...) {
  if (utils::packageVersion("readr") >= "2.0.0") {
    readr::read_csv(file, show_col_types = FALSE, ...)
  } else {
    suppressMessages(readr::read_csv(file, ...))
  }
}

## ── 1) β-matrix (cached) ────────────────────────────────────────────────────
## cache under Results
cache_dir <- file.path(results_root, "ewas_cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
beta_rds      <- file.path(cache_dir, "beta_comb_qc.rds")
manifest_csv  <- file.path(cache_dir, "sample_manifest_v1_v2.csv")

read_pp <- function(path, tag) {
  rg  <- read.metharray.exp(path, recursive = TRUE, extended = TRUE, force = TRUE)
  bet <- getBeta(preprocessNoob(rg, dyeMethod = "single"))
  if (tag == "v2") bet <- betasCollapseToPfx(bet)
  bad <- ENmix::QCinfo(rg)$badCpG
  bet <- bet[ setdiff(rownames(bet), bad), , drop = FALSE]
  bet
}

## Save sample IDs per platform so we can label Array_Version later
if (!file.exists(beta_rds)) {
  msg("Reading / normalising IDATs …")
  beta1 <- read_pp(idat_v1, "v1")
  beta2 <- read_pp(idat_v2, "v2")

  sample_ids_v1 <- colnames(beta1)
  sample_ids_v2 <- colnames(beta2)

  common <- intersect(rownames(beta1), rownames(beta2))
  stopifnot(length(common) > 0)
  beta_comb <- cbind(beta1[common, , drop = FALSE],
                     beta2[common, , drop = FALSE])

  saveRDS(list(beta = beta_comb,
               sample_ids_v1 = sample_ids_v1,
               sample_ids_v2 = sample_ids_v2),
          beta_rds)

  tibble(
    SampleID = c(sample_ids_v1, sample_ids_v2),
    Array_Version = c(rep("v1", length(sample_ids_v1)),
                      rep("v2", length(sample_ids_v2)))
  ) %>%
    distinct() %>%
    write_csv(manifest_csv)

  msg("β-matrix cached.")
} else {
  cache_obj <- readRDS(beta_rds)
  if (is.list(cache_obj) && !is.null(cache_obj$beta)) {
    beta_comb     <- cache_obj$beta
    sample_ids_v1 <- cache_obj$sample_ids_v1
    sample_ids_v2 <- cache_obj$sample_ids_v2
  } else {
    ## backward compat if the RDS used to be only the matrix
    beta_comb     <- cache_obj
    sample_ids_v1 <- character(0)
    sample_ids_v2 <- character(0)
  }
  msg("β-matrix loaded from cache.")
}

## --- Remove sex-linked probes ----------------------------------------------
get_annotation_for <- function(beta_mat) {
  if (requireNamespace("IlluminaHumanMethylationEPICanno.ilm10b4.hg19", quietly = TRUE)) {
    suppressPackageStartupMessages(library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19))
    a <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  } else if (requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    suppressPackageStartupMessages(library(IlluminaHumanMethylation450kanno.ilmn12.hg19))
    a <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  } else {
    stop("Install an Illumina methylation annotation package (EPIC or 450k).")
  }
  a[rownames(a) %in% rownames(beta_mat), , drop = FALSE]
}

anno <- get_annotation_for(beta_comb)

# Try multiple possible chromosome fields across annotation versions
chr_vec <- NULL
if ("chr" %in% colnames(anno))             chr_vec <- anno$chr
if (is.null(chr_vec) && "CHR" %in% colnames(anno))      chr_vec <- anno$CHR
if (is.null(chr_vec) && "seqnames" %in% colnames(anno)) chr_vec <- as.character(anno$seqnames)
if (is.null(chr_vec)) stop("Could not find chromosome column in annotation.")

is_sex_chr <- chr_vec %in% c("chrX","chrY","X","Y")
sex_probes <- rownames(anno)[is_sex_chr]

n_before <- nrow(beta_comb)
if (length(sex_probes) > 0) {
  beta_comb <- beta_comb[setdiff(rownames(beta_comb), sex_probes), , drop = FALSE]
}
msg(length(sex_probes), " chrX/chrY probes removed; ",
    n_before, " → ", nrow(beta_comb), " CpGs remain.")

## ── Variance filter (optional) ──────────────────────────────────────────────
M.mat <- log2(beta_comb / pmax(1 - beta_comb, 1e-6) + 1e-6)
if (variance_percentile > 0) {
  v <- rowVars(M.mat, na.rm = TRUE)
  var_thr <- quantile(v, variance_percentile, na.rm = TRUE)
  keep    <- v > var_thr
  beta_comb <- beta_comb[keep, , drop = FALSE]
  M.mat     <- M.mat[keep, , drop = FALSE]
  msg(sum(keep), " CpGs kept (top ", 100 * (1 - variance_percentile), "% var).")
}

## ── 2a) EpiScore for fibroblasts (RPC) ──────────────────────────────────────
data(centEpiFibIC.m)  # built-in colon-fibroblast reference

run_episcore <- function(beta) {
  common <- intersect(rownames(beta), rownames(centEpiFibIC.m))
  est    <- epidish(beta[common, ],
                    centEpiFibIC.m[common, ],
                    method = "RPC")$estF
  as.data.frame(est) * 100    # columns = Epi, Fib, IC
}

msg("Running EpiScore (RPC) for fibroblast …")
fib_scores <- run_episcore(beta_comb) %>%
  rownames_to_column("SampleID")

## ── 2b) HiTIMED deconvolution (h = 6) ───────────────────────────────────────
msg("Running HiTIMED  h = 6 …")
frac_h6 <- HiTIMED_deconvolution(beta_comb, "COAD", h = 6) %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(fib_scores, by = "SampleID")   # adds Fib from EpiScore

## ── 3) Collapse / normalise fractions ───────────────────────────────────────
collapse_frac <- function(df) {
  get_ <- function(x) if (x %in% names(df)) df[[x]] else 0
  imm_cols <- c("CD8mem","CD8nv","CD4mem","CD4nv","Treg","Bmem","Bnv","DC","Mono","NK")
  totalImm <- rowSums(df[imm_cols], na.rm = TRUE); totalImm[totalImm == 0] <- NA
  tibble(
    SampleID     = df$SampleID,
    Tumor        = get_("Tumor"),
    Fibroblast   = get_("Fib"),               # from EpiScore
    StromalOther = get_("Endothelial") + get_("Epithelial"),
    CD8_T        = (get_("CD8mem") + get_("CD8nv")) / totalImm,
    LymphOther   = (get_("CD4mem") + get_("CD4nv") +
                    get_("Treg")   + get_("Bmem") + get_("Bnv")) / totalImm
  ) %>% mutate(across(where(is.numeric), as.numeric))
}

## ── 4) Tier templates (unchanged design) ────────────────────────────────────
tier_template <- list(
  Tier1 = c("age","sex_num","MLH1_num"),
  Tier2 = c("age","sex_num","MLH1_num","Tumor"),
  Tier3 = c("age","sex_num","MLH1_num","Tumor",
            "Fibroblast","StromalOther",
            "CD8_T","LymphOther")
)
trim_tiers <- function(tiers, avail) lapply(tiers, intersect, avail)

## ── 5) Loop over phenotypes (NO ComBat) ─────────────────────────────────────
for (pheno_tag in names(phenotypes)) {
  pheno_var <- phenotypes[[pheno_tag]]
  msg("\n===== ", pheno_var, " =====")

  meta <- read_csv_silent(meta_csv) %>%
    mutate(
      SampleID      = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
      any_mets      = as.integer(any_mets),
      distant       = as.integer(Distant_Mets),
      ln_only       = as.integer(ln_only),
      sex_num       = as.integer(sex == "M"),
      MLH1_num      = as.integer(MLH1 == 1)
    )

  ## Use saved vectors to tag array version (for reference only)
  meta <- meta %>%
    mutate(
      Array_Version = case_when(
        SampleID %in% sample_ids_v1 ~ "v1",
        SampleID %in% sample_ids_v2 ~ "v2",
        TRUE                        ~ NA_character_
      )
    )

  if (any(is.na(meta$Array_Version))) {
    miss <- meta$SampleID[is.na(meta$Array_Version)]
    msg("WARNING: ", length(miss), " meta SampleIDs not matched to v1/v2: ",
        paste(head(miss, 8), collapse = ", "),
        if (length(miss) > 8) " …" else "")
  }

  meta$pheno_bin <- dplyr::case_when(
    pheno_var == "any_mets" ~ meta$any_mets,
    pheno_var == "distant"  ~ meta$distant,
    pheno_var == "ln_only"  ~ (meta$ln_only & !meta$distant),
    TRUE                    ~ NA_integer_
  )
  stopifnot(all(meta$pheno_bin %in% 0:1, na.rm = TRUE))

  frac_df  <- collapse_frac(frac_h6)
  meta_all <- left_join(meta, frac_df, by = "SampleID")
  ok       <- complete.cases(meta_all[, tier_template$Tier3])
  meta_f   <- meta_all[ok, , drop = FALSE]
  M_f      <- M.mat[, meta_f$SampleID, drop = FALSE]

  ## NO ComBat: same-cohort v1/v2 + intersected probes is sufficient
  M_cb <- M_f

  tier_list <- trim_tiers(tier_template, names(meta_f))
  fits <- purrr::map(tier_list, function(covs) {
    # Only numeric covariates with non-zero variance; use stats::sd explicitly
    is_var <- vapply(
      meta_f[covs],
      function(x) is.numeric(x) && is.finite(stats::sd(x, na.rm = TRUE)) && stats::sd(x, na.rm = TRUE) > 0,
      logical(1)
    )
    nz <- covs[is_var]

    form <- if (length(nz) == 0) {
      ~ pheno_bin
    } else {
      as.formula(paste("~ pheno_bin +", paste(nz, collapse = " + ")))
    }
    des <- model.matrix(form, meta_f)
    eBayes(lmFit(M_cb, des), trend = TRUE, robust = TRUE)
  })

  ## write results under Results root
  out_dir <- file.path(results_root, glue("results_ewas_{pheno_tag}"), "H6")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(fits, file.path(out_dir, "ewas_fits.rds"))
  msg("✓  fits saved → ", out_dir)
}

msg("\n✅ Finished: Any, LN-only, Distant  (HiTIMED-h6, immune-normalised + EpiScore; NO ComBat)")
```


Viewing our CpG's using TCGA methylation data: TGCA normal and tumor
```{r}
###############################################################################
# TCGA_Validation_Pipeline_V7_Array_Aware_pdf_no_titles_with_axis_legend.R
# - Vector PDFs for Illustrator
# - NO legends/titles on the main heatmap
# - Separate column-annotation legend PDF
# - Separate colorbar legend PDF
# - Optional row-labels PDF (matching clustered order)
# - Also writes row/column order CSVs for Illustrator alignment
###############################################################################

## -- 1) USER SETTINGS --
path_tcga_full_idats <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/idats/"
meta_data_path       <- "/Users/adiallo/Desktop/The_Big_Project/Data/TCGA/idats"
path_tcga_full_meta  <- file.path(meta_data_path, "GSE68838_tidy_metadata.csv")
path_ewas_results    <- "/Users/adiallo/Desktop/Thesis/Results"

h_level         <- "H6"
tier            <- "Tier3"
phenotypes      <- c("any", "ln", "dist")
top_n_cpgs_list <- c(25, 50, 100)

base_output_dir <- "results_TCGA_validation_heatmaps_V7"
dir.create(base_output_dir, showWarnings = FALSE, recursive = TRUE)

reuse_if_exists         <- TRUE
export_row_labels_axis  <- FALSE   # set TRUE to also export a row-label strip
set.seed(20250809)

## -- 2) LIBRARIES & HELPERS --
suppressPackageStartupMessages({
  library(minfi); library(limma); library(pheatmap); library(sesame)
  library(dplyr); library(readr); library(glue); library(tibble); library(stringr)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  library(grid)
})

msg <- \(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

use_if_exists <- function(name, fun) {
  if (reuse_if_exists && exists(name, envir = .GlobalEnv) && !is.null(get(name, envir = .GlobalEnv))) {
    msg("Using ", name, " from RAM."); get(name, envir = .GlobalEnv)
  } else { val <- fun(); assign(name, val, envir = .GlobalEnv); val }
}

process_idats <- function(idat_dir, cache_path, sample_name_cleaner = NULL) {
  if (reuse_if_exists && file.exists(cache_path)) {
    msg("Loading cached beta matrix: ", basename(cache_path)); return(readRDS(cache_path))
  }
  msg("Processing raw IDATs in: ", idat_dir)
  idat_files <- list.files(idat_dir, pattern="\\.idat$", full.names=TRUE, recursive=TRUE)
  stopifnot(length(idat_files) > 0)
  basenames <- unique(sub("_Grn.idat|_Red.idat","", idat_files))
  rg_set    <- read.metharray(basenames = basenames, force = TRUE)
  if (!is.null(sample_name_cleaner)) colnames(rg_set) <- sample_name_cleaner(colnames(rg_set))
  mset <- preprocessNoob(rg_set)
  beta <- getBeta(mset)
  saveRDS(beta, cache_path)
  beta
}

## -- 3) draw_heatmap(): main (no legends) + separate legends + optional row labels + order CSVs
draw_heatmap <- function(beta_matrix,
                         meta_df,
                         cpg_ids,
                         out_stub,        # path WITHOUT extension; we’ll append suffixes
                         gap_pos = NULL,
                         palette = c("#313695","#4575b4","#74add1","#abd9e9","#e0f3f8",
                                     "#ffffbf","#fee090","#fdae61","#f46d43","#d73027","#a50026"),
                         n_breaks = 99) {
  cat("\n--- draw_heatmap() → STUB:", basename(out_stub), "---\n")
  if (is.null(meta_df)) stop("meta_df is NULL — cannot proceed.\n")

  ## 1) Intersect CpGs and subset beta
  cpgs <- intersect(cpg_ids, rownames(beta_matrix))
  if (length(cpgs) == 0) { msg("   ! No common CpGs with beta matrix. Skipping."); return(invisible(NULL)) }

  ## 2) Order samples by metadata groups (preserve meta order)
  pretty_group <- c(
    TCGA_Normal        = "TCGA Normal",
    No_Metastasis      = "TCGA No Metastasis",
    LN_Metastasis      = "TCGA LN Metastasis",
    Distant_Metastasis = "TCGA Distant Metastasis"
  )
  meta_df <- meta_df %>% filter(SampleID %in% colnames(beta_matrix))
  if (nrow(meta_df) == 0) { msg("   ! No metadata overlap."); return(invisible(NULL)) }

  meta_df$Group <- factor(meta_df$Group, levels = names(pretty_group), ordered = TRUE)
  meta_df <- meta_df %>% filter(!is.na(Group)) %>% arrange(Group)

  ordered_samples <- meta_df$SampleID[meta_df$SampleID %in% colnames(beta_matrix)]
  if (length(ordered_samples) == 0) { msg("   ! No ordered samples remain."); return(invisible(NULL)) }

  ann_col <- data.frame(
    `TCGA Status` = factor(
      pretty_group[ meta_df$Group[ match(ordered_samples, meta_df$SampleID) ] ],
      levels = pretty_group
    )
  )
  rownames(ann_col) <- ordered_samples

  beta_sub <- beta_matrix[cpgs, ordered_samples, drop = FALSE]

  ## 3) Row labels (Gene + CpG)
  epic_anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  sub_anno  <- epic_anno[cpgs, , drop = FALSE]
  gene_nm   <- sapply(strsplit(as.character(sub_anno$UCSC_RefGene_Name), ";"), `[`, 1)
  labels    <- ifelse(is.na(gene_nm) | gene_nm == "", cpgs, paste(gene_nm, cpgs))

  ## 4) Robust color breaks on row-scaled values
  row_scale <- function(m) { m2 <- t(scale(t(m))); m2[!is.finite(m2)] <- 0; m2 }
  z <- row_scale(beta_sub)
  rng <- range(z[is.finite(z)], na.rm = TRUE)
  if (!is.finite(rng[1]) || !is.finite(rng[2]) || rng[1] == rng[2]) {
    eps <- ifelse(all(rng == 0), 1e-6, max(1e-6, abs(rng[1])*1e-6))
    rng <- c(rng[1] - eps, rng[2] + eps)
  }
  breaks <- seq(rng[1], rng[2], length.out = n_breaks + 1)
  colors <- colorRampPalette(palette)(n_breaks)

  ## 5) Height heuristic
  plot_h <- max(3, 3 + length(cpgs) * 0.15)

  ## Filenames
  main_pdf      <- paste0(out_stub, ".pdf")
  collegend_pdf <- paste0(out_stub, "_legend.pdf")
  barlegend_pdf <- paste0(out_stub, "_colorbar.pdf")
  rowlabels_pdf <- paste0(out_stub, "_rowlabels.pdf")
  roworder_csv  <- paste0(out_stub, "_row_order.csv")
  colorder_csv  <- paste0(out_stub, "_col_order.csv")

  ## 6) MAIN HEATMAP PDF (NO TITLE, NO LEGENDS)
  pdf(main_pdf, width = 10, height = plot_h)
  ph <- pheatmap(
    z,                              # already row-scaled matrix
    color              = colors,
    breaks             = breaks,
    annotation_col     = ann_col,
    annotation_legend  = FALSE,     # <- hide annotation legend on main
    legend             = FALSE,     # <- hide heatmap colorbar on main
    show_colnames      = FALSE,
    cluster_rows       = TRUE,
    cluster_cols       = FALSE,
    treeheight_row     = 0,
    scale              = "none",
    fontsize_row       = 8,
    labels_row         = labels,
    gaps_row           = gap_pos,
    border_color       = NA,
    main               = NA         # <- no title
  )
  dev.off()
  msg(glue("✓ Saved PDF: {basename(main_pdf)}"))

  ## 7) COLUMN LEGEND (“axis”) PDF — match colnames exactly to annotation rows
  legend_dummy <- matrix(0, nrow = 1, ncol = length(ordered_samples))
  colnames(legend_dummy) <- ordered_samples
  rownames(legend_dummy) <- " "
  pdf(collegend_pdf, width = 3, height = 2)
  pheatmap(
    legend_dummy,
    color              = "white",
    breaks             = c(-1, 0, 1),
    annotation_col     = ann_col,
    annotation_legend  = TRUE,
    legend             = FALSE,
    cluster_rows       = FALSE,
    cluster_cols       = FALSE,
    show_rownames      = FALSE,
    show_colnames      = FALSE,
    cellheight         = 0.05,
    border_color       = NA,
    main               = NA
  )
  dev.off()
  msg(glue("✓ Saved Legend PDF: {basename(collegend_pdf)}"))

  ## 8) COLORBAR LEGEND PDF (for z-score scale)
  pdf(barlegend_pdf, width = 2.2, height = 3.0)
  pheatmap(
    matrix(seq_len(10), nrow = 10),
    color   = colors,
    breaks  = breaks,
    legend  = TRUE,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    show_rownames = FALSE,
    show_colnames = FALSE,
    annotation_legend = FALSE,
    border_color = NA,
    main = NA
  )
  dev.off()
  msg(glue("✓ Saved Colorbar PDF: {basename(barlegend_pdf)}"))

  ## 9) OPTIONAL: ROW LABELS PDF (ordered like the clustered heatmap)
  row_order_idx <- if (!is.null(ph$tree_row)) ph$tree_row$order else seq_len(nrow(z))
  if (isTRUE(export_row_labels_axis)) {
    labels_ordered <- labels[row_order_idx]
    pdf(rowlabels_pdf, width = 3, height = plot_h)
    grid::grid.newpage()
    vp <- grid::viewport(width = unit(1, "npc"), height = unit(1, "npc"))
    grid::pushViewport(vp)
    n <- length(labels_ordered)
    y_positions <- rev(seq(0.08, 0.92, length.out = n))
    for (i in seq_along(labels_ordered)) {
      grid::grid.text(labels_ordered[i], x = unit(0.98, "npc"), y = unit(y_positions[i], "npc"),
                      just = "right", gp = grid::gpar(cex = 0.7))
    }
    grid::popViewport()
    dev.off()
    msg(glue("✓ Saved Row Labels PDF: {basename(rowlabels_pdf)}"))
  }

  ## 10) WRITE ORDER CSVs (for Illustrator alignment) —— PATCHED
  # Build Group labels directly from meta_df to ensure matching length/order
  group_vec <- meta_df$Group[ match(ordered_samples, meta_df$SampleID) ]
  group_pretty_vec <- pretty_group[ as.character(group_vec) ]
  group_pretty_vec <- as.character(group_pretty_vec)  # ensure character

  # Guard Direction when gap_pos is NULL
  dir_vec <- if (!is.null(gap_pos)) {
    ifelse(seq_along(row_order_idx) <= gap_pos, "Hyper", "Hypo")
  } else {
    rep(NA_character_, length(row_order_idx))
  }

  row_order_tbl <- tibble(
    CpG_ID      = cpgs[row_order_idx],
    Gene        = ifelse(is.na(gene_nm[row_order_idx]) | gene_nm[row_order_idx] == "",
                         NA_character_, gene_nm[row_order_idx]),
    Label       = labels[row_order_idx],
    Direction   = dir_vec,
    RowPosition = seq_along(row_order_idx)
  )
  write_csv(row_order_tbl, roworder_csv)

  col_order_tbl <- tibble(
    SampleID    = ordered_samples,
    Group       = group_pretty_vec,
    ColPosition = seq_along(ordered_samples)
  )
  write_csv(col_order_tbl, colorder_csv)

  msg(glue("✓ Saved order CSVs: {basename(roworder_csv)}, {basename(colorder_csv)}"))

  invisible(list(
    main_pdf      = main_pdf,
    collegend_pdf = collegend_pdf,
    barlegend_pdf = barlegend_pdf,
    rowlabels_pdf = if (isTRUE(export_row_labels_axis)) rowlabels_pdf else NA
  ))
}

## -- 4) LOAD / CACHE TCGA DATA --
msg("--- Pre-loading TCGA beta matrix ---")
cache_path <- file.path(base_output_dir, "tcga_full_betas.rds")
beta_tcga <- if (file.exists(cache_path)) {
  msg("Loading cache: ", cache_path); readRDS(cache_path)
} else {
  msg("Cache missing — re-processing IDATs…")
  process_idats(path_tcga_full_idats, cache_path, sample_name_cleaner = \(s) sub("_.*", "", s))
}

# Load metadata
meta_tcga <- use_if_exists("meta_tcga", function() {
  raw <- read_csv(path_tcga_full_meta, show_col_types = FALSE)
  raw %>%
    filter(pathologic_t_staging == "T3") %>%
    rename(SampleID = geo_accession) %>%
    mutate(
      Group = case_when(
        str_detect(source_name_ch1, "Solid normal tissue")                  ~ "TCGA_Normal",
        clinical_m_staging == "M0" & pathologic_n_staging == "N0"           ~ "No_Metastasis",
        clinical_m_staging == "M0" & pathologic_n_staging != "N0"           ~ "LN_Metastasis",
        clinical_m_staging == "M1"                                          ~ "Distant_Metastasis",
        TRUE                                                                 ~ NA_character_
      )
    ) %>%
    filter(!is.na(Group)) %>%
    select(SampleID, Group)
})

## -- 5) MAIN ANALYSIS LOOP --
for (pheno in phenotypes) {
  msg(glue("\n=== VALIDATION: {toupper(pheno)} ==="))
  od <- file.path(base_output_dir, pheno)
  dir.create(od, showWarnings = FALSE, recursive = TRUE)

  fit_file <- file.path(path_ewas_results, glue("results_ewas_{pheno}"), h_level, "ewas_fits.rds")
  if (!file.exists(fit_file)) { msg("Skipping — no fits at ", fit_file); next }

  msg("--- Loading DH EWAS fits ---")
  fits   <- readRDS(fit_file)
  fo     <- fits[[tier]]
  fitobj <- if (inherits(fo, "MArrayLM")) fo else fo$fit
  tt     <- topTable(fitobj, coef = "pheno_bin", number = Inf, sort.by = "p")

  msg("--- Filtering to 450k probes ---")
  anno450k  <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  common_pr <- intersect(rownames(tt), rownames(anno450k))
  filt_tt   <- tt[common_pr, ]
  msg(glue(" -> {nrow(filt_tt)} probes on 450k array"))

  dt <- as.data.frame(filt_tt)
  for (n in top_n_cpgs_list) {
    n_half <- floor(n/2)
    hyper  <- rownames(utils::head(dt[dt$logFC >  0, ], n_half))
    hypo   <- rownames(utils::head(dt[dt$logFC <  0, ], n - n_half))
    combo  <- c(hyper, hypo)
    gap    <- length(hyper)

    msg(glue("Drawing Top {n} heatmap"))

    # Filename stub includes phenotype, H-level, tier, and N
    stub <- file.path(od, glue("heatmap_{pheno}_{h_level}_{tier}_top{n}"))

    draw_heatmap(
      beta_matrix = beta_tcga,
      meta_df     = meta_tcga,
      cpg_ids     = combo,
      out_stub    = stub,
      gap_pos     = gap
    )
  }
}

msg("\n✅ All TCGA validation analyses complete.")
```


Running EWAS on the TCGA Dataset
```{r}
###############################################################################
# TCGA (GEO→TCGA rename) + Tier-3 EWAS pipeline [FINAL VERSION]
# - Centralized results root + 300-dpi exports
# - Caching preserved; core logic unchanged
###############################################################################

suppressPackageStartupMessages({
  library(readr); library(dplyr); library(stringr); library(tibble)
  library(limma); library(HiTIMED); library(EpiDISH)
  library(minfi)
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  ## NEW: for plots/exports
  library(ggplot2)
  library(Cairo)   # Illustrator-friendly vector PDF
})

## --- CONFIG ------------------------------------------------------------------
path_tcga_full_idats <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/idats/"
meta_data_path       <- "/Users/adiallo/Desktop/The_Big_Project/Data/TCGA/idats"
path_tcga_full_meta  <- file.path(meta_data_path, "GSE68838_tidy_metadata.csv")

## NEW: Centralize all outputs here
results_root <- "/Users/adiallo/Desktop/Thesis/Results"
base_out_dir <- file.path(results_root, "TCGA_Tier3")
dir.create(base_out_dir, showWarnings = FALSE, recursive = TRUE)

### --- processed data cache (remains under base_out_dir) --- ###
path_beta_cache <- file.path(base_out_dir, "tcga_full_betas_from_idats.rds")

## --- helpers -----------------------------------------------------------------
msg <- \(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

### NEW: Illustrator PDF + 300-dpi raster helper (TIFF by default) -------------
save_pdf_tiff <- function(plot_obj, out_dir, stem,
                          width = 7, height = 5, dpi = 300,
                          raster_ext = "tiff") {
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  # 1) Vector PDF (for Illustrator)
  pdf_path <- file.path(out_dir, paste0(stem, ".pdf"))
  CairoPDF(file = pdf_path, width = width, height = height)
  print(plot_obj)
  dev.off()

  # 2) 300-dpi raster (journal compliance)
  raster_path <- file.path(out_dir, paste0(stem, ".", raster_ext))
  if (tolower(raster_ext) %in% c("tif","tiff")) {
    ggsave(raster_path, plot_obj,
           width = width, height = height, units = "in",
           dpi = dpi, device = "tiff", compression = "lzw")
  } else {
    ggsave(raster_path, plot_obj,
           width = width, height = height, units = "in",
           dpi = dpi, device = "png")
  }
  invisible(list(pdf = pdf_path, raster = raster_path))
}

### --- Efficient, caching-based iDat processing function --- ###
process_idats <- function(idat_dir, cache_path, sample_name_cleaner = NULL) {
  if (file.exists(cache_path)) {
    msg("Loading cached beta matrix: ", basename(cache_path))
    return(readRDS(cache_path))
  }
  msg("Cache not found. Processing raw IDATs from: ", idat_dir)
  idat_files <- list.files(idat_dir, pattern="\\.idat$", full.names=TRUE, recursive=TRUE)
  if (length(idat_files) == 0) stop("No .idat files found in ", idat_dir)

  basenames <- unique(sub("_Grn.idat|_Red.idat","", idat_files))
  rg_set    <- read.metharray(basenames = basenames, force = TRUE)

  if (!is.null(sample_name_cleaner)) {
    colnames(rg_set) <- sample_name_cleaner(colnames(rg_set))
  }

  mset <- preprocessNoob(rg_set)
  beta <- getBeta(mset)

  msg("Saving processed beta matrix to cache: ", basename(cache_path))
  saveRDS(beta, cache_path)
  beta
}

read_csv_silent <- function(file, ...) {
  if (utils::packageVersion("readr") >= "2.0.0") readr::read_csv(file, show_col_types = FALSE, ...)
  else suppressMessages(readr::read_csv(file, ...))
}
short16 <- function(x) substr(x, 1, 16)
extract_tcga_barcode <- function(x) {
  m <- stringr::str_match(x, "(TCGA-[A-Z0-9]{2}-[A-Z0-9]{4}-[0-9]{2}[A-Z])")
  m[,2]
}
as_M <- function(beta, eps=1e-6) log2((beta + eps)/(1 - beta + eps))
`%||%` <- function(a,b) if (is.null(a)) b else a

collapse_frac_Tier3 <- function(df) {
  get_ <- function(x) if (x %in% names(df)) df[[x]] else 0
  imm_cols <- intersect(c("CD8mem","CD8nv","CD4mem","CD4nv","Treg","Bmem","Bnv","DC","Mono","NK"),
                        names(df))
  totalImm <- if (length(imm_cols)) rowSums(df[imm_cols], na.rm = TRUE) else NA_real_
  totalImm[totalImm == 0] <- NA
  tibble(
    SampleID     = df$SampleID,
    Tumor        = get_("Tumor"),
    Fibroblast   = get_("Fib"),
    StromalOther = get_("Endothelial") + get_("Epithelial"),
    CD8_T        = (get_("CD8mem") + get_("CD8nv")) / totalImm,
    LymphOther   = (get_("CD4mem") + get_("CD4nv") + get_("Treg") + get_("Bmem") + get_("Bnv")) / totalImm
  ) %>% mutate(across(where(is.numeric), as.numeric))
}

TIER3_COVS <- c("age","sex_num","Tumor","Fibroblast","StromalOther","CD8_T","LymphOther")
nz_covs <- function(df, covs=TIER3_COVS) {
  keep <- intersect(covs, names(df))
  has_variance <- sapply(df[keep], function(x) {
    s <- sd(as.numeric(x), na.rm = TRUE)
    !is.na(s) && s > 0
  })
  keep[has_variance]
}

## --- 1) Load or Create beta_tcga object using the caching function ---------
msg("--- Checking for/Loading TCGA beta matrix ---")
beta_tcga <- process_idats(
  idat_dir = path_tcga_full_idats,
  cache_path = path_beta_cache,
  sample_name_cleaner = \(s) sub("_.*", "", s)
)

### --- 2) Copy for safe processing -------------------------------------------
beta_processed <- beta_tcga

## --- 3) Tier-3 deconvolution (HiTIMED + EpiDISH) ----------------------------
msg("Running HiTIMED (COAD, h=6) on original matrix…")
hitimed_results <- HiTIMED::HiTIMED_deconvolution(tumor_beta = beta_tcga, tumor_type = "COAD", h = 6)
frac_h6 <- hitimed_results$fractions %>%
  as.data.frame() %>%
  rownames_to_column("SampleID")

msg("EpiDISH Fib (RPC) on original matrix…")
data(centEpiFibIC.m)
common_fib <- intersect(rownames(beta_tcga), rownames(centEpiFibIC.m))
fib_scores <- if (length(common_fib) >= 50) {
  est <- EpiDISH::epidish(beta_tcga[common_fib, , drop=FALSE],
                          centEpiFibIC.m[common_fib, ], method="RPC")$estF
  as.data.frame(est) %>% rownames_to_column("SampleID") %>% transmute(SampleID, Fib = Fib * 100)
} else {
  warning("Few common probes for EpiDISH; setting Fib to NA.")
  tibble(SampleID = colnames(beta_tcga), Fib = NA_real_)
}

tier3_covars_gsm <- collapse_frac_Tier3(frac_h6 %>% left_join(fib_scores, by="SampleID"))

## --- 4) Metadata & GEO→TCGA mapping -----------------------------------------
if (!file.exists(path_tcga_full_meta)) stop("Metadata CSV not found: ", path_tcga_full_meta, call. = FALSE)
meta_raw <- read_csv_silent(path_tcga_full_meta)

meta_map <- meta_raw %>%
  mutate(
    gsm              = geo_accession,
    tcga_from_sample = short16(sample),
    tcga_from_title  = short16(extract_tcga_barcode(title)),
    tcga_barcode     = ifelse(!is.na(tcga_from_sample) & tcga_from_sample != "",
                              tcga_from_sample, tcga_from_title)
  ) %>%
  dplyr::select(gsm, tcga_barcode, everything())

if (sum(!is.na(meta_map$tcga_barcode)) == 0) {
  stop("Could not derive TCGA barcodes from 'sample' or 'title' in the CSV.", call. = FALSE)
}

## --- 5) Rename deconv results & beta columns to TCGA -------------------------
gsm_ids_in_results    <- tier3_covars_gsm$SampleID
tcga_ids_for_results  <- meta_map$tcga_barcode[match(gsm_ids_in_results, meta_map$gsm)]

tier3_covars <- tier3_covars_gsm %>%
  mutate(SampleID = tcga_ids_for_results) %>%
  filter(!is.na(SampleID) & SampleID != "")

msg(sprintf("Converted %d/%d SampleIDs in deconvolution results from GEO→TCGA.",
            nrow(tier3_covars), nrow(tier3_covars_gsm)))

old_cols <- colnames(beta_processed)
gsm_from_col <- sub("^((GSM[0-9]+)).*$", "\\1", old_cols)
map_vec <- meta_map$tcga_barcode[match(gsm_from_col, meta_map$gsm)]
new_cols <- ifelse(is.na(map_vec) | map_vec=="", old_cols, map_vec)
colnames(beta_processed) <- new_cols
mapped_n <- sum(!is.na(map_vec) & map_vec!="")
msg(sprintf("Renamed %d/%d beta columns from GEO→TCGA.", mapped_n, length(old_cols)))

## --- 6) Build tumor metadata (age/sex/Group) in TCGA space -------------------
meta_std <- meta_raw %>%
  transmute(
    SampleID = short16(sample),
    age      = suppressWarnings(as.numeric(age)),
    sex_num  = case_when(
      tolower(sex) %in% c("male","m")   ~ 1L,
      tolower(sex) %in% c("female","f") ~ 0L,
      TRUE ~ NA_integer_
    ),
    is_normal = str_detect(source_name_ch1 %||% "", regex("^\\s*Solid\\s+normal\\s+tissue$", ignore_case = TRUE)),
    M_clean   = toupper(trimws(clinical_m_staging %||% "")),
    N_clean   = toupper(trimws(pathologic_n_staging %||% "")),
    Group = case_when(
      is_normal ~ "TCGA_Normal",
      M_clean == "M1" ~ "Distant_Metastasis",
      M_clean == "M0" & N_clean %in% c("N1","N1A","N1B","N1C","N2","N2A","N2B") ~ "LN_Metastasis",
      M_clean == "M0" & N_clean %in% c("N0") ~ "No_Metastasis",
      TRUE ~ NA_character_
    )
  )

missing_tcga <- which(is.na(meta_std$SampleID) | meta_std$SampleID=="")
if (length(missing_tcga)) {
  fill <- short16(extract_tcga_barcode(meta_raw$title[missing_tcga]))
  meta_std$SampleID[missing_tcga] <- fill
}

is_tumor <- meta_std$Group %in% c("No_Metastasis","LN_Metastasis","Distant_Metastasis")
meta_tumor <- meta_std %>%
  filter(is_tumor, SampleID %in% colnames(beta_processed)) %>%
  distinct(SampleID, .keep_all = TRUE)

msg(sprintf("Tumor samples matched to beta_processed: %d", nrow(meta_tumor)))
print(meta_tumor %>% dplyr::count(Group))

## --- 7) Final Tier-3 metadata ------------------------------------------------
meta_tcga_Tier3 <- meta_tumor %>%
  left_join(tier3_covars, by = "SampleID") %>%
  filter(SampleID %in% colnames(beta_processed)) %>%
  arrange(match(SampleID, colnames(beta_processed))) %>%
  mutate(Array_Version = "450k")

usable_covs <- nz_covs(meta_tcga_Tier3)
msg("Tier-3 covariates usable: ", paste(usable_covs, collapse = ", "))

## --- 8) Remove sex-chromosome probes ----------------------------------------
anno <- minfi::getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno <- anno[rownames(anno) %in% rownames(beta_processed), , drop = FALSE]
chr_vec <- NULL
if ("chr" %in% colnames(anno))       chr_vec <- anno$chr
if (is.null(chr_vec) && "CHR" %in% colnames(anno)) chr_vec <- anno$CHR
if (is.null(chr_vec) && "seqnames" %in% colnames(anno)) chr_vec <- as.character(anno$seqnames)
if (is.null(chr_vec)) stop("Could not find a chromosome column in 450k annotation.")
is_sex_chr <- chr_vec %in% c("chrX","chrY","X","Y")
sex_probes <- rownames(anno)[is_sex_chr]
n_before <- nrow(beta_processed)
if (length(sex_probes)) {
  beta_processed <- beta_processed[setdiff(rownames(beta_processed), sex_probes), , drop = FALSE]
}
msg(length(sex_probes), " chrX/chrY probes removed for TCGA EWAS; ",
    n_before, " → ", nrow(beta_processed), " CpGs remain.")

## --- 9) EWAS (Tier-3 only) + figure exports ---------------------------------
M_processed <- as_M(beta_processed)

meta_tcga_Tier3 <- meta_tcga_Tier3 %>%
  mutate(
    any_mets = as.integer(Group %in% c("LN_Metastasis","Distant_Metastasis")),
    ln_only  = as.integer(Group == "LN_Metastasis"),
    distant  = as.integer(Group == "Distant_Metastasis")
  )

run_ewas_tier3 <- function(M, meta, pheno_var, out_dir) {
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  figs_dir <- file.path(out_dir, "figures")
  dir.create(figs_dir, showWarnings = FALSE, recursive = TRUE)

  meta$pheno_bin <- dplyr::case_when(
    pheno_var == "any_mets" ~ meta$any_mets,
    pheno_var == "ln_only"  ~ meta$ln_only,
    pheno_var == "distant"  ~ meta$distant,
    TRUE ~ NA_integer_
  )
  meta_f <- meta %>% filter(!is.na(pheno_bin))
  keep_ids <- meta_f$SampleID[meta_f$SampleID %in% colnames(M)]
  M_f      <- M[, keep_ids, drop=FALSE]
  meta_f   <- meta_f[match(keep_ids, meta_f$SampleID), ]

  covs3 <- nz_covs(meta_f)
  form3 <- if (length(covs3)) as.formula(paste("~ pheno_bin +", paste(covs3, collapse=" + "))) else ~ pheno_bin
  des3  <- model.matrix(form3, meta_f)

  fit3  <- limma::eBayes(limma::lmFit(M_f, des3), trend=TRUE, robust=TRUE)
  tt    <- limma::topTable(fit3, coef="pheno_bin", number=Inf, sort.by="P")

  # Persist stats
  saveRDS(list(fit=fit3, design=des3, covs_used=covs3, meta=meta_f),
          file.path(out_dir, "fit.rds"))
  readr::write_csv(tibble::rownames_to_column(as.data.frame(tt), "CpG"),
                   file.path(out_dir, "topTable.csv"))

  # --- NEW: quick volcano → PDF (vector) + TIFF (300 dpi) --------------------
  if (all(c("logFC","P.Value") %in% colnames(tt))) {
    dfv <- tibble(CpG = rownames(tt),
                  logFC = tt$logFC,
                  neglog10P = -log10(tt$P.Value))
    vol <- ggplot(dfv, aes(x = logFC, y = neglog10P)) +
      geom_point(alpha = 0.6, size = 0.8) +
      labs(title = paste0("Volcano (", pheno_var, ")"),
           x = "logFC (metastasis effect)",
           y = expression(-log[10](italic(P)))) +
      theme_minimal(base_size = 11)

    save_pdf_tiff(vol, figs_dir, stem = paste0("volcano_", pheno_var),
                  width = 7, height = 5, dpi = 300, raster_ext = "tiff")
  }

  msg(sprintf("EWAS %s → n=%d | covs: %s | top CpG: %s (P=%.3g)",
              pheno_var, nrow(meta_f), paste(covs3, collapse=", "),
              rownames(tt)[1], tt$P.Value[1]))
  invisible(tt)
}

for (ph in c("any_mets","ln_only","distant")) {
  run_ewas_tier3(M_processed, meta_tcga_Tier3, ph, file.path(base_out_dir, ph))
}

msg("\n✅ TCGA Tier-3 validation complete. Outputs → ", base_out_dir)
```

Venn Diagram
```{r}
###############################################################################
# DH vs TCGA EWAS validation — Venn-only (with 450k harmonization)
# Discovery: DH p < 1e-3 (nominal)
# Replication: TCGA one-sided p < 0.05 in DH-implied direction
# - DH fits (Tier3):
#   /Users/adiallo/Desktop/Thesis/code/results_ewas_any/H6/ewas_fits.rds
#   /Users/adiallo/Desktop/Thesis/code/results_ewas_ln/H6/ewas_fits.rds
#   /Users/adiallo/Desktop/Thesis/code/results_ewas_dist/H6/ewas_fits.rds
# - TCGA fits (Tier3):
#   results_tcga_tier3/{any_mets, ln_only, distant}/fit.rds
# - Output: results_validation_DH_vs_TCGA/venn_only/venn_{dh}_{tcga}.pdf
###############################################################################

suppressPackageStartupMessages({
  library(limma); library(dplyr); library(tibble); library(readr); library(glue)
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
})

# Ensure VennDiagram is installed/loaded (for Venn-only output)
if (!requireNamespace("VennDiagram", quietly = TRUE)) {
  install.packages("VennDiagram", repos = "https://cloud.r-project.org")
}
library(VennDiagram); library(grid)

## --- CONFIG ------------------------------------------------------------------
dh_root   <- "/Users/adiallo/Desktop/Thesis/Results"
tcga_root <- "results_tcga_tier3"
out_dir   <- "results_validation_DH_vs_TCGA"
venn_dir  <- file.path(out_dir, "venn_only")
dir.create(venn_dir, showWarnings = FALSE, recursive = TRUE)

# Corrected mapping
pheno_map <- tibble::tribble(
  ~dh_tag, ~tcga_tag,   ~pretty,
  "any",   "any_mets",  "Any metastasis",
  "ln",    "ln_only",   "LN-only",
  "dist",  "distant",   "Distant"
)

# Thresholds
DH_P_CUT            <- 1e-3   # DH discovery threshold (nominal p)
TCGA_P_ONE_SIDED    <- 0.05   # TCGA replication threshold (one-sided)

## --- Helpers -----------------------------------------------------------------
msg <- \(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")
`%||%` <- function(a,b) if (is.null(a)) b else a

top_table_robust <- function(fit) {
  cols <- colnames(fit$coefficients)
  coef_to_use <- if ("pheno_bin" %in% cols) "pheno_bin" else tail(cols, 1)
  limma::topTable(fit, coef = coef_to_use, number = Inf, sort.by = "P")
}

load_dh <- function(dh_tag) {
  f <- file.path(dh_root, glue("results_ewas_{dh_tag}"), "H6", "ewas_fits.rds")
  if (!file.exists(f)) stop("Missing DH fit: ", f, call. = FALSE)
  fits <- readRDS(f)
  fit  <- fits$Tier3 %||% fits[[3]]
  tt   <- top_table_robust(fit)
  tibble::rownames_to_column(as.data.frame(tt), "CpG") %>%
    transmute(CpG, logFC_dh = logFC, P_dh = P.Value, adjP_dh = adj.P.Val)
}

load_tcga <- function(tcga_tag) {
  f <- file.path(tcga_root, tcga_tag, "fit.rds")
  if (!file.exists(f)) stop("Missing TCGA fit: ", f, call. = FALSE)
  obj <- readRDS(f)
  fit <- if (inherits(obj, "MArrayLM")) obj else obj$fit
  tt  <- top_table_robust(fit)
  tibble::rownames_to_column(as.data.frame(tt), "CpG") %>%
    transmute(CpG, logFC_tcga = logFC, P_tcga = P.Value, adjP_tcga = adj.P.Val)
}

# Convert TCGA two-sided p to one-sided p in the direction predicted by DH
p_one_sided_in_DH_direction <- function(p_two, sign_tcga, sign_dh) {
  same <- sign_tcga == sign_dh
  p1   <- ifelse(same, p_two/2, 1 - p_two/2)  # conservative one-sided from two-sided
  p1
}

draw_pair_venn <- function(setA, setB, labels, main_title) {
  grid::grid.newpage()
  v <- VennDiagram::draw.pairwise.venn(
    area1       = length(setA),
    area2       = length(setB),
    cross.area  = length(intersect(setA, setB)),
    category    = labels,
    fill        = c("#6baed6", "#fd8d3c"),
    alpha       = c(0.5, 0.5),
    lty         = "blank",
    cat.cex     = 1.2,
    cex         = 1.5,
    main        = main_title
  )
  grid::grid.draw(v)
}

## --- 450k probe universe (platform harmonization) ----------------------------
anno450k  <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
cpgs_450k <- rownames(anno450k)

## --- Generate Venn-only PDFs -------------------------------------------------
for (i in seq_len(nrow(pheno_map))) {
  dh_tag   <- pheno_map$dh_tag[i]
  tcga_tag <- pheno_map$tcga_tag[i]
  pretty   <- pheno_map$pretty[i]

  msg(glue("— {pretty} — loading & harmonizing"))
  dh   <- load_dh(dh_tag)     %>% filter(CpG %in% cpgs_450k)
  tcga <- load_tcga(tcga_tag) %>% filter(CpG %in% cpgs_450k)

  # DH discovery set: nominal p < 1e-3
  dh$sign_dh <- sign(dh$logFC_dh)
  dh_disc_all <- dh %>% filter(P_dh < DH_P_CUT)

  # Compute TCGA one-sided p in DH direction (for CpGs present in both)
  both <- inner_join(
    dh_disc_all %>% dplyr::select(CpG, sign_dh),
    tcga %>% dplyr::select(CpG, logFC_tcga, P_tcga),
    by = "CpG"
  ) %>%
    mutate(
      sign_tcga = sign(logFC_tcga),
      p1_tcga   = p_one_sided_in_DH_direction(P_tcga, sign_tcga, sign_dh)
    )

  # TCGA replication set = CpGs with one-sided p < 0.05 in DH direction
  tcga_rep_all <- both %>% filter(p1_tcga < TCGA_P_ONE_SIDED)

  # Direction-specific subsets (hyper/hypo defined by DH sign)
  dh_hyper <- dh_disc_all %>% filter(sign_dh > 0)
  dh_hypo  <- dh_disc_all %>% filter(sign_dh < 0)

  tcga_rep_hyper <- both %>% filter(sign_dh > 0, p1_tcga < TCGA_P_ONE_SIDED)
  tcga_rep_hypo  <- both %>% filter(sign_dh < 0, p1_tcga < TCGA_P_ONE_SIDED)

  # Prepare sets for Venns
  set_dh_all   <- dh_disc_all$CpG
  set_tcga_all <- tcga_rep_all$CpG

  set_dh_hyper   <- dh_hyper$CpG
  set_tcga_hyper <- tcga_rep_hyper$CpG

  set_dh_hypo   <- dh_hypo$CpG
  set_tcga_hypo <- tcga_rep_hypo$CpG

  # PDF (3 pages): All / Hyper / Hypo
  pdf_file <- file.path(venn_dir, glue("venn_{dh_tag}_{tcga_tag}.pdf"))
  pdf(pdf_file, width = 6.5, height = 5.5)

  # 1) All significant (DH nominal vs TCGA one-sided replication)
  draw_pair_venn(
    set_dh_all, set_tcga_all,
    labels = c("DH (p<1e-3)", "TCGA (one-sided p<0.05)"),
    main_title = glue("{pretty}: All (DH nominal, TCGA one-sided)")
  )

  # 2) Hyper only (DH>0; TCGA same direction one-sided)
  draw_pair_venn(
    set_dh_hyper, set_tcga_hyper,
    labels = c("DH hyper (p<1e-3)", "TCGA hyper (one-sided p<0.05)"),
    main_title = glue("{pretty}: Hyper only (direction-aware)")
  )

  # 3) Hypo only (DH<0; TCGA same direction one-sided)
  draw_pair_venn(
    set_dh_hypo, set_tcga_hypo,
    labels = c("DH hypo (p<1e-3)", "TCGA hypo (one-sided p<0.05)"),
    main_title = glue("{pretty}: Hypo only (direction-aware)")
  )

  dev.off()
  msg("✓ Wrote Venn PDF: ", pdf_file)
}

msg("Done. Venn-only PDFs are in: ", normalizePath(venn_dir))
```


Maing a table of comparisons
```{r}
###############################################################################
# DH vs TCGA EWAS validation — Table output (centralized results + 300-dpi figs)
# - For each phenotype, list DH discovery CpGs (p < 1e-3)
# - Annotate with TCGA logFC, p-value, one-sided p, replication flag
# - Outputs:
#     <results_root>/Validation_DH_vs_TCGA/tables/{tag}_validation.csv
#   Optional (NEW):
#     <results_root>/Validation_DH_vs_TCGA/figures/replication_counts.(pdf|tiff)
###############################################################################

suppressPackageStartupMessages({
  library(limma); library(dplyr); library(tibble); library(readr); library(glue)
  ## NEW (optional figs)
  library(ggplot2); library(Cairo)
})

## --- CENTRALIZED PATHS -------------------------------------------------------
results_root <- "/Users/adiallo/Desktop/Thesis/Results"

# Consistent with your updated TCGA Tier-3 script:
tcga_root <- file.path(results_root, "TCGA_Tier3")

# Validation outputs live here:
out_dir <- file.path(results_root, "Validation_DH_vs_TCGA")
tab_dir <- file.path(out_dir, "tables")
fig_dir <- file.path(out_dir, "figures")  # NEW (optional)
dir.create(tab_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

## If your DH discovery fits live under Thesis/Results as well, set this:
dh_root <- results_root
# If your DH discovery fits still live under your code tree, uncomment:
# dh_root <- "/Users/adiallo/Desktop/Thesis/code"

pheno_map <- tribble(
  ~dh_tag, ~tcga_tag,   ~pretty,
  "any",   "any_mets",  "Any metastasis",
  "ln",    "ln_only",   "LN-only",
  "dist",  "distant",   "Distant"
)

DH_P_CUT         <- 1e-3   # discovery cutoff
TCGA_P_ONE_SIDED <- 0.05   # replication cutoff

msg <- \(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")
`%||%` <- function(a,b) if (is.null(a)) b else a

top_table_robust <- function(fit) {
  cols <- colnames(fit$coefficients)
  coef_to_use <- if ("pheno_bin" %in% cols) "pheno_bin" else tail(cols, 1)
  limma::topTable(fit, coef = coef_to_use, number = Inf, sort.by = "P")
}

load_dh <- function(dh_tag) {
  # Looks for: <results_root>/results_ewas_<tag>/H6/ewas_fits.rds  (or in dh_root)
  f <- file.path(dh_root, glue("results_ewas_{dh_tag}"), "H6", "ewas_fits.rds")
  if (!file.exists(f)) stop("DH fits not found: ", f, call. = FALSE)
  fits <- readRDS(f)
  fit  <- fits$Tier3 %||% fits[[3]]
  tt   <- top_table_robust(fit)
  rownames_to_column(as.data.frame(tt), "CpG") %>%
    transmute(CpG, logFC_dh = logFC, P_dh = P.Value)
}

load_tcga <- function(tcga_tag) {
  # Looks for: <results_root>/TCGA_Tier3/<tcga_tag>/fit.rds
  f <- file.path(tcga_root, tcga_tag, "fit.rds")
  if (!file.exists(f)) stop("TCGA fit not found: ", f, call. = FALSE)
  obj <- readRDS(f)
  fit <- if (inherits(obj, "MArrayLM")) obj else obj$fit
  tt  <- top_table_robust(fit)
  rownames_to_column(as.data.frame(tt), "CpG") %>%
    transmute(CpG, logFC_tcga = logFC, P_tcga = P.Value)
}

# one-sided p in DH direction
p_one_sided <- function(p_two, sign_tcga, sign_dh) {
  same <- sign_tcga == sign_dh
  ifelse(same, p_two/2, 1 - p_two/2)
}

## --- Main loop ---------------------------------------------------------------
rep_counts <- list()  # NEW (optional summary for plotting)

for (i in seq_len(nrow(pheno_map))) {
  dh_tag   <- pheno_map$dh_tag[i]
  tcga_tag <- pheno_map$tcga_tag[i]
  pretty   <- pheno_map$pretty[i]

  msg(glue("Processing {pretty}"))

  dh   <- load_dh(dh_tag)     %>% filter(P_dh < DH_P_CUT)
  tcga <- load_tcga(tcga_tag)

  tbl <- dh %>%
    left_join(tcga, by = "CpG") %>%
    mutate(
      sign_dh        = sign(logFC_dh),
      sign_tcga      = sign(logFC_tcga),
      one_sided_tcga = p_one_sided(P_tcga, sign_tcga, sign_dh),
      Replicated     = ifelse(!is.na(one_sided_tcga) & one_sided_tcga < TCGA_P_ONE_SIDED, "Yes", "No")
    )

  out_csv <- file.path(tab_dir, glue("{dh_tag}_{tcga_tag}_validation.csv"))
  write_csv(tbl, out_csv)
  msg("✓ Wrote ", out_csv, " with ", nrow(tbl), " DH discovery CpGs")

  # Collect counts for optional summary figure
  rep_counts[[tcga_tag]] <- tibble(
    phenotype = tcga_tag,
    n_discovery = nrow(dh),
    n_rep = sum(tbl$Replicated == "Yes", na.rm = TRUE)
  )
}

## --- OPTIONAL: Save a quick replication summary figure (PDF + 300-dpi TIFF) --
if (length(rep_counts)) {
  dfc <- bind_rows(rep_counts)
  dfc$phenotype <- factor(dfc$phenotype, levels = pheno_map$tcga_tag)

  p <- ggplot(dfc, aes(x = phenotype, y = n_rep)) +
    geom_col() +
    geom_text(aes(label = paste0(n_rep, "/", n_discovery)),
              vjust = -0.4, size = 3) +
    labs(title = "Replication counts in TCGA (one-sided p < 0.05)",
         x = NULL, y = "Replicated CpGs") +
    theme_minimal(base_size = 11)

  # Illustrator-ready vector PDF
  pdf_path <- file.path(fig_dir, "replication_counts.pdf")
  CairoPDF(file = pdf_path, width = 6.5, height = 4.5)
  print(p); dev.off()

  # 300-dpi TIFF (journal-friendly)
  tiff_path <- file.path(fig_dir, "replication_counts.tiff")
  ggsave(tiff_path, p, width = 6.5, height = 4.5, units = "in",
         dpi = 300, device = "tiff", compression = "lzw")

  msg("✓ Saved summary figure → ", pdf_path, " and ", tiff_path)
}
```


Generating Boxplot to compare significant CpG's which were shared between experiments DH and TCGA
```{r}
# --- SCRIPT 1: 01_prepare_data.R ---

suppressPackageStartupMessages({
  library(dplyr); library(readr); library(tidyr); library(stringr)
  library(glue); library(tibble); library(minfi)
})

## ---- Paths (Update as needed) ------------------------------------------------
results_root <- "/Users/adiallo/Desktop/Thesis/Results"
val_root     <- file.path(results_root, "Validation_DH_vs_TCGA")
tab_dir      <- file.path(val_root, "tables")
fig_dir      <- file.path(val_root, "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

dh_beta_rds <- file.path(results_root, "ewas_cache", "beta_comb_qc.rds")
dh_meta_csv <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"
path_tcga_full_idats <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/DNA_Methylation/dm_data/TCGA/idats/"
path_tcga_full_meta  <- "/Users/adiallo/Desktop/The_Big_Project/Data/TCGA/idats/GSE68838_tidy_metadata.csv"

## ---- Phenotype map & Tunables -----------------------------------------------
pheno_map <- tribble(
  ~dh_tag, ~tcga_tag,   ~pretty,
  "any",   "any_mets",  "Any metastasis",
  "ln",    "ln_only",   "LN-only",
  "dist",  "distant",   "Distant"
)
MAX_CPGS_PER_PHENO  <- 24

## ---- Helper Functions -------------------------------------------------------
msg <- \(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

load_dh_beta <- function(path) {
  if (!file.exists(path)) stop("DH β RDS not found: ", path, call. = FALSE)
  obj <- readRDS(path)
  if (is.list(obj) && !is.null(obj$beta)) return(obj$beta)
  if (is.matrix(obj)) return(obj)
  stop("Unexpected DH β object in: ", path)
}

load_dh_meta <- function(csv_path) {
  if (!file.exists(csv_path)) stop("DH meta CSV not found: ", csv_path, call. = FALSE)
  read_csv(csv_path, show_col_types = FALSE) %>%
    transmute(
      SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
      any_mets = as.integer(any_mets),
      ln_only  = as.integer(ln_only),
      distant  = as.integer(Distant_Mets)
    )
}

load_tcga_beta_from_idats <- function(idat_dir) {
  msg("Processing TCGA raw IDATs from: ", idat_dir)
  idat_files <- list.files(idat_dir, pattern="\\.idat$", full.names=TRUE, recursive=TRUE)
  if (length(idat_files) == 0) stop("No .idat files found in ", idat_dir)

  basenames <- unique(sub("_Grn.idat|_Red.idat","", idat_files))
  rg_set    <- read.metharray(basenames = basenames, force = TRUE)
  
  colnames(rg_set) <- sub("_.*", "", colnames(rg_set))

  mset <- preprocessNoob(rg_set)
  beta <- getBeta(mset)
  
  msg(glue("Successfully processed {ncol(beta)} TCGA samples from IDATs."))
  return(beta)
}

build_group_dh <- function(meta_dh, dh_tag) {
  if (dh_tag == "any") {
    meta_dh %>% mutate(Grp2 = case_when(
      any_mets == 1 ~ "Metastasis",
      any_mets == 0 ~ "No_Metastasis",
      TRUE ~ NA_character_
    )) %>% filter(!is.na(Grp2)) %>%
      mutate(Grp2 = factor(Grp2, levels = c("No_Metastasis","Metastasis")))
  } else if (dh_tag == "ln") {
    meta_dh %>% mutate(Grp2 = case_when(
      ln_only == 1 ~ "LN_Metastasis",
      ln_only == 0 ~ "No_Metastasis",
      TRUE ~ NA_character_
    )) %>% filter(!is.na(Grp2)) %>%
      mutate(Grp2 = factor(Grp2, levels = c("No_Metastasis","LN_Metastasis")))
  } else if (dh_tag == "dist") {
    meta_dh %>% mutate(Grp2 = case_when(
      distant == 1 ~ "Distant_Metastasis",
      distant == 0 ~ "No_Metastasis",
      TRUE ~ NA_character_
    )) %>% filter(!is.na(Grp2)) %>%
      mutate(Grp2 = factor(Grp2, levels = c("No_Metastasis","Distant_Metastasis")))
  } else stop("Unknown dh_tag: ", dh_tag)
}

build_group_tcga <- function(meta_tcga, tcga_tag) {
  if (tcga_tag == "any_mets") {
    meta_tcga %>% mutate(Grp2 = case_when(
      Group %in% c("LN_Metastasis","Distant_Metastasis") ~ "Metastasis",
      Group == "No_Metastasis" ~ "No_Metastasis",
      TRUE ~ NA_character_
    )) %>% filter(!is.na(Grp2)) %>%
      mutate(Grp2 = factor(Grp2, levels = c("No_Metastasis","Metastasis")))
  } else if (tcga_tag == "ln_only") {
    meta_tcga %>% mutate(Grp2 = case_when(
      Group == "LN_Metastasis" ~ "LN_Metastasis",
      Group == "No_Metastasis" ~ "No_Metastasis",
      TRUE ~ NA_character_
    )) %>% filter(!is.na(Grp2)) %>%
      mutate(Grp2 = factor(Grp2, levels = c("No_Metastasis","LN_Metastasis")))
  } else if (tcga_tag == "distant") {
    meta_tcga %>% mutate(Grp2 = case_when(
      Group == "Distant_Metastasis" ~ "Distant_Metastasis",
      Group == "No_Metastasis" ~ "No_Metastasis",
      TRUE ~ NA_character_
    )) %>% filter(!is.na(Grp2)) %>%
      mutate(Grp2 = factor(Grp2, levels = c("No_Metastasis","Distant_Metastasis")))
  } else stop("Unknown tcga_tag: ", tcga_tag)
}

beta_long <- function(beta_mat, sample_ids, cpg_ids, dataset_label) {
  cpgs <- intersect(cpg_ids, rownames(beta_mat))
  sids <- intersect(sample_ids, colnames(beta_mat))
  if (!length(cpgs) || !length(sids)) return(tibble())
  as_tibble(beta_mat[cpgs, sids, drop = FALSE], rownames = "CpG") |>
    pivot_longer(-CpG, names_to = "SampleID", values_to = "Beta") |>
    mutate(Dataset = dataset_label, .before = 1)
}

## ---- Load and Align DH Data -------------------------------------------------
msg("Loading DH β and metadata …")
beta_dh <- load_dh_beta(dh_beta_rds)
meta_dh <- load_dh_meta(dh_meta_csv)

msg("Aligning DH beta and metadata...")
common_dh_ids <- intersect(meta_dh$SampleID, colnames(beta_dh))
beta_dh <- beta_dh[, common_dh_ids, drop = FALSE]
meta_dh <- meta_dh %>% filter(SampleID %in% common_dh_ids)
msg(glue("Found {length(common_dh_ids)} common samples for DH analysis."))

## ---- Load and Align TCGA Data -----------------------------------------------
beta_tcga <- load_tcga_beta_from_idats(path_tcga_full_idats)
msg("Loading and preparing TCGA metadata...")
if (!file.exists(path_tcga_full_meta)) {
  stop("TCGA metadata file not found: ", path_tcga_full_meta)
}
meta_raw_tcga <- read_csv(path_tcga_full_meta, show_col_types = FALSE)
meta_tcga <- meta_raw_tcga %>%
  rename(SampleID = geo_accession) %>%
  mutate(
    Group = case_when(
      clinical_m_staging == "M0" & pathologic_n_staging == "N0"       ~ "No_Metastasis",
      clinical_m_staging == "M0" & pathologic_n_staging %in% c("N1", "N1a", "N1b", "N1c", "N2", "N2a", "N2b") ~ "LN_Metastasis",
      clinical_m_staging == "M1"                                      ~ "Distant_Metastasis",
      TRUE                                                            ~ NA_character_
    )
  ) %>%
  select(SampleID, Group) %>%
  filter(!is.na(Group))
msg(glue("Prepared TCGA metadata for {nrow(meta_tcga)} samples."))

common_tcga_ids <- intersect(colnames(beta_tcga), meta_tcga$SampleID)
beta_tcga <- beta_tcga[, common_tcga_ids, drop = FALSE]
meta_tcga <- meta_tcga %>% filter(SampleID %in% common_tcga_ids)
msg(glue("Found {length(common_tcga_ids)} final common samples for TCGA."))

## ---- Main Data Preparation Loop ---------------------------------------------
all_pheno_data <- list()

for (i in seq_len(nrow(pheno_map))) {
  dh_tag   <- pheno_map$dh_tag[i]
  tcga_tag <- pheno_map$tcga_tag[i]
  pretty   <- pheno_map$pretty[i]
  msg(glue("\n--- Preparing data for phenotype: {pretty} ---"))

  val_csv <- file.path(tab_dir, glue("{dh_tag}_{tcga_tag}_validation.csv"))
  if (!file.exists(val_csv)) { msg("Skipping, no validation file."); next }
  val_tbl <- read_csv(val_csv, show_col_types = FALSE)
  rep_cpgs <- val_tbl %>% filter(Replicated == "Yes") %>% arrange(P_tcga) %>% pull(CpG) %>% unique() %>% head(MAX_CPGS_PER_PHENO)
  if (!length(rep_cpgs)) { msg("Skipping, no replicated CpGs."); next }

  meta_dh2   <- build_group_dh(meta_dh, dh_tag)
  meta_tcga2 <- build_group_tcga(meta_tcga, tcga_tag)

  dh_long   <- beta_long(beta_dh, meta_dh2$SampleID, rep_cpgs, "DH") %>%
    left_join(meta_dh2 %>% select(SampleID, Grp2), by = "SampleID")
  tcga_long <- beta_long(beta_tcga, meta_tcga2$SampleID, rep_cpgs, "TCGA") %>%
    left_join(meta_tcga2 %>% select(SampleID, Grp2), by = "SampleID")

  both <- bind_rows(dh_long, tcga_long) %>%
    filter(!is.na(Grp2), !is.na(Beta)) %>%
    mutate(
      phenotype_tag = tcga_tag,
      pretty_pheno = pretty,
      CpG = factor(CpG, levels = rep_cpgs)
    )
  
  all_pheno_data[[tcga_tag]] <- both
}

## ---- Combine and Save Final Data File ---------------------------------------
final_data_for_plotting <- bind_rows(all_pheno_data)
output_csv_path <- file.path(fig_dir, "combined_data_for_plotting.csv")
write_csv(final_data_for_plotting, output_csv_path)

msg(glue("\n✅ Successfully created the combined data file with {nrow(final_data_for_plotting)} rows."))
msg(glue("Please inspect the file here: {output_csv_path}"))
```

```{r}
# --- SCRIPT 10: 10_final_single_comprehensive_plot.R ---

suppressPackageStartupMessages({
  library(dplyr); library(readr); library(tidyr); library(stringr)
  library(ggplot2); library(glue); library(tibble); library(Cairo); library(purrr)
  library(ggpubr) 
})

## ---- Paths and Tunables -----------------------------------------------------
fig_dir      <- "/Users/adiallo/Desktop/Thesis/Results/Validation_DH_vs_TCGA/figures"
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

# Path to the data file created by Script 1
input_csv    <- file.path(fig_dir, "combined_data_for_plotting.csv")

## ---- Helper Functions -------------------------------------------------------
msg <- \(...) cat("[", format(Sys.time(), "%H:%M:%S"), "] ", ..., "\n")

grp_levels_for <- function(tcga_tag) {
  switch(tcga_tag,
    "any_mets" = c("No_Metastasis","Metastasis"),
    "ln_only"  = c("No_Metastasis","LN_Metastasis"),
    "distant"  = c("No_Metastasis","Distant_Metastasis"),
    stop("Unknown tcga_tag: ", tcga_tag)
  )
}

save_pdf_tiff <- function(plot_obj, out_stub, width, height, dpi = 300) {
  dir.create(dirname(out_stub), showWarnings = FALSE, recursive = TRUE)
  CairoPDF(file = paste0(out_stub, ".pdf"), width = width, height = height)
  print(plot_obj); dev.off()
  ggsave(paste0(out_stub, ".tiff"), plot_obj,
         width = width, height = height, units = "in",
         dpi = dpi, device = "tiff", compression = "lzw")
}

## ---- Load the prepared data -------------------------------------------------
if (!file.exists(input_csv)) {
  stop("Clean data file not found. Please run Script 1 first: ", input_csv)
}
combined_data <- read_csv(input_csv, show_col_types = FALSE)
msg(glue("Successfully loaded prepared data with {nrow(combined_data)} rows."))

## ---- Main Loop for Single, All-in-One Plots ---------------------------------
phenotypes_to_run <- unique(combined_data$phenotype_tag)

for (current_pheno in phenotypes_to_run) {
  
  pheno_data <- combined_data %>% filter(phenotype_tag == current_pheno)
  pretty_name <- unique(pheno_data$pretty_pheno)
  msg(glue("\n--- Generating single plots for: {pretty_name} ---"))
  
  levels_vec <- grp_levels_for(current_pheno)
  
  # Create a new interaction variable for the x-axis
  pheno_data <- pheno_data %>%
    mutate(
      Grp2 = factor(Grp2, levels = levels_vec),
      x_group = interaction(Dataset, Grp2, sep = "\n")
    )
  
  cpgs_to_plot <- unique(pheno_data$CpG)

  for (current_cpg in cpgs_to_plot) {
    
    cpg_plot_data <- pheno_data %>% filter(CpG == current_cpg)
    msg(glue("  -> Plotting {current_cpg}"))

    # Define all four comparisons for the brackets
    comparison_list <- list(
      c(paste("DH", levels_vec[1], sep="\n"), paste("DH", levels_vec[2], sep="\n")),
      c(paste("TCGA", levels_vec[1], sep="\n"), paste("TCGA", levels_vec[2], sep="\n")),
      c(paste("DH", levels_vec[1], sep="\n"), paste("TCGA", levels_vec[1], sep="\n")),
      c(paste("DH", levels_vec[2], sep="\n"), paste("TCGA", levels_vec[2], sep="\n"))
    )

    p <- ggplot(cpg_plot_data, aes(x = x_group, y = Beta, fill = Dataset)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
      # Add brackets for all four comparisons
      stat_compare_means(
        comparisons = comparison_list, 
        method = "t.test", 
        label = "p.signif", 
        hide.ns = TRUE,
        symnum.args = list(cutpoints = c(0, 0.001, 0.01, 0.05, 1), symbols = c("***", "**", "*", "ns"))
      ) +
      labs(
        title = glue("{pretty_name}: {current_cpg}"),
        subtitle = "Comprehensive Comparison",
        x = NULL, y = "Beta (DNA methylation)", fill = "Dataset"
      ) +
      # Increase y-axis expansion to make space for stacked brackets
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.35))) +
      theme_classic(base_size = 12) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        plot.subtitle = element_text(size=9)
      )

    pheno_file_tag <- gsub(" ", "_", gsub("[^A-Za-z0-9 ]", "", tolower(pretty_name)))
    out_stub <- file.path(fig_dir, glue("SinglePlot_{current_cpg}_{pheno_file_tag}"))
    save_pdf_tiff(p, out_stub, width = 6, height = 5)
  }
}

msg("\n✅ All single comprehensive plots have been saved.")
```


Validation Single Cell
```{r}
options(future.globals.maxSize = 8000 * 1024^2)

###############################################################################
# atlas_DH_end_to_end_exports_v2.R
#  - EWAS top-N CpGs → genes
#  - Atlas label mapping (cl295v11SubFull) → detailtype/supertype (barcode-aligned)
#  - Subset to T cells / Fibroblasts / Tumor cells
#  - Export PDFs:
#       * UMAP_overview.pdf (supertype + detailtype)
#       * ModuleScore_DHsig1_UMAP.pdf
#       * FeaturePlots_paged_pXXX.pdf
#       * DotPlots_by_supertype_pXXX.pdf
#  - Save enriched atlas RDS
###############################################################################

## 0) CONFIG (edit paths as needed)
path_ewas_results   <- "/Users/adiallo/Desktop/Thesis/code"
h_level             <- "H6"
tier                <- "Tier3"
phenotypes          <- c("any","ln","dist")
top_n_cpgs_list     <- c(25, 50, 100)
max_n_total         <- max(top_n_cpgs_list)
n_half              <- floor(max_n_total/2)

atlas_rds           <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/Visium_samples/Visium_Data/atlas_data.rds"
out_dir             <- "results_TCGA_validation_heatmaps_V7"
overview_pdf        <- file.path(out_dir, "UMAP_overview.pdf")
module_pdf          <- file.path(out_dir, "ModuleScore_DHsig1_UMAP.pdf")
feature_pdf_prefix  <- file.path(out_dir, "FeaturePlots_paged")      # will create _p001.pdf, _p002.pdf, ...
dot_pdf_prefix      <- file.path(out_dir, "DotPlots_by_supertype")   # will create _p001.pdf, _p002.pdf, ...
save_atlas_rds      <- file.path(out_dir, "atlas_with_supertype.rds")

# FeaturePlot / DotPlot pagination controls
features_per_page   <- 24
downsample_n        <- 0                  # set >0 (e.g., 80000) to downsample cells for FeaturePlots only
raster_dpi_vec      <- c(200, 200)        # must be two-length numeric vector

## 1) LIBRARIES
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(dplyr)
  library(glue)
  library(ggplot2)
  library(patchwork)
  library(grid)
  library(limma)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
})

# Larger seed for reproducibility (affects AddModuleScore background bins, etc.)
set.seed(20250809)

msg <- function(...) cat("[", format(Sys.time(), "%H:%M:%S"), "]", ..., "\n")

# ---- helper: safely set default layer on Assay5 (no-op on others) ----------
set_default_layer_safe <- function(obj, assay = "RNA", layer = "data") {
  if (!assay %in% names(obj@assays)) return(obj)
  aa <- obj[[assay]]
  if (inherits(aa, "Assay5")) {
    # Only SeuratObject v5 Assay5 supports DefaultLayer<-
    try({
      SeuratObject::DefaultLayer(aa) <- layer
      obj[[assay]] <- aa
    }, silent = TRUE)
  }
  obj
}

## 2) CONSOLIDATED LABELS (exact matches to cl295v11SubFull)
consolidated_labels <- c(
  "cB2 (B GC-like)" = "B cells",
  "cE02 (Stem/TA-like/Immature Goblet)" = "Tumor cells",
  "cE04 (Enterocyte 1)" = "Tumor cells",
  "cE03 (Stem/TA-like prolif)" = "Tumor cells",
  "cM01 (Monocyte)" = "Monocytes",
  "cS28 (GREM1+ CAF)" = "Fibroblasts",
  "cM02 (Macrophage-like)" = "Macrophages",
  "cS06 (Endo lymphatic)" = "Endothelial cells",
  "cP2 (Plasma IgG)" = "Plasma cells",
  "cS29 (MMP3+ CAF)" = "Fibroblasts",
  "cMA01 (Mast)" = "Mast cells",
  "cP1 (Plasma IgA)" = "Plasma cells",
  "cS33 (Schwann)" = "Schwann cells",
  "cTNI08 (CD4+ Treg)" = "Treg",
  "cE06 (Immature Goblet)" = "Tumor cells",
  "cS04 (Endo)" = "Endothelial cells",
  "cS17 (Pericyte)" = "Pericytes",
  "cS30 (CAF CCL8 Fibro-like)" = "Fibroblasts",
  "cS31 (CAF stem niche Fibro-like)" = "Fibroblasts",
  "cTNI13 (CD8+ T IL17+)" = "CD8+ T-cells",
  "cTNI17 (gd-like T)" = "CD8+ T-cells",
  "cE01 (Stem/TA-like)" = "Tumor cells",
  "cS27 (CXCL14+ CAF)" = "Fibroblasts",
  "cTNI01 (CD4+ IL7R+)" = "CD4+ T-cells",
  "cE05 (Enterocyte 2)" = "Tumor cells",
  "cB3 (B CD40+ GC-like)" = "B cells",
  "cM07 (pDC)" = "Dendritic cells",
  "cB1 (B IGD+IgM+)" = "B cells",
  "cTNI04 (CD4+ IL7R+CCL5+)" = "CD4+ T-cells",
  "cS26 (Myofibro)" = "Fibroblasts",
  "cS13 (Endo venous-like)" = "Endothelial cells",
  "cTNI18 (gd-like T PDCD1+)" = "T cells",
  "cTNI21 (PLZF+ T prolif)" = "T cells",
  "cS08 (Endo arterial-like)" = "Endothelial cells",
  "cP3 (Plasma IgG prolif)" = "Plasma cells",
  "cM09 (mregDC)" = "Dendritic cells",
  "cTNI14 (CD8+ CXCL13+)" = "CD8+ T-cells",
  "cTNI26 (ILC3)" = "Innate lymphoid cells",
  "cM10 (Granulocyte)" = "Granulocytes",
  "cM04 (DC2)" = "Dendritic cells",
  "cTNI05 (CD4+ IL17+)" = "CD4+ T-cells",
  "cTNI03 (CD4+ IL7R+HSP+)" = "CD4+ T-cells",
  "cS05 (Endo venous)" = "Endothelial cells",
  "cS19 (Pericyte)" = "Pericytes",
  "cTNI02 (CD4+ IL7R+SELL+)" = "CD4+ T-cells",
  "cTNI20 (PLZF+ T)" = "T cells",
  "cTNI10 (CD8+ IL7R+)" = "CD8+ T-cells",
  "cS09 (Endo)" = "Endothelial cells",
  "cTNI09 (CD4+ Treg prolif)" = "Treg",
  "cS16 (Pericyte)" = "Pericytes",
  "cTNI11 (CD8+GZMK+)" = "CD8+ T-cells",
  "cS12 (Endo)" = "Endothelial cells",
  "cE08 (Goblet)" = "Goblet cells",
  "cTNI06 (CD4+ TFH)" = "TFH cells",
  "cS32 (Smooth Muscle)" = "Smooth muscle cells",
  "cS14 (Endo)" = "Endothelial cells",
  "cS25 (Fibro CCL8+)" = "Fibroblasts",
  "cS22 (Fibro stem cell niche)" = "Fibroblasts",
  "cS10 (Endo tip cells)" = "Endothelial cells",
  "cE11 (Enteroendocrine)" = "Enteroendocrine cells",
  "cM05 (DC2 C1Q+)" = "Dendritic cells",
  "cTNI15 (CD8+ CXCL13+ HSP+)" = "CD8+ T-cells",
  "cTNI07 (CD4+ CXCL13+)" = "CD4+ T-cells",
  "cTNI16 (CD8+ CXCL13+ prolif)" = "CD8+ T-cells",
  "cTNI22 (cTNI22)" = "T cells",
  "cE10 (Tuft)" = "Tuft cells",
  "cS15 (Pericyte)" = "Pericytes",
  "cTNI12 (CD8+ IL7R+)" = "CD8+ T-cells",
  "cS20 (Pericyte prolif)" = "Pericytes",
  "cS18 (Pericyte)" = "Pericytes",
  "cS11 (Endo proif)" = "Endothelial cells",
  "cM06 (DC IL22RA2)" = "Dendritic cells",
  "cS02 (Endo capillary)" = "Endothelial cells",
  "cS01 (Endo arterial)" = "Endothelial cells",
  "cE09 (Best4)" = "Goblet cells",
  "cS03 (Endo capillary)" = "Endothelial cells",
  "cS07 (Endo capillary-like)" = "Endothelial cells",
  "cS24 (Fibro BMP-producing)" = "Fibroblasts",
  "cM08 (AS-DC)" = "Dendritic cells",
  "cTNI23 (NK CD16A+)" = "NK cells",
  "cS21 (Fibro stem cell niche)" = "Fibroblasts",
  "cTNI24 (NK GZMK+)" = "NK cells",
  "cM03 (DC1)" = "Dendritic cells",
  "cE07 (Goblet/Enterocyte)" = "Goblet cells",
  "cTNI25 (NK XCL1+)" = "NK cells",
  "cTNI19 (gd-like T prolif)" = "T cells",
  "cS23 (Fibro BMP-producing)" = "Fibroblasts"
)

## 3) EWAS → top-N DH probes (balanced hyper/hypo) → genes
msg(glue("Extracting top-{max_n_total} DH CpGs across phenotypes…"))
dh_probes <- character()
for (pheno in phenotypes) {
  fit_file <- file.path(path_ewas_results, paste0("results_ewas_", pheno),
                        h_level, "ewas_fits.rds")
  if (!file.exists(fit_file)) { msg("Skipping: no fits at ", fit_file); next }
  fits <- readRDS(fit_file)
  fo   <- fits[[tier]]
  fit  <- if (inherits(fo, "MArrayLM")) fo else fo$fit
  tt   <- limma::topTable(fit, coef = "pheno_bin", number = Inf, sort.by = "P") %>% as.data.frame()
  hyper <- rownames(utils::head(tt[tt$logFC >  0, ], n_half))
  hypo  <- rownames(utils::head(tt[tt$logFC <  0, ], max_n_total - n_half))
  dh_probes <- union(dh_probes, c(hyper, hypo))
}
msg("Total unique DH CpGs:", length(dh_probes))

msg("Mapping DH probes to genes…")
anno_df <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
common_probes <- intersect(dh_probes, rownames(anno_df))
refgenes_chr  <- as.character(anno_df[common_probes, "UCSC_RefGene_Name"])
genes_first   <- sapply(strsplit(refgenes_chr, ";"), `[`, 1)
dh_sig_genes  <- sort(unique(genes_first[nzchar(genes_first) & !is.na(genes_first)]))
msg("Total DH-significant genes:", length(dh_sig_genes))

## 4) LOAD ATLAS & MAP LABELS → detail/super (barcode-aligned)
msg("Loading atlas…")
atlas <- readRDS(atlas_rds)
label_col <- "cl295v11SubFull"
stopifnot(label_col %in% colnames(atlas@meta.data))

full_labels <- as.character(atlas@meta.data[[label_col]])
short_codes <- sub(" .*", "", full_labels)

mapped_detail <- ifelse(full_labels %in% names(consolidated_labels),
                        consolidated_labels[full_labels],
                        NA_character_)

# Fallbacks for broad buckets if not in dictionary
fallback_detail <- ifelse(is.na(mapped_detail) & grepl("(CAF|Fibro)", full_labels, ignore.case = TRUE), "Fibroblasts",
                   ifelse(is.na(mapped_detail) & grepl("^cTNI", short_codes), "T cells",
                   ifelse(is.na(mapped_detail) & grepl("^cE",   short_codes), "Tumor cells",
                   NA_character_)))

detailtype <- ifelse(is.na(mapped_detail), fallback_detail, mapped_detail)
detailtype[is.na(detailtype)] <- full_labels  # last resort: keep original label

collapse_to_super <- function(x) {
  if (x %in% c("Fibroblasts")) return("Fibroblasts")
  if (x %in% c("Tumor cells")) return("Tumor cells")
  if (x %in% c("CD4+ T-cells","CD8+ T-cells","Treg","TFH cells","T cells",
               "NK cells","Innate lymphoid cells")) return("T cells")
  return("Other")
}
supertype <- vapply(detailtype, collapse_to_super, character(1))

stopifnot(length(detailtype) == ncol(atlas), length(supertype) == ncol(atlas))
names(detailtype) <- colnames(atlas)
names(supertype)  <- colnames(atlas)

atlas <- Seurat::AddMetaData(
  atlas,
  metadata = data.frame(detailtype = as.character(detailtype),
                        supertype  = as.character(supertype),
                        row.names  = colnames(atlas))
)

msg("Superclass counts:\n")
print(table(atlas$supertype))

## 5) SUBSET TO TARGET SUPERTYPES; PREP GENES
keep_types <- c("T cells","Fibroblasts","Tumor cells")
atlas_sub  <- subset(atlas, subset = supertype %in% keep_types)
msg("Subset cell count:", ncol(atlas_sub))

DefaultAssay(atlas_sub) <- "RNA"
atlas_sub <- set_default_layer_safe(atlas_sub, assay = "RNA", layer = "data")

# FIX 1: intersect genes with RNA assay features
dh_sig_genes_in <- intersect(dh_sig_genes, rownames(atlas_sub[["RNA"]]))
msg("Genes found in atlas:", length(dh_sig_genes_in))

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

## 6) UMAP OVERVIEW (supertype + detailtype)
msg("Saving overview PDF → ", overview_pdf)
pdf(overview_pdf, width = 12, height = 6)
print(
  DimPlot(atlas_sub, reduction = "umap", group.by = "supertype",
          pt.size = 0.25, raster = TRUE) +
    ggtitle("C295 Atlas UMAP — T cells / Fibroblasts / Tumor cells") +
    theme(plot.title = element_text(face = "bold"))
)
print(
  DimPlot(atlas_sub, reduction = "umap", group.by = "detailtype",
          pt.size = 0.2, raster = TRUE) +
    ggtitle("Detailed types (dictionary + fallbacks)") +
    theme(plot.title = element_text(face = "bold"))
)
dev.off()

###############################################################################
## 6A) Build gene-level direction (hyper / hypo) from limma logFC
##     (same fits you already generated; aggregates to gene level)
###############################################################################

stopifnot(exists("anno_df"), exists("dh_probes"))

# Map probes → first gene (same rule as above)
probe_genes <- {
  common_probes <- intersect(dh_probes, rownames(anno_df))
  refgenes_chr  <- as.character(anno_df[common_probes, "UCSC_RefGene_Name"])
  genes_first   <- sapply(strsplit(refgenes_chr, ";"), `[`, 1)
  tibble::tibble(CpG = common_probes, Gene = genes_first) |>
    dplyr::filter(!is.na(Gene), Gene != "")
}

# Collect per-probe logFC across phenotypes
logfc_rows <- list()
for (pheno in phenotypes) {
  fit_file <- file.path(path_ewas_results, paste0("results_ewas_", pheno),
                        h_level, "ewas_fits.rds")
  if (!file.exists(fit_file)) next
  fits <- readRDS(fit_file)
  fo   <- fits[[tier]]
  fit  <- if (inherits(fo, "MArrayLM")) fo else fo$fit
  tt   <- limma::topTable(fit, coef = "pheno_bin", number = Inf, sort.by = "P") |>
          tibble::rownames_to_column("CpG") |>
          dplyr::select(CpG, logFC)
  tt_dh <- dplyr::semi_join(tt, tibble::tibble(CpG = dh_probes), by = "CpG") |>
           dplyr::left_join(probe_genes, by = "CpG") |>
           dplyr::filter(!is.na(Gene) & Gene != "")
  tt_dh$phenotype <- pheno
  logfc_rows[[pheno]] <- tt_dh
}
logfc_df <- dplyr::bind_rows(logfc_rows)

# Aggregate to gene-level mean logFC across probes and phenotypes
gene_delta <- logfc_df |>
  dplyr::group_by(Gene) |>
  dplyr::summarise(mean_logFC = mean(logFC, na.rm = TRUE),
                   n_probes   = dplyr::n()) |>
  dplyr::ungroup()

# Intersect with RNA assay features for stability
genes_in_atlas <- intersect(gene_delta$Gene, rownames(atlas[["RNA"]]))
gene_delta <- dplyr::filter(gene_delta, Gene %in% genes_in_atlas)

# Split into hypo / hyper lists
hypo_genes  <- gene_delta$Gene[gene_delta$mean_logFC < 0]
hyper_genes <- gene_delta$Gene[gene_delta$mean_logFC > 0]

msg(glue("Gene-level direction: {length(hypo_genes)} hypo, {length(hyper_genes)} hyper (",
         "{length(setdiff(gene_delta$Gene, union(hypo_genes, hyper_genes)))} zero)."))

###############################################################################
## 6B) Score ALL cells + one PDF with: context, net (hyper−hypo), hypo, hyper
###############################################################################

stopifnot("supertype" %in% colnames(atlas@meta.data))
DefaultAssay(atlas) <- "RNA"
atlas <- set_default_layer_safe(atlas, assay = "RNA", layer = "data")

# Safety for empty lists
if (length(hypo_genes)  == 0) hypo_genes  <- character()
if (length(hyper_genes) == 0) hyper_genes <- character()

# Compute module scores (Seurat always appends "1")
if (length(hypo_genes) > 0) {
  atlas <- AddModuleScore(atlas, features = list(hypo_genes),
                          name = "DHhypo", nbin = 24, ctrl = 100, assay = "RNA")
}
if (length(hyper_genes) > 0) {
  atlas <- AddModuleScore(atlas, features = list(hyper_genes),
                          name = "DHhyper", nbin = 24, ctrl = 100, assay = "RNA")
}

hypo_col  <- if (any(grepl("^DHhypo", colnames(atlas@meta.data))))
               grep("^DHhypo",  colnames(atlas@meta.data), value = TRUE)[1] else NA_character_
hyper_col <- if (any(grepl("^DHhyper", colnames(atlas@meta.data))))
               grep("^DHhyper", colnames(atlas@meta.data), value = TRUE)[1] else NA_character_

# Net score = Hyper - Hypo (NA-safe)
atlas$net_score <- NA_real_
if (!is.na(hyper_col) && !is.na(hypo_col)) {
  atlas$net_score <- atlas@meta.data[[hyper_col]] - atlas@meta.data[[hypo_col]]
} else if (!is.na(hyper_col)) {
  atlas$net_score <- atlas@meta.data[[hyper_col]]
} else if (!is.na(hypo_col)) {
  atlas$net_score <- -atlas@meta.data[[hypo_col]]
}

# Output file (single, combined)
allcells_combo_pdf <- file.path(out_dir, "UMAP_allcells_net_hypo_hyper.pdf")

# Color palettes (colorblind-friendly)
pal_div <- c("#2c7bb6", "#ffffbf", "#d7191c")                # blue - white - red (net)
pal_byr  <- c("#2c7bb6", "#fee08b", "#d73027")   # blue → yellow → red
pal_hypo <- c("#f7fbff", "#6baed6", "#08306b")   # keep if you still want hypo-only
pal_hyper<- c("#fff5f0", "#fb6a4a", "#67000d")   # keep if you still want hyper-only

msg("Saving ALL-cells combined UMAP (context + net + hypo + hyper) → ", allcells_combo_pdf)
pdf(allcells_combo_pdf, width = 14, height = 8)

# A) Context: all cells by supertype
p_ctx <- DimPlot(atlas, reduction = "umap", group.by = "supertype",
                 pt.size = 0.1, raster = TRUE) +
  ggtitle("All cell types — supertype") +
  theme(plot.title = element_text(face = "bold"))

# B) Net (hyper − hypo) with divergent palette
p_net <- FeaturePlot(atlas, features = "net_score", reduction = "umap",
                     order = TRUE, pt.size = 0.12, raster = TRUE,
                     raster.dpi = raster_dpi_vec) +
  scale_color_gradientn(colors = pal_byr, name = "Net (Hyper − Hypo)") +
  ggtitle("EWAS gene direction — NET score") +
  theme(plot.title = element_text(face = "bold"))

# C) Hypo-only
p_hypo <- if (!is.na(hypo_col)) {
  FeaturePlot(atlas, features = hypo_col, reduction = "umap",
              order = TRUE, pt.size = 0.12, raster = TRUE,
              raster.dpi = raster_dpi_vec) +
    scale_color_gradientn(colors = pal_hypo, name = "Hypo score") +
    ggtitle("Hypo-methylated gene set (expression module)") +
    theme(plot.title = element_text(face = "bold"))
} else {
  ggplot() + ggtitle("No hypo gene set") + theme_void()
}

# D) Hyper-only
p_hyper <- if (!is.na(hyper_col)) {
  FeaturePlot(atlas, features = hyper_col, reduction = "umap",
              order = TRUE, pt.size = 0.12, raster = TRUE,
              raster.dpi = raster_dpi_vec) +
    scale_color_gradientn(colors = pal_hyper, name = "Hyper score") +
    ggtitle("Hyper-methylated gene set (expression module)") +
    theme(plot.title = element_text(face = "bold"))
} else {
  ggplot() + ggtitle("No hyper gene set") + theme_void()
}

print((p_ctx | p_net) / (p_hypo | p_hyper))
dev.off()

###############################################################################
# Bivariate UMAP — raster points, 100k cells, no title, separate legend PDF
###############################################################################

suppressPackageStartupMessages({
  library(ggrastr)
  library(grid)
  library(scales)
})

# ---- User settings ----
plot_n_cells <- 100000      # number of cells to plot
raster_dpi   <- 150         # point raster resolution
pt_size      <- 0.2
seed_plot    <- 20250809
clip_lo_q    <- 0.05        # quantile clipping for color scale low
clip_hi_q    <- 0.95        # quantile clipping for color scale high
out_pdf_bi   <- file.path(out_dir, "UMAP_allcells_bivariate_100k.pdf")
out_pdf_leg  <- file.path(out_dir, "UMAP_bivariate_legend.pdf")
# -----------------------

# Get score columns
hypo_col  <- grep("^DHhypo",  colnames(atlas@meta.data), value = TRUE)[1]
hyper_col <- grep("^DHhyper", colnames(atlas@meta.data), value = TRUE)[1]
stopifnot(!is.na(hypo_col), !is.na(hyper_col))

# Build plotting dataframe
emb <- as.data.frame(atlas@reductions$umap@cell.embeddings)[, 1:2, drop = FALSE]
colnames(emb) <- c("UMAP_1","UMAP_2")
emb$hypo  <- atlas@meta.data[[hypo_col]]
emb$hyper <- atlas@meta.data[[hyper_col]]

# Downsample
set.seed(seed_plot)
if (nrow(emb) > plot_n_cells) {
  emb <- emb[sample.int(nrow(emb), plot_n_cells), ]
}

# Clip and rescale
rescale_clip <- function(x, lo_q, hi_q) {
  lo <- quantile(x, probs = lo_q, na.rm = TRUE)
  hi <- quantile(x, probs = hi_q, na.rm = TRUE)
  scales::rescale(pmin(pmax(x, lo), hi), to = c(0, 1), from = c(lo, hi))
}
emb$r <- rescale_clip(emb$hyper, clip_lo_q, clip_hi_q)  # red
emb$b <- rescale_clip(emb$hypo,  clip_lo_q, clip_hi_q)  # blue
emb$g <- 0
emb$col_hex <- rgb(emb$r, emb$g, emb$b)

# Legend matrix (for separate export)
legend_res <- 200
leg_mat <- array(0, dim = c(legend_res, legend_res, 3))
red_vals  <- matrix(rep(seq(0, 1, length.out = legend_res), each = legend_res),
                    nrow = legend_res, byrow = FALSE)
blue_vals <- matrix(rep(seq(0, 1, length.out = legend_res), times = legend_res),
                    nrow = legend_res, byrow = TRUE)
leg_mat[,,1] <- red_vals
leg_mat[,,3] <- blue_vals
leg_grob <- grid::rasterGrob(leg_mat, interpolate = TRUE)

# --- Main plot ---
p_bi <- ggplot(emb, aes(x = UMAP_1, y = UMAP_2)) +
  ggrastr::geom_point_rast(aes(color = col_hex),
                           size = pt_size, raster.dpi = raster_dpi, shape = 16) +
  scale_color_identity() +
  coord_equal() +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")  # no legend, no title

# Save main plot PDF
pdf(out_pdf_bi, width = 9, height = 7)
print(p_bi)
dev.off()

# --- Legend export ---
pdf(out_pdf_leg, width = 3, height = 3)
grid::grid.newpage()
grid::grid.draw(leg_grob)
grid::grid.text("Hyper ↑ (red)  →", x = unit(0.5, "npc"), y = unit(0.0, "npc"),
                gp = grid::gpar(cex = 0.8))
grid::grid.text("Hypo ↑ (blue) ↑", x = unit(0.0, "npc"), y = unit(0.5, "npc"),
                rot = 90, gp = grid::gpar(cex = 0.8))
dev.off()

message("✅ Saved main plot: ", out_pdf_bi)
message("✅ Saved legend: ", out_pdf_leg)

## 10) SAVE ENRICHED ATLAS
saveRDS(atlas, file = save_atlas_rds)
msg("✅ Done: overview, module score, FeaturePlots (paged), and DotPlots (paged) exported.")
```

```{r}
###############################################################################
# UMAP figures with readable palettes (blue → yellow → red for NET)
# - Expects: atlas (Seurat), out_dir (string), raster_dpi_vec (num[2]),
#            hypo_col, hyper_col (column names in atlas@meta.data),
#            and atlas$net_score already computed.
###############################################################################

suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(RColorBrewer)
  library(scales)
  library(ggrastr)
  library(grid)
})

# ---------- Sanity checks ----------
stopifnot(inherits(atlas, "Seurat"))
stopifnot("net_score" %in% colnames(atlas@meta.data))
stopifnot(dir.exists(out_dir))

# If these weren’t set earlier, define a conservative default:
if (!exists("raster_dpi_vec")) raster_dpi_vec <- c(200, 200)

# Try to locate score columns if not provided
if (!exists("hypo_col") || is.na(hypo_col)) {
  hypo_col  <- grep("^DHhypo",  colnames(atlas@meta.data), value = TRUE)[1]
}
if (!exists("hyper_col") || is.na(hyper_col)) {
  hyper_col <- grep("^DHhyper", colnames(atlas@meta.data), value = TRUE)[1]
}

# ---------- Palettes ----------
# Blue → Yellow → Red for NET (reverse of RdYlBu)
pal_byr      <- colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(255)
pal_blue_seq <- colorRampPalette(brewer.pal(9, "Blues"))(255)  # hypo
pal_red_seq  <- colorRampPalette(brewer.pal(9, "Reds"))(255)   # hyper

# Clip extremes (2–98%) for readable contrast
clip_limits <- function(x, lo = 0.02, hi = 0.98) {
  c(quantile(x, lo, na.rm = TRUE), quantile(x, hi, na.rm = TRUE))
}

# ---------- Panels ----------
# A) Context by supertype
p_ctx <- DimPlot(atlas, reduction = "umap", group.by = "supertype",
                 pt.size = 0.10, raster = TRUE) +
  ggtitle("All cell types — supertype") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# B) NET (hyper − hypo) with blue→yellow→red
net_limits <- clip_limits(atlas$net_score)
p_net <- FeaturePlot(atlas, features = "net_score", reduction = "umap",
                     order = TRUE, pt.size = 0.12, raster = TRUE,
                     raster.dpi = raster_dpi_vec) +
  scale_color_gradientn(colors = pal_byr,
                        limits = net_limits, oob = squish,
                        name = "Net (Hyper - Hypo)") +
  ggtitle("EWAS gene direction — NET score") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# C) Hypo-only
if (!is.na(hypo_col)) {
  hypo_vals   <- atlas@meta.data[[hypo_col]]
  hypo_limits <- clip_limits(hypo_vals)
  p_hypo <- FeaturePlot(atlas, features = hypo_col, reduction = "umap",
                        order = TRUE, pt.size = 0.12, raster = TRUE,
                        raster.dpi = raster_dpi_vec) +
    scale_color_gradientn(colors = pal_blue_seq,
                          limits = hypo_limits, oob = squish,
                          name = "Hypo score") +
    ggtitle("Hypo-methylated gene set") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))
} else {
  p_hypo <- ggplot() + ggtitle("No hypo gene set") + theme_void()
}

# D) Hyper-only
if (!is.na(hyper_col)) {
  hyper_vals   <- atlas@meta.data[[hyper_col]]
  hyper_limits <- clip_limits(hyper_vals)
  p_hyper <- FeaturePlot(atlas, features = hyper_col, reduction = "umap",
                         order = TRUE, pt.size = 0.12, raster = TRUE,
                         raster.dpi = raster_dpi_vec) +
    scale_color_gradientn(colors = pal_red_seq,
                          limits = hyper_limits, oob = squish,
                          name = "Hyper score") +
    ggtitle("Hyper-methylated gene set") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))
} else {
  p_hyper <- ggplot() + ggtitle("No hyper gene set") + theme_void()
}

# ---------- Save 4-panel figure ----------
allcells_combo_pdf <- file.path(out_dir, "UMAP_allcells_net_hypo_hyper.pdf")
pdf(allcells_combo_pdf, width = 14, height = 8)
print((p_ctx | p_net) / (p_hypo | p_hyper))
dev.off()
message("Saved: ", allcells_combo_pdf)

# ---------- NET-only export (clean, paper-ready) ----------
out_pdf_net_only <- file.path(out_dir, "UMAP_allcells_NET_only.pdf")
pdf(out_pdf_net_only, width = 9, height = 7, family = "Helvetica")
print(p_net + theme(legend.position = "right"))
dev.off()
message("Saved: ", out_pdf_net_only)

# ---------- OPTIONAL: tuned RGB bivariate composite (kept for comparison) ---
use_bivariate <- FALSE  # set TRUE if you still want the RGB blend

if (use_bivariate && !is.na(hypo_col) && !is.na(hyper_col)) {
  plot_n_cells <- 100000
  raster_dpi   <- 200
  pt_size      <- 0.20
  seed_plot    <- 20250809
  clip_lo_q    <- 0.05
  clip_hi_q    <- 0.95
  out_pdf_bi   <- file.path(out_dir, "UMAP_allcells_bivariate_RGB_tuned.pdf")
  out_pdf_leg  <- file.path(out_dir, "UMAP_bivariate_RGB_legend.pdf")

  emb <- as.data.frame(atlas@reductions$umap@cell.embeddings)[, 1:2, drop = FALSE]
  colnames(emb) <- c("UMAP_1","UMAP_2")
  emb$hypo  <- atlas@meta.data[[hypo_col]]
  emb$hyper <- atlas@meta.data[[hyper_col]]

  set.seed(seed_plot)
  if (nrow(emb) > plot_n_cells) {
    emb <- emb[sample.int(nrow(emb), plot_n_cells), ]
  }

  rescale_clip <- function(x, lo_q, hi_q) {
    lo <- quantile(x, probs = lo_q, na.rm = TRUE)
    hi <- quantile(x, probs = hi_q, na.rm = TRUE)
    scales::rescale(pmin(pmax(x, lo), hi), to = c(0, 1), from = c(lo, hi))
  }

  emb$r <- rescale_clip(emb$hyper, clip_lo_q, clip_hi_q) ^ 0.9
  emb$b <- rescale_clip(emb$hypo,  clip_lo_q, clip_hi_q) ^ 0.9
  emb$g <- 0
  emb$col_hex <- rgb(emb$r, emb$g, emb$b)

  p_bi <- ggplot(emb, aes(UMAP_1, UMAP_2)) +
    ggrastr::geom_point_rast(aes(color = col_hex),
                             size = pt_size, raster.dpi = raster_dpi, shape = 16) +
    scale_color_identity() +
    coord_equal() +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none")

  pdf(out_pdf_bi, width = 9, height = 7, family = "Helvetica")
  print(p_bi)
  dev.off()
  message("Saved: ", out_pdf_bi)

  # ASCII legend (no UTF-8 arrows)
  legend_res <- 220
  leg_mat <- array(0, dim = c(legend_res, legend_res, 3))
  red_vals  <- matrix(rep(seq(0, 1, length.out = legend_res), each = legend_res),
                      nrow = legend_res, byrow = FALSE)
  blue_vals <- matrix(rep(seq(0, 1, length.out = legend_res), times = legend_res),
                      nrow = legend_res, byrow = TRUE)
  leg_mat[,,1] <- red_vals
  leg_mat[,,3] <- blue_vals
  leg_grob <- grid::rasterGrob(leg_mat, interpolate = TRUE)

  pdf(out_pdf_leg, width = 3.2, height = 3.2, family = "Helvetica")
  grid::grid.newpage()
  grid::grid.draw(leg_grob)
  grid::grid.text("Hyper up (red)  ->", x = unit(0.56, "npc"), y = unit(0.05, "npc"),
                  gp = grid::gpar(cex = 0.8))
  grid::grid.text("Hypo up (blue) ^", x = unit(0.05, "npc"), y = unit(0.56, "npc"),
                  rot = 90, gp = grid::gpar(cex = 0.8))
  dev.off()
  message("Saved: ", out_pdf_leg)
}
```


Expanding single gene analysis to include metastasis related genes

```{r}
###############################################################################
# HiTIMED h=6 + immune collapse/normalization + merge with PDCD1 Δβ/ΔM (scores)
# - Uses your cached β-matrix: Results/ewas_cache/beta_comb_qc.rds
# - Runs HiTIMED_deconvolution(..., tumor_type="COAD", h=6, tissue_type="tumor")
# - Collapses CD8nv+CD8mem → CD8_total; keeps Treg explicitly
# - Builds Immune_total from available immune components; normalizes CD8/Treg
# - Adds Fib via EpiDISH if missing in HiTIMED output
# - Merges with `scores` (must already exist) to yield `delta_tbl_h6`
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(tibble)
  library(HiTIMED); library(EpiDISH)
})

## ---- Paths ------------------------------------------------------------------
results_root <- "/Users/adiallo/Desktop/Thesis/Results"
beta_rds     <- file.path(results_root, "ewas_cache", "beta_comb_qc.rds")

## ---- Load β from cache ------------------------------------------------------
beta_obj  <- readRDS(beta_rds)
beta_comb <- if (is.list(beta_obj) && !is.null(beta_obj$beta)) beta_obj$beta else beta_obj
stopifnot(is.matrix(beta_comb))

## ---- Align to `scores` samples if `scores` exists ---------------------------
## Expect `scores` with: SampleID, Region (TSS/Body), BetaCollapsed, Mmean, mets_progression
if (!exists("scores")) stop("`scores` (PDCD1 TSS/Body collapsed table) must already exist in the session.")
keep_ids <- intersect(colnames(beta_comb), unique(scores$SampleID))
if (length(keep_ids) == 0) stop("No overlapping SampleID between beta_comb and scores.")
beta_comb <- beta_comb[, keep_ids, drop = FALSE]

## ---- Run HiTIMED h=6 (COAD, tumor tissue) ----------------------------------
frac_h6 <- HiTIMED_deconvolution(
  tumor_beta  = beta_comb,
  tumor_type  = "COAD",
  h           = 6,
  tissue_type = "tumor"
) %>%
  as.data.frame() %>%
  rownames_to_column("SampleID")

## ---- Add Fib via EpiDISH (if not provided by HiTIMED) ----------------------
if (!"Fib" %in% names(frac_h6)) {
  data("centEpiFibIC.m", package = "EpiDISH", envir = environment())
  common <- intersect(rownames(beta_comb), rownames(centEpiFibIC.m))
  fib_est <- epidish(beta_comb[common, , drop = FALSE],
                     centEpiFibIC.m[common, , drop = FALSE],
                     method = "RPC")$estF %>%
    as.data.frame() %>% rownames_to_column("SampleID") %>% select(SampleID, Fib)
  frac_h6 <- frac_h6 %>% left_join(fib_est, by = "SampleID")
}

## ---- Helpers ----------------------------------------------------------------
present  <- function(df, candidates) intersect(candidates, names(df))
sum_cols <- function(df, cols) if (length(cols)) rowSums(df[cols], na.rm = TRUE) else rep(0, nrow(df))

## ---- Identify immune components at h=6 --------------------------------------
# Common h=6 immune components (adjusted defensively)
cd8_parts <- present(frac_h6, c("CD8nv","CD8mem"))
cd4_parts <- present(frac_h6, c("CD4nv","CD4mem"))  # Treg handled separately
b_parts   <- present(frac_h6, c("Bnv","Bmem","Bcell","B"))
innate    <- present(frac_h6, c("NK","Mono","Monocyte","DC","Neu","Neutrophil","Eos","Eosinophil","Bas","Basophil"))
treg_col  <- present(frac_h6, c("Treg","Tregs","Regulatory_T"))

# Build totals
CD8_total <- if (length(cd8_parts)) sum_cols(frac_h6, cd8_parts) else {
  # graceful fallback if a single CD8 column exists
  cd8_fallback <- present(frac_h6, c("CD8T","CD8"))
  if (length(cd8_fallback)) frac_h6[[cd8_fallback[1]]] else rep(NA_real_, nrow(frac_h6))
}
CD4_total_noTreg <- if (length(cd4_parts)) sum_cols(frac_h6, cd4_parts) else rep(0, nrow(frac_h6))
Treg_vals        <- if (length(treg_col)) frac_h6[[treg_col[1]]] else rep(NA_real_, nrow(frac_h6))

# Immune_total = CD8_total + CD4_total_noTreg + Treg + B + innate
Immune_total <- CD8_total +
                CD4_total_noTreg +
                ifelse(is.na(Treg_vals), 0, Treg_vals) +
                sum_cols(frac_h6, b_parts) +
                sum_cols(frac_h6, innate)
Immune_total[!is.finite(Immune_total) | Immune_total <= 1e-6] <- NA_real_

## ---- StromalOther (optional covariate) --------------------------------------
stromal_cols <- present(frac_h6, c("Endothelial","Epithelial"))
StromalOther <- if (length(stromal_cols)) rowSums(frac_h6[stromal_cols], na.rm = TRUE) else rep(NA_real_, nrow(frac_h6))

## ---- Normalized fractions ---------------------------------------------------
frac_h6_norm <- frac_h6 %>%
  mutate(
    CD8_total     = CD8_total,
    CD4_total     = CD4_total_noTreg + ifelse(is.na(Treg_vals), 0, Treg_vals),
    Treg          = Treg_vals,
    Immune_total  = Immune_total,
    CD8_total_norm = CD8_total / Immune_total,
    Treg_norm      = Treg      / Immune_total,
    StromalOther   = StromalOther
  ) %>%
  select(SampleID, CD8_total, Treg, Immune_total,
         CD8_total_norm, Treg_norm, StromalOther, Fib, everything())

## ---- Build Δβ / ΔM from `scores` and merge ---------------------------------
delta_tbl_h6 <- scores %>%
  select(SampleID, Region, BetaCollapsed, Mmean, mets_progression) %>%
  pivot_wider(names_from = Region, values_from = c(BetaCollapsed, Mmean)) %>%
  mutate(
    Delta_Beta = BetaCollapsed_Body - BetaCollapsed_TSS,
    Delta_M    = Mmean_Body        - Mmean_TSS
  ) %>%
  left_join(frac_h6_norm, by = "SampleID")

## ---- Sanity outputs ---------------------------------------------------------
cat("\n[HiTIMED h=6] Detected immune parts:\n",
    "  CD8 parts: ", paste(if(length(cd8_parts)) cd8_parts else "(none; used fallback if available)", collapse="+"), "\n",
    "  CD4 parts (no Treg): ", paste(if(length(cd4_parts)) cd4_parts else "(none)", collapse="+"), "\n",
    "  Treg col: ", if(length(treg_col)) treg_col[1] else "(none)", "\n",
    "  B parts: ", paste(if(length(b_parts)) b_parts else "(none)", collapse="+"), "\n",
    "  Innate: ", paste(if(length(innate)) innate else "(none)", collapse=", "), "\n\n", sep = "")

print(delta_tbl_h6 %>%
        select(SampleID, mets_progression,
               Delta_Beta, Delta_M,
               CD8_total_norm, Treg_norm, Fib, StromalOther) %>%
        head(12))

# Result: delta_tbl_h6 (merged table ready for correlations/plots/models)

###############################################################################
# PDCD1 Δβ (Body – TSS) vs CD8 (HiTIMED h=6) — clean plot, no grid, no color by mets
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(rlang)
})

# ---- Helper: find the h=6 fractions table -----------------------------------
get_frac_h6 <- function() {
  if (exists("fractions_h6", inherits = TRUE)) return(get("fractions_h6", inherits = TRUE))
  if (exists("frac_h6",     inherits = TRUE)) return(get("frac_h6",     inherits = TRUE))
  if (exists("frac_6",      inherits = TRUE)) return(get("frac_6",      inherits = TRUE))
  if (exists("HIT_DH",      inherits = TRUE)) {
    HIT <- get("HIT_DH", inherits = TRUE)
    if (is.list(HIT) && !is.null(HIT[["h6"]])) return(HIT[["h6"]])
  }
  stop("Could not find fractions table. Expected one of: fractions_h6, frac_h6, frac_6, or HIT_DH[['h6']].")
}

frac_raw <- get_frac_h6()

if (!"SampleID" %in% names(frac_raw)) {
  if (!is.null(rownames(frac_raw)) && all(nzchar(rownames(frac_raw)))) {
    frac_raw <- tibble::rownames_to_column(as.data.frame(frac_raw), "SampleID")
  } else {
    stop("Fractions table has no 'SampleID' column and no usable rownames.")
  }
}

# ---- Normalize immune features ----------------------------------------------
immune_candidates <- c(
  "Bas","Eos","Neu","Mono","DC","NK",
  "Bnv","Bmem","CD4nv","CD4mem","Treg",
  "CD8nv","CD8mem"
)
present_immune <- intersect(immune_candidates, names(frac_raw))

frac_h6_norm <- frac_raw %>%
  mutate(
    CD8_total     = CD8nv + CD8mem,
    Immune_total  = rowSums(across(all_of(present_immune)), na.rm = TRUE),
    CD8_total_norm = if_else(Immune_total > 0, CD8_total / Immune_total, NA_real_),
    Treg_norm      = if ("Treg" %in% names(.)) if_else(Immune_total > 0, Treg / Immune_total, NA_real_) else NA_real_,
    StromalOther   = (if ("Endothelial" %in% names(.)) Endothelial else 0) +
                     (if ("Epithelial"  %in% names(.)) Epithelial  else 0)
  )

frac_keep <- frac_h6_norm %>%
  select(SampleID, CD8_total_norm, Treg_norm,
         Fib = dplyr::any_of("Fib"),
         StromalOther)

# ---- Build Δβ table ---------------------------------------------------------
scores_df <- get("scores", inherits = TRUE)

delta_tbl_h6 <- scores_df %>%
  select(SampleID, Region, BetaCollapsed, mets_progression) %>%
  pivot_wider(names_from = Region, values_from = BetaCollapsed) %>%
  mutate(Delta_Beta = (Body %||% NA_real_) - (TSS %||% NA_real_)) %>%
  left_join(frac_keep, by = "SampleID") %>%
  filter(!is.na(Delta_Beta), !is.na(CD8_total_norm))

# ---- Linear model -----------------------------------------------------------
covars <- c("CD8_total_norm",
            if ("mets_progression" %in% names(delta_tbl_h6)) "mets_progression",
            if ("Fib"             %in% names(delta_tbl_h6)) "Fib",
            if ("StromalOther"    %in% names(delta_tbl_h6)) "StromalOther")

form <- as.formula(paste("Delta_Beta ~", paste(covars, collapse = " + ")))
fit  <- lm(form, data = delta_tbl_h6)
print(summary(fit))

cd8_term <- summary(fit)$coefficients[rownames(summary(fit)$coefficients) %in% "CD8_total_norm", , drop = FALSE]
cd8_lab  <- if (nrow(cd8_term) == 1) {
  paste0("LM slope (CD8_norm) = ", signif(cd8_term[1, "Estimate"], 3),
         ", p = ", signif(cd8_term[1, "Pr(>|t|)"], 3))
} else {
  "LM fitted"
}

# ---- Plot Δβ vs CD8 (no color by mets, no grid) ------------------------------
p <- ggplot(delta_tbl_h6, aes(x = CD8_total_norm, y = Delta_Beta)) +
  geom_point(size = 2, alpha = 0.85, color = "#1f78b4") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(x = "CD8 fraction (HiTIMED h=6, normalized by total immune)",
       y = expression(Delta*beta~"(Body – TSS)"),
       title = "PDCD1: Δβ vs CD8 infiltration",
       subtitle = cd8_lab) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# ---- Save -------------------------------------------------------------------
out_dir <- "/Users/adiallo/Desktop/Thesis/Results/PDCD1_analysis"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(file.path(out_dir, "PDCD1_DeltaBeta_vs_CD8norm_h6_LM_clean.pdf"),
       p, width = 6.2, height = 5.0)
```

```{r}
###############################################################################
# gene_region_beta_plots_clean.R
# Purpose: Clean per-gene plots for Illustrator
# - β-values (TSS vs Body) across metastasis groups
# - No titles, no stats labels, no facet strip backgrounds, no grid lines
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(glue)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  library(ggplot2)
})

## ---- Inputs -----------------------------------------------------------------
results_root <- "/Users/adiallo/Desktop/Thesis/Results"
out_dir      <- file.path(results_root, "GeneRegionSummaries_clean")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

beta_rds     <- file.path(results_root, "ewas_cache", "beta_comb_qc.rds")
meta_csv     <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

genes_of_interest <- c(
  "PDCD1",
  "SLIT2", "SOCS1", "COL11A1", "TCF8", "INHBB",
  "WNT7A", "WNT6", "ROBO1", "LCP1", "EPHB1", "ADAMTS18"
)

## ---- Load β-matrix + meta ---------------------------------------------------
beta_comb <- {
  obj <- readRDS(beta_rds)
  if (is.list(obj) && !is.null(obj$beta)) obj$beta else obj
}
meta <- read_csv(meta_csv, show_col_types = FALSE) %>%
  mutate(
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    mets_progression = case_when(
      Distant_Mets == 1 ~ "Distant Mets",
      ln_only == 1      ~ "LN-only Mets",
      TRUE              ~ "No Mets"
    ),
    mets_progression = factor(mets_progression,
                              levels = c("No Mets","LN-only Mets","Distant Mets"))
  ) %>% select(SampleID, mets_progression)

common_ids <- intersect(colnames(beta_comb), meta$SampleID)
stopifnot(length(common_ids) > 0)
beta_comb <- beta_comb[, common_ids, drop = FALSE]
meta      <- filter(meta, SampleID %in% common_ids)

## ---- Annotation helper ------------------------------------------------------
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19) %>% as.data.frame()

get_gene_cpgs <- function(gene_symbol, beta_mat, anno_df){
  df <- anno_df %>%
    filter(UCSC_RefGene_Name == gene_symbol,
           UCSC_RefGene_Group %in% c("TSS200","TSS1500","Body")) %>%
    select(Name, UCSC_RefGene_Group) %>%
    filter(Name %in% rownames(beta_mat))
  list(
    tss  = df %>% filter(UCSC_RefGene_Group %in% c("TSS200","TSS1500")) %>% pull(Name),
    body = df %>% filter(UCSC_RefGene_Group == "Body") %>% pull(Name)
  )
}

## ---- Transform + collapse ---------------------------------------------------
to_m <- function(b) log2((b + 1e-6) / (1 - b + 1e-6))  # safe logit2

collapse_region <- function(cpgs, beta_mat, label){
  if (!length(cpgs)) return(NULL)
  B <- beta_mat[cpgs, , drop = FALSE]
  M <- t(apply(B, 1, to_m))
  tibble(
    SampleID      = colnames(B),
    Region        = label,
    BetaCollapsed = colMeans(B, na.rm = TRUE),
    Mmean         = colMeans(M, na.rm = TRUE)
  )
}

## ---- Core function: one gene -> clean plot ---------------------------------
analyze_gene_clean <- function(gene_symbol, beta_mat, meta_df, anno_df, out_dir){
  cpgs <- get_gene_cpgs(gene_symbol, beta_mat, anno_df)
  if (!length(cpgs$tss) && !length(cpgs$body)) {
    message("Skipping ", gene_symbol, ": no TSS or Body CpGs present in matrix.")
    return(NULL)
  }
  df_tss  <- collapse_region(cpgs$tss,  beta_mat, "TSS")
  df_body <- collapse_region(cpgs$body, beta_mat, "Body")
  scores <- bind_rows(df_tss, df_body) %>%
    left_join(meta_df, by = "SampleID") %>%
    filter(!is.na(mets_progression))

  plot_df <- scores %>% select(SampleID, Region, BetaCollapsed, mets_progression)

p <- ggplot(plot_df, aes(x = Region, y = BetaCollapsed, fill = Region)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.75, color = "gray30") +
  geom_jitter(aes(group = SampleID), width = 0.15, size = 0.8, alpha = 0.6) +
  facet_wrap(~ mets_progression) +
  { if (!is.null(lab_df)) geom_text(data = lab_df, aes(x = 1.5, y = y, label = label),
                                    size = 3.2, vjust = 0, inherit.aes = FALSE) } +
  scale_fill_manual(values = c("TSS" = "#1f78b4", "Body" = "#e31a1c")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Genomic Position (Region)",
    y = "β-value (mean across CpGs per region)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),                # remove grids
    panel.border = element_rect(color = "black", fill = NA), # keep box outline
    strip.background = element_blank(),          # remove grey facet boxes
    axis.line = element_line(color = "black"),   # keep axis lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )

  ggsave(file.path(out_dir, glue("{gene_symbol}_TSSvsBody_betaPlot_clean.pdf")),
         p, width = 6.5, height = 4.8, device = cairo_pdf)
}

## ---- Run over all genes -----------------------------------------------------
purrr::walk(genes_of_interest, ~ analyze_gene_clean(.x, beta_comb, meta, anno, out_dir))

cat("✓ Saved cleaned per-gene PDF plots to:\n  ", out_dir, "\n")

###############################################################################
# All genes combined plot, broken out by metastasis status, with p-value stars
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(ggpubr); library(purrr)
})

# Build one big data frame for all genes
all_plot_df <- purrr::map_dfr(genes_of_interest, function(gene_symbol) {
  cpgs <- get_gene_cpgs(gene_symbol, beta_comb, anno)
  df_tss  <- collapse_region(cpgs$tss,  beta_comb, "TSS")
  df_body <- collapse_region(cpgs$body, beta_comb, "Body")
  bind_rows(df_tss, df_body) %>%
    mutate(Gene = gene_symbol)
})

all_plot_df <- all_plot_df %>%
  left_join(meta, by = "SampleID") %>%
  filter(!is.na(mets_progression))

# Comparisons for TSS vs Body
comparisons <- list(c("TSS", "Body"))

# ---- Corrected Combined Plot: Faceted by Gene + Metastasis Group ----
p_all_genes_mets <- ggplot(all_plot_df,
                           aes(x = Region, y = BetaCollapsed, fill = Region)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.6, color = "gray30") +
  geom_jitter(width = 0.15, size = 0.7, alpha = 0.5) +
  stat_compare_means(comparisons = comparisons, method = "wilcox.test",
                     paired = TRUE, label = "p.signif",
                     hide.ns = TRUE, size = 5) +
  # Facet by both metastasis status and gene to ensure correct comparisons
  facet_wrap(~ mets_progression + Gene, nrow = 3) + 
  scale_fill_manual(values = c("TSS" = "#1f78b4", "Body" = "#e31a1c")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Genomic Region",
    y = "β-value (mean across CpGs per region)",
    fill = "Genomic Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    panel.border = element_rect(color = "gray80", fill = NA),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold", size = 8),
    legend.position = "none"
  )

ggsave(file.path(out_dir, "AllGenes_TSSvsBody_Faceted_withPstars.pdf"),
       p_all_genes_mets, width = 10, height = 8, device = cairo_pdf)
```

Testing if markers exist
```{r}
###############################################################################
# exhaustion_gene_check.R
# Purpose: Check which exhaustion genes have annotated CpGs in EPIC array
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(readr); library(stringr); library(tidyr)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
})

## ---- Inputs -----------------------------------------------------------------
results_root <- "/Users/adiallo/Desktop/Thesis/Results"
beta_rds     <- file.path(results_root, "ewas_cache", "beta_comb_qc.rds")

genes_of_interest <- c(
  "PDCD1",     # PD-1
  "HAVCR2",    # TIM-3
  "CTLA4",     # CTLA-4
  "TIGIT",     # TIGIT
  "LAG3",      # LAG-3
  "VSIR",      # VISTA
  "LAYN",      # LAYN
  "TOX", "TOX2", "TOX3", "TOX4",   # TOX family
  "CCT6A", "TUBA1B",
  "CXCL13",
  "TCF7"       # TCF-1
)

## ---- Load beta to get CpG universe -----------------------------------------
beta_comb <- {
  obj <- readRDS(beta_rds)
  if (is.list(obj) && !is.null(obj$beta)) obj$beta else obj
}
all_cpgs <- rownames(beta_comb)

## ---- Load + preprocess annotation ------------------------------------------
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19) %>%
  as.data.frame() %>%
  # Expand multi-gene entries
  separate_rows(UCSC_RefGene_Name, sep = ";") %>%
  # Expand multi-group entries (e.g., "TSS200;TSS200")
  separate_rows(UCSC_RefGene_Group, sep = ";")

## ---- Check each gene --------------------------------------------------------
check_results <- lapply(genes_of_interest, function(g) {
  df <- anno %>%
    filter(UCSC_RefGene_Name == g,
           UCSC_RefGene_Group %in% c("TSS200","TSS1500","Body"),
           Name %in% all_cpgs)

  tibble(
    Gene    = g,
    n_total = nrow(df),
    n_TSS   = sum(df$UCSC_RefGene_Group %in% c("TSS200","TSS1500")),
    n_Body  = sum(df$UCSC_RefGene_Group == "Body")
  )
}) %>% bind_rows()

print(check_results)

## ---- Save to CSV for reference ----------------------------------------------
out_csv <- file.path(results_root, "ExhaustionGenes_CpGcheck.csv")
write_csv(check_results, out_csv)
cat("✓ Saved check results to:", out_csv, "\n")
```

Looking for exhasution genes
```{r}
###############################################################################
# exhaustion_gene_region_plots.R
# Purpose: Generate clean per-gene plots for exhaustion markers
# - β-values (TSS vs Body) across metastasis groups
# - No titles, no stats labels, no facet strip backgrounds, no grid lines
# - Ensures correct matching of gene names in annotation
###############################################################################

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(glue)
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
  library(ggplot2); library(ggpubr); library(purrr)
})

## ---- Inputs -----------------------------------------------------------------
results_root <- "/Users/adiallo/Desktop/Thesis/Results"
out_dir      <- file.path(results_root, "ExhaustionGenePlots_clean")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

beta_rds     <- file.path(results_root, "ewas_cache", "beta_comb_qc.rds")
meta_csv     <- "/Users/adiallo/Desktop/Thesis/Data_Documents/data_all.csv"

# Exhaustion gene set
genes_of_interest <- c(
  "PDCD1", "HAVCR2", "CTLA4", "TIGIT", "LAG3", "VSIR",
  "LAYN", "TOX", "TOX2", "TOX3", "TOX4",
  "CXCL13", "TCF7",
  "CCT6A", "TUBA1B"
)

## ---- Load β-matrix + meta ---------------------------------------------------
beta_comb <- {
  obj <- readRDS(beta_rds)
  if (is.list(obj) && !is.null(obj$beta)) obj$beta else obj
}
meta <- read_csv(meta_csv, show_col_types = FALSE) %>%
  mutate(
    SampleID = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    mets_progression = case_when(
      Distant_Mets == 1 ~ "Distant Mets",
      ln_only == 1      ~ "LN-only Mets",
      TRUE              ~ "No Mets"
    ),
    mets_progression = factor(mets_progression,
                              levels = c("No Mets","LN-only Mets","Distant Mets"))
  ) %>% dplyr::select(SampleID, mets_progression)

common_ids <- intersect(colnames(beta_comb), meta$SampleID)
stopifnot(length(common_ids) > 0)
beta_comb <- beta_comb[, common_ids, drop = FALSE]
meta      <- filter(meta, SampleID %in% common_ids)

## ---- Annotation helper ------------------------------------------------------
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19) %>%
  as.data.frame() %>%
  separate_rows(UCSC_RefGene_Name, sep = ";") %>%
  separate_rows(UCSC_RefGene_Group, sep = ";")

get_gene_cpgs <- function(gene_symbol, beta_mat, anno_df){
  df <- anno_df %>%
    filter(UCSC_RefGene_Name == gene_symbol,
           UCSC_RefGene_Group %in% c("TSS200","TSS1500","Body")) %>%
    filter(Name %in% rownames(beta_mat))
  list(
    tss  = df %>% filter(UCSC_RefGene_Group %in% c("TSS200","TSS1500")) %>% pull(Name),
    body = df %>% filter(UCSC_RefGene_Group == "Body") %>% pull(Name)
  )
}

## ---- Transform + collapse ---------------------------------------------------
to_m <- function(b) log2((b + 1e-6) / (1 - b + 1e-6))  # safe logit2

collapse_region <- function(cpgs, beta_mat, label){
  if (!length(cpgs)) return(NULL)
  B <- beta_mat[cpgs, , drop = FALSE]
  M <- t(apply(B, 1, to_m))
  tibble(
    SampleID      = colnames(B),
    Region        = label,
    BetaCollapsed = colMeans(B, na.rm = TRUE),
    Mmean         = colMeans(M, na.rm = TRUE)
  )
}

## ---- Core function: one gene -> clean plot ---------------------------------
analyze_gene_clean <- function(gene_symbol, beta_mat, meta_df, anno_df, out_dir){
  cpgs <- get_gene_cpgs(gene_symbol, beta_mat, anno_df)
  if (!length(cpgs$tss) && !length(cpgs$body)) {
    message("Skipping ", gene_symbol, ": no TSS or Body CpGs present in matrix.")
    return(NULL)
  }
  df_tss  <- collapse_region(cpgs$tss,  beta_mat, "TSS")
  df_body <- collapse_region(cpgs$body, beta_mat, "Body")
  scores <- bind_rows(df_tss, df_body) %>%
    left_join(meta_df, by = "SampleID") %>%
    filter(!is.na(mets_progression))

  plot_df <- scores %>% dplyr::select(SampleID, Region, BetaCollapsed, mets_progression)

  p <- ggplot(plot_df, aes(x = Region, y = BetaCollapsed, fill = Region)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.75, color = "gray30") +
    geom_jitter(aes(group = SampleID), width = 0.15, size = 0.8, alpha = 0.6) +
    facet_wrap(~ mets_progression) +
    scale_fill_manual(values = c("TSS" = "#1f78b4", "Body" = "#e31a1c")) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      x = "Genomic Position (Region)",
      y = "β-value (mean across CpGs per region)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      panel.border = element_rect(color = "black", fill = NA),
      strip.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      legend.position = "none"
    )

  ggsave(file.path(out_dir, glue("{gene_symbol}_TSSvsBody_betaPlot_clean.pdf")),
         p, width = 6.5, height = 4.8, device = cairo_pdf)
}

## ---- Run over all genes -----------------------------------------------------
walk(genes_of_interest, ~ analyze_gene_clean(.x, beta_comb, meta, anno, out_dir))

cat("✓ Saved cleaned per-gene PDF plots to:\n  ", out_dir, "\n")

###############################################################################
# Combined exhaustion panel plot
###############################################################################

all_plot_df <- map_dfr(genes_of_interest, function(gene_symbol) {
  cpgs <- get_gene_cpgs(gene_symbol, beta_comb, anno)
  df_tss  <- collapse_region(cpgs$tss,  beta_comb, "TSS")
  df_body <- collapse_region(cpgs$body, beta_comb, "Body")
  bind_rows(df_tss, df_body) %>%
    mutate(Gene = gene_symbol)
})

all_plot_df <- all_plot_df %>%
  left_join(meta, by = "SampleID") %>%
  filter(!is.na(mets_progression))

comparisons <- list(c("TSS", "Body"))

p_all_genes_mets <- ggplot(all_plot_df,
                           aes(x = Region, y = BetaCollapsed, fill = Region)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.6, color = "gray30") +
  geom_jitter(width = 0.15, size = 0.7, alpha = 0.5) +
  stat_compare_means(comparisons = comparisons, method = "wilcox.test",
                     paired = TRUE, label = "p.signif",
                     hide.ns = TRUE, size = 5) +
  facet_wrap(~ mets_progression + Gene, nrow = 3) + 
  scale_fill_manual(values = c("TSS" = "#1f78b4", "Body" = "#e31a1c")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Genomic Region",
    y = "β-value (mean across CpGs per region)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    panel.border = element_rect(color = "gray80", fill = NA),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold", size = 8),
    legend.position = "none"
  )

ggsave(file.path(out_dir, "AllExhaustionGenes_TSSvsBody_Faceted_withPstars.pdf"),
       p_all_genes_mets, width = 12, height = 9, device = cairo_pdf)

cat("✓ Saved combined exhaustion gene panel plot to:\n  ", out_dir, "\n")
```

Differential expression analysis using MAST on the single cell data
```{r}
# ---
# SCRIPT: Harmony Integration and MAST DE Analysis (Mets vs. Non-Mets)
# ---
#
# PURPOSE:
# This script performs a full single-cell analysis workflow:
# 1. Loads and harmonizes Seurat data.
# 2. Merges patient-level covariates (age, sex, MLH1).
# 3. Runs Harmony to correct for patient-specific batch effects for visualization.
# 4. Saves a corrected UMAP plot to verify the integration.
# 5. Performs differential gene expression analysis using MAST, statistically
#    controlling for patient effects and other specified covariates.
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(harmony)
  library(MAST)
  library(data.table)
  library(dplyr)
  library(tibble)
  library(readr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
})

# --- 2. Define Inputs, Outputs, and Parameters ---
seurat_object_file <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/Single_cell/Single_cell_samples/Single Cell/combined_data.rds"
analysis_outputs <- "/Users/adiallo/Desktop/Thesis/Results"

tumor_atlas_labels <- c(
  "cE01 (Stem/TA-like)", "cE02 (Stem/TA-like/Immature Goblet)",
  "cE03 (Stem/TA-like prolif)", "cE04 (Enterocyte 1)", "cE05 (Enterocyte 2)",
  "cE06 (Immature Goblet)", "cE07 (Goblet/Enterocyte)", "cE08 (Goblet)",
  "cE09 (Best4)", "cE10 (Tuft)", "cE11 (Enteroendocrine)"
)

# --- 3. Load Data ---
message(paste("Loading Seurat object from:", seurat_object_file))
srt <- readRDS(file = seurat_object_file)
message("...Data loaded successfully.")

# --- 4. Create Patient Metadata Lookup Table ---
# This table is now filled with your provided data.
patient_info_df <- tibble::tribble(
  ~patient_id, ~age, ~sex_char, ~MLH1_num,
  "3",   76, "M", 1,
  "106", 93, "F", 0,
  "11",  61, "F", 1,
  "116", 64, "M", 0,
  "13",  90, "F", 1,
  "18",  52, "M", 1,
  "40",  71, "F", 0,
  "46",  81, "M", 1,
  "72",  80, "M", 0,
  "86",  59, "F", 1
) %>%
  mutate(sex_num = ifelse(sex_char == "M", 1, 0)) # Create numeric sex variable

# --- 5. Data Harmonization & Merging ---
message("[INFO] Harmonizing metadata columns and merging patient info...")
srt$atlas_label <- srt$predicted.id
srt$patient_id <- gsub("^patient_", "", srt$orig.ident)
srt$mets <- ifelse(tolower(srt$dataset_origin) == "metastasis", 1, 0)

# Merge the hard-coded patient info into the Seurat metadata
srt@meta.data <- srt@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  left_join(patient_info_df, by = "patient_id") %>%
  tibble::column_to_rownames("cell_id")
message("...Harmonization and merging complete.")


# --- 6. Harmony Integration & Visualization ---
message("[HARMONY] Starting integration workflow...")
srt <- NormalizeData(srt, verbose = FALSE)
srt <- FindVariableFeatures(srt, verbose = FALSE)
srt <- ScaleData(srt, verbose = FALSE)
srt <- RunPCA(srt, verbose = FALSE)
srt <- RunHarmony(srt, group.by.vars = "patient_id", reduction.save = "harmony", verbose = FALSE)
srt <- RunUMAP(srt, reduction = "harmony", dims = 1:30, verbose = FALSE)

# Create and save a diagnostic UMAP plot
message("[HARMONY] Creating and saving diagnostic UMAP plot...")
umap_coords <- as.data.frame(srt@reductions$umap@cell.embeddings)
x_col <- colnames(umap_coords)[1]
y_col <- colnames(umap_coords)[2]
plot_df <- cbind(umap_coords, srt@meta.data)
set.seed(123)
shuffled_df <- plot_df[sample(nrow(plot_df)), ]

p_corrected <- ggplot(shuffled_df, aes(x = .data[[x_col]], y = .data[[y_col]], color = patient_id)) +
  geom_point(size = 0.5, alpha = 0.8) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_classic() +
  ggtitle("Corrected UMAP (Shuffled, Colored by Patient)")

output_dir_plots <- file.path(analysis_outputs, "UMAP_Plots")
dir.create(output_dir_plots, showWarnings = FALSE, recursive = TRUE)
ggsave(file.path(output_dir_plots, "UMAP_Corrected_by_Patient.pdf"), p_corrected, width = 8, height = 6)
message(paste("...Corrected UMAP saved to:", output_dir_plots))


# --- 7. Differential Expression with MAST (on original data) ---
message("[MAST] Preparing data for Tumor cell subtypes defined by the atlas.")
cells_focus <- WhichCells(srt, expression = atlas_label %in% tumor_atlas_labels)

if (length(cells_focus) < 100) {
  stop(paste("Fewer than 100 cells found for the specified tumor atlas labels - stopping analysis."))
}

expr <- GetAssayData(srt, slot = "data", assay = "RNA")[, cells_focus, drop = FALSE]
meta <- srt@meta.data[cells_focus, , drop = FALSE]
meta$mets <- factor(meta$mets, levels = c(0, 1), labels = c("No_Mets", "Mets"))

min_detect_frac <- 0.05
min_detect_abs  <- 50
detect_counts <- Matrix::rowSums(expr > 0)
min_detect <- max(min_detect_abs, ceiling(min_detect_frac * ncol(expr)))
keep_genes <- names(which(detect_counts >= min_detect))
expr <- expr[keep_genes, , drop = FALSE]
message(sprintf("[MAST] Testing %d genes in %d cells.", length(keep_genes), length(cells_focus)))

sca <- FromMatrix(
  exprsArray = as.matrix(expr),
  cData      = data.frame(row.names = rownames(meta), meta)
)

colData(sca)$cngeneson <- scale(colSums(assay(sca) > 0))

message("[MAST] Scaling available QC covariates...")
# **UPDATED**: Added age, sex_num, and MLH1_num to the list of potential covariates
optional_covariates <- c("nCount_RNA", "percent.mt", "S.Score", "G2M.Score", "age", "sex_num", "MLH1_num")
found_covariates <- c() 

for (var in optional_covariates) {
  if (var %in% colnames(colData(sca))) {
    if (is.numeric(colData(sca)[[var]]) && var(colData(sca)[[var]], na.rm = TRUE) > 0) {
      colData(sca)[[var]] <- scale(colData(sca)[[var]])
      found_covariates <- c(found_covariates, var)
      message(sprintf("  - Found and scaled '%s'.", var))
    } else {
      message(sprintf("  - [Warning] %s found but has zero variance or is non-numeric — skipping.", var))
    }
  } else {
    message(sprintf("  - [Warning] %s not found in metadata — skipping.", var))
  }
}

message("[MAST] Building model formula...")
colData(sca)$patient_id <- factor(colData(sca)$patient_id)

use_patient_block <- FALSE
if (nlevels(colData(sca)$patient_id) > 1) {
  use_patient_block <- TRUE
  message("[MAST] More than one patient found. Including patient_id as a blocking factor.")
  contrasts(colData(sca)$patient_id) <- contr.sum(nlevels(colData(sca)$patient_id))
} else {
  message("[MAST] [Warning] Only one patient found. Analysis will proceed WITHOUT correcting for patient-specific effects.")
}

base_formula_str <- "~mets"
if (use_patient_block) {
  base_formula_str <- paste(base_formula_str, "+ patient_id")
}
base_formula_str <- paste(base_formula_str, "+ cngeneson")

final_formula_str <- base_formula_str
if (length(found_covariates) > 0) {
  final_formula_str <- paste(base_formula_str, "+", paste(found_covariates, collapse = " + "))
}
fml <- as.formula(final_formula_str)
message(sprintf("[MAST] Using model formula: %s", final_formula_str))

fit <- zlm(formula = fml, sca = sca, method = "bayesglm", ebayes = TRUE)

message("[MAST] Extracting results using the likelihood-ratio test...")
summary_cond <- summary(fit, doLRT = 'metsMets')
summary_dt <- summary_cond$datatable

fc_hurdle <- merge(
  summary_dt[contrast == 'metsMets' & component == 'H', .(primerid, `Pr(>Chisq)`)],
  summary_dt[contrast == 'metsMets' & component == 'logFC', .(primerid, coef)],
  by = 'primerid'
)

colnames(fc_hurdle) <- c("gene", "p_value", "log2FC")
res_final <- fc_hurdle %>%
  as_tibble() %>%
  mutate(FDR = p.adjust(p_value, method = "fdr")) %>%
  arrange(FDR, p_value) %>%
  relocate(gene)

output_dir_de <- file.path(analysis_outputs, "DE_MAST_Mets_vs_NonMets")
dir.create(output_dir_de, showWarnings = FALSE, recursive = TRUE)

file_out <- file.path(output_dir_de, "DEGs_Tumor_Cells_from_Atlas.csv")
write_csv(res_final, file_out)

top_up   <- res_final %>% filter(FDR < 0.05, log2FC > 0.25)
top_down <- res_final %>% filter(FDR < 0.05, log2FC < -0.25)

message("\n-------------------------------------------------")
message("✅ WORKFLOW COMPLETE")
message("-------------------------------------------------")
message(sprintf(
  "Cell Types Analyzed: Tumor Cells (defined by atlas labels)\nDifferentially Expressed Genes (FDR < 0.05 & |log2FC| > 0.25):\n  - Upregulated in Mets: %d\n  - Downregulated in Mets: %d\n\nResults saved to:\n%s",
  nrow(top_up), nrow(top_down), file_out
))
message("-------------------------------------------------\n")
```

Still using mast but a mixed effects model in place of the other one. 
```{r}
# ---
# SCRIPT: MAST Mixed-Effects Model DE Analysis (Mets vs. Non-Mets)
# ---
#
# PURPOSE:
# This script uses MAST's mixed-effects modeling capabilities to perform
# differential expression analysis. It treats each patient as a random effect
# to properly account for the non-independence of cells from the same
# individual, which corrects for pseudoreplication bias.
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(harmony)
  library(MAST)
  library(data.table)
  library(dplyr)
  library(tibble)
  library(readr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
})

# --- 2. Define Inputs, Outputs, and Parameters ---
seurat_object_file <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/Single_cell/Single_cell_samples/Single Cell/combined_data.rds"
analysis_outputs <- "/Users/adiallo/Desktop/Thesis/Results"

tumor_atlas_labels <- c(
  "cE01 (Stem/TA-like)", "cE02 (Stem/TA-like/Immature Goblet)",
  "cE03 (Stem/TA-like prolif)", "cE04 (Enterocyte 1)", "cE05 (Enterocyte 2)",
  "cE06 (Immature Goblet)", "cE07 (Goblet/Enterocyte)", "cE08 (Goblet)",
  "cE09 (Best4)", "cE10 (Tuft)", "cE11 (Enteroendocrine)"
)

# --- 3. Create Patient Metadata Lookup Table ---
patient_info_df <- tibble::tribble(
  ~patient_id, ~age, ~sex_char, ~MLH1_num,
  "3",   76, "M", 1,
  "106", 93, "F", 0,
  "11",  61, "F", 1,
  "116", 64, "M", 0,
  "13",  90, "F", 1,
  "18",  52, "M", 1,
  "40",  71, "F", 0,
  "46",  81, "M", 1,
  "72",  80, "M", 0,
  "86",  59, "F", 1
) %>%
  mutate(sex_num = ifelse(sex_char == "M", 1, 0))

# --- 4. Load and Prepare Seurat Object ---
message(paste("Loading Seurat object from:", seurat_object_file))
srt <- readRDS(file = seurat_object_file)
message("...Data loaded successfully.")

# --- 5. Data Harmonization & Merging ---
message("[INFO] Harmonizing metadata columns and merging patient info...")
srt$atlas_label <- srt$predicted.id
srt$patient_id <- gsub("^patient_", "", srt$orig.ident)
srt$mets <- ifelse(tolower(srt$dataset_origin) == "metastasis", 1, 0)

srt@meta.data <- srt@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  left_join(patient_info_df, by = "patient_id") %>%
  tibble::column_to_rownames("cell_id")
message("...Harmonization and merging complete.")

# --- 6. Data Preparation for MAST ---
# This section is modified to run on the desired cell types and prepare the sca object.
message("[MAST] Preparing data for Tumor cell subtypes defined by the atlas.")
cells_focus <- WhichCells(srt, expression = atlas_label %in% tumor_atlas_labels)

if (length(cells_focus) < 100) {
  stop(paste("Fewer than 100 cells found for the specified tumor atlas labels - stopping analysis."))
}

# NOTE: For MAST, we use the log-normalized 'data' slot
expr <- GetAssayData(srt, slot = "data", assay = "RNA")[, cells_focus, drop = FALSE]
meta <- srt@meta.data[cells_focus, , drop = FALSE]
meta$mets <- factor(meta$mets, levels = c(0, 1), labels = c("No_Mets", "Mets"))

min_detect_frac <- 0.05
min_detect_abs  <- 50
detect_counts <- Matrix::rowSums(expr > 0)
min_detect <- max(min_detect_abs, ceiling(min_detect_frac * ncol(expr)))
keep_genes <- names(which(detect_counts >= min_detect))
expr <- expr[keep_genes, , drop = FALSE]
message(sprintf("[MAST] Testing %d genes in %d cells.", length(keep_genes), length(cells_focus)))

sca <- FromMatrix(
  exprsArray = as.matrix(expr),
  cData      = data.frame(row.names = rownames(meta), meta)
)

colData(sca)$cngeneson <- scale(colSums(assay(sca) > 0))

message("[MAST] Scaling available QC covariates...")
optional_covariates <- c("nCount_RNA", "percent.mt", "S.Score", "G2M.Score", "age", "sex_num", "MLH1_num")
found_covariates <- c() 

for (var in optional_covariates) {
  if (var %in% colnames(colData(sca))) {
    if (is.numeric(colData(sca)[[var]]) && var(colData(sca)[[var]], na.rm = TRUE) > 0) {
      colData(sca)[[var]] <- scale(colData(sca)[[var]])
      found_covariates <- c(found_covariates, var)
      message(sprintf("  - Found and scaled '%s'.", var))
    }
  }
}


# --- 7. Fit the MAST Mixed-Effects Model ---
# WARNING: This method can be extremely slow and memory-intensive.
message("[MAST] Building and fitting mixed-effects model (this will take a long time)...")

# Create a vector of the fixed-effect terms (the variables we are testing or correcting for)
fixed_effects <- c("mets", "cngeneson", found_covariates)

# Assemble the final formula with a random effect for patient_id: `(1|patient_id)`
# This term tells the model that cells from the same patient are related.
final_formula_str <- paste("~", paste(fixed_effects, collapse = " + "), "+ (1|patient_id)")
fml <- as.formula(final_formula_str)
message(sprintf("[MAST] Using mixed-effects model formula: %s", final_formula_str))

# Fit the model using the 'glmer' engine, which is required for mixed models in MAST.
fit <- zlm(formula = fml, 
           sca = sca, 
           method = 'glmer', 
           ebayes = FALSE,
           fitArgsD = list(nAGQ = 0)) # A setting for faster (but less precise) estimation

# --- 8. Extract and Process Results ---
message("[MAST] Extracting results using the likelihood-ratio test...")
summary_cond <- summary(fit, doLRT = 'metsMets') # Test the 'mets' fixed effect
summary_dt <- summary_cond$datatable

fc_hurdle <- merge(
  summary_dt[contrast == 'metsMets' & component == 'H', .(primerid, `Pr(>Chisq)`)],
  summary_dt[contrast == 'metsMets' & component == 'logFC', .(primerid, coef)],
  by = 'primerid'
)

colnames(fc_hurdle) <- c("gene", "p_value", "log2FC")
res_final <- fc_hurdle %>%
  as_tibble() %>%
  mutate(FDR = p.adjust(p_value, method = "fdr")) %>%
  arrange(FDR, p_value) %>%
  relocate(gene)

# --- 9. Save Output ---
output_dir_de <- file.path(analysis_outputs, "DE_MAST_MixedModel_Mets_vs_NonMets")
dir.create(output_dir_de, showWarnings = FALSE, recursive = TRUE)

file_out <- file.path(output_dir_de, "DEGs_MAST_MixedModel_Tumor_Cells.csv")
write_csv(res_final, file_out)

top_up   <- res_final %>% filter(FDR < 0.05, log2FC > 0.25)
top_down <- res_final %>% filter(FDR < 0.05, log2FC < -0.25)

message("\n-------------------------------------------------")
message("✅ MAST MIXED-MODEL WORKFLOW COMPLETE")
message("-------------------------------------------------")
message(sprintf(
  "Cell Types Analyzed: Tumor Cells\nDifferentially Expressed Genes (FDR < 0.05 & |log2FC| > 0.25):\n  - Upregulated in Mets: %d\n  - Downregulated in Mets: %d\n\nResults saved to:\n%s",
  nrow(top_up), nrow(top_down), file_out
))
message("-------------------------------------------------\n")
```

```{r}
# ---
# SCRIPT: Heatmap of Top Differentially Expressed Genes
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(pheatmap)
  library(dplyr)
  library(tibble)
  library(readr)
  library(Matrix)
})

# --- 2. Load and Prepare Data ---
analysis_outputs <- "/Users/adiallo/Desktop/Thesis/Results"
mast_results_file <- file.path(analysis_outputs, "DE_MAST_MixedModel_Mets_vs_NonMets", "DEGs_MAST_MixedModel_Tumor_Cells.csv")
res_final <- readr::read_csv(mast_results_file)

# This assumes the 'sca' (SingleCellAssay) and 'srt' (full Seurat object)
# are available in your R environment from the previous script.

# --- 3. Select the Top Genes for Plotting ---
significant_genes_df <- res_final %>% filter(FDR < 0.05)
top_up <- significant_genes_df %>% arrange(desc(log2FC)) %>% head(25)
top_down <- significant_genes_df %>% arrange(log2FC) %>% head(25)
genes_to_plot <- bind_rows(top_up, top_down) %>% pull(gene)

# --- 4. Create a Pseudobulk Matrix of Average Expression ---
log_norm_counts <- assay(sca)[genes_to_plot, ]
cell_metadata <- as.data.frame(colData(sca))
counts_t <- t(as.matrix(log_norm_counts))
pseudobulk_avg_t <- aggregate(counts_t, by = list(patient = cell_metadata$patient_id), FUN = mean)
pseudobulk_avg <- t(pseudobulk_avg_t[, -1])
colnames(pseudobulk_avg) <- pseudobulk_avg_t$patient
plot_data <- t(scale(t(pseudobulk_avg)))

# --- 5. Prepare Annotations and Plot Heatmap ---
# **CORRECTED SECTION**: Build the annotation directly from the full Seurat object
# to ensure all patient data is correct.
annotation_df <- srt@meta.data %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  dplyr::select(patient_id, mets, sex_num) %>%
  remove_rownames() %>%
  column_to_rownames("patient_id")

# Reorder the annotation rows to match the column order of the heatmap data
annotation_df <- annotation_df[colnames(plot_data), , drop = FALSE]

# Create readable character columns for the plot's legend
annotation_df$Metastasis <- ifelse(annotation_df$mets == 1, "Mets", "No Mets")
annotation_df$Sex <- ifelse(annotation_df$sex_num == 1, "M", "F")
annotation_df$mets <- NULL
annotation_df$sex_num <- NULL

# Define the output file path
output_dir_plots <- file.path(analysis_outputs, "Final_Plots")
dir.create(output_dir_plots, showWarnings = FALSE, recursive = TRUE)
heatmap_file <- file.path(output_dir_plots, "Top50_DEG_Heatmap_Corrected.pdf")

# Generate the heatmap object
heatmap_plot <- pheatmap(plot_data,
                         annotation_col = annotation_df,
                         cluster_rows = TRUE,
                         cluster_cols = TRUE,
                         show_colnames = FALSE,
                         main = "Top 50 DEGs (Mets vs. No Mets)",
                         silent = TRUE)

# Save to a cairo PDF
cairo_pdf(heatmap_file, width = 8, height = 10)
print(heatmap_plot)
dev.off()

message(paste("...Corrected heatmap saved to:", heatmap_file))
```

```{r}
# ---
# SCRIPT: UMAP Feature Plot for Individual Overlapping Genes
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
})

# --- 2. Define Your High-Confidence Genes ---
# The genes that overlapped between your MAST DEG and DMG lists
genes_to_plot <- c("CCDC60", "PRMT6", "RASGRP2")

# --- 3. Generate the Feature Plot ---
# This will create a grid of UMAPs, one for each gene.
# 'order = TRUE' ensures that cells with higher expression are plotted on top.
feature_plot_panel <- FeaturePlot(
  object = srt,
  features = genes_to_plot,
  pt.size = 0.5,
  order = TRUE,
  combine = TRUE # This will arrange the plots into a single grid
) &
  scale_color_viridis_c() # Apply a colorblind-friendly palette to all plots

# --- 4. Save the Plot to a PDF ---
output_dir_plots <- file.path(analysis_outputs, "Final_Plots")
dir.create(output_dir_plots, showWarnings = FALSE, recursive = TRUE)
plot_file <- file.path(output_dir_plots, "FeaturePlot_Overlapping_Genes.pdf")

ggsave(plot_file, feature_plot_panel, width = 12, height = 4)

message(paste("...Feature plot panel saved to:", plot_file))

# Print the plot to the screen
print(feature_plot_panel)
```


Psudobulk Analysis using DESeq
```{r}
# ---
# SCRIPT: Pseudobulk DE Analysis on ALL CELL TYPES
# ---
#
# PURPOSE:
# This script performs a pseudobulk analysis by aggregating counts from ALL cells
# within each patient to create a single sample. This approach mirrors a bulk
# tissue analysis. It then uses DESeq2 to find DEGs associated with metastasis
# status while controlling for patient-level covariates.
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tibble)
  library(readr)
  library(stringr)
  library(Matrix)
  library(DESeq2)
})

# --- 2. Define Inputs and Outputs ---
seurat_object_file <- "/Users/adiallo/Desktop/Thesis/Data_Documents/All_Data/Single_cell/Single_cell_samples/Single Cell/combined_data.rds"
analysis_outputs <- "/Users/adiallo/Desktop/Thesis/Results"

# --- 3. Create Patient Metadata Lookup Table ---
patient_info_df <- tibble::tribble(
  ~patient_id, ~age, ~sex_char, ~MLH1_num,
  "3",   76, "M", 1,
  "106", 93, "F", 0,
  "11",  61, "F", 1,
  "116", 64, "M", 0,
  "13",  90, "F", 1,
  "18",  52, "M", 1,
  "40",  71, "F", 0,
  "46",  81, "M", 1,
  "72",  80, "M", 0,
  "86",  59, "F", 1
) %>%
  mutate(sex_num = ifelse(sex_char == "M", 1, 0))

# --- 4. Load and Prepare Seurat Object ---
message(paste("Loading Seurat object from:", seurat_object_file))
srt <- readRDS(file = seurat_object_file)

message("[INFO] Harmonizing metadata columns and merging patient info...")
srt$patient_id <- gsub("^patient_", "", srt$orig.ident)
srt$mets <- ifelse(tolower(srt$dataset_origin) == "metastasis", 1, 0)

srt@meta.data <- srt@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  left_join(patient_info_df, by = "patient_id") %>%
  tibble::column_to_rownames("cell_id")
message("...Data preparation complete.")

# --- 5. Aggregate Counts to Pseudobulk Level (ALL CELLS) ---
message("[PSEUDOBULK] Aggregating counts from ALL cells to patient level...")

# Get the raw counts from the RNA assay for the ENTIRE object
raw_counts <- GetAssayData(srt, slot = "counts", assay = "RNA")

# Get the corresponding metadata
cell_metadata <- srt@meta.data

# Aggregate counts by patient_id
counts_t <- t(raw_counts)
pseudobulk_t <- rowsum(counts_t, group = cell_metadata$patient_id, reorder = TRUE)
pseudobulk_counts <- t(pseudobulk_t)

message("...Aggregation complete.")

# --- 6. Prepare Metadata for DESeq2 ---
# Create a metadata data frame with one unique row per patient
patient_metadata <- cell_metadata %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  select(patient_id, mets, age, sex_num, MLH1_num) %>%
  remove_rownames() %>%
  column_to_rownames("patient_id")

# Ensure the column order of the counts matrix matches the row order of the metadata
pseudobulk_counts <- pseudobulk_counts[, rownames(patient_metadata)]

# --- 7. Run DESeq2 ---
message("[DESeq2] Running differential expression analysis...")

dds <- DESeqDataSetFromMatrix(countData = round(pseudobulk_counts),
                              colData = patient_metadata,
                              design = ~ age + sex_num + MLH1_num + mets)
dds$mets <- factor(dds$mets, levels = c("0", "1"))
dds$mets <- relevel(dds$mets, ref = "0")
dds <- DESeq(dds)
res <- results(dds, name="mets_1_vs_0")
res <- as.data.frame(res) %>%
  rownames_to_column("gene") %>%
  as_tibble() %>%
  arrange(padj)

message("...DESeq2 analysis complete.")

# --- 8. Save Output ---
output_dir_de <- file.path(analysis_outputs, "DE_Pseudobulk_All_Cells")
dir.create(output_dir_de, showWarnings = FALSE, recursive = TRUE)

file_out <- file.path(output_dir_de, "DEGs_Pseudobulk_All_Cells.csv")
write_csv(res, file_out)

top_hits <- res %>% filter(padj < 0.05)

message("\n-------------------------------------------------")
message("✅ PSEUDOBULK WORKFLOW COMPLETE")
message("-------------------------------------------------")
message(sprintf(
  "Analysis of: ALL Cells (aggregated by patient)\nDifferentially Expressed Genes (FDR < 0.05):\n  - Found %d significant genes.\n\nResults saved to:\n%s",
  nrow(top_hits), file_out
))
message("-------------------------------------------------\n")
```

```{r}
# ---
# SCRIPT: Heatmap of Overlapping Genes
# ---
#
# PURPOSE:
# Generate and save a publication-quality heatmap as a PDF, showing the
# expression pattern of genes that overlap between methylation and
# expression analyses. The heatmap is annotated by metastasis and sex.
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(pheatmap)
  library(DESeq2)
  library(dplyr)
  library(tibble)
  library(readr)
})

# --- 2. Define Paths and Gene Lists ---
analysis_outputs <- "/Users/adiallo/Desktop/Thesis/Results"
degs_results_file <- file.path(analysis_outputs, "DE_Pseudobulk_All_Cells", "DEGs_Pseudobulk_All_Cells.csv")
degs_df <- readr::read_csv(degs_results_file)

degs_list <- degs_df %>% filter(padj < 0.05) %>% pull(gene)

any_mets_dmgs <- c("CAPN13", "JMJD8", "CA10", "LINC00424", "GALNT14", "ADRA2C", "ANK1", "FER1L6-AS2", "SLIT2", "ABL1", "PAK1", "C10orf11", "CCDC60", "PTPRM", "HLA-DOA", "LOC644145", "H2AFY2", "TFAP2B", "C8orf33", "KCNQ2", "ATP7B", "DSCR3", "SNORA39", "SPTBN4")
ln_mets_dmgs <- c("CENPB", "GPR1-AS", "ZMYM2", "FAM126A", "ABCA7", "LOC100190940", "GOLT1A")
distant_mets_dmgs <- c("MARK1", "BSN", "FNDC4", "FAM150B", "HLA-L", "ELOVL5", "EPHB1", "FAM69B", "COL5A1", "TNXB", "LRRC14B", "TMEM87B", "RASA3", "SOCS1", "MAPK4", "ADRA1D", "GALNT9", "SLC38A3", "NAT8L", "PVALB", "C12orf68", "PRDM8", "GPR123", "ZDHHC22", "LHX4", "C3orf50", "BAI3", "C14orf39", "NXN", "SLC1A2", "RASGRP2", "TCHH", "KLF7", "CORO2B", "LINC01152", "CSMD2", "SHC4", "NRN1", "TET1", "ABR", "RGS7", "ZDBF2", "FLJ44817", "BRSK2", "BNC2", "C1orf65", "FAM163A", "LMO2", "ZNF382", "ISM1", "MIXL1", "GRHL3", "KIF1A", "GABRB2", "DTX1", "CHST1", "ECE2", "HTR4", "GRK5", "DPYSL4", "STMN2", "INHBB", "NEUROG1", "S1PR1", "CDYL", "SNCAIP", "SLC30A3", "LOC387763", "SLC30A10", "C3orf55", "MLN", "MCC", "TMOD1", "NTF3", "WNT1", "TACR1", "WNT5A", "AP3B2", "PRMT6", "TFPT", "FGR", "OTX2-AS1", "3-Mar", "4-Mar", "FJX1", "MIR548H4", "BEND6", "ZNF775", "STXBP5L", "NAV2", "SOX14", "PRPH", "LMF1", "NEURL", "ADAMTS19-AS1", "KREMEN2", "GRIN2C", "CR2", "SAMD11", "PITX2", "MACROD1", "STMN3", "MFSD2B", "EPHB6", "CNIH2", "TRPM1", "CACNB4", "ALDH1A3", "NRP2", "WSCD1", "SIX3", "SESN1", "CCDC42B", "PRPF31", "UQCC")

# --- 3. Generate and Save Heatmap ---
message("Generating heatmap...")

# Get all unique overlapping genes
all_dmgs <- unique(c(any_mets_dmgs, ln_mets_dmgs, distant_mets_dmgs))
overlapping_genes <- intersect(degs_list, all_dmgs)

if (length(overlapping_genes) > 0) {
  # This assumes 'dds' is your DESeq2 object from the previous analysis
  vst_counts <- varianceStabilizingTransformation(dds, blind = FALSE)
  plot_data <- assay(vst_counts)[overlapping_genes, , drop = FALSE]

  # Z-score scale the data by row (gene)
  plot_data <- t(scale(t(plot_data)))

  # Create an annotation data frame for the heatmap columns
  annotation_df <- as.data.frame(colData(dds)[, c("mets", "sex_num")])
  annotation_df$Metastasis <- ifelse(annotation_df$mets == 1, "Mets", "No Mets")
  annotation_df$Sex <- ifelse(annotation_df$sex_num == 1, "M", "F")
  annotation_df$mets <- NULL
  annotation_df$sex_num <- NULL

  # Create the heatmap object
  heatmap_plot <- pheatmap(plot_data,
                           annotation_col = annotation_df,
                           cluster_rows = TRUE,
                           cluster_cols = TRUE,
                           show_colnames = FALSE,
                           main = "Expression of Overlapping Genes (All Cells Pseudobulk)",
                           silent = TRUE)

  # Define output path and save the plot to a PDF
  output_dir_plots <- file.path(analysis_outputs, "Final_Plots")
  dir.create(output_dir_plots, showWarnings = FALSE, recursive = TRUE)
  heatmap_file <- file.path(output_dir_plots, "Overlap_Heatmap.pdf")
  
  cairo_pdf(heatmap_file, width = 8, height = 6)
  print(heatmap_plot)
  dev.off()
  
  message(paste("...Heatmap saved to:", heatmap_file))
  
} else {
  message("No overlapping genes found between DEGs and DMGs. Skipping heatmap.")
}

```

```{r}
# ---
# SCRIPT: UMAP Feature Plot for Differentially Methylated Genes (DMGs)
# ---

# --- 1. Load Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(ggplot2)
  library(patchwork)
})

# --- 2. Load Your Processed Seurat Object ---
# This should be the 'srt' object that has already had Harmony run on it.
# If it's not in your environment, load it:
# srt <- readRDS("path/to/your/harmony_processed_seurat_object.rds")

# --- 3. Create 'cell_class' Metadata Column ---
# This mapping creates the broad cell type categories from your detailed atlas labels.
message("Creating 'cell_class' metadata column...")
label_to_class <- c("cE01 (Stem/TA-like)"="Tumor cells", "cE02 (Stem/TA-like/Immature Goblet)"="Tumor cells", "cE03 (Stem/TA-like prolif)"="Tumor cells", "cE04 (Enterocyte 1)"="Tumor cells", "cE05 (Enterocyte 2)"="Tumor cells", "cE06 (Immature Goblet)"="Tumor cells", "cE07 (Goblet/Enterocyte)"="Tumor cells", "cE08 (Goblet)"="Tumor cells", "cE09 (Best4)"="Tumor cells", "cE10 (Tuft)"="Tumor cells", "cE11 (Enteroendocrine)"="Tumor cells", "cS25 (Fibro CCL8+)"="Fibroblasts", "cS26 (Myofibro)"="Fibroblasts", "cS27 (CXCL14+ CAF)"="Fibroblasts", "cS28 (GREM1+ CAF)"="Fibroblasts", "cS29 (MMP3+ CAF)"="Fibroblasts", "cS30 (CAF CCL8 Fibro-like)"="Fibroblasts", "cS21 (Fibro stem cell niche)"="Fibroblasts", "cS22 (Fibro stem cell niche)"="Fibroblasts", "cS23 (Fibro BMP-producing)"="Fibroblasts", "cS24 (Fibro BMP-producing)"="Fibroblasts", "cS31 (CAF stem niche Fibro-like)"="Fibroblasts", "cTNI10 (CD8+ IL7R+)"="CD8+ T-cells", "cTNI11 (CD8+GZMK+)"="CD8+ T-cells", "cTNI12 (CD8+ IL7R+)"="CD8+ T-cells", "cTNI13 (CD8+ T IL17+)"="CD8+ T-cells", "cTNI14 (CD8+ CXCL13+)"="CD8+ T-cells", "cTNI15 (CD8+ CXCL13+ HSP+)"="CD8+ T-cells", "cTNI16 (CD8+ CXCL13+ prolif)"="CD8+ T-cells", "cM02 (Macrophage-like)"="Macrophages", "cM01 (Monocyte)"="Macrophages", "cTNI01 (CD4+ IL7R+)"="CD4+ T-cells", "cTNI02 (CD4+ IL7R+SELL+)"="CD4+ T-cells","cTNI03 (CD4+ IL7R+HSP+)"="CD4+ T-cells","cTNI04 (CD4+ IL7R+CCL5+)"="CD4+ T-cells","cTNI05 (CD4+ IL17+)"="CD4+ T-cells","cTNI06 (CD4+ TFH)"="TFH cells","cTNI07 (CD4+ CXCL13+)"="CD4+ T-cells","cTNI08 (CD4+ Treg)"="Treg","cTNI09 (CD4+ Treg prolif)"="Treg")

srt$cell_class <- unname(label_to_class[srt$predicted.id])
srt$cell_class[is.na(srt$cell_class)] <- "Other" # Assign 'Other' to any unmapped labels
message("...Done.")

# --- 4. Define Your Full List of Differentially Methylated Genes ---
any_mets_dmgs <- c("CAPN13", "JMJD8", "CA10", "LINC00424", "GALNT14", "ADRA2C", "ANK1", "FER1L6-AS2", "SLIT2", "ABL1", "PAK1", "C10orf11", "CCDC60", "PTPRM", "HLA-DOA", "LOC644145", "H2AFY2", "TFAP2B", "C8orf33", "KCNQ2", "ATP7B", "DSCR3", "SNORA39", "SPTBN4")
ln_mets_dmgs <- c("CENPB", "GPR1-AS", "ZMYM2", "FAM126A", "ABCA7", "LOC100190940", "GOLT1A")
distant_mets_dmgs <- c("MARK1", "BSN", "FNDC4", "FAM150B", "HLA-L", "ELOVL5", "EPHB1", "FAM69B", "COL5A1", "TNXB", "LRRC14B", "TMEM87B", "RASA3", "SOCS1", "MAPK4", "ADRA1D", "GALNT9", "SLC38A3", "NAT8L", "PVALB", "C12orf68", "PRDM8", "GPR123", "ZDHHC22", "LHX4", "C3orf50", "BAI3", "C14orf39", "NXN", "SLC1A2", "RASGRP2", "TCHH", "KLF7", "CORO2B", "LINC01152", "CSMD2", "SHC4", "NRN1", "TET1", "ABR", "RGS7", "ZDBF2", "FLJ44817", "BRSK2", "BNC2", "C1orf65", "FAM163A", "LMO2", "ZNF382", "ISM1", "MIXL1", "GRHL3", "KIF1A", "GABRB2", "DTX1", "CHST1", "ECE2", "HTR4", "GRK5", "DPYSL4", "STMN2", "INHBB", "NEUROG1", "S1PR1", "CDYL", "SNCAIP", "SLC30A3", "LOC387763", "SLC30A10", "C3orf55", "MLN", "MCC", "TMOD1", "NTF3", "WNT1", "TACR1", "WNT5A", "AP3B2", "PRMT6", "TFPT", "FGR", "OTX2-AS1", "3-Mar", "4-Mar", "FJX1", "MIR548H4", "BEND6", "ZNF775", "STXBP5L", "NAV2", "SOX14", "PRPH", "LMF1", "NEURL", "ADAMTS19-AS1", "KREMEN2", "GRIN2C", "CR2", "SAMD11", "PITX2", "MACROD1", "STMN3", "MFSD2B", "EPHB6", "CNIH2", "TRPM1", "CACNB4", "ALDH1A3", "NRP2", "WSCD1", "SIX3", "SESN1", "CCDC42B", "PRPF31", "UQCC")

all_dmgs_list <- unique(c(any_mets_dmgs, ln_mets_dmgs, distant_mets_dmgs))

# --- 5. Calculate the Module Score ---
srt <- AddModuleScore(
  object = srt,
  features = list(all_dmgs_list),
  name = 'DMG_Score'
)

# --- 6. Generate the Feature Plot ---
p_score <- FeaturePlot(
  object = srt,
  features = 'DMG_Score1',
  pt.size = 0.5,
  order = TRUE,
  min.cutoff = 'q10',
  max.cutoff = 'q90'
) +
  ggtitle("Expression of DMG Signature") +
  scale_color_viridis_c()

# For comparison, plot the cell types
p_clusters <- DimPlot(srt, group.by = "cell_class") + ggtitle("Cell Types")

# Display the two plots side-by-side
print(p_clusters + p_score)
```

